### 2023-01-24 presentation/index.qmd
prepare presentation in revealjs

---
title: "Machine Learned Models for Estimation of Sleep Quality in Free-Living Accelerometer Data"
subtitle: "(working title)"
author: "Esben Lykke, PhD student"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
format:
  revealjs: 
    scrollable: true
    smaller: true
    theme: moon
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    logo: images/SDU_WHITE_RGB-png.png
    footer: https://github.com/esbenlykke/sleep_study
    css: my_styles.css
editor_options: 
  chunk_output_type: console
smooth-scroll: true
---

## Background {.scrollable transition="convex"}

Lists can optionally be displayed incrementally:

::: incremental
-   First item
-   Second item
-   Third item
:::

. . .

Insert pauses to make other types of content display incrementally.

::: footer
Learn more: [Incremental Lists](https://quarto.org/docs/presentations/revealjs/#incremental-lists)
:::

## Methods {.scrollable transition="zoom"}

::: {.fragment .fade-in-then-semi-out}
Hello there
:::

:::{.fragment .fade-in-then-semi-out}
hi there
:::

---

## {.smaller .scrollable transition="convex" .column}

```{r}
source("~/sleep_study/presentation/code/tables_performance.R")

tab_1
```

Performance of the models to predict each class <span style="color:gold">**seperately**</span>, i.e., "sleep" and "in-bed".

## {.smaller .scrollable transition="convex" .column}

```{r}
tab_2
```

Performance of the models to predict each <span style="color:gold">**combined**</span> class, i.e., "sleep" + "in-bed".

---

## Bland-Altman Plots {.smaller transition="convex"}

::: {.panel-tabset}

### Decision Tree

```{r}
#| echo: false
source("~/sleep_study/presentation/code/ba_plots.R")
plots$decision_tree[[1]]

plots$decision_tree[[2]]

plots$decision_tree[[3]]

plots$decision_tree[[4]]

plots$decision_tree[[5]]
```

### Logistic Regression

```{r}
plots$logistic_regression[[1]]

plots$logistic_regression[[2]]

plots$logistic_regression[[3]]

plots$logistic_regression[[4]]

plots$logistic_regression[[5]]
```

### XGBoost

```{r}
plots$xgboost[[1]]

plots$xgboost[[2]]

plots$xgboost[[3]]

plots$xgboost[[4]]

plots$xgboost[[5]]
```

### Neural Net

```{r}
plots$neural_net[[1]]

plots$neural_net[[2]]

plots$neural_net[[3]]

plots$neural_net[[4]]

plots$neural_net[[5]]
```

:::

---

### Tree Plots {.smaller .scrollable transition="zoom" .column}

```{r}
#| echo: false
library(arrow)
library(visNetwork)
library(tidymodels)

in_bed_tree <-
  read_rds("~/sleep_study/data/models/fitted_models/in_bed_simple_CART_fit.rds")

in_bed_tree |>
  extract_fit_engine() |>
  visTree(
    height = "800px",
    nodesPopSize = TRUE,
    minNodeSize = 10,
    maxNodeSize = 30,
    fallenLeaves = TRUE,
    legend = FALSE,
    digits = 3,
    direction = "LR"
  ) |>
  visNodes(font = list(
    color = "#EEE8D5",
    strokeWidth = 1
  )) |>
  visEdges(font = list(
    color = "#EEE8D5",
    strokeWidth = 1
  )) |> 
  visLegend(addEdges = list(font = list(color = "#EEE8D5")))
```

### 2023-01-24 presentation/code/ba_plots.R

@@ -1,0 +1,159 @@
+library(tidyverse)
+library(arrow)
+library(patchwork)
+library(showtext)
+
+stats_files <-
+  list.files("~/sleep_study/data/processed/stats_predictions", full.names = TRUE)
+
+all_stats <-
+  map(stats_files, read_parquet) |>
+  set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
+
+get_diff_stats <-
+  function(tbl) {
+    tbl |>
+      mutate(
+        diff_spt_hrs = spt_hrs - zm_spt_hrs,
+        diff_tst_hrs = tst_hrs - zm_tst_hrs,
+        diff_se_percent = se_percent - zm_se_percent,
+        diff_lps_min = lps_min - zm_lps_min,
+        diff_waso_min = waso_min - zm_waso_min
+      ) |>
+      rowwise() |>
+      mutate(
+        avg_spt_hrs = mean(c(spt_hrs, zm_spt_hrs)),
+        avg_tst_hrs = mean(c(tst_hrs, zm_tst_hrs)),
+        avg_se_percent = mean(c(se_percent, zm_se_percent)),
+        avg_lps_min = mean(c(lps_min, zm_lps_min)),
+        avg_waso_min = mean(c(waso_min, zm_waso_min))
+      ) |>
+      ungroup()
+    # filter(se_percent > 0 & lps_min > 0) # TODO why are these numbers fucked?
+  }
+
+get_summary_stats <-
+  function(tbl) {
+    tbl |>
+      # filter(se_percent > 0) |>
+      summarise(
+        across(contains("diff"), list(
+          upper = ~ mean(.x) + 1.96 * sd(.x),
+          lower = ~ mean(.x) - 1.96 * sd(.x),
+          mean = ~ mean(.x)
+        ))
+      )
+  }
+
+all_diffs <-
+  all_stats |>
+  map(get_diff_stats)
+
+all_summaries <-
+  all_stats |>
+  map(get_diff_stats) |>
+  map(get_summary_stats)
+
+# BA plots for simple trees -----------------------------------------------
+
+
+font_add_google("Mukta", family = "mukta")
+font_add_google("IBM Plex Serif", family = "ibm")
+showtext_auto()
+
+all_plots <-
+  function(diffs, summaries) {
+    ba_plot <- function(diffs, avg, diff, mean_diff, lower, upper, title, x_axis = "", y_axis = "") {
+      diffs |>
[…]

### 2023-01-24 presentation/code/tables_performance.R

@@ -1,0 +1,90 @@
+library(tidyverse)
+library(gt)
+library(showtext)
+
+metric_data <- read_csv("~/sleep_study/data/processed/performance_metrics.csv")
+
+
+# Create table ------------------------------------------------------------
+
+
+font_add_google("IBM Plex Serif", family = "ibm")
+showtext_auto()
+
+tab_1 <- 
+  metric_data |> 
+  mutate(
+    .metric = factor(.metric, 
+                     levels = c("f_meas", "accuracy", "sensitivity", "specificity"),
+                     labels = c("F1 Score", "Accuracy", "Sensitivity", "Specificity"))
+  ) |> 
+  group_by(event) |>
+  gt(rowname_col = ".metric") |>
+  fmt_percent(
+    columns = c(logistic_regression:xgboost),
+    decimals = 2
+  ) |>
+  cols_label(
+    logistic_regression = "Logistic Regression",
+    neural_net = "Neural Network",
+    decision_tree = "Decision Tree",
+    xgboost = "XGboost"
+  ) |>
+  tab_style(
+    style = cell_text(size = px(12)),
+    locations = cells_body(
+      columns = c(logistic_regression:xgboost)
+    )
+  ) |>
+  tab_header(
+    title = md("Performance Metrics"),
+    subtitle = "Grouped by Event Prediction"
+  ) |>
+  cols_width(everything() ~ px(100)) |> 
+  cols_align(align = "center") |> 
+  data_color(columns = decision_tree, colors = "#4A666C", alpha = .6) |>
+  tab_options(table.font.names = "ibm",
+              table.font.color.light = "#EEE8D5",
+              table.background.color = "#002B36",
+  ) 
+
+comb_metric_data <- read_csv("~/sleep_study/data/processed/combined_preds_performance_metrics.csv")
+
+tab_2 <- 
+  comb_metric_data |> 
+  mutate(
+    .metric = factor(.metric, 
+                     levels = c("f_meas", "accuracy", "sensitivity", "specificity"),
+                     labels = c("F1 Score", "Accuracy", "Sensitivity", "Specificity"))
+  ) |> 
+  group_by(event) |>
+  gt(rowname_col = ".metric") |>
+  fmt_percent(
+    columns = c(logistic_regression:xgboost),
+    decimals = 2
+  ) |>
+  cols_label(
+    logistic_regression = "Logistic Regression",
+    neural_net = "Neural Network",
[…]

### 2023-01-26 presentation/index.qmd
proxy clock feature figure

@@
   revealjs: 
-    scrollable: true
     smaller: true
@@
 smooth-scroll: true
+bibliography: references.bib
 ---
@@
 ## Background {.scrollable transition="convex"}
+
+```{r}
+#| echo: false
+library(gt)
+library(arrow)
+library(visNetwork)
+library(tidymodels)
+```
 
@@
 
-## Methods {.scrollable transition="zoom"}
+## Methods {transition="convex"}
+
+Overview
+Network plot of data preparation steps
+
+. . .
+
+network plot
+
+## Features {transition="fade" .smaller}
+
+:::: {.columns}
+
+::: {.column width="30%"}
+::: {.fragment}
+**Basic Features**
+
+-   Weekday 
+-   Time of Day
+-   Placement
+-   Temperature
+:::
+:::
+
+::: {.column width="30%"}
+::: {.fragment}
+**ACC derived features**^[@skotte_detection_2014]
+
+-   Mean ACC X 
+-   Mean ACC Y
+-   Mean ACC Z
+-   Standard Deviation X
+-   Standard Deviation Y
+-   Standard Deviation Z
+-   Max Standard Deviation
+-   Inclination
+:::
+:::
+
+::: {.column width="40%"}
+::: {.fragment}
+**Sensor-Independent Features**^[@walch_sleep_2019, @forger_simpler_1999]
+
+-   Clock Proxy Linear
+-   Clock Proxy Cosinus
+:::
+:::
+
+::::
+
+:::aside
+:::
+---
+
+## Human Circadian Clock {.smaller transition="convex"}
+
+@forger_simpler_1999: a so-called cubic van der Pol equation
+
+::: {.fragment}
+$$\frac{dx_c}{dt}=\frac{\pi}{12}\begin{cases}\mu(x_c-\frac{4x^3}{3})-x\begin{bmatrix}(\frac{24}{0.99669\tau_x})^2+kB\end{bmatrix}\end{cases}$$
+:::
 
 ::: {.fragment .fade-in-then-semi-out}
-Hello there
+This thing is dependent on ambient light and body temperature!
 :::
 
-:::{.fragment .fade-in-then-semi-out}
-hi there
+::: {.fragment .fade-in-then-semi-out}
+@walch_sleep_2019 incorporated this feature using step counts from the Apple Watch
 :::
 
----
+::: {.fragment}
+But as demonstrated by @walch_sleep_2019, a simple cosine function does the tricks just as well :)
+:::
 
-## {.smaller .scrollable transition="convex" .column}
+## Circadian Proxy Features {.smaller transition="convex"}
+
+```{r}
+source(here::here("presentation/code/plot_sensor-independent.R"))
+
+plot_proxy
+```
+
+## {.smaller transition="convex" .column}
 
@@
 
-## {.smaller .scrollable transition="convex" .column}
+## {.smaller transition="convex" .column}
 
@@
 
-## Bland-Altman Plots {.smaller transition="convex"}
+## Bland-Altman Plots {.smaller transition="convex" .slide}
 
@@
 #| echo: false
-source("~/sleep_study/presentation/code/ba_plots.R")
+source(here::here("presentation/code/ba_plots.R"))
 plots$decision_tree[[1]]
@@
 
-### Tree Plots {.smaller .scrollable transition="zoom" .column}
+## In-bed classification flow {.smaller transition="zoom" .column}
 
 ```{r}
-#| echo: false
-library(arrow)
-library(visNetwork)
-library(tidymodels)
-
 in_bed_tree <-
@@
   visTree(
-    height = "800px",
+    height = "550px",
     nodesPopSize = TRUE,
@@
     digits = 3,
-    direction = "LR"
+    direction = "UD",
+    colorVar = c(
+      "#CC5500", "#8B0000", "#FFD700",
+      "#228B22", "#654321", "#8B0000", "#FFDB58"
+    ),
+    export = FALSE
+  ) |>
+  visNodes(font = list(
+    color = "#EEE8D5",
+    size = 20,
+    strokeWidth = 1
+  )) |>
+  visEdges(font = list(
+    color = "#EEE8D5",
+    size = 20,
+    strokeWidth = 1
+  )) |> 
+  visInteraction(dragNodes = FALSE, 
+                 dragView = FALSE, 
+                 zoomView = FALSE)
+```
+
+---
+
+## Sleep classification flow {.smaller transition="zoom" .column}
+
+```{r}
+sleep_tree <-
+  read_rds("~/sleep_study/data/models/fitted_models/sleep_simple_CART_fit.rds")
+
+sleep_tree |>
+  extract_fit_engine() |>
+  visTree(
+    height = "550px",
+    nodesPopSize = TRUE,
+    minNodeSize = 10,
+    maxNodeSize = 30,
+    fallenLeaves = TRUE,
+    legend = FALSE,
+    digits = 3,
+    direction = "UD",
+    colorVar = c(
+      "#CC5500", "#8B0000", "#FFD700",
+      "#228B22", "#654321", "#8B0000", "#FFDB58"
+    ),
+    export = FALSE
   ) |>
@@
   )) |> 
-  visLegend(addEdges = list(font = list(color = "#EEE8D5")))
+  visInteraction(dragNodes = FALSE, 
+                 dragView = FALSE, 
+                 zoomView = FALSE)
 ```

### 2023-01-26 code/prepare_data/sensor_independent_features.R

@@ -8,6 +8,8 @@ cat("Creating sensor-independent features. This won't take long...")
 
 data <-
   read_parquet("data/processed/data_for_modelling/bsl_thigh.parquet")
+# data <-
+#   read_parquet("data/processed/data_for_modelling/fup_thigh.parquet")
 
 
 # Subject-level in-bed time -----------------------------------------------
@@ -109,7 +111,7 @@ clock_proxy |>
   select(
     id, datetime, unix_time, day, noon_day, age, placement,
     clock_proxy_cos, clock_proxy_linear, incl, theta, temp:time_day
-  ) |>
+  ) |> count(noon_day, id)
   write_parquet("data/processed/data_for_modelling/bsl_thigh_sensor_independent_features.parquet")
 
 
@@ -123,7 +125,7 @@ clock_proxy |>
   select(
     id, datetime, unix_time, day, noon_day, age, placement,
     clock_proxy_cos, clock_proxy_linear, incl, theta, temp:time_day
-  ) |>
-  filter(id == 757104) |> 
+  ) |> 
+  filter(id == 1587606) |> 
   ggplot(aes(datetime, clock_proxy_cos)) +
-  geom_step()
+  geom_step() 

### 2023-01-26 presentation/code/build_tree_plots.R

@@ -1,182 +1,0 @@
-library(tidyverse)
-library(igraph)
-library(showtext)
-
-in_bed_tree_data <-
-  tibble(
-    from = c(
-      "clock_proxy_cos < 31e-18",
-      "clock_proxy_cos < 31e-18",
-      "sdmax >= 0.0031",
-      "sdmax >= 0.0031",
-      "clock_proxy_cos < 0.27",
-      "clock_proxy_cos < 0.27",
-      "temp < 33",
-      "temp < 33",
-      "incl >= 119",
-      "incl >= 119",
-      "sdmax >= 0.0046",
-      "sdmax >= 0.0046",
-      "macc_y >= 0.64",
-      "macc_y >= 0.64"
-    ),
-    to = c(
-      "sdmax >= 0.0031",
-      "clock_proxy_cos < 0.27",
-      "out-bed 53%",
-      "temp < 33",
-      "sdmax >= 0.0046",
-      "in-bed 3%",
-      "out-bed 1%",
-      "incl >= 119",
-      "out-bed 5%",
-      "in-bed ~0%",
-      "macc_y >= 0.64",
-      "in-bed 1%",
-      "out-bed 5%",
-      "in-bed 32%"
-    )
-  )
-
-g <-
-  graph_from_data_frame(in_bed_tree_data, directed = TRUE)
-
-coords <-
-  layout_as_tree(g) |>
-  as_tibble(.name_repair = ~ c("x", "y"))
-
-output_df <-
-  coords |>
-  mutate(
-    step = vertex_attr(g, "name"),
-    label = step,
-    x = x * -1
-  )
-
-plot_nodes <-
-  output_df %>%
-  mutate(
-    xmin = x - 0.35,
-    xmax = x + 0.35,
-    ymin = y - 0.15,
-    ymax = y + 0.15
-  )
-
-plot_edges <-
-  in_bed_tree_data %>%
-  mutate(id = row_number()) %>%
-  pivot_longer(
-    cols = c("from", "to"),
-    names_to = "s_e",
-    values_to = "step"
-  ) %>%
-  left_join(plot_nodes, by = "step") %>%
-  select(-c(label, y, xmin, xmax)) %>%
-  mutate(y = ifelse(s_e == "from", ymin, ymax)) %>%
-  select(-c(ymin, ymax))
-
-font_add_google(name = "IBM Plex Serif", family = "ibm")
-showtext_auto()
-
-# p <-
[…]

### 2023-01-26 presentation/code/network_tree_plots.R

@@ -1,0 +1,174 @@
+library(tidyverse)
+library(igraph)
+library(showtext)
+
+in_bed_tree_data <-
+  tibble(
+    from = c(
+      "clock_proxy_cos < 31e-18",
+      "clock_proxy_cos < 31e-18",
+      "sdmax >= 0.0031",
+      "sdmax >= 0.0031",
+      "clock_proxy_cos < 0.27",
+      "clock_proxy_cos < 0.27",
+      "temp < 33",
+      "temp < 33",
+      "incl >= 119",
+      "incl >= 119",
+      "sdmax >= 0.0046",
+      "sdmax >= 0.0046",
+      "macc_y >= 0.64",
+      "macc_y >= 0.64"
+    ),
+    to = c(
+      "sdmax >= 0.0031",
+      "clock_proxy_cos < 0.27",
+      "out-bed 53%",
+      "temp < 33",
+      "sdmax >= 0.0046",
+      "in-bed 3%",
+      "out-bed 1%",
+      "incl >= 119",
+      "out-bed 5%",
+      "in-bed ~0%",
+      "macc_y >= 0.64",
+      "in-bed 1%",
+      "out-bed 5%",
+      "in-bed 32%"
+    )
+  )
+
+g <-
+  graph_from_data_frame(in_bed_tree_data, directed = TRUE)
+
+coords <-
+  layout_as_tree(g) |>
+  as_tibble(.name_repair = ~ c("x", "y"))
+
+output_df <-
+  coords |>
+  mutate(
+    step = vertex_attr(g, "name"),
+    label = step,
+    x = x * -1
+  )
+
+plot_nodes <-
+  output_df %>%
+  mutate(
+    xmin = x - 0.35,
+    xmax = x + 0.35,
+    ymin = y - 0.15,
+    ymax = y + 0.15
+  )
+
+plot_edges <-
+  in_bed_tree_data %>%
+  mutate(id = row_number()) %>%
+  pivot_longer(
+    cols = c("from", "to"),
+    names_to = "s_e",
+    values_to = "step"
+  ) %>%
+  left_join(plot_nodes, by = "step") %>%
+  select(-c(label, y, xmin, xmax)) %>%
+  mutate(y = ifelse(s_e == "from", ymin, ymax)) %>%
+  select(-c(ymin, ymax))
+
+font_add_google(name = "IBM Plex Serif", family = "ibm")
+showtext_auto()
+
+# p <-
+plot_nodes |>
+  ggplot(
+    aes(
[…]

### 2023-01-26 presentation/code/plot_sensor-independent.R

@@ -1,0 +1,176 @@
+library(tidyverse)
+library(lubridate)
+library(showtext)
+library(ggtext)
+
+proxy <-
+  read_csv(here::here("presentation/data/__clock_proxy_example.csv"))
+
+font_add_google("IBM Plex Serif", family = "ibm")
+showtext_auto()
+
+static_start <-
+  proxy |>
+  slice(min(row_number()[clock_proxy_cos > 0 & noon_day == 27])) |>
+  pull(datetime)
+
+static_end <-
+  proxy |>
+  slice(max(row_number()[clock_proxy_cos > 0 & noon_day == 27])) |>
+  pull(datetime)
+
+clock_proxy_start27 <-
+  proxy |>
+  slice(min(row_number()[clock_proxy_cos > 0 & noon_day == 25])) |>
+  pull(datetime)
+
+clock_proxy_end27 <-
+  proxy |>
+  slice(max(row_number()[clock_proxy_cos > 0 & noon_day == 25])) |>
+  pull(datetime)
+
+# clock_proxy_start26 <-
+#   proxy |>
+#   slice(min(row_number()[clock_proxy_cos > 0 & noon_day == 26])) |>
+#   pull(datetime)
+#
+# clock_proxy_end26 <-
+#   proxy |>
+#   slice(max(row_number()[clock_proxy_cos > 0 & noon_day == 26])) |>
+#   pull(datetime)
+
+plot_proxy <-
+  proxy |>
+  ggplot(aes(datetime, clock_proxy_cos)) +
+  geom_step(color = "#EEE8D5", linewidth = .8, alpha = .8) +
+  geom_step(color = "#4D8C57", linewidth = .3, alpha = .8) +
+  geom_vline(xintercept = as_datetime(static_start), color = "darkorange", lty = 2) +
+  geom_vline(xintercept = as_datetime(static_end), color = "darkorange", lty = 2) +
+  geom_vline(xintercept = as_datetime(clock_proxy_start27), color = "lightblue", lty = 2) +
+  geom_vline(xintercept = as_datetime(clock_proxy_end27), color = "lightblue", lty = 2) +
+  # geom_vline(xintercept = as_datetime(clock_proxy_start26), color = "lightblue", lty = 2) +
+  # geom_vline(xintercept = as_datetime(clock_proxy_end26), color = "lightblue", lty = 2) +
+  annotate(
+    geom = "text",
+    x = as_datetime(static_start) - hours(1),
[…]

### 2023-01-26 presentation/index.qmd
make html file self-contained

@@
     slide-number: true
-    chalkboard: 
-      buttons: false
     preview-links: auto
@@
     css: my_styles.css
+embed-resources: true
 editor_options: 

### 2023-01-28 presentation/index.qmd
create flowchart

@@
 
+## Exclusion Criteria {transition="fade"}
+
+```{r}
+#| fig-align: center
+source(here::here("presentation/code/flowchart_eligible_nights.R"))
+
+flow
+```
+
+::: aside
+1. @hirshkowitz_national_2015
+:::
+
 ## Features {transition="fade" .smaller}
@@
 ::: {.column width="30%"}
-::: {.fragment}
+
 **Basic Features**
@@
 -   Temperature
-:::
+
 :::
@@
 ::: {.column width="30%"}
-::: {.fragment}
+
 **ACC derived features**^[@skotte_detection_2014]
@@
 -   Inclination
-:::
+
 :::
@@
 ::: {.column width="40%"}
-::: {.fragment}
+
 **Sensor-Independent Features**^[@walch_sleep_2019, @forger_simpler_1999]
@@
 -   Clock Proxy Cosinus
-:::
+
 :::
@@
 
-:::aside
-:::
 ---
@@
 
-plot_proxy
+plot_proxy_cos
+```
+
+## Circadian Proxy Features {.smaller transition="convex"}
+
+```{r}
+source(here::here("presentation/code/plot_sensor-independent.R"))
+
+plot_proxy_lin
 ```

### 2023-01-28 code/prepare_data/zm_join_sceens_acc.R

@@ -5,7 +5,7 @@ library(lubridate)
 library(hms)
 library(arrow)
 
-# path <- "data/processed/screens_baseline.parquet"
+path <- "data/processed/screens_baseline.parquet"
 
 args <- R.utils::commandArgs(trailingOnly = TRUE)
 

### 2023-01-28 presentation/code/network_tree_plots.R

@@ -1,146 +1,9 @@
 library(tidyverse)
-library(igraph)
-library(showtext)
-
-in_bed_tree_data <-
-  tibble(
-    from = c(
-      "clock_proxy_cos < 31e-18",
-      "clock_proxy_cos < 31e-18",
-      "sdmax >= 0.0031",
-      "sdmax >= 0.0031",
-      "clock_proxy_cos < 0.27",
-      "clock_proxy_cos < 0.27",
-      "temp < 33",
-      "temp < 33",
-      "incl >= 119",
-      "incl >= 119",
-      "sdmax >= 0.0046",
-      "sdmax >= 0.0046",
-      "macc_y >= 0.64",
-      "macc_y >= 0.64"
-    ),
-    to = c(
-      "sdmax >= 0.0031",
-      "clock_proxy_cos < 0.27",
-      "out-bed 53%",
-      "temp < 33",
-      "sdmax >= 0.0046",
-      "in-bed 3%",
-      "out-bed 1%",
-      "incl >= 119",
-      "out-bed 5%",
-      "in-bed ~0%",
-      "macc_y >= 0.64",
-      "in-bed 1%",
-      "out-bed 5%",
-      "in-bed 32%"
-    )
-  )
-
-g <-
-  graph_from_data_frame(in_bed_tree_data, directed = TRUE)
-
-coords <-
-  layout_as_tree(g) |>
-  as_tibble(.name_repair = ~ c("x", "y"))
-
-output_df <-
-  coords |>
-  mutate(
-    step = vertex_attr(g, "name"),
-    label = step,
-    x = x * -1
-  )
-
-plot_nodes <-
-  output_df %>%
-  mutate(
-    xmin = x - 0.35,
-    xmax = x + 0.35,
-    ymin = y - 0.15,
-    ymax = y + 0.15
-  )
-
-plot_edges <-
-  in_bed_tree_data %>%
-  mutate(id = row_number()) %>%
-  pivot_longer(
-    cols = c("from", "to"),
-    names_to = "s_e",
-    values_to = "step"
-  ) %>%
-  left_join(plot_nodes, by = "step") %>%
-  select(-c(label, y, xmin, xmax)) %>%
-  mutate(y = ifelse(s_e == "from", ymin, ymax)) %>%
-  select(-c(ymin, ymax))
-
-font_add_google(name = "IBM Plex Serif", family = "ibm")
-showtext_auto()
-
-# p <-
-plot_nodes |>
-  ggplot(
-    aes(
[…]

### 2023-02-06 presentation/index.qmd
create hyperparameter tuning plot, ROC curve plots, and correct combined performance metrics

@@
 ---
-title: "Machine Learned Models for Estimation of Sleep Quality in Free-Living Accelerometer Data"
+title: "Sleep Quality Estimation using Machine Learning and Thigh-Mounted Accelerometer Data in Free Living Conditions"
 subtitle: "(working title)"
@@
 library(tidymodels)
+library(tidymodels)
 ```
 
-Lists can optionally be displayed incrementally:
+- Sleep plays a vital role in health, thus, improving the assessment of sleep–wake outside of a laboratory environment is critical
+- The gold standard (PSG) is costly and inconvenient.
+- Methods for estimating sleep/wake based on accelerometry exist, primarily from wrist-worn devices
+- Cole-Kripke and Sadeh algorithms are commonly used
+- determine in-bed time is difficult, usually set by sleep log and/or human scorers
+- detect wakefulness is difficult, worse performance in populations with sleep disorders
+- typically a two level analysis: epoch based and summarized across night(s)
+- Zmachine-derived sleep stats
+-
 
-::: incremental
--   First item
--   Second item
--   Third item
-:::
+Purpose...
 
-. . .
+## But Esben, what about them sleep stages!?
 
-Insert pauses to make other types of content display incrementally.
+* I did free-living PSG recordings of sleep but...
+  - Super fragile -> shitty data
+  - Combersome and time consuming
+  - free-living?
+  - would surface skin temperature + acc be enough? Most likely needs HR
 
-::: footer
-Learn more: [Incremental Lists](https://quarto.org/docs/presentations/revealjs/#incremental-lists)
-:::
+It was likely a dead end from the get-go :(
 
@@
 
-Overview
-Network plot of data preparation steps
+data preparation, big time-consumer is handling raw acc data
 
-. . .
+only thigh data used. HSBC and other is only thigh data...
 
-network plot
+could be interesting to build models on thigh and hip ocmbined.
+
+all zm recording is considered as in-bed (sensor problem?)
+
+no sleep stages, only sleep/awake
+
+sensor problems during sleep, up to 20 consecutive epochs (200 sec) are treated as sleep
 
@@
 
-::: {.column width="30%"}
+::: {.fragment .column width="30%"}
 
@@
 
-::: {.column width="40%"}
+::: {.fragment .column width="40%"}
 
@@
 
-## {.smaller transition="convex" .column}
+## building Models {.smaller transition="convex"}
 
 ```{r}
-source("~/sleep_study/presentation/code/tables_performance.R")
+source(here::here("presentation/code/plot_hyperparameter_tuning.R"))
+
+hyper_plot
+```
+
+## Results {transition="fade"}
+
+:::: {.columns}
+
+::: {.fragment .column width="50%"}
+
+* Performance Metrics
+  - F1 Score
+  - Accuracy
+  - Sensitivity
+  - Specificity
+  - ROC curves
+
+:::
+
+::: {.fragment .column width="50%"}
+
+* Agreement With Zmachine Sleep Stats
+  - Sleep Period Time
+  - Total Sleep Time
+  - Sleep Efficiency
+  - Latency Until Persistent Sleep
+  - Wake After Sleep Onset
+
+:::
+
+::::
+
+## ROC Curves {.smaller transition="fade" .column-left}
+
+```{r}
+source(here::here("presentation/code/plot_roc_curves.R"))
+
+roc_plots
+```
+
+## {.smaller transition="fade" .column-left}
+
+```{r}
+source(here::here("presentation/code/tables_performance.R"))
 
@@
 
-## {.smaller transition="convex" .column}
+## {.smaller transition="fade" .column}
 
@@
 Performance of the models to predict each <span style="color:gold">**combined**</span> class, i.e., "sleep" + "in-bed".
-
----
 
@@
 in_bed_tree <-
-  read_rds("~/sleep_study/data/models/fitted_models/in_bed_simple_CART_fit.rds")
+  read_rds(here::here("data/models/fitted_models/in_bed_simple_CART_fit.rds"))
 
@@
   )) |>
-  visInteraction(dragNodes = FALSE, 
+  visInteraction(
+    dragNodes = FALSE,
     dragView = FALSE,
-                 zoomView = FALSE)
+    zoomView = FALSE
+  )
 ```
@@
 sleep_tree <-
-  read_rds("~/sleep_study/data/models/fitted_models/sleep_simple_CART_fit.rds")
+  read_rds(here::here("data/models/fitted_models/sleep_simple_CART_fit.rds"))
 
@@
   )) |>
-  visInteraction(dragNodes = FALSE, 
+  visInteraction(
+    dragNodes = FALSE,
     dragView = FALSE,
-                 zoomView = FALSE)
+    zoomView = FALSE
+  )
 ```
+
+## Discussion {.smaller transition="zoom"}
+
+- heteroscedasticity
+- Cheung 2018 table 4: actigraphy provides a sufficiently narrow range of possible mean differences (CI 95%) clinical significant thresholds
+
+## References {transition="convex"}

### 2023-02-06 code/evaluate_models/performance_all_models.R

@@ -18,21 +18,19 @@ in_bed_fits <- map(in_bed_fits_files, read_rds) |>
 sleep_fits <- map(sleep_fits_files, read_rds) |>
   set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
 
-test <- read_parquet("data/processed/testing_data.parquet") |>
-  group_by(id) |>
-  slice_sample(n = 100)
+test <- read_parquet("data/processed/testing_data.parquet") 
 
 # Metrics and plots -------------------------------------------------------
 
 
 my_metrics <-
-  metric_set(f_meas, accuracy, sensitivity, specificity)
+  metric_set(f_meas, accuracy, sensitivity, precision, specificity)
 
 get_metrics <-
   function(fit, truth, estimate) {
     fit |>
       augment(test) |>
-      my_metrics(truth = {{ truth }}, estimate = {{ estimate }})
+      my_metrics(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
   }
 
 in_bed_metrics <-
@@ -74,7 +72,7 @@ get_scores <-
         zm_in_bed_awake = as_factor(if_else(in_bed == 1 & sleep == 0, 1, 0)),
         zm_in_bed_sleep = as_factor(if_else(in_bed == 1 & sleep == 1, 1, 0))
       ) |>
-      my_metrics(truth = {{ truth }}, estimate = {{ estimate }})
+      my_metrics(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
   }
 
 
@@ -85,7 +83,7 @@ map2_dfr(in_bed_fits, sleep_fits,
   select(-.estimator) |>
   pivot_wider(names_from = model, values_from = .estimate) |>
   mutate(
-    event = "In-Bed Awake"
+    event = "In-Bed Awake Prediction"
   ) |>
   bind_rows(
     map2_dfr(in_bed_fits, sleep_fits,
@@ -95,7 +93,7 @@ map2_dfr(in_bed_fits, sleep_fits,
       select(-.estimator) |>
       pivot_wider(names_from = model, values_from = .estimate) |>
       mutate(
-        event = "In-Bed Sleep"
+        event = "In-Bed Sleep Prediction"
       )
   ) |> 
   write_csv("data/processed/combined_preds_performance_metrics.csv")

### 2023-02-06 code/evaluate_models/plot_roc_curves.R

@@ -1,0 +1,63 @@
+#!/usr/bin/env
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+
+in_bed_fits_files <- list.files("data/models/fitted_models", full.names = TRUE) |>
+  str_subset("mars", negate = TRUE) |>
+  str_subset("in_bed")
+
+sleep_fits_files <- list.files("data/models/fitted_models", full.names = TRUE) |>
+  str_subset("mars", negate = TRUE) |>
+  str_subset("sleep")
+
+in_bed_fits <- map(in_bed_fits_files, read_rds) |>
+  set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
+
+sleep_fits <- map(sleep_fits_files, read_rds) |>
+  set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
+
+test <- read_parquet("data/processed/testing_data.parquet") 
+
+
+
+# Get ROC -----------------------------------------------------------------
+
+
+get_roc <-
+  function(fit, truth, estimate) {
+    fit |>
+      augment(test) |>
+      roc_curve(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
+  }
+
+map_dfr(in_bed_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |> 
+    write_parquet("data/processed/roc_data_in_bed.parquet")
+
+map_dfr(sleep_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |> 
+    write_parquet("data/processed/roc_data_sleep.parquet")
+
+
+# ROC curve plots ---------------------------------------------------------
+
+
+in_bed_roc |>
+  ggplot(
+    aes(1 - specificity, sensitivity, color = model)
+  ) + # plot with 2 ROC curves for each model
+  geom_line(linewidth = .8, alpha = .7) +
+  geom_abline(slope = 1, intercept = 0, linewidth = 0.4, lty = 2) +
+  scale_color_brewer(palette = "Dark2") +
+  coord_fixed() +
+  theme_light()
+
+sleep_roc |>
+  ggplot(
+    aes(1 - specificity, sensitivity, color = model)
+  ) + # plot with 2 ROC curves for each model
+  geom_line(linewidth = .8, alpha = .7) +
+  geom_abline(slope = 1, intercept = 0, linewidth = 0.4, lty = 2) +
+  scale_color_brewer(palette = "Dark2") +
+  coord_fixed() +
+  theme_light()

### 2023-02-06 code/evaluate_models/table_performance_metrics.R

@@ -5,8 +5,8 @@ library(gt)
 library(showtext)
 
 
-metric_data <- read_csv("data/processed/performance_metrics.csv")
-comb_metric_data <- read_csv("data/processed/combined_preds_performance_metrics.csv")
+separate_metrics <- read_csv("data/processed/performance_metrics.csv")
+combined_metrics <- read_csv("data/processed/combined_preds_performance_metrics.csv")
 
 # Create table ------------------------------------------------------------
 
@@ -14,38 +14,47 @@ comb_metric_data <- read_csv("data/processed/combined_preds_performance_metrics.
 font_add_google("IBM Plex Serif", family = "ibm")
 showtext_auto()
 
-tab_performance <-
-  comb_metric_data |> 
-  mutate(
-    .metric = factor(.metric, 
-                     levels = c("f_meas", "accuracy", "sensitivity", "specificity"),
-                     labels = c("F1 Score", "Accuracy", "Sensitivity", "Specificity"))
-  ) |> 
-  group_by(event) |>
-  gt(rowname_col = ".metric") |>
-  fmt_percent(
-    columns = c(logistic_regression:xgboost),
-    decimals = 2
-  ) |>
-  cols_label(
-    logistic_regression = "Logistic Regression",
-    neural_net = "Neural Network",
-    decision_tree = "Decision Tree",
-    xgboost = "XGboost"
-  ) |>
-  tab_style(
-    style = cell_text(size = px(12)),
-    locations = cells_body(
-      columns = c(logistic_regression:xgboost)
-    )
-  ) |>
-  tab_header(
-    title = md("Performance Metrics"),
-    subtitle = "Grouped by Event Prediction"
-  ) |>
-  tab_options(table.font.names = "ibm") |>
-  cols_width(everything() ~ px(100)) |> 
-  cols_align(align = "center") |> 
-  data_color(columns = decision_tree, colors = "lightblue", alpha = .6)
+tab_fun <- function(metric_data) {
+  metric_data |>
+    mutate(
+      .metric = factor(.metric,
+        levels = c("f_meas", "accuracy", "sensitivity", "precision", "specificity"),
+        labels = c("F1 Score", "Accuracy", "Sensitivity", "Precision", "Specificity")
+      )
+    ) |>
+    group_by(event) |>
+    gt(rowname_col = ".metric") |>
+    fmt_percent(
+      columns = c(logistic_regression:xgboost),
+      decimals = 2
+    ) |>
+    cols_label(
[…]

### 2023-02-06 presentation/code/plot_hyperparameter_tuning.R

@@ -1,0 +1,43 @@
+library(tidyverse)
+library(tidymodels)
+
+data <- 
+  read_rds(here::here("presentation/data/sleep_workflowsets_results.rds"))
+
+hyper_plot <- 
+  data |> 
+  autoplot(
+    rank_metric = "f_meas",  # <- how to order models
+    metric = "f_meas",       # <- which metric to visualize
+    select_best = FALSE     # <- one point per workflow
+  ) +
+  # geom_text(aes(y = mean - .003, label = wflow_id), angle = 90, hjust = 1) +
+  lims(y = c(.96, 1)) +
+  labs(
+    title = "Hyperparameter Tuning",
+    subtitle = "10-fold cross validation using 10-level grid search",
+    color = "Model Type",
+    shape = NULL,
+    y = "F1 Metric",
+    caption = "This figure contains only data from 1 person and, hence, is only for illustrative purposes."
+  ) +
+  scale_color_brewer(palette = "Dark2",
+                     label = c("XGboost", "Logistic Regression", "MARS", "Neural Net", "Random Forest")) +
+  theme(
+    text = element_text(color = "#EEE8D5", family = "ibm"),
+    axis.title = element_text(family = "ibm", size = 20, color = "#EEE8D5"),
+    axis.text = element_text(size = 20, color = "#EEE8D5"),
+    plot.title = element_text(family = "ibm", size = 30, face = "bold", hjust = .5),
+    plot.subtitle = element_text(size = 20, hjust = .5),
+    plot.caption = element_text(size = 14),
+    panel.grid.minor = element_blank(),
+    panel.grid = element_line(color = "#293D42"),
+    plot.background = element_rect(fill = "#002B36", color = "#EEE8D5"),
+    panel.background = element_rect(color = "#002B36", fill = "#002B36"),
+    legend.background = element_rect(fill = "#002B36"),
+    legend.text = element_text(size = 16),
+    legend.title = element_text(size = 18, face = "bold"),
+    legend.key = element_rect(fill = "#002B36"),
+    legend.position = "bottom"
+  ) +
+  guides(shape = "none")

### 2023-02-06 presentation/code/tables_performance.R

@@ -11,13 +11,14 @@ metric_data <- read_csv("~/sleep_study/data/processed/performance_metrics.csv")
 font_add_google("IBM Plex Serif", family = "ibm")
 showtext_auto()
 
-tab_1 <- 
-  metric_data |> 
+tab_1 <-
+  metric_data |>
   mutate(
-    .metric = factor(.metric, 
-                     levels = c("f_meas", "accuracy", "sensitivity", "specificity"),
-                     labels = c("F1 Score", "Accuracy", "Sensitivity", "Specificity"))
-  ) |> 
+    .metric = factor(.metric,
+      levels = c("f_meas", "accuracy", "sensitivity", "precision", "specificity"),
+      labels = c("F1 Score", "Accuracy", "Sensitivity", "Precision", "Specificity")
+    )
+  ) |>
   group_by(event) |>
   gt(rowname_col = ".metric") |>
   fmt_percent(
@@ -40,23 +41,25 @@ tab_1 <-
     title = md("Performance Metrics"),
     subtitle = "Grouped by Event Prediction"
   ) |>
-  cols_width(everything() ~ px(100)) |> 
-  cols_align(align = "center") |> 
+  cols_width(everything() ~ px(100)) |>
+  cols_align(align = "center") |>
   data_color(columns = decision_tree, colors = "#4A666C", alpha = .6) |>
-  tab_options(table.font.names = "ibm",
-              table.font.color.light = "#EEE8D5",
-              table.background.color = "#002B36",
-  ) 
+  tab_options(
+    table.font.names = "ibm",
+    table.font.color.light = "#EEE8D5",
+    table.background.color = "#002B36",
+  )
 
 comb_metric_data <- read_csv("~/sleep_study/data/processed/combined_preds_performance_metrics.csv")
 
-tab_2 <- 
-  comb_metric_data |> 
+tab_2 <-
+  comb_metric_data |>
   mutate(
-    .metric = factor(.metric, 
-                     levels = c("f_meas", "accuracy", "sensitivity", "specificity"),
-                     labels = c("F1 Score", "Accuracy", "Sensitivity", "Specificity"))
-  ) |> 
+    .metric = factor(.metric,
+      levels = c("f_meas", "accuracy", "sensitivity", "precision", "specificity"),
+      labels = c("F1 Score", "Accuracy", "Sensitivity", "Precision", "Specificity")
+    )
+  ) |>
   group_by(event) |>
[…]

### 2023-02-17 presentation/index.qmd
assumptions for BA not met, see QQ-plots. Perform repeated measures, nonparametric BA.

@@
 ---
-title: "Sleep Quality Estimation using Machine Learning and Thigh-Mounted Accelerometer Data in Free Living Conditions"
+title: "Sleep Quality Estimation using Thigh-Mounted Accelerometer Data in Free Living Conditions"
 subtitle: "(working title)"
@@
     css: my_styles.css
+    scrollable: true
 embed-resources: true
@@
   - Combersome and time consuming
-  - free-living?
+  - free-living when wired up like a robot?
   - would surface skin temperature + acc be enough? Most likely needs HR
@@
 
-data preparation, big time-consumer is handling raw acc data
+- data preparation, big time-consumer is handling raw acc data
+- only thigh data used. HSBC and other is only thigh data...
+- all zm recording is considered as in-bed (sensor problem?)
+- no sleep stages, only sleep/awake
+- sensor problems during sleep, up to 20 consecutive epochs (200 sec) are treated as sleep
 
-only thigh data used. HSBC and other is only thigh data...
-
-could be interesting to build models on thigh and hip ocmbined.
-
-all zm recording is considered as in-bed (sensor problem?)
-
-no sleep stages, only sleep/awake
-
-sensor problems during sleep, up to 20 consecutive epochs (200 sec) are treated as sleep
-
-## Exclusion Criteria {transition="fade"}
+## Exclusion Criteria {transition="fade" .H2}
 
@@
 ::: {.fragment}
-But as demonstrated by @walch_sleep_2019, a simple cosine function does the tricks just as well :)
+But as demonstrated by @walch_sleep_2019, a simple cosine function does the trick just as well :)
 :::
@@
 
-## ROC Curves {.smaller transition="fade" .column-left}
+## ROC Curves {.smaller transition="fade"}
 
@@
 
-## {.smaller transition="fade" .column-left}
+## Lots of Metrics {.smaller transition="zoom" .H2}
 
+:::: {.columns}
+
+::: {.column}
 ```{r}
@@
 
-tab_1
+tab1
 ```
+:::
 
-Performance of the models to predict each class <span style="color:gold">**seperately**</span>, i.e., "sleep" and "in-bed".
+::: {.column}
+```{r}
+tab2
+```
+:::
 
-## {.smaller transition="fade" .column}
+::::
 
-```{r}
-tab_2
-```
-
-Performance of the models to predict each <span style="color:gold">**combined**</span> class, i.e., "sleep" + "in-bed".
-
-## Bland-Altman Plots {.smaller transition="convex" .slide}
+## Bland-Altman Plots {.smaller transition="fade"}
 
@@
 #| echo: false
-source(here::here("presentation/code/ba_plots.R"))
+source(here::here("presentation/code/plot_ba.R"))
 plots$decision_tree[[1]]
@@
 
----
+## Bland-Altman Analysis {.smaller transition="fade" .H2}
 
-## In-bed classification flow {.smaller transition="zoom" .column}
+```{r}
+source(here::here("presentation/code/table_ba.R"))
+
+tab_ba
+```
+
+## In-bed classification flow {.smaller transition="zoom" .H2}
 
@@
 
-## Sleep classification flow {.smaller transition="zoom" .column}
+## Sleep classification flow {.smaller transition="zoom" .H2}
 
@@
 - Cheung 2018 table 4: actigraphy provides a sufficiently narrow range of possible mean differences (CI 95%) clinical significant thresholds
+- could be interesting to build models on thigh and hip ocmbined.
 

### 2023-02-17 code/EDA/plot_qqs.R

@@ -1,0 +1,58 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(arrow)
+
+stats_files <-
+  list.files("~/sleep_study/data/processed/stats_predictions", full.names = TRUE)
+
+all_stats <-
+  map(stats_files, read_parquet) |>
+  set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
+
+
+get_diff_stats <-
+  function(tbl) {
+    tbl |>
+      mutate(
+        diff_spt_hrs = spt_hrs - zm_spt_hrs,
+        diff_tst_hrs = tst_hrs - zm_tst_hrs,
+        diff_se_percent = se_percent - zm_se_percent,
+        diff_lps_min = lps_min - zm_lps_min,
+        diff_waso_min = waso_min - zm_waso_min
+      ) |>
+      rowwise() |>
+      mutate(
+        avg_spt_hrs = mean(c(spt_hrs, zm_spt_hrs)),
+        avg_tst_hrs = mean(c(tst_hrs, zm_tst_hrs)),
+        avg_se_percent = mean(c(se_percent, zm_se_percent)),
+        avg_lps_min = mean(c(lps_min, zm_lps_min)),
+        avg_waso_min = mean(c(waso_min, zm_waso_min))
+      ) |>
+      ungroup()
+  }
+
+
+all_diffs <-
+  all_stats |>
+  map(get_diff_stats)
+
+
+plot_qqs <- function(df) {
+  df |>
+    select(id, noon_day, contains("diff")) |>
+    pivot_longer(-c(id:noon_day)) |>
+    ggplot(aes(sample = value)) +
+    geom_qq(shape = 21, fill = "darkorange", alpha = .5) +
+    geom_qq_line() +
+    facet_wrap(~name, scales = "free") +
+    theme_light()
+}
+
+qq_plots <-
+  map(all_diffs, plot_qqs)
+
+filenames <- 
+  paste0("visuals/qqplots_", names(qq_plots), ".png")
+
+walk2(qq_plots, filenames, ~ ggsave(plot = .x, filename = .y))

### 2023-02-17 code/evaluate_models/mixed_effect_LOA.R

@@ -1,0 +1,79 @@
+library(tidyverse)
+library(arrow)
+library(patchwork)
+library(showtext)
+
+
+# TODO consider calculating 95% CI for bias
+
+
+stats_files <-
+  list.files("~/sleep_study/data/processed/stats_predictions", full.names = TRUE)
+
+all_stats <-
+  map(stats_files, read_parquet) |>
+  set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
+# |>
+#   map(
+#     ~ filter(.x, (id != 1742705 | noon_day != 24) &
+#         (id != 1044005 | noon_day != 16) &
+#         (id != 1750904 | noon_day != 21) &
+#         (id != 1718904 | noon_day != 8)
+#     )
+#   ) # TODO look into these lps_min outliers
+
+get_diff_stats <-
+  function(tbl) {
+    tbl |>
+      mutate(
+        diff_spt_hrs = spt_hrs - zm_spt_hrs,
+        diff_tst_hrs = tst_hrs - zm_tst_hrs,
+        diff_se_percent = se_percent - zm_se_percent,
+        diff_lps_min = lps_min - zm_lps_min,
+        diff_waso_min = waso_min - zm_waso_min
+      ) |>
+      rowwise() |>
+      mutate(
+        avg_spt_hrs = mean(c(spt_hrs, zm_spt_hrs)),
+        avg_tst_hrs = mean(c(tst_hrs, zm_tst_hrs)),
+        avg_se_percent = mean(c(se_percent, zm_se_percent)),
+        avg_lps_min = mean(c(lps_min, zm_lps_min)),
+        avg_waso_min = mean(c(waso_min, zm_waso_min))
+      ) |>
+      ungroup()
+  }
+
+
+all_diffs <-
+  all_stats |>
+  map(get_diff_stats)
+
+test <- all_diffs$logistic_regression
+
+plot_qqs <- funtion(df){
+df |> 
+  select(id, noon_day, contains("diff")) |> 
+  pivot_longer(-c(id:noon_day)) |> 
+  ggplot(aes(sample = value)) +
+  geom_qq(shape = 21, fill = "darkorange", alpha = .5) + 
+  geom_qq_line() +
+  facet_wrap(~ name, scales = "free") +
+  theme_light()
+}
+
+# Differences are not normal. Use nonparametric approach for LOA. 
+
+# test |> 
+#   loa_lme(data = _,
+#           diff = "diff_spt_hrs",
[…]

### 2023-02-17 presentation/code/plot_sensor-independent.R

@@ -151,7 +151,7 @@ plot_proxy_lin <-
     geom = "text",
     x = as_datetime(static_end) + hours(1),
     y = .5,
-    label = hms::as_hms(static_start - seconds(10)),
+    label = hms::as_hms(static_end + seconds(10)),
     color = "#EEE8D5",
     angle = 90,
     size = 8

### 2023-02-17 presentation/code/tables_performance.R

@@ -1,6 +1,7 @@
 library(tidyverse)
 library(gt)
 library(showtext)
+library(ggtext)
 
 metric_data <- read_csv("~/sleep_study/data/processed/performance_metrics.csv")
 
@@ -11,7 +12,7 @@ metric_data <- read_csv("~/sleep_study/data/processed/performance_metrics.csv")
 font_add_google("IBM Plex Serif", family = "ibm")
 showtext_auto()
 
-tab_1 <-
+tab1 <-
   metric_data |>
   mutate(
     .metric = factor(.metric,
@@ -37,22 +38,52 @@ tab_1 <-
       columns = c(logistic_regression:xgboost)
     )
   ) |>
-  tab_header(
-    title = md("Performance Metrics"),
-    subtitle = "Grouped by Event Prediction"
+  tab_style(
+    style = cell_fill("#2C4E57"),
+    locations = cells_body(rows = c(1, 3, 5, 6, 8, 10))
   ) |>
+  tab_header(
+    title = "",
+    subtitle = md("Performance of the models to predict each class
+                             <span style='color:gold'>**seperately**</span>")
+  ) |> 
   cols_width(everything() ~ px(100)) |>
   cols_align(align = "center") |>
-  data_color(columns = decision_tree, colors = "#4A666C", alpha = .6) |>
+  # data_color(columns = decision_tree, colors = "#4A666C", alpha = .6) |>
   tab_options(
     table.font.names = "ibm",
     table.font.color.light = "#EEE8D5",
     table.background.color = "#002B36",
+    heading.border.bottom.style = "none",
+    table.border.top.width = px(3),
+    table.border.top.style = "none", # transparent
+    table.border.bottom.style = "none",
+    source_notes.border.bottom.color = "#002B36",
+    column_labels.font.weight = "normal",
+    column_labels.border.top.style = "none",
+    column_labels.border.bottom.width = px(2),
+    column_labels.border.bottom.color = "#2C4E57",
+    column_labels.padding = px(1),
+    row_group.border.top.style = "none",
+    row_group.border.top.color = "#2C4E57",
+    row_group.border.bottom.width = px(0),
+    row_group.border.bottom.color = "#2C4E57",
+    stub.border.color = "#2C4E57",
+    stub.border.width = px(0),
+    stub.font.size = 14,
+    table_body.border.bottom.color = "#002B36",
+    table_body.hlines.color = "#2C4E57",
+    # table_body.hlines.style = "none",
[…]

### 2023-02-20 code/evaluate_models/mixed_effect_LOA.R
Update README.md

@@ -38,17 +38,20 @@ all_diffs <-
   map(get_diff_stats)
 
 
-# Differences are not normal. Use nonparametric approach for LOA.
-# But is this correct for repeated measures? I think not.
+# This method is nonparametric and accounts for repeated measures. Double check with Jan!
 
-get_agree <- function(df = all_diffs, pred, ref, delta, type) {
-  map(df, ~ agree_nest(
-    x = pred,
-    y = ref,
+# Mixed Effects Limits of Agreement
+# This function allows for the calculation of bootstrapped limits of agreement 
+# when there are multiple observations per subject
+
+get_agree <- function(df = all_diffs, diff, condition = "noon_day", delta) {
+  map(df, ~ loa_mixed(
+    diff = diff,
+    condition = condition,
     id = "id",
     data = .,
     delta = delta,
-    prop_bias = FALSE
+    replicates = 1000
   )) |>
     map_dfr("loa", .id = "model") |>
     rownames_to_column(var = "ba_metric") |>
@@ -68,7 +71,7 @@ tst <- get_agree(all_diffs, "tst_hrs", "zm_tst_hrs", 2)
 se_percent <- get_agree(all_diffs, "se_percent", "zm_se_percent", 20)
 
 # lps_min
-lps <- get_agree(all_diffs, "se_percent", "zm_se_percent", 20)
+lps <- get_agree(all_diffs, "lps_min", "zm_lps_min", 20)
 
 # waso
 waso <- get_agree(all_diffs, "waso_min", "zm_waso_min", 10)
@@ -76,14 +79,4 @@ waso <- get_agree(all_diffs, "waso_min", "zm_waso_min", 10)
 ba_metrics <-
   as_tibble(bind_rows(spt, tst, se_percent, lps, waso))
 
-
-# This method is nonparametric and accounts for repeated measures. Double check with Jan!
-
-# Mixed Effects Limits of Agreement
-# This function allows for the calculation of bootstrapped limits of agreement when there are multiple
-# observations per subject
-
-# lps_mixed <- 
-  loa_mixed(diff = "diff_lps_min", condition = "noon_day", id = "id", data = test,
-            delta = 20,
-            replicates = 100)
+beepr::beep(3)

### 2023-02-23 presentation/index.qmd
Update README.md
95% CI and mixed effects LOA+bias on plots

@@
 ---
-title: "Sleep Quality Estimation using Thigh-Mounted Accelerometer Data in Free Living Conditions"
+title: "Sleep Quality Estimation using Raw Data from Thigh-Mounted Accelerometers during Free Living Conditions"
 subtitle: "(working title)"
@@
 
-## Background {.scrollable transition="convex"}
+## Background {transition="convex"}
 
@@
 
-## But Esben, what about them sleep stages!?
+## But Esben, what about them sleep stages!? {transition="convex"}
 
@@
 
-## Exclusion Criteria {transition="fade" .H2}
+## Exclusion Criteria {transition="convex" .H2}
 
@@
 
-## Features {transition="fade" .smaller}
+## Features {transition="convex" .H2}
 
@@
 
-## Human Circadian Clock {.smaller transition="convex"}
+## Human Circadian Clock {transition="convex"}
 
@@
 
-## Circadian Proxy Features {.smaller transition="convex"}
+## Circadian Proxy Features {transition="convex"}
 
@@
 
-## building Models {.smaller transition="convex"}
+## building Models {transition="convex"}
 
@@
 
-## Results {transition="fade"}
+## Results {transition="convex" .H2}
 
@@
 ::: {.fragment .column width="50%"}
+
+### Epoch-Based
 
@@
 
+### Summarized across nights {.H3}
+
 * Agreement With Zmachine Sleep Stats 
@@
 
-## ROC Curves {.smaller transition="fade"}
+## ROC Curves {transition="convex"}
 
@@
 
-## Lots of Metrics {.smaller transition="zoom" .H2}
+## Lots of Metrics {transition="convex" .H2}
 
@@
 
-## Bland-Altman Plots {.smaller transition="fade"}
+## Bland-Altman Plots {transition="convex"}
 
@@
 source(here::here("presentation/code/plot_ba.R"))
-plots$decision_tree[[1]]
 
-plots$decision_tree[[2]]
+dc_plots[[1]]
 
-plots$decision_tree[[3]]
+dc_plots[[2]]
 
-plots$decision_tree[[4]]
+dc_plots[[3]]
 
-plots$decision_tree[[5]]
+dc_plots[[4]]
+
+dc_plots[[5]]
 ```
@@
 ```{r}
-plots$logistic_regression[[1]]
+lr_plots[[1]]
 
-plots$logistic_regression[[2]]
+lr_plots[[2]]
 
-plots$logistic_regression[[3]]
+lr_plots[[3]]
 
-plots$logistic_regression[[4]]
+lr_plots[[4]]
 
-plots$logistic_regression[[5]]
+lr_plots[[5]]
+```
+
+### Neural Net
+
+```{r}
+nn_plots[[1]]
+
+nn_plots[[2]]
+
+nn_plots[[3]]
+
+nn_plots[[4]]
+
+nn_plots[[5]]
 ```
@@
 ```{r}
-plots$xgboost[[1]]
+xg_plots[[1]]
 
-plots$xgboost[[2]]
+xg_plots[[2]]
 
-plots$xgboost[[3]]
+xg_plots[[3]]
 
-plots$xgboost[[4]]
+xg_plots[[4]]
 
-plots$xgboost[[5]]
-```
-
-### Neural Net
-
-```{r}
-plots$neural_net[[1]]
-
-plots$neural_net[[2]]
-
-plots$neural_net[[3]]
-
-plots$neural_net[[4]]
-
-plots$neural_net[[5]]
+xg_plots[[5]]
 ```
@@
 
-## Bland-Altman Analysis {.smaller transition="fade" .H2}
+## Bland-Altman Analysis {transition="convex" .H2}
 
@@
 
-## In-bed classification flow {.smaller transition="zoom" .H2}
+## In-bed classification flow {transition="convex" .H2}
 
@@
 
-## Sleep classification flow {.smaller transition="zoom" .H2}
+## Sleep classification flow {transition="convex" .H2}
 
@@
 
-## Discussion {.smaller transition="zoom"}
+## Discussion {transition="convex"}
 
@@
 - could be interesting to build models on thigh and hip ocmbined.
+- multiclass vs multilabel classification
 

### 2023-02-23 presentation/code/plot_ba.R

@@ -4,8 +4,8 @@ library(patchwork)
 library(showtext)
 
 
-# TODO consider calculating 95% CI for bias
-
+ba_metrics <-
+  read_csv(here::here("data/processed/mixed_effect_ba.csv"))
 
 stats_files <-
   list.files("~/sleep_study/data/processed/stats_predictions", full.names = TRUE)
@@ -43,50 +43,77 @@ get_diff_stats <-
       ungroup()
   }
 
-
-get_summary_stats <-
-  function(tbl) {
-    tbl |>
-      # filter(se_percent > 0) |>
-      summarise(
-        across(contains("diff"), list(
-          upper = ~ mean(.x) + 1.96 * sd(.x),
-          lower = ~ mean(.x) - 1.96 * sd(.x),
-          mean = ~ mean(.x)
-        ))
-      )
-  }
-
-all_diffs <-
-  all_stats |>
-  map(get_diff_stats)
-
-all_summaries <-
+diff_avg <-
   all_stats |>
   map(get_diff_stats) |>
-  map(get_summary_stats)
-# TODO calculate nonparametric metrics for BA using 'SimplyAgree'
-# source("code/evaluate_models/mixed_effect_LOA.R")
+  map(~ .x |> select(id, contains(c("diff", "avg")))) |>
+  map(~ .x |> pivot_longer(contains("diff"), names_to = "diffs", values_to = "diff_values")) |>
+  map(~ .x |> pivot_longer(contains("avg"), names_to = "avgs", values_to = "avg_values")) |>
+  map(~ .x |> filter(
+    (str_detect(diffs, "spt") & str_detect(avgs, "spt")) |
+      (str_detect(diffs, "tst") & str_detect(avgs, "tst")) |
+      (str_detect(diffs, "se") & str_detect(avgs, "se")) |
+      (str_detect(diffs, "lps") & str_detect(avgs, "lps")) |
+      (str_detect(diffs, "waso") & str_detect(avgs, "waso"))
+  )) |>
+  map(~ .x |> group_split(diffs, avgs))
 
 
-# BA plots for simple trees -----------------------------------------------
+h_lines <-
+  ba_metrics |>
+  group_split(model, .keep = FALSE) |>
+  setNames(c("decision_tree", "logistic_regression", "neural_net", "xgboost")) |>
+  map(~ .x |>
+    select(ba_metric, estimate, type) |>
[…]

### 2023-02-23 presentation/code/table_ba.R

@@ -6,8 +6,8 @@ library(gtExtras)
 library(here)
 
 
-# TODO consider calculating 95% CI for bias
-
+ba_metrics <-
+  read_csv(here::here("data/processed/mixed_effect_ba.csv"))
 
 stats_files <-
   list.files(here("data/processed/stats_predictions"), full.names = TRUE)
@@ -63,11 +63,9 @@ all_diffs <-
   all_stats |>
   map(get_diff_stats)
 
-all_summaries <-
-  all_stats |>
-  map(get_diff_stats) |>
-  map(get_summary_stats) |>
-  bind_rows(.id = "model")
+
+ba_metrics |>
+  pivot_wider(names_from = c(model, ba_metric), values_from = c(estimate:upper_ci))
 
 
 # create table ------------------------------------------------------------
@@ -76,60 +74,53 @@ font_add_google("IBM Plex Serif", family = "ibm")
 showtext_auto()
 
 tab_ba <-
-all_summaries |>
-  pivot_longer(-model) |>
+ba_metrics |>
   mutate(
     ba_metric = case_when(
-      str_detect(name, "mean") ~ "Bias",
-      str_detect(name, "upper") ~ "LOA Upper",
-      str_detect(name, "lower") ~ "LOA Lower"
+      str_detect(ba_metric, "bias") ~ "Bias (95% CI)",
+      str_detect(ba_metric, "lower") ~ "Lower LOA (95% CI)",
+      str_detect(ba_metric, "upper") ~ "Upper LOA (95% CI)"
     ),
-    name = str_remove_all(name, "diff_|_upper|_lower|_mean"),
-    name = case_when(
-      name == "spt_hrs" ~ "Sleep Period Time",
-      name == "tst_hrs" ~ "Total Sleep Time",
-      name == "se_percent" ~ "Sleep Efficiency",
-      name == "lps_min" ~ "Latency Until Persistent Sleep",
-      name == "waso_min" ~ "Wake After Sleep Onset"
+    model = case_when(
+      str_detect(model, "logistic") ~ "Logistic Regression",
+      str_detect(model, "decision") ~ "Decision Tree",
+      str_detect(model, "neural") ~ "Neural Net",
+      TRUE ~ "XGboost"
     ),
-    name = fct_relevel(name, c("Sleep Period Time", "Total Sleep Time", 
-                               "Sleep Efficiency", "Latency Until Persistent Sleep",
-                               "Wake After Sleep Onset"))
[…]

### 2023-02-24 presentation/index.qmd
update README.md

@@
 ---
-title: "Sleep Quality Estimation using Raw Data from Thigh-Mounted Accelerometers during Free Living Conditions"
+title: "Sleep Quality Estimation using Accelerometer Data from Thigh-Mounted Devices During in Free Living Conditions"
 subtitle: "(working title)"

### 2023-02-24 presentation/code/plot_ba.R

@@ -88,13 +88,6 @@ font_add_google("IBM Plex Serif", family = "ibm")
 showtext_auto()
 
 
-titles <-
-  c(
-    "Sleep Period Time (hrs)", "Total Sleep Time (hrs)",
-    "Sleep Efficiency (%)", "Latency Until Persisten Sleep (min)",
-    "Wake After Sleep Onset (min)"
-  )
-
 plot_ba <- function(...) {
   ggplot(..1, aes(avg_values, diff_values)) +
     annotate(
@@ -170,6 +163,13 @@ plot_ba <- function(...) {
     ) 
 }
 
+titles <-
+  c(
+    "Sleep Period Time (hrs)", "Total Sleep Time (hrs)",
+    "Sleep Efficiency (%)", "Latency Until Persisten Sleep (min)",
+    "Wake After Sleep Onset (min)"
+  )
+
 
 create_titles <- function(x) {
   x %>%
@@ -186,8 +186,7 @@ dc_plots <-
   pmap(
     list(diff_avg$decision_tree, h_lines$decision_tree, ci_lines$decision_tree), plot_ba
   ) |> 
-  create_titles()
-
+  create_titles() 
 
 lr_plots <-
   pmap(

### 2023-02-28 presentation/index.qmd
binary relevance setup

@@
 
+::: {.fragment}
+
 - data preparation, big time-consumer is handling raw acc data
@@
 - sensor problems during sleep, up to 20 consecutive epochs (200 sec) are treated as sleep
+- Combine binary classifiers to produce multilabel outcomes (i.e., binary Relevance)
+- Evaluated combined classifiers using weighted micro-averaging.
+
+:::
 
@@
 
-## Circadian Proxy Features {transition="convex"}
+## Circadian Proxy Features {transition="fade"}
 
@@
 hyper_plot
+```
+
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
+source(here::here("presentation/code/plot_pred_example.R"))
+
+plot_ex
+```
+
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
+source(here::here("presentation/code/plot_pred_example.R"))
+
+plot_lps
+```
+
+::: aside 
+This methedology is called binary relevance.
+:::
+
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
+source(here::here("presentation/code/plot_pred_example.R"))
+
+plot_tst
 ```
@@
 roc_plots
+```
+
+## ROC Curves {transition="convex"}
+
+```{r}
+roc_combined_plots
 ```
@@
 - multiclass vs multilabel classification
+- in-bed awake/sleep is highly imbalanced -> maybe train a new classifier accounting for imbalanced data (SMOTE)
+- model combined preds instead?
 

### 2023-02-28 code/evaluate_models/create_roc_data.R

@@ -18,7 +18,7 @@ in_bed_fits <- map(in_bed_fits_files, read_rds) |>
 sleep_fits <- map(sleep_fits_files, read_rds) |>
   set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
 
-test <- read_parquet("data/processed/testing_data.parquet") 
+test <- read_parquet("data/processed/testing_data.parquet")
 
 
 
@@ -32,9 +32,108 @@ get_roc <-
       roc_curve(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
   }
 
-map_dfr(in_bed_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |> 
-    write_parquet("data/processed/roc_data_in_bed.parquet")
+map_dfr(in_bed_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
+  write_parquet("data/processed/roc_data_in_bed.parquet")
 
-map_dfr(sleep_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |> 
-    write_parquet("data/processed/roc_data_sleep.parquet")
+map_dfr(sleep_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
+  write_parquet("data/processed/roc_data_sleep.parquet")
+
+
+
+# multiclass ROC ----------------------------------------------------------
+
+
+# get_combined_roc <- function(fit_sleep, fit_inbed, truth, estimate) {
+#   fit_sleep |>
+#     predict(test, type = "prob") |>
+#     bind_cols(
+#       fit_inbed |>
+#         predict(test, type = "prob")
+#     ) |>
+#     rename(in_bed_0 = 1, in_bed_1 = 2, sleep_0 = 3, sleep_1 = 4) |>
+#     bind_cols(test |> select(in_bed, sleep)) |>
+#     transmute(
+#       in_bed_sleep_pred = ((sleep_1 * in_bed_1) - min(sleep_1 * in_bed_1)) / (max(sleep_1 * in_bed_1) - min(sleep_1 * in_bed_1)),
+#       in_bed_awake_pred = ((sleep_0 * in_bed_1) - min(sleep_0 * in_bed_1)) / (max(sleep_0 * in_bed_1) - min(sleep_0 * in_bed_1)),
[…]

### 2023-02-28 presentation/code/plot_pred_example.R

@@ -1,0 +1,203 @@
+library(tidyverse)
+library(lubridate)
+library(ggtext)
+
+font_add_google("Mukta", family = "mukta")
+font_add_google("IBM Plex Serif", family = "ibm")
+showtext_auto()
+
+plot_ex <-
+  tibble(
+    datetime = seq(
+      as_datetime("2020-01-15 12:00:00"),
+      as_datetime("2020-01-16 12:00:00"), seconds(10)
+    ),
+    in_bed = if_else(
+      datetime > as_datetime("2020-01-15 21:00:00") &
+        datetime < as_datetime("2020-01-16 07:00:00"), 1, 0
+    ),
+    sleep = if_else(
+      datetime > as_datetime("2020-01-15 21:30:00") &
+        datetime < as_datetime("2020-01-16 06:50:00"), 1, 0
+    )
+  ) |>
+  ggplot() +
+  geom_step(aes(datetime, in_bed, color = "1")) +
+  geom_step(aes(datetime, sleep - .05, color = "2")) +
+  scale_y_continuous(
+    breaks = c(0, 1),
+    labels = c("No", "Yes"),
+    limits = c(-.2, 1.5)
+  ) +
+  scale_x_datetime(
+    breaks = c(
+      as_datetime("2020-01-15 21:00:00"),
+      as_datetime("2020-01-16 00:00:00"),
+      as_datetime("2020-01-16 07:00:00")
+    ),
+    labels = c("Evening", "Midnight", "Morning"),
+    limits = c(as_datetime("2020-01-15 16:00:00"), as_datetime("2020-01-16 09:00:00"))
+  ) +
+  scale_color_manual(
+    values = c("palegreen", "darkorange"),
+    labels = c("In-Bed", "Asleep")
+  ) +
+  labs(
+    title = "Example from Single-Night Predictions",
+    x = NULL,
+    y = NULL,
+    color = NULL,
+    caption = "Note: simulated data"
+  ) +
+  theme(
+    text = element_text(color = "#EEE8D5"),
+    axis.title = element_text(family = "ibm", size = 20, color = "#EEE8D5"),
+    axis.text = element_text(size = 20, color = "#EEE8D5"),
+    plot.title = element_text(family = "ibm", size = 30, face = "bold"),
+    panel.grid.minor = element_blank(),
+    panel.grid = element_line(color = "#293D42"),
+    plot.background = element_rect(fill = "#002B36", color = "#EEE8D5"),
[…]

### 2023-02-28 presentation/code/plot_roc_curves.R

@@ -58,20 +58,20 @@ roc_sleep <-
 
 roc_plots <-
   roc_in_bed + roc_sleep + plot_layout(guides = "collect") &
-    theme(
-      text = element_text(color = "#EEE8D5", family = "ibm"),
-      axis.title = element_text(family = "ibm", size = 20, color = "#EEE8D5"),
-      axis.text = element_text(size = 20, color = "#EEE8D5"),
-      plot.title = element_text(family = "ibm", size = 30, face = "bold", hjust = .5),
-      plot.subtitle = element_text(size = 20, hjust = .5),
-      plot.caption = element_text(size = 14),
-      panel.grid.minor = element_blank(),
-      panel.grid = element_line(color = "#293D42"),
-      plot.background = element_rect(fill = "#002B36", color = "#002B36"),
-      panel.background = element_rect(color = "#002B36", fill = "#002B36"),
-      legend.background = element_rect(fill = "#002B36"),
-      legend.text = element_text(size = 16),
-      legend.title = element_text(size = 18, face = "bold"),
-      legend.key = element_rect(fill = "#002B36"),
-      legend.position = "bottom"
-    )
+  theme(
+    text = element_text(color = "#EEE8D5", family = "ibm"),
+    axis.title = element_text(family = "ibm", size = 20, color = "#EEE8D5"),
+    axis.text = element_text(size = 20, color = "#EEE8D5"),
+    plot.title = element_text(family = "ibm", size = 30, face = "bold", hjust = .5),
+    plot.subtitle = element_text(size = 20, hjust = .5),
+    plot.caption = element_text(size = 14),
+    panel.grid.minor = element_blank(),
+    panel.grid = element_line(color = "#293D42"),
+    plot.background = element_rect(fill = "#002B36", color = "#002B36"),
+    panel.background = element_rect(color = "#002B36", fill = "#002B36"),
+    legend.background = element_rect(fill = "#002B36"),
+    legend.text = element_text(size = 16),
+    legend.title = element_text(size = 18, face = "bold"),
+    legend.key = element_rect(fill = "#002B36"),
+    legend.position = "bottom"
+  )

### 2023-03-08 code/evaluate_models/binary_relevance_create_predictions.R
working on correct tie breaking

@@ -1,0 +1,66 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+
+test <-
+  read_parquet("data/processed/binary_relevance_testing_data.parquet")
+
+# Load models -------------------------------------------------------------
+
+# remember to mount disk
+
+in_bed_asleep_fit_filenames <-
+  list.files("/media/esbenlykke/My Passport/fitted_models/axed_models/", full.names = TRUE) |>
+  str_subset("in_bed_asleep")
+
+in_bed_awake_fit_filenames <-
+  list.files("/media/esbenlykke/My Passport/fitted_models/axed_models/", full.names = TRUE) |>
+  str_subset("in_bed_awake")
+
+out_bed_awake_fit_filenames <-
+  list.files("/media/esbenlykke/My Passport/fitted_models/axed_models/", full.names = TRUE) |>
+  str_subset("out_bed_awake")
+
+
+in_bed_asleep_fits <-
+  in_bed_asleep_fit_filenames |>
+  map(read_rds)
+
+in_bed_awake_fits <-
+  in_bed_awake_fit_filenames |>
+  map(read_rds)
+
+out_bed_awake_fits <-
+  out_bed_awake_fit_filenames |>
+  map(read_rds)
+
+
+create_preds <- function(in_bed_asleep_fits, in_bed_awake_fits, out_bed_awake_fits) {
+  in_bed_asleep_fits |>
+    augment(test) |>
+    rename_with(.cols = contains("pred"), ~ paste0("in_bed_asleep_", .x)) |>
+    bind_cols(
+      in_bed_awake_fits |>
+        predict(test, type = "prob") |>
+        rename_with(.cols = contains("pred"), ~ paste0("in_bed_awake_", .x))
+    ) |>
+    bind_cols(out_bed_awake_fits |>
+      predict(test, type = "prob") |>
+      rename_with(.cols = contains("pred"), ~ paste0("out_bed_awake_", .x))) |>
+    mutate(
+      in_bed_awake_.pred_class = as_factor(if_else(in_bed_awake_.pred_1 >= .5, 1, 0)),
+      out_bed_awake_.pred_class = as_factor(if_else(out_bed_awake_.pred_1 >= .5, 1, 0)),
+      .after = in_bed_asleep_.pred_class
+    )
+}
+
+dfs <- 
[…]

### 2023-03-08 code/evaluate_models/binary_relevance_create_stats_from_fits.R

@@ -1,41 +1,9 @@
 #!/usr/bin/env Rscript
 
 library(tidyverse)
-library(tidymodels)
 library(arrow)
 library(slider)
 library(lubridate)
-
-test <-
-  read_parquet("data/processed/binary_relevance_testing_data.parquet") 
-# TODO what causes these outliers?
-
-# Load models -------------------------------------------------------------
-
-# in_bed_fits <-
-in_bed_asleep_fit_filenames <-
-  list.files("/media/esbenlykke/My Passport/fitted_models", full.names = TRUE) |>
-  str_subset("in_bed_asleep") 
-  
-in_bed_awake_fit_filenames <-
-  list.files("data/binary_relevance_models/fitted_models", full.names = TRUE) |>
-  str_subset("in_bed_awake") 
-
-out_bed_awake_fit_filenames <- 
-  list.files("data/binary_relevance_models/fitted_models", full.names = TRUE) |>
-  str_subset("out_bed_awake") 
-
-in_bed_asleep_fits <-
-  in_bed_asleep_fit_filenames |>
-  map(read_rds)
-
-in_bed_awake_fits <-
-  in_bed_awake_fit_filenames |>
-  map(read_rds)
-
-out_bed_awake_fits <-
-  out_bed_awake_fit_filenames |>
-  map(read_rds)
 
 
 zm_stats <-
@@ -48,41 +16,40 @@ zm_stats <-
   select(id, day, month, spt_hrs, tst_hrs, se_percent, lps_min, waso_min) |>
   rename_with(.cols = -c(id, day, month), ~ paste0("zm_", .))
 
+dfs <-
+  map(list.files("/media/esbenlykke/My Passport/binary_relevance_preds",
+    full.names = TRUE,
+    pattern = "no_ties"
+  ), read_parquet)
+
 # create stats ------------------------------------------------------------
 
-create_stats <-
-  function(in_be_asleep_fit, in_bed_awake_fit, out_bed_awake_fit) {
-    in_be_asleep_fit |>
-      predict(test, type = "prob") 
-      bind_cols(
-        in_bed_awake_fit |>
-          predict(test, type = "prob")
-      ) |>
-        bind_cols(out_bed_awake_fit |> 
-                    predict(test, type = "prob")) # TODO fix colnames
[…]

### 2023-03-08 code/evaluate_models/binary_releveance_break_ties.R

@@ -1,0 +1,56 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+
+dfs <- 
+  map(list.files("/media/esbenlykke/My Passport/binary_relevance_preds", full.names = TRUE), 
+      read_parquet)
+
+# Break ties --------------------------------------------------------------
+
+
+break_ties <- function(tbl) {
+  preds_no_ties <-
+    tbl %>% 
+    mutate(across(where(is.factor), as.integer) - 1) |>
+    filter(
+      !(in_bed_asleep_.pred_class == 1 & in_bed_awake_.pred_class == 1) &
+        !(in_bed_asleep_.pred_class == 1 & out_bed_awake_.pred_class == 1) &
+        !(in_bed_awake_.pred_class == 1 & out_bed_awake_.pred_class == 1)
+    )
+
+
+  preds_ties_broken <-
+    tbl |>
+    filter(
+      (in_bed_asleep_.pred_class == 1 & in_bed_awake_.pred_class == 1) |
+        (in_bed_asleep_.pred_class == 1 & out_bed_awake_.pred_class == 1) |
+        (in_bed_awake_.pred_class == 1 & out_bed_awake_.pred_class == 1)
+    ) %>%
+    mutate(
+      max = pmax(in_bed_asleep_.pred_1, in_bed_awake_.pred_1, out_bed_awake_.pred_1),
+      win = case_when(
+        in_bed_asleep_.pred_1 == max ~ "x",
+        in_bed_awake_.pred_1 == max ~ "y",
+        out_bed_awake_.pred_1 == max ~ "z"
+      ),
+      in_bed_asleep_.pred_class = if_else(win == "x", 1, 0),
+      in_bed_awake_.pred_class = if_else(win == "y", 1, 0),
+      out_bed_awake_.pred_class = if_else(win == "z", 1, 0)
+    ) %>%
+    mutate(across(where(is.factor), as.integer) - 1) |>
+    select(-c(max, win))
+
+  preds_no_ties %>%
+    bind_rows(preds_ties_broken)
+}
+
+dfs_no_ties <-
+  map(dfs, break_ties)
+
+names(dfs_no_ties) <- c("decision_tree", "logistic_regression", "neural_net", "xgboost") %>%
+  str_c("/media/esbenlykke/My Passport/binary_relevance_preds/preds_no_ties_", ., ".parquet")
+
+walk2(dfs, names(dfs), ~ write_parquet(.x, .y))

### 2023-03-10 code/create_models/binary_relevance_logistic_regression.R
setup for three workflows: crude, binary relevance, and multiclass

@@ -31,7 +31,7 @@ train <- training(spl)
 test <- testing(spl)
 
 # write test data to file for use in later stages
-# write_parquet(train, "data/data_for_modelling/binary_relevance_training_data.parquet")
+# write_parquet(test, "data/processed/binary_relevance_testing_data.parquet")
 
 folds <-
   group_vfold_cv(train, group = id, v = 5, balance = "groups")
@@ -45,7 +45,8 @@ in_bed_asleep_rec <-
       sdacc_x + sdacc_y + sdacc_z + sdmax + temp_sd + clock_proxy_cos + clock_proxy_linear,
     data = train
   ) |>
-  step_zv(all_predictors())
+  step_zv(all_predictors()) %>%
+  step_normalize(all_numeric_predictors())
 
 in_bed_awake_rec <-
   recipe(
@@ -53,7 +54,8 @@ in_bed_awake_rec <-
       sdacc_x + sdacc_y + sdacc_z + sdmax + temp_sd + clock_proxy_cos + clock_proxy_linear,
     data = train
   ) |>
-  step_zv(all_predictors()) |> 
+  step_zv(all_predictors()) %>%
+  step_normalize(all_numeric_predictors()) %>%
   step_smote(in_bed_awake)
 
 out_bed_awake_rec <-
@@ -62,33 +64,33 @@ out_bed_awake_rec <-
       sdacc_x + sdacc_y + sdacc_z + sdmax + temp_sd + clock_proxy_cos + clock_proxy_linear,
     data = train
   ) |>
-  step_zv(all_predictors())
+  step_zv(all_predictors()) %>%
+  step_normalize(all_numeric_predictors())
 
-# Create CART model spec --------------------------------------------------
+# Create log model spec --------------------------------------------------
 
-cart_spec <-
-  decision_tree(cost_complexity = tune(), min_n = tune(), tree_depth = tune()) |>
-  set_engine("rpart") |>
+log_spec <-
+  logistic_reg(penalty = tune(), mixture = tune()) |>
+  set_engine("glmnet") |>
   set_mode("classification")
 
-tree_grid <-
-  grid_latin_hypercube(cost_complexity(),
-                       min_n(),
-                       tree_depth(range = c(3, 7)),
-                       size = 15
+log_grid <-
+  crossing(
[…]

### 2023-03-10 code/create_models/crude_workflowsets.R

@@ -1,5 +1,5 @@
 #!/usr/bin/env Rscript
-#
+
 library(tidyverse)
 library(tidymodels)
 library(arrow)
@@ -9,6 +9,8 @@ tidymodels_prefer()
 options(tidymodels.dark = TRUE)
 
 cat("Packages loaded...\n")
+
+path <- "/media/esbenlykke/My Passport/crude/"
 
 # Spend data budget -------------------------------------------------------
 cat("Spending data budget...\n")
@@ -92,7 +94,7 @@ xgb_spec <-
   boost_tree(
     tree_depth = tune(), learn_rate = tune(),
     loss_reduction = tune(), min_n = tune(),
-    sample_size = tune(), trees = tune()
+    sample_size = tune(), trees = seq(100, 700, 200)
   ) |>
   set_engine("xgboost", verbose = TRUE) |>
   set_mode("classification")
@@ -175,33 +177,33 @@ doParallel::registerDoParallel(cores = 6)
 # Tuning models with regular grid search ----------------------------------
 # In bed
 
-cat("Tuning the following workflows:\n")
-all_in_bed_workflows
-
-tictoc::tic()
-grid_ctrl <-
-  control_grid(
-    verbose = TRUE,
-    allow_par = TRUE,
-    parallel_over = "everything",
-    save_pred = FALSE,
-    save_workflow = FALSE
-  )
-
-in_bed_grid_results <-
-  all_in_bed_workflows |>
-  workflow_map(
-    seed = 123,
-    "tune_grid",
-    resamples = folds,
-    grid = 5,
-    control = grid_ctrl,
-    metrics = metric_set(f_meas, roc_auc),
-    verbose = TRUE
-  )
-tictoc::toc()
-
-write_rds(in_bed_grid_results, "data/models/in_bed_workflowsets_results.rds")
+# cat("Tuning the following workflows:\n")
+# all_in_bed_workflows
+# 
+# tictoc::tic()
+# grid_ctrl <-
+#   control_grid(
+#     verbose = TRUE,
+#     allow_par = TRUE,
+#     parallel_over = "everything",
+#     save_pred = FALSE,
+#     save_workflow = FALSE
+#   )
+# 
+# in_bed_grid_results <-
+#   all_in_bed_workflows |>
+#   workflow_map(
+#     seed = 123,
+#     "tune_grid",
+#     resamples = folds,
+#     grid = 5,
+#     control = grid_ctrl,
[…]

### 2023-03-10 code/create_models/multiclass_workflowsets.R

@@ -1,0 +1,151 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+
+
+tidymodels_prefer()
+options(tidymodels.dark = TRUE)
+
+cat("Packages loaded...\n")
+
+path <- "/media/esbenlykke/My Passport/multiclass/"
+
+# Spend data budget -------------------------------------------------------
+cat("Spending data budget...\n")
+set.seed(123)
+
+data <-
+  read_parquet("data/data_for_modelling/all_data_incl_sensor_independent_features.parquet") |>
+  mutate(
+    in_bed = as_factor(in_bed),
+    sleep = as_factor(sleep),
+    multiclass = case_when(
+      in_bed == 1 & sleep == 1 ~ "in_bed_asleep",
+      in_bed == 1 & sleep == 0 ~ "in_bed_awake",
+      in_bed == 0 & sleep == 0 ~ "out_bed_awake"
+    )
+  )
+
+# ids <- data %>% distinct(id) %>% slice(1:10)
+#   
+# data <- data %>% 
+#   filter(id %in% ids$id) %>% 
+#   group_by(id, multiclass) %>% 
+#   slice_sample(n = 1000) %>% 
+#   ungroup()
+
+set.seed(123)
+spl <-
+  data |>
+  group_initial_split(group = id, prop = .5)
+
+train <- training(spl)
+test <- testing(spl)
+
+# write out train data for later fitting of optimized workflows
+# write_parquet(train, "data/data_for_modelling/multiclass_training_data.parquet")
+
+# write out test data for later evaluation
+# write_parquet(test, "data/data_for_modelling/multiclass_testing_data.parquet")
+
+folds <-
+  group_vfold_cv(train, group = id, v = 5, balance = "groups")
+
+
+# Recipes -----------------------------------------------------------------
+cat("Creating preprocessing recipe...\n")
+
+multiclass_rec <-
+  recipe(
+    multiclass ~ age + incl + temp + macc_x + macc_y + macc_z +
+      sdacc_x + sdacc_y + sdacc_z + sdmax + temp_sd + clock_proxy_cos + clock_proxy_linear,
+    data = train
+  ) |>
+  step_zv(all_predictors()) |>
+  step_normalize(all_numeric_predictors())
+
+
+
+# Model specifications ----------------------------------------------------
+
+glmnet_spec <-
[…]

### 2023-03-10 code/evaluate_models/binary_releveance_break_ties.R

@@ -1,16 +1,20 @@
 #!/usr/bin/env Rscript
 
 library(tidyverse)
-library(tidymodels)
 library(arrow)
 
+filenames <- list.files("/media/esbenlykke/My Passport/binary_relevance_preds", 
+           full.names = TRUE) %>% 
+  str_subset("no_ties", negate = TRUE)
+
 dfs <- 
-  map(list.files("/media/esbenlykke/My Passport/binary_relevance_preds", full.names = TRUE), 
-      read_parquet)
+  map(filenames, read_parquet) %>% 
+  setNames(str_remove_all(filenames, "/media/esbenlykke/My Passport/binary_relevance_preds/"))
 
 # Break ties --------------------------------------------------------------
 
-
+### Threshold-based tie breaking
+ 
 break_ties <- function(tbl) {
   preds_no_ties <-
     tbl %>% 
@@ -44,7 +48,7 @@ break_ties <- function(tbl) {
     select(-c(max, win))
 
   preds_no_ties %>%
-    bind_rows(preds_ties_broken)
+    full_join(preds_ties_broken)
 }
 
 dfs_no_ties <-
@@ -53,4 +57,10 @@ dfs_no_ties <-
 names(dfs_no_ties) <- c("decision_tree", "logistic_regression", "neural_net", "xgboost") %>%
   str_c("/media/esbenlykke/My Passport/binary_relevance_preds/preds_no_ties_", ., ".parquet")
 
-walk2(dfs, names(dfs), ~ write_parquet(.x, .y))
+walk2(dfs_no_ties, names(dfs_no_ties), ~ write_parquet(.x, .y))
+
+### Random tie-breaking
+
+### Deterministic tie-breaking based on F1 score
+
+### Cost-based, i.e., give advantage to in-bed awake?

### 2023-03-15 code/create_models/crude_finalize_and_fit_workflowsets.R
create roc data for all approaches

@@ -32,31 +32,31 @@ all_in_bed_wf <-
 # Logistic regression
 cat("Finalizing and fitting logistic regression model\n")
 
-# best_in_bed_logistic_result <-
-#   all_in_bed_wf %>%
-#   extract_workflow_set_result("in_bed_logistic_regression") |>
-#   select_best(metric = "f_meas")
-# 
-# all_in_bed_wf |>
-#   extract_workflow("in_bed_logistic_regression") |>
-#   finalize_workflow(best_in_bed_logistic_result) %>%
-#   fit(train) %>%
-#   write_rds(str_c(path, "fitted_models/crude_in_bed_log_reg_fit.rds"))
-# 
-# 
-# # Neural network
-# cat("Finalizing and fitting neural network model\n")
-# 
-# best_in_bed_nn_result <-
-#   all_in_bed_wf %>%
-#   extract_workflow_set_result("in_bed_neural_network") |>
-#   select_best(metric = "f_meas")
-# 
-# all_in_bed_wf |>
-#   extract_workflow("in_bed_neural_network") |>
-#   finalize_workflow(best_in_bed_nn_result)  %>%
-#   fit(train) %>%
-#   write_rds(str_c(path, "fitted_models/crude_in_bed_nnet_fit.rds"))
+best_in_bed_logistic_result <-
+  all_in_bed_wf %>%
+  extract_workflow_set_result("in_bed_logistic_regression") |>
+  select_best(metric = "f_meas")
+
+all_in_bed_wf |>
+  extract_workflow("in_bed_logistic_regression") |>
+  finalize_workflow(best_in_bed_logistic_result) %>%
+  fit(train) %>%
+  write_rds(str_c(path, "fitted_models/crude_in_bed_log_reg_fit.rds"))
+
+
+# Neural network
+cat("Finalizing and fitting neural network model\n")
+
+best_in_bed_nn_result <-
+  all_in_bed_wf %>%
+  extract_workflow_set_result("in_bed_neural_network") |>
+  select_best(metric = "f_meas")
+
+all_in_bed_wf |>
+  extract_workflow("in_bed_neural_network") |>
+  finalize_workflow(best_in_bed_nn_result)  %>%
+  fit(train) %>%
+  write_rds(str_c(path, "fitted_models/crude_in_bed_nnet_fit.rds"))
 
 
 # XGboost
@@ -80,46 +80,46 @@ cat("All done\n")
 
[…]

### 2023-03-15 code/create_models/crude_workflowsets.R

@@ -23,12 +23,12 @@ data <-
   )
 
 ### for testing purposes
-# ids <- data %>% distinct(id) %>% slice(1:10)
+# ids <- data %>% distinct(id) %>% slice(1:15)
 # 
 # data <- data %>%
 #   filter(id %in% ids$id) %>%
-#   group_by(id, in_bed, sleep) %>%
-#   slice_sample(n = 100) %>%
+#   group_by(id) %>%
+#   slice_sample(n = 1000) %>%
 #   ungroup()
 ###
 
@@ -103,7 +103,7 @@ xgb_spec <-
   boost_tree(
     tree_depth = tune(), learn_rate = tune(),
     loss_reduction = tune(), min_n = tune(),
-    sample_size = tune(), trees = seq(100, 700, 200)
+    sample_size = tune(), trees = tune()
   ) |>
   set_engine("xgboost", verbose = TRUE) |>
   set_mode("classification")
@@ -201,7 +201,7 @@ all_in_bed_workflows
 
 tictoc::tic()
 in_bed_grid_results <-
-  all_in_bed_workflows[3,] |>
+  all_in_bed_workflows |>
   workflow_map(
     seed = 123,
     "tune_grid",

### 2023-03-15 presentation/code/plot_roc_curves.R

@@ -8,12 +8,12 @@ library(here)
 library(patchwork)
 
 in_bed_roc <-
-  read_parquet(here("data/processed/roc_data_in_bed.parquet")) |>
+  read_parquet(here("data/processed/crude_roc_data_in_bed.parquet")) |>
   group_by(model) |>
   slice_sample(n = 1000, replace = TRUE)
 
 sleep_roc <-
-  read_parquet(here("data/processed/roc_data_sleep.parquet")) |>
+  read_parquet(here("data/processed/crude_roc_data_sleep.parquet")) |>
   group_by(model) |>
   slice_sample(n = 1000, replace = TRUE)
 

### 2023-03-16 presentation/index.qmd
working on correct tie breaking
setup for three workflows: crude, binary relevance, and multiclass
create roc data for all approaches
work on roc curves

@@
 - Evaluated combined classifiers using weighted micro-averaging.
+- adjust threshold for in bed awake or give edge in tie breakers??
+- zm sleep stats inclusion criteria
+- three approaches: tw- binary (risk of ghost classes), binary relevance (choosing the right tie breaker), and multiclass
 
@@
 
+## Present three modeling approaches
+
+::: {.fragment}
+- Multiclass classifier with three labels: in-bed-asleep, in-bed-awake, and out-bed-awake
+  - pros: simplicity, speed, joint modeling, consistency across classes
+  - cons: struggle with class imbalance, performance
+:::
+
+::: {.fragment}
+- Two binary classifiers: in-bed/out-bed and asleep/awake
+  - pros: 
+  - cons: ghost classes
+:::
+
+::: {.fragment}
+- Three binary classifiers: in-bed-asleep 1/0, in-bed-awake 1/0, and out-bed-awake 1/0
+  - pros: more flexible, can better handle class imbalance
+  - cons: complex training, low speed, feature redundancy
+:::
+
 ## building Models {transition="convex"}
@@
 
+multiclass_ex
+```
+
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
+multiclass_lps
+```
+
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
+multiclass_tst
+```
+
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
 plot_ex
@@
 #| fig-asp: .4
-source(here::here("presentation/code/plot_pred_example.R"))
-
 plot_lps
 ```
-
-::: aside 
-This methedology is called binary relevance.
-:::
 
@@
 #| fig-asp: .4
-source(here::here("presentation/code/plot_pred_example.R"))
-
 plot_tst
@@
 
-## ROC Curves {transition="convex"}
+<!-- ## ROC Curves {transition="convex"} -->
+<!-- ### Two binary models (in-bed/asleep) -->
 
-```{r}
-source(here::here("presentation/code/plot_roc_curves.R"))
+<!-- ```{r} -->
+<!-- source(here::here("presentation/code/plot_roc_curves.R")) -->
 
-roc_plots
-```
+<!-- roc_plots -->
+<!-- ``` -->
 
-## ROC Curves {transition="convex"}
+<!-- ## ROC Curves {transition="convex"} -->
+<!-- ### Binary relevance -->
 
-```{r}
-roc_combined_plots
-```
+<!-- ```{r} -->
+<!-- source(here::here) -->
+<!-- ``` -->
 
-## Lots of Metrics {transition="convex" .H2}
+<!-- ## Lots of Metrics {transition="convex" .H2} -->
 
-:::: {.columns}
+<!-- :::: {.columns} -->
 
-::: {.column}
-```{r}
-source(here::here("presentation/code/tables_performance.R"))
+<!-- ::: {.column} -->
+<!-- ```{r} -->
+<!-- source(here::here("presentation/code/tables_performance.R")) -->
 
-tab1
-```
-:::
+<!-- tab1 -->
+<!-- ``` -->
+<!-- ::: -->
 
-::: {.column}
-```{r}
-tab2
-```
-:::
+<!-- ::: {.column} -->
+<!-- ```{r} -->
+<!-- tab2 -->
+<!-- ``` -->
+<!-- ::: -->
 
-::::
+<!-- :::: -->
 
-## Bland-Altman Plots {transition="convex"}
+<!-- ## Bland-Altman Plots {transition="convex"} -->
 
-::: {.panel-tabset}
+<!-- ::: {.panel-tabset} -->
 
-### Decision Tree
+<!-- ### Decision Tree -->
 
-```{r}
-#| echo: false
-source(here::here("presentation/code/plot_ba.R"))
+<!-- ```{r} -->
+<!-- #| echo: false -->
+<!-- source(here::here("presentation/code/plot_ba.R")) -->
 
-dc_plots[[1]]
+<!-- dc_plots[[1]] -->
 
-dc_plots[[2]]
+<!-- dc_plots[[2]] -->
 
-dc_plots[[3]]
+<!-- dc_plots[[3]] -->
 
-dc_plots[[4]]
+<!-- dc_plots[[4]] -->
 
-dc_plots[[5]]
-```
+<!-- dc_plots[[5]] -->
+<!-- ``` -->
 
-### Logistic Regression
+<!-- ### Logistic Regression -->
 
-```{r}
-lr_plots[[1]]
+<!-- ```{r} -->
+<!-- lr_plots[[1]] -->
 
-lr_plots[[2]]
+<!-- lr_plots[[2]] -->
 
-lr_plots[[3]]
+<!-- lr_plots[[3]] -->
 
-lr_plots[[4]]
+<!-- lr_plots[[4]] -->
 
-lr_plots[[5]]
-```
+<!-- lr_plots[[5]] -->
+<!-- ``` -->
 
-### Neural Net
+<!-- ### Neural Net -->
 
-```{r}
-nn_plots[[1]]
+<!-- ```{r} -->
+<!-- nn_plots[[1]] -->
 
-nn_plots[[2]]
+<!-- nn_plots[[2]] -->
 
-nn_plots[[3]]
+<!-- nn_plots[[3]] -->
 
-nn_plots[[4]]
+<!-- nn_plots[[4]] -->
 
-nn_plots[[5]]
-```
+<!-- nn_plots[[5]] -->
+<!-- ``` -->
 
-### XGBoost
+<!-- ### XGBoost -->
 
-```{r}
-xg_plots[[1]]
+<!-- ```{r} -->
+<!-- xg_plots[[1]] -->
 
-xg_plots[[2]]
+<!-- xg_plots[[2]] -->
 
-xg_plots[[3]]
+<!-- xg_plots[[3]] -->
 
-xg_plots[[4]]
+<!-- xg_plots[[4]] -->
 
-xg_plots[[5]]
-```
+<!-- xg_plots[[5]] -->
+<!-- ``` -->
 
-:::
+<!-- ::: -->
 
-## Bland-Altman Analysis {transition="convex" .H2}
+<!-- ## Bland-Altman Analysis {transition="convex" .H2} -->
 
-```{r}
-source(here::here("presentation/code/table_ba.R"))
+<!-- ```{r} -->
+<!-- source(here::here("presentation/code/table_ba.R")) -->
 
-tab_ba
-```
+<!-- tab_ba -->
+<!-- ``` -->
 
-## In-bed classification flow {transition="convex" .H2}
+<!-- ## In-bed classification flow {transition="convex" .H2} -->
 
-```{r}
-in_bed_tree <-
-  read_rds(here::here("data/models/fitted_models/in_bed_simple_CART_fit.rds"))
+<!-- ```{r} -->
+<!-- in_bed_tree <- -->
+<!--   read_rds(here::here("data/models/fitted_models/in_bed_simple_CART_fit.rds")) -->
 
-in_bed_tree |>
-  extract_fit_engine() |>
-  visTree(
-    height = "550px",
-    nodesPopSize = TRUE,
-    minNodeSize = 10,
-    maxNodeSize = 30,
-    fallenLeaves = TRUE,
-    legend = FALSE,
-    digits = 3,
-    direction = "UD",
-    colorVar = c(
-      "#CC5500", "#8B0000", "#FFD700",
-      "#228B22", "#654321", "#8B0000", "#FFDB58"
-    ),
-    export = FALSE
-  ) |>
-  visNodes(font = list(
-    color = "#EEE8D5",
-    size = 20,
-    strokeWidth = 1
-  )) |>
-  visEdges(font = list(
-    color = "#EEE8D5",
-    size = 20,
-    strokeWidth = 1
-  )) |>
-  visInteraction(
-    dragNodes = FALSE,
-    dragView = FALSE,
-    zoomView = FALSE
-  )
-```
+<!-- in_bed_tree |> -->
+<!--   extract_fit_engine() |> -->
+<!--   visTree( -->
+<!--     height = "550px", -->
+<!--     nodesPopSize = TRUE, -->
+<!--     minNodeSize = 10, -->
+<!--     maxNodeSize = 30, -->
+<!--     fallenLeaves = TRUE, -->
+<!--     legend = FALSE, -->
+<!--     digits = 3, -->
+<!--     direction = "UD", -->
+<!--     colorVar = c( -->
+<!--       "#CC5500", "#8B0000", "#FFD700", -->
+<!--       "#228B22", "#654321", "#8B0000", "#FFDB58" -->
+<!--     ), -->
+<!--     export = FALSE -->
+<!--   ) |> -->
+<!--   visNodes(font = list( -->
+<!--     color = "#EEE8D5", -->
+<!--     size = 20, -->
+<!--     strokeWidth = 1 -->
+<!--   )) |> -->
+<!--   visEdges(font = list( -->
+<!--     color = "#EEE8D5", -->
+<!--     size = 20, -->
+<!--     strokeWidth = 1 -->
+<!--   )) |> -->
+<!--   visInteraction( -->
+<!--     dragNodes = FALSE, -->
+<!--     dragView = FALSE, -->
+<!--     zoomView = FALSE -->
+<!--   ) -->
+<!-- ``` -->
 
----
+<!-- --- -->
 
-## Sleep classification flow {transition="convex" .H2}
+<!-- ## Sleep classification flow {transition="convex" .H2} -->
 
-```{r}
-sleep_tree <-
-  read_rds(here::here("data/models/fitted_models/sleep_simple_CART_fit.rds"))
+<!-- ```{r} -->
+<!-- sleep_tree <- -->
+<!--   read_rds(here::here("data/models/fitted_models/sleep_simple_CART_fit.rds")) -->
 
-sleep_tree |>
-  extract_fit_engine() |>
-  visTree(
-    height = "550px",
-    nodesPopSize = TRUE,
-    minNodeSize = 10,
-    maxNodeSize = 30,
-    fallenLeaves = TRUE,
-    legend = FALSE,
-    digits = 3,
-    direction = "UD",
-    colorVar = c(
-      "#CC5500", "#8B0000", "#FFD700",
-      "#228B22", "#654321", "#8B0000", "#FFDB58"
-    ),
-    export = FALSE
-  ) |>
-  visNodes(font = list(
-    color = "#EEE8D5",
-    strokeWidth = 1
-  )) |>
-  visEdges(font = list(
-    color = "#EEE8D5",
-    strokeWidth = 1
-  )) |>
-  visInteraction(
-    dragNodes = FALSE,
-    dragView = FALSE,
-    zoomView = FALSE
-  )
-```
+<!-- sleep_tree |> -->
+<!--   extract_fit_engine() |> -->
+<!--   visTree( -->
+<!--     height = "550px", -->
+<!--     nodesPopSize = TRUE, -->
+<!--     minNodeSize = 10, -->
+<!--     maxNodeSize = 30, -->
+<!--     fallenLeaves = TRUE, -->
+<!--     legend = FALSE, -->
+<!--     digits = 3, -->
+<!--     direction = "UD", -->
+<!--     colorVar = c( -->
+<!--       "#CC5500", "#8B0000", "#FFD700", -->
+<!--       "#228B22", "#654321", "#8B0000", "#FFDB58" -->
+<!--     ), -->
+<!--     export = FALSE -->
+<!--   ) |> -->
+<!--   visNodes(font = list( -->
+<!--     color = "#EEE8D5", -->
+<!--     strokeWidth = 1 -->
+<!--   )) |> -->
+<!--   visEdges(font = list( -->
+<!--     color = "#EEE8D5", -->
+<!--     strokeWidth = 1 -->
+<!--   )) |> -->
+<!--   visInteraction( -->
+<!--     dragNodes = FALSE, -->
+<!--     dragView = FALSE, -->
+<!--     zoomView = FALSE -->
+<!--   ) -->
+<!-- ``` -->
 
-## Discussion {transition="convex"}
+<!-- ## Discussion {transition="convex"} -->
 
-- heteroscedasticity
-- Cheung 2018 table 4: actigraphy provides a sufficiently narrow range of possible mean differences (CI 95%) clinical significant thresholds
-- could be interesting to build models on thigh and hip ocmbined.
-- multiclass vs multilabel classification
-- in-bed awake/sleep is highly imbalanced -> maybe train a new classifier accounting for imbalanced data (SMOTE)
-- model combined preds instead?
+<!-- - heteroscedasticity -->
+<!-- - Cheung 2018 table 4: actigraphy provides a sufficiently narrow range of possible mean differences (CI 95%) clinical significant thresholds -->
+<!-- - could be interesting to build models on thigh and hip ocmbined. -->
+<!-- - multiclass vs multilabel classification -->
+<!-- - in-bed awake/sleep is highly imbalanced -> maybe train a new classifier accounting for imbalanced data (SMOTE) -->
+<!-- - model combined preds instead? -->
 
-## References {transition="convex"}
+<!-- ## References {transition="convex"} -->

### 2023-03-16 presentation/code/plot_hyperparameter_tuning.R

@@ -19,7 +19,7 @@ hyper_plot <-
     color = "Model Type",
     shape = NULL,
     y = "F1 Metric",
-    caption = "This figure contains only data from 1 person and, hence, is only for illustrative purposes."
+    caption = "Note: this is based on simulated data and is not representative of the actual training data from the study"
   ) +
   scale_color_brewer(palette = "Dark2",
                      label = c("XGboost", "Neural Net", "Multinomial Logistic Regression", "Support Vector Machine")) +

### 2023-03-16 presentation/code/plot_roc_curves.R

@@ -7,21 +7,92 @@ library(ggtext)
 library(here)
 library(patchwork)
 
+multi_roc <-
+  read_parquet("data/processed/multiclass_OvsALL_roc_data.parquet") %>%
+  group_by(model) %>%
+  slice_sample(n = 112, replace = FALSE)
+
+in_bed_awake <-
+  read_parquet("data/processed/binary_relevance_roc_data_in_bed_awake.parquet") %>%
+  group_by(model) %>%
+  slice_sample(n = 1000, replace = TRUE)
+
+in_bed_asleep <-
+  read_parquet("data/processed/binary_relevance_roc_data_in_bed_asleep.parquet") %>%
+  group_by(model) %>%
+  slice_sample(n = 1000, replace = TRUE)
+
+out_bed_awake <-
+  read_parquet("data/processed/binary_relevance_roc_data_out_bed_awake.parquet") %>%
+  group_by(model) %>%
+  slice_sample(n = 1000, replace = TRUE)
+
 in_bed_roc <-
-  read_parquet(here("data/processed/crude_roc_data_in_bed.parquet")) |>
-  group_by(model) |>
+  read_parquet(here("data/processed/crude_roc_data_in_bed.parquet")) %>%
+  group_by(model) %>%
   slice_sample(n = 1000, replace = TRUE)
 
 sleep_roc <-
-  read_parquet(here("data/processed/crude_roc_data_sleep.parquet")) |>
-  group_by(model) |>
+  read_parquet(here("data/processed/crude_roc_data_sleep.parquet")) %>%
+  group_by(model) %>%
   slice_sample(n = 1000, replace = TRUE)
 
 
+plot_roc <- function(roc_data) {
+  in_bed_roc %>%
+    ggplot(
+      aes(1 - specificity, sensitivity, color = model)
+    ) + # plot with 2 ROC curves for each model
+    geom_line(linewidth = .8, alpha = .7) +
+    geom_abline(slope = 1, intercept = 0, linewidth = 0.4, lty = 2, color = "lightblue") +
+    scale_color_brewer(palette = "Dark2") +
+    labs(
+      title = "Prediction of In-Bed",
+      x = "1 - Specificity",
+      y = "Sensitivity"
+    ) +
+    theme(
+      plot.margin = margin(r = 50, unit = "pt")
+    ) +
+    theme(
+      text = element_text(color = "#EEE8D5", family = "ibm"),
[…]

### 2023-03-16 presentation/index.qmd
roc curves are done. Next up is tables with all performance metrics.

@@
 
-<!-- ## ROC Curves {transition="convex"} -->
-<!-- ### Two binary models (in-bed/asleep) -->
+## ROC Curves {transition="convex"}
 
-<!-- ```{r} -->
-<!-- source(here::here("presentation/code/plot_roc_curves.R")) -->
+```{r}
+source(here::here("presentation/code/plot_roc_curves.R"))
 
-<!-- roc_plots -->
-<!-- ``` -->
+multi_roc_plot
+```
 
-<!-- ## ROC Curves {transition="convex"} -->
-<!-- ### Binary relevance -->
+## ROC Curves {transition="convex"}
 
-<!-- ```{r} -->
-<!-- source(here::here) -->
-<!-- ``` -->
+```{r}
+three_classifiers_roc_plot
+```
+
+## ROC Curves {transition="convex"}
+
+```{r}
+#| fig-width: 8
+#| fig-align: center
+two_classifiers_roc_plot
+```
 
@@
 
-<!-- ## References {transition="convex"} -->
+## References {transition="convex"}

### 2023-03-21 presentation/index.qmd
create new ba plots

@@
   revealjs:
-    smaller: true
+    smaller: false
     theme: moon
@@
 
+```{r setup, include=FALSE} 
+knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
+```
+
+# Background {transition="fade"}
+
 ## Background {transition="convex"}
-
-```{r}
-#| echo: false
-library(gt)
-library(arrow)
-library(visNetwork)
-library(tidymodels)
-library(tidymodels)
-```
 
@@
 - Zmachine-derived sleep stats
--
-
-Purpose...
+- Purpose...
 
@@
 
+# Methods {transition="fade"}
+
 ## Methods {transition="convex"}
-
-::: {.fragment}
 
@@
 - three approaches: tw- binary (risk of ghost classes), binary relevance (choosing the right tie breaker), and multiclass
-
-:::
 
@@
 #| fig-asp: .4
+
 multiclass_lps
@@
 #| fig-asp: .4
+
 multiclass_tst
+```
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
+
+br_ex
+```
+
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
+br_lps
+```
+
+## Estimate Sleep Quality Metrics {transition="fade"}
+
+```{r}
+#| fig-asp: .4
+br_tst
 ```
@@
 
-## Results {transition="convex" .H2}
+# Results {transition="fade"}
+
+## {.H2 transition="convex"}
 
 :::: {.columns}
-
-::: {.fragment .column width="50%"}
+::: {.column width="50%"}
 
@@
 
-::: {.fragment .column width="50%"}
+::: {.column width="50%"}
 
@@
 :::
-
 ::::
 
-## ROC Curves {transition="convex"}
+## Many ROC Curves {transition="convex"}
+
+## {transition="convex"}
 
@@
 
-## ROC Curves {transition="convex"}
+## {transition="convex"}
 
@@
 
-<!-- ## Lots of Metrics {transition="convex" .H2} -->
+## Lots of Metrics {transition="convex"}
 
-<!-- :::: {.columns} -->
+## {transition="convex"}
+::: {layout="[[-1], [1], [-1]]"}
+```{r}
+source(here::here("presentation/code/tables_performance.R"))
 
-<!-- ::: {.column} -->
-<!-- ```{r} -->
-<!-- source(here::here("presentation/code/tables_performance.R")) -->
+tab_multi_metrics
+```
+:::
 
-<!-- tab1 -->
-<!-- ``` -->
-<!-- ::: -->
+## {transition="convex" .H2}
 
-<!-- ::: {.column} -->
-<!-- ```{r} -->
-<!-- tab2 -->
-<!-- ``` -->
-<!-- ::: -->
+::: {layout="[[-1], [1], [-1]]"}
+```{r}
+tab_crude_metrics
+```
+:::
 
-<!-- :::: -->
+## {transition="convex" .H2}
+
+```{r}
+tab_br_metrics
+```
 

### 2023-03-21 code/create_test_predictions.R

@@ -1,0 +1,128 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+
+
+### Crude
+
+test_crude <-
+  read_parquet("data/data_for_modelling/crude_testing_data.parquet")
+# filter((id != 1377205) &
+#          (id != 1377204 | !noon_day %in% c(9, 25)) &
+#          (id != 1377204 | noon_day != 22))
+# TODO what causes these outliers?
+
+
+in_bed_filenames <-
+  list.files("/media/esbenlykke/My Passport/crude/fitted_models/axed_models",
+    full.names = TRUE
+  ) |>
+  str_subset("in_bed")
+
+sleep_filenames <-
+  list.files("/media/esbenlykke/My Passport/crude/fitted_models/axed_models",
+    full.names = TRUE
+  ) |>
+  str_subset("sleep")
+
+in_bed_fits <- map(in_bed_filenames, read_rds) |>
+  set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
+
+sleep_fits <- map(sleep_filenames, read_rds) |>
+  set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
+
+create_crude_preds <- function(fit1, fit2) {
+  fit1 %>%
+    augment(test_crude) %>%
+    rename(pred_in_bed = .pred_class, pred_in_bed_0 = .pred_0, pred_in_bed_1 = .pred_1) %>%
+    bind_cols(predict(fit2, test_crude, type = "prob")) %>%
+    bind_cols(predict(fit2, test_crude)) %>%
+    rename(pred_sleep = .pred_class, pred_sleep_0 = .pred_0, pred_sleep_1 = .pred_1)
+}
+
+preds_crude <-
+  map2_dfr(in_bed_fits, sleep_fits, create_crude_preds, 
+           .progress = TRUE,
+           .id = "model")
+
+write_parquet(preds_crude, "data/processed/crude_test_predictions.parquet")
+
+
+### Binary relevance
+test_br <-
+  read_parquet("data/data_for_modelling/binary_relevance_testing_data.parquet")
+
+in_bed_asleep_filenames <-
+  list.files("/media/esbenlykke/My Passport/binary_relevance/fitted_models/axed_models",
+    full.names = TRUE
+  ) |>
+  str_subset("in_bed_asleep")
+
+in_bed_awake_filenames <-
[…]

### 2023-03-21 code/evaluate_models/create_roc_data_all_models.R

@@ -1,0 +1,106 @@
+#!/usr/bin/env
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+
+
+get_roc <-
+  function(fit, truth, estimate) {
+    fit |>
+      augment(test) |>
+      roc_curve(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
+  }
+
+
+# CRUDE -------------------------------------------------------------------
+
+
+in_bed_fits_files <- list.files("/media/esbenlykke/My Passport/crude/fitted_models/axed_models",
+  full.names = TRUE
+) %>%
+  str_subset("in_bed")
+
+
+sleep_fits_files <- list.files("/media/esbenlykke/My Passport/crude/fitted_models/axed_models",
+  full.names = TRUE
+) %>%
+  str_subset("sleep")
+
+in_bed_fits <- map(in_bed_fits_files, read_rds) |>
+  set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
+
+sleep_fits <- map(sleep_fits_files, read_rds) |>
+  set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
+
+test <- read_parquet("data/data_for_modelling/crude_testing_data.parquet")
+
+
+map_dfr(in_bed_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
+  write_parquet("data/processed/crude_roc_data_in_bed.parquet")
+
+map_dfr(sleep_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
+  write_parquet("data/processed/crude_roc_data_sleep.parquet")
+
+
+# Binary relevance --------------------------------------------------------
+
+
+in_bed_asleep_files <- list.files("/media/esbenlykke/My Passport/binary_relevance/fitted_models/axed_models",
+  full.names = TRUE
+) %>%
+  str_subset("asleep")
+
+in_bed_awake_files <- list.files("/media/esbenlykke/My Passport/binary_relevance/fitted_models/axed_models",
+  full.names = TRUE
+) %>%
+  str_subset("in_bed_awake")
+
+out_bed_awake_files <- list.files("/media/esbenlykke/My Passport/binary_relevance/fitted_models/axed_models",
+  full.names = TRUE
+) %>%
+  str_subset("out_bed_awake")
+
+in_bed_asleep_fits <- map(in_bed_asleep_files, read_rds) |>
[…]

### 2023-03-21 code/evaluate_models/create_stats_from_fits.R

@@ -1,85 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(tidymodels)
-library(arrow)
-library(slider)
-library(lubridate)
-
-test <-
-  read_parquet("data/processed/testing_data.parquet") |>
-  filter((id != 1377205) &
-           (id != 1377204 | !noon_day %in% c(9, 25)) &
-           (id != 1377204 | noon_day != 22))
-# TODO what causes these outliers?
-
-# Load models -------------------------------------------------------------
-
-# in_bed_fits <-
-in_bed_fit_files <-
-  list.files("data/models/fitted_models", full.names = TRUE) |>
-  str_subset("in_bed") |>
-  str_subset("MARS", negate = TRUE) # for some reason the MARS did not train properly, exclude it.
-
-sleep_fit_files <-
-  list.files("data/models/fitted_models", full.names = TRUE) |>
-  str_subset("sleep") |>
-  str_subset("MARS", negate = TRUE)
-
-in_bed_fits <-
-  in_bed_fit_files |>
-  map(read_rds)
-
-sleep_fits <-
-  sleep_fit_files |>
-  map(read_rds)
-
-
-zm_stats <-
-  read_csv("data/processed/all_zm_stats.csv") |>
-  janitor::clean_names() |>
-  mutate(
-    day = day(start_date),
-    month = month(start_date)
-  ) |>
-  select(id, day, month, spt_hrs, tst_hrs, se_percent, lps_min, waso_min) |>
-  rename_with(.cols = -c(id, day, month), ~ paste0("zm_", .))
-
-# create stats ------------------------------------------------------------
-
-create_stats <-
-  function(in_bed_fit, sleep_fit) {
-    in_bed_fit |>
-      augment(test) |>
-      select(id, datetime, noon_day, in_bed_pred = .pred_class) |>
-      bind_cols(
-        sleep_fit |>
-          predict(test) |>
-          rename(sleep_pred = .pred_class)
-      ) |>
-      mutate(across(in_bed_pred:sleep_pred, as.integer) - 1) |>
-      mutate(
-        month = month(datetime - hours(12)),
-        in_bed_72 = slide_sum(in_bed_pred, before = 36, after = 36),
-        sleep_72 = slide_sum(sleep_pred, before = 36, after = 36),
[…]

### 2023-03-21 code/evaluate_models/performance_all_models.R

@@ -1,101 +1,0 @@
-#!/usr/bin/env RscriptsuppressMessages(library(tidyverse))
-
-library(tidyverse)
-library(tidymodels)
-library(arrow)
-
-in_bed_fits_files <- list.files("data/models/fitted_models", full.names = TRUE) |>
-  str_subset("mars", negate = TRUE) |>
-  str_subset("in_bed")
-
-sleep_fits_files <- list.files("data/models/fitted_models", full.names = TRUE) |>
-  str_subset("mars", negate = TRUE) |>
-  str_subset("sleep")
-
-in_bed_fits <- map(in_bed_fits_files, read_rds) |>
-  set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
-
-sleep_fits <- map(sleep_fits_files, read_rds) |>
-  set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
-
-test <- read_parquet("data/processed/testing_data.parquet") 
-
-
-# calculate metrics -------------------------------------------------------
-
-
-
-my_metrics <-
-  metric_set(f_meas, accuracy, sensitivity, precision, specificity)
-
-get_metrics <-
-  function(fit, truth, estimate) {
-    fit |>
-      augment(test) |>
-      my_metrics(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
-  }
-
-in_bed_metrics <-
-  map_dfr(in_bed_fits, ~ get_metrics(.x, in_bed, .pred_class), .id = "model")
-
-sleep_metrics <-
-  map_dfr(sleep_fits, ~ get_metrics(.x, sleep, .pred_class), .id = "model")
-
-in_bed_metrics |>
-  select(-.estimator) |>
-  pivot_wider(names_from = model, values_from = .estimate) |>
-  mutate(event = "In-bed Prediction") |>
-  bind_rows(
-    sleep_metrics |>
-      select(-.estimator) |>
-      pivot_wider(names_from = model, values_from = .estimate) |>
-      mutate(event = "Sleep Prediction")
-  ) |>
-  write_csv("data/processed/performance_metrics.csv")
-
-
-
-# combined metrics --------------------------------------------------------
-
-
-get_scores <-
-  function(in_bed_fit, sleep_fit, truth, estimate) {
-    in_bed_fit |>
-      augment(test, type = "prob") |>
[…]

### 2023-03-21 code/evaluate_models/table_performance_metrics.R

@@ -5,8 +5,7 @@ library(gt)
 library(showtext)
 
 
-separate_metrics <- read_csv("data/processed/performance_metrics.csv")
-combined_metrics <- read_csv("data/processed/combined_preds_performance_metrics.csv")
+crude_metrics <- read_csv("data/processed/crude_performance_metrics.csv")
 
 # Create table ------------------------------------------------------------
 
@@ -50,7 +49,7 @@ tab_fun <- function(metric_data) {
     data_color(columns = decision_tree, colors = "lightblue", alpha = .6)
 }
 
-tab_fun(separate_metrics)
+tab_fun(crude_metrics)
 
 gtsave(tab_performance, "visuals/table_separate_preds_performance_metrics.html")
 

### 2023-03-21 presentation/code/plot_pred_example.R

@@ -6,7 +6,7 @@ font_add_google("Mukta", family = "mukta")
 font_add_google("IBM Plex Serif", family = "ibm")
 showtext_auto()
 
-my_theme <- 
+my_theme <-
   theme(
     text = element_text(color = "#EEE8D5"),
     axis.title = element_text(family = "ibm", size = 20, color = "#EEE8D5"),
@@ -27,41 +27,197 @@ my_theme <-
 
 multiclass_ex <-
   tibble(
-  datetime = seq(
-    as_datetime("2020-01-15 12:00:00"),
-    as_datetime("2020-01-16 12:00:00"), seconds(10)
-  ),
-  in_bed_awake = if_else(
-    (datetime > as_datetime("2020-01-15 21:00:00") &
-      datetime < as_datetime("2020-01-15 21:30:00")) |
-      (datetime > as_datetime("2020-01-16 06:00:00") &
-        datetime < as_datetime("2020-01-16 06:50:00")), 1, 0
-  ),
-  in_bed_asleep = if_else(
-    datetime > as_datetime("2020-01-15 21:30:00") &
-      datetime < as_datetime("2020-01-16 06:00:00"), 1, 0
-  ),
-  out_bed_awake = if_else(
-    datetime > as_datetime("2020-01-15 12:00:00") &
-      datetime < as_datetime("2020-01-15 21:00:00") |
-      datetime > as_datetime("2020-01-16 06:50:00") &
-        datetime < as_datetime("2020-01-16 12:00:00"), 1, 0
-  )
-) %>% 
-  # pivot_longer(-datetime) %>% 
-    ggplot() +
+    datetime = seq(
+      as_datetime("2020-01-15 12:00:00"),
+      as_datetime("2020-01-16 12:00:00"), seconds(10)
+    ),
+    multiclass = factor(
+      case_when(
+        datetime >= as_datetime("2020-01-15 12:00:00") &
+          datetime < as_datetime("2020-01-15 21:00:00") |
+          datetime > as_datetime("2020-01-16 06:50:00") &
+            datetime <= as_datetime("2020-01-16 12:00:00") ~ "Out-Bed Awake",
+        (datetime >= as_datetime("2020-01-15 21:00:00") &
+          datetime <= as_datetime("2020-01-15 21:30:00")) |
+          (datetime >= as_datetime("2020-01-16 06:00:00") &
[…]

### 2023-03-24 presentation/index.qmd
calculate MDC and corresponding CIs

@@
 knitr::opts_chunk$set(warning = FALSE, message = FALSE)
+
+library(visNetwork)
+library(tidymodels)
+library(tidyverse)
 ```
@@
 
-## Background {transition="convex"}
+## Background {transition="convex" .smaller}
 
@@
 
-## But Esben, what about them sleep stages!? {transition="convex"}
+## But Esben, what about them sleep stages!? {transition="convex" .smaller}
 
@@
 
-## Methods {transition="convex"}
+## Methods {transition="convex" .smaller}
 
@@
 
-## Features {transition="convex" .H2}
+## Features {transition="convex" .H2 .smaller}
 
@@
 
-## Human Circadian Clock {transition="convex"}
+## Human Circadian Clock {transition="convex" .smaller}
 
@@
 
-## Present three modeling approaches
+## Present three modeling approaches  {.smaller}
 
@@
 
-## {.H2 transition="convex"}
+## {.H2 transition="convex" .smaller}
 
@@
 
-### Epoch-Based
+### Epoch-Based {.smaller}
 
@@
 
-### Summarized across nights {.H3}
+### Summarized across nights {.H3 .smaller}
 
@@
   - Wake After Sleep Onset
+* Minimal Detectable Change
 
@@
 
-## Many ROC Curves {transition="convex"}
+## Many ROC Curves {transition="convex" .H2}
 
@@
 
-## Lots of Metrics {transition="convex"}
+## Lots of Metrics {transition="convex" .H2}
 
@@
 
-<!-- ## Bland-Altman Plots {transition="convex"} -->
+## Bland-Altman Plots {transition="convex" .smaller}
 
-<!-- ::: {.panel-tabset} -->
+::: {.panel-tabset}
 
-<!-- ### Decision Tree -->
+### Sleep Period Time {.smaller}
 
-<!-- ```{r} -->
-<!-- #| echo: false -->
-<!-- source(here::here("presentation/code/plot_ba.R")) -->
+```{r}
+#| echo: false
+source(here::here("presentation/code/plot_ba_metric-wise.R"))
 
-<!-- dc_plots[[1]] -->
+spt_plots[[1]]
 
-<!-- dc_plots[[2]] -->
+spt_plots[[2]]
 
-<!-- dc_plots[[3]] -->
+spt_plots[[3]]
 
-<!-- dc_plots[[4]] -->
+spt_plots[[4]]
+```
 
-<!-- dc_plots[[5]] -->
-<!-- ``` -->
+### Total Sleep Time {.smaller}
 
-<!-- ### Logistic Regression -->
+```{r}
+tst_plots[[1]]
 
-<!-- ```{r} -->
-<!-- lr_plots[[1]] -->
+tst_plots[[2]]
 
-<!-- lr_plots[[2]] -->
+tst_plots[[3]]
 
-<!-- lr_plots[[3]] -->
+tst_plots[[4]]
+```
 
-<!-- lr_plots[[4]] -->
+### Sleep Effeciency {.smaller}
 
-<!-- lr_plots[[5]] -->
-<!-- ``` -->
+```{r}
+se_plots[[1]]
 
-<!-- ### Neural Net -->
+se_plots[[2]]
 
-<!-- ```{r} -->
-<!-- nn_plots[[1]] -->
+se_plots[[3]]
 
-<!-- nn_plots[[2]] -->
+se_plots[[4]]
+```
 
-<!-- nn_plots[[3]] -->
+### Latency Until Persistant Sleep {.smaller}
 
-<!-- nn_plots[[4]] -->
+```{r}
+lps_plots[[1]]
 
-<!-- nn_plots[[5]] -->
-<!-- ``` -->
+lps_plots[[2]]
 
-<!-- ### XGBoost -->
+lps_plots[[3]]
 
-<!-- ```{r} -->
-<!-- xg_plots[[1]] -->
+lps_plots[[4]]
+```
 
-<!-- xg_plots[[2]] -->
+### Wake After Sleep Onset {.smaller}
 
-<!-- xg_plots[[3]] -->
+```{r}
+waso_plots[[1]]
 
-<!-- xg_plots[[4]] -->
+waso_plots[[2]]
 
-<!-- xg_plots[[5]] -->
-<!-- ``` -->
+waso_plots[[3]]
 
-<!-- ::: -->
+waso_plots[[4]]
+```
 
-<!-- ## Bland-Altman Analysis {transition="convex" .H2} -->
+:::
 
-<!-- ```{r} -->
-<!-- source(here::here("presentation/code/table_ba.R")) -->
+## Bland-Altman Analysis {transition="convex" .H2}
 
-<!-- tab_ba -->
-<!-- ``` -->
+```{r}
+source(here::here("presentation/code/table_ba.R"))
 
-<!-- ## In-bed classification flow {transition="convex" .H2} -->
+tab_ba_crude
+```
 
-<!-- ```{r} -->
-<!-- in_bed_tree <- -->
-<!--   read_rds(here::here("data/models/fitted_models/in_bed_simple_CART_fit.rds")) -->
+## In-bed classification flow {transition="convex" .H2}
 
-<!-- in_bed_tree |> -->
-<!--   extract_fit_engine() |> -->
-<!--   visTree( -->
-<!--     height = "550px", -->
-<!--     nodesPopSize = TRUE, -->
-<!--     minNodeSize = 10, -->
-<!--     maxNodeSize = 30, -->
-<!--     fallenLeaves = TRUE, -->
-<!--     legend = FALSE, -->
-<!--     digits = 3, -->
-<!--     direction = "UD", -->
-<!--     colorVar = c( -->
-<!--       "#CC5500", "#8B0000", "#FFD700", -->
-<!--       "#228B22", "#654321", "#8B0000", "#FFDB58" -->
-<!--     ), -->
-<!--     export = FALSE -->
-<!--   ) |> -->
-<!--   visNodes(font = list( -->
-<!--     color = "#EEE8D5", -->
-<!--     size = 20, -->
-<!--     strokeWidth = 1 -->
-<!--   )) |> -->
-<!--   visEdges(font = list( -->
-<!--     color = "#EEE8D5", -->
-<!--     size = 20, -->
-<!--     strokeWidth = 1 -->
-<!--   )) |> -->
-<!--   visInteraction( -->
-<!--     dragNodes = FALSE, -->
-<!--     dragView = FALSE, -->
-<!--     zoomView = FALSE -->
-<!--   ) -->
-<!-- ``` -->
+```{r}
+read_rds(here::here("/media/esbenlykke/My Passport/crude/fitted_models/axed_models/decision_tree_in_bed_AXED.rds")) %>%
+  extract_fit_engine() |>
+  visTree(
+    height = "550px",
+    nodesPopSize = TRUE,
+    minNodeSize = 10,
+    maxNodeSize = 30,
+    fallenLeaves = TRUE,
+    legend = FALSE,
+    digits = 3,
+    direction = "UD",
+    colorVar = c(
+      "#CC5500", "#8B0000", "#FFD700",
+      "#228B22", "#654321", "#8B0000", "#FFDB58"
+    ),
+    export = FALSE
+  ) |>
+  visNodes(font = list(
+    color = "#EEE8D5",
+    size = 20,
+    strokeWidth = 1
+  )) |>
+  visEdges(font = list(
+    color = "#EEE8D5",
+    size = 20,
+    strokeWidth = 1
+  )) |>
+  visInteraction(
+    dragNodes = FALSE,
+    dragView = FALSE,
+    zoomView = FALSE
+  )
+```
 
-<!-- --- -->
+---
 
-<!-- ## Sleep classification flow {transition="convex" .H2} -->
+## Sleep classification flow {transition="convex" .H2}
 
-<!-- ```{r} -->
-<!-- sleep_tree <- -->
-<!--   read_rds(here::here("data/models/fitted_models/sleep_simple_CART_fit.rds")) -->
+```{r}
+read_rds(here::here("/media/esbenlykke/My Passport/crude/fitted_models/axed_models/decision_tree_sleep_AXED.rds")) %>% 
+  extract_fit_engine() |>
+  visTree(
+    height = "550px",
+    nodesPopSize = TRUE,
+    minNodeSize = 10,
+    maxNodeSize = 30,
+    fallenLeaves = TRUE,
+    legend = FALSE,
+    digits = 3,
+    direction = "UD",
+    colorVar = c(
+      "#CC5500", "#8B0000", "#FFD700",
+      "#228B22", "#654321", "#8B0000", "#FFDB58"
+    ),
+    export = FALSE
+  ) |>
+  visNodes(font = list(
+    color = "#EEE8D5",
+    strokeWidth = 1
+  )) |>
+  visEdges(font = list(
+    color = "#EEE8D5",
+    strokeWidth = 1
+  )) |>
+  visInteraction(
+    dragNodes = FALSE,
+    dragView = FALSE,
+    zoomView = FALSE
+  )
+```
 
-<!-- sleep_tree |> -->
-<!--   extract_fit_engine() |> -->
-<!--   visTree( -->
-<!--     height = "550px", -->
-<!--     nodesPopSize = TRUE, -->
-<!--     minNodeSize = 10, -->
-<!--     maxNodeSize = 30, -->
-<!--     fallenLeaves = TRUE, -->
-<!--     legend = FALSE, -->
-<!--     digits = 3, -->
-<!--     direction = "UD", -->
-<!--     colorVar = c( -->
-<!--       "#CC5500", "#8B0000", "#FFD700", -->
-<!--       "#228B22", "#654321", "#8B0000", "#FFDB58" -->
-<!--     ), -->
-<!--     export = FALSE -->
-<!--   ) |> -->
-<!--   visNodes(font = list( -->
-<!--     color = "#EEE8D5", -->
-<!--     strokeWidth = 1 -->
-<!--   )) |> -->
-<!--   visEdges(font = list( -->
-<!--     color = "#EEE8D5", -->
-<!--     strokeWidth = 1 -->
-<!--   )) |> -->
-<!--   visInteraction( -->
-<!--     dragNodes = FALSE, -->
-<!--     dragView = FALSE, -->
-<!--     zoomView = FALSE -->
-<!--   ) -->
-<!-- ``` -->
+# Discussion {transition="convex"}
 
-<!-- ## Discussion {transition="convex"} -->
+## Discussion {transition="convex" .smaller}
 
-<!-- - heteroscedasticity -->
-<!-- - Cheung 2018 table 4: actigraphy provides a sufficiently narrow range of possible mean differences (CI 95%) clinical significant thresholds -->
-<!-- - could be interesting to build models on thigh and hip ocmbined. -->
-<!-- - multiclass vs multilabel classification -->
-<!-- - in-bed awake/sleep is highly imbalanced -> maybe train a new classifier accounting for imbalanced data (SMOTE) -->
-<!-- - model combined preds instead? -->
+- heteroscedasticity
+- Cheung 2018 table 4: actigraphy provides a sufficiently narrow range of possible mean differences (CI 95%) clinical significant thresholds
+- could be interesting to build models on thigh and hip ocmbined.
+- multiclass vs multilabel classification
+- in-bed awake/sleep is highly imbalanced -> maybe train a new classifier accounting for imbalanced data (SMOTE)
+- model combined preds instead?
 

### 2023-03-24 code/evaluate_models/create_stats_all_models.R

@@ -30,8 +30,10 @@ zm_stats <-
   select(id, day, month, spt_hrs, tst_hrs, se_percent, lps_min, waso_min) |>
   rename_with(.cols = -c(id, day, month), ~ paste0("zm_", .))
 
+
 # create stats ------------------------------------------------------------
 
+### Crude
 create_crude_stats <- function(tbl) {
   tbl %>%
     mutate(across(c(pred_in_bed, pred_sleep), as.integer) - 1) |>
@@ -54,9 +56,53 @@ create_crude_stats <- function(tbl) {
     inner_join(zm_stats, by = c("id", "noon_day" = "day", "month"))
 }
 
-preds_crude %>% 
-  group_by(model) %>% 
-  group_modify(~ create_crude_stats(.x), .keep = TRUE) %>% 
+preds_crude %>%
+  group_by(model) %>%
+  group_modify(~ create_crude_stats(.x), .keep = TRUE) %>%
   write_parquet("data/processed/crude_stats.parquet")
 
 ### binary relevance
+
+### Multiclass
+create_multi_stats <- function(tbl){
+  tbl %>%
+  mutate(
+    month = month(datetime - hours(12)),
+    in_bed_asleep_72 = slide_sum(.pred_class == "in_bed_asleep", before = 36, after = 36),
+    in_bed_awake_72 = slide_sum(.pred_class == "in_bed_awake", before = 36, after = 36),
+    out_bed_awake_72 = slide_sum(.pred_class == "out_bed_awake", before = 36, after = 36),
+    in_bed = if_else(in_bed_asleep_72 >= 60 | in_bed_awake_72 >= 60, 1, 0)
+  ) %>%
+  group_by(id, noon_day, month) %>% 
+  summarise(
+    spt_hrs = ((max(row_number()[in_bed == 1] + 30) - min(row_number()[in_bed == 1] - 30)) * 10) / 60 / 60,
+    tst_hrs = (sum(in_bed_asleep_72 >= 60) * 10) / 60 / 60,
+    se_percent = 100 * (tst_hrs / spt_hrs),
+    lps_min = (((min(row_number()[in_bed_asleep_72 == 60] + 30)) - min(row_number()[in_bed_awake_72 == 60] - 30)) * 10) / 60,
+    waso_min = ((sum(.pred_class == "in_bed_awake") - lps_min) * 10) / 60,
+    .groups = "drop"
+  ) %>% 
[…]

### 2023-03-24 code/evaluate_models/minimal_detectable_change.R

@@ -1,0 +1,58 @@
+library(tidyverse)
+library(arrow)
+library(lme4)
+library(car)
+
+crude_stats <-
+  read_parquet("data/processed/crude_stats.parquet") %>%
+  mutate(
+    id = as_factor(id)
+  ) %>%
+  split(.$model) %>%
+  map(drop_na)
+
+crude_dc_data <-
+  crude_stats[[1]] %>% 
+  ungroup()
+
+library(lme4)
+library(car)
+
+
+sleep_data <- 
+  crude_dc_data %>% 
+  select(id, spt_hrs)
+
+# Fit a linear mixed effects model to estimate the within-subject standard deviation for each subject
+sleep_model <- lmer(spt_hrs ~ (1 | id), data = sleep_data)
+
+# Extract sd from model
+sd <- sigma(sleep_model)
+
+# Perform Levene's test
+levene_test <- leveneTest(resid(sleep_model) ~ sleep_data$id) %>% broom::tidy()
+
+# Levenes test is non sign. No need to use ICC
+
+# Calculate SEM
+n <- length(unique(sleep_data$id)) # number of groups
+k <- length(sleep_data$spt_hrs) / n # sample size per group
+SEM <- sd / sqrt(n * k)
+
+# MDC
+MDC <- 1.96 * SEM * sqrt(2)
+MDC * 60
+
+# 95%CI
+# Calculate degrees of freedom
+df <- n - 1
+
+# Calculate critical value for alpha = 0.05
+t_critical <- qt(1 - 0.05/2, df)
+
+# Calculate confidence interval
+CI <- c(MDC - t_critical * SEM * sqrt(2) * 60, MDC + t_critical * SEM * sqrt(2) * 60)
+
+# Print MDC and CI
+cat("MDC:", round(MDC, 2), "\n")
+cat("CI:", round(CI[1], 2), "-", round(CI[2], 2))

### 2023-03-24 presentation/code/plot_ba_metric-wise.R

@@ -1,0 +1,233 @@
+library(tidyverse)
+library(arrow)
+library(patchwork)
+library(showtext)
+
+
+crude_me_ba <-
+  read_csv(here::here("data/processed/crude_mixed_effect_ba.csv"))
+
+crude_stats <-
+  read_parquet(here::here("data/processed/crude_stats.parquet"))
+
+
+get_diff_stats <-
+  function(tbl) {
+    tbl |>
+      mutate(
+        diff_spt_hrs = spt_hrs - zm_spt_hrs,
+        diff_tst_hrs = tst_hrs - zm_tst_hrs,
+        diff_se_percent = se_percent - zm_se_percent,
+        diff_lps_min = lps_min - zm_lps_min,
+        diff_waso_min = waso_min - zm_waso_min
+      ) |>
+      rowwise() |>
+      mutate(
+        avg_spt_hrs = mean(c(spt_hrs, zm_spt_hrs)),
+        avg_tst_hrs = mean(c(tst_hrs, zm_tst_hrs)),
+        avg_se_percent = mean(c(se_percent, zm_se_percent)),
+        avg_lps_min = mean(c(lps_min, zm_lps_min)),
+        avg_waso_min = mean(c(waso_min, zm_waso_min))
+      ) |>
+      ungroup()
+  }
+
+x <- get_diff_stats(crude_stats) %>%
+  pivot_longer(contains("diff"), names_to = "type_diff", values_to = "all_diffs") %>%
+  select(model, type_diff, all_diffs) %>%
+  split(.$type_diff)
+
+y <- get_diff_stats(crude_stats) %>%
+  pivot_longer(contains("avg"), names_to = "type_avg", values_to = "all_avgs") %>%
+  select(type_avg, all_avgs) %>%
+  split(.$type_avg)
+
+crude_all_diffs <-
+  map2(x, y, ~ bind_cols(.x, .y)) %>%
+  map(~ group_split(.x, model)) %>%
+  set_names(c("lps", "se", "spt", "tst", "waso"))
+
+
+
+h_lines <-
+  crude_me_ba %>%
+  split(.$type) %>%
+  map(~ .x %>%
+    select(ba_metric, estimate, model) |>
+    pivot_wider(names_from = ba_metric, values_from = estimate) |>
+    group_split(model))
+
+ci_lines <-
+  crude_me_ba %>%
+  split(.$type) %>%
+  map(~ .x |>
+    select(ba_metric, model, lower_ci:upper_ci) |>
[…]

### 2023-03-31 code/evaluate_models/table_descriptives_of_sleep_stats.R
we do rm cors instead

@@ -1,9 +1,9 @@
 library(lme4)
-library(simr)
 library(tidyverse)
 library(arrow)
 library(gt)
-
+library(showtext)
+library(gtExtras)
 
 crude_stats <-
   read_parquet(here::here("data/processed/crude_stats.parquet"))
@@ -23,6 +23,9 @@ get_diff_stats <-
   }
 
 diffs <- get_diff_stats(crude_stats)
+
+
+### Means and SDs ###
 
 spt <-
   diffs %>%
@@ -134,73 +137,286 @@ waso <-
 
 
 ### Covariance between repeated measures ###
-model_outcomes <- c(
-  "spt_hrs", "tst_hrs", "se_percent", "lps_min", "waso_min"
-)
-
-zm_outcomes <-
-  c("zm_spt_hrs", "zm_tst_hrs", "zm_se_percent", "zm_lps_min", "zm_waso_min")
+# model_outcomes <- c(
+#   "spt_hrs", "tst_hrs", "se_percent", "lps_min", "waso_min"
+# )
+#
+# zm_outcomes <-
+#   c("zm_spt_hrs", "zm_tst_hrs", "zm_se_percent", "zm_lps_min", "zm_waso_min")
 
 # covariance values between the repeated measures for each outcome, grouped by model.
-covariance <-
-  diffs %>%
-  select(model, id, noon_day, all_of(model_outcomes)) %>%
-  drop_na() %>%
-  pivot_longer(cols = all_of(model_outcomes), names_to = "outcome", values_to = "value") %>%
-  group_by(id, outcome, model) %>%
-  arrange(model, id, outcome, noon_day) %>%
-  mutate(value_diff = value - lag(value)) %>%
-  filter(!value_diff == "NaN" & !value == "Inf" & !value == "-Inf") %>%
-  group_by(model, outcome) %>%
-  filter((outcome == "lps_min" & value_diff < 100) |
-    (outcome == "waso_min" & value_diff < 100) |
-    (outcome == "spt_hrs" & value_diff < 5) |
-    (outcome == "tst_hrs" & value_diff < 5) |
-    (outcome == "se_percent" & value_diff < 30)) %>%
-  summarize(
-    mean_cov = cov(noon_day, value_diff, use = "pairwise.complete.obs"),
-    .groups = "drop"
-  ) %>%
-  pivot_wider(names_from = outcome, values_from = mean_cov) %>%
-  rename(method = model) %>%
[…]

### 2023-04-28 presentation/index.qmd
we do rm cors instead
plot roc for in-bed

@@
     preview-links: auto
-    logo: images/SDU_WHITE_RGB-png.png
+    logo: visuals/SDU_WHITE_RGB-png.png
     footer: https://github.com/esbenlykke/sleep_study
@@
 
-## building Models {transition="convex"}
-
-```{r}
-source(here::here("presentation/code/plot_hyperparameter_tuning.R"))
-
-hyper_plot
-```
-
 ## Estimate Sleep Quality Metrics {transition="fade"}
@@
 
+## Sequential Classifiers
+
+![](visuals/chained_models_data_flow.png)
+
 # Results {transition="fade"}
@@
 
-## Many ROC Curves {transition="convex" .H2}
+## In-bed models {transition="convex"}
 
-## {transition="convex"}
+![](visuals/plot_hyper_10.png)
 
+## In-bed performance of decision tree model {transition="convex"}
+
+:::: {.columns}
+
+::: {.column width="60%"}
+![](visuals/plot_in_bed_roc.png)
+:::
+
+::: {.column width="40%"}
 ```{r}
-source(here::here("presentation/code/plot_roc_curves.R"))
+source(here::here("presentation/code/tables_performance.R"))
 
-multi_roc_plot
+tab_in_bed
 ```
+:::
 
-## {transition="convex"}
-
-```{r}
-three_classifiers_roc_plot
-```
-
-## {transition="convex"}
-
-```{r}
-#| fig-width: 8
-#| fig-align: center
-two_classifiers_roc_plot
-```
+::::
 
@@
 
-### Sleep Period Time {.smaller}
+<!-- ### Sleep Period Time {.smaller} -->
 
-```{r}
-#| echo: false
-source(here::here("presentation/code/plot_ba_metric-wise.R"))
+<!-- ```{r} -->
+<!-- #| echo: false -->
+<!-- source(here::here("presentation/code/plot_ba_metric-wise.R")) -->
 
-spt_plots[[1]]
+<!-- spt_plots[[1]] -->
 
-spt_plots[[2]]
+<!-- spt_plots[[2]] -->
 
-spt_plots[[3]]
+<!-- spt_plots[[3]] -->
 
-spt_plots[[4]]
-```
+<!-- spt_plots[[4]] -->
+<!-- ``` -->
 
-### Total Sleep Time {.smaller}
+<!-- ### Total Sleep Time {.smaller} -->
 
-```{r}
-tst_plots[[1]]
+<!-- ```{r} -->
+<!-- tst_plots[[1]] -->
 
-tst_plots[[2]]
+<!-- tst_plots[[2]] -->
 
-tst_plots[[3]]
+<!-- tst_plots[[3]] -->
 
-tst_plots[[4]]
-```
+<!-- tst_plots[[4]] -->
+<!-- ``` -->
 
-### Sleep Effeciency {.smaller}
+<!-- ### Sleep Effeciency {.smaller} -->
 
-```{r}
-se_plots[[1]]
+<!-- ```{r} -->
+<!-- se_plots[[1]] -->
 
-se_plots[[2]]
+<!-- se_plots[[2]] -->
 
-se_plots[[3]]
+<!-- se_plots[[3]] -->
 
-se_plots[[4]]
-```
+<!-- se_plots[[4]] -->
+<!-- ``` -->
 
-### Latency Until Persistant Sleep {.smaller}
+<!-- ### Latency Until Persistant Sleep {.smaller} -->
 
-```{r}
-lps_plots[[1]]
+<!-- ```{r} -->
+<!-- lps_plots[[1]] -->
 
-lps_plots[[2]]
+<!-- lps_plots[[2]] -->
 
-lps_plots[[3]]
+<!-- lps_plots[[3]] -->
 
-lps_plots[[4]]
-```
+<!-- lps_plots[[4]] -->
+<!-- ``` -->
 
-### Wake After Sleep Onset {.smaller}
+<!-- ### Wake After Sleep Onset {.smaller} -->
 
-```{r}
-waso_plots[[1]]
+<!-- ```{r} -->
+<!-- waso_plots[[1]] -->
 
-waso_plots[[2]]
+<!-- waso_plots[[2]] -->
 
-waso_plots[[3]]
+<!-- waso_plots[[3]] -->
 
-waso_plots[[4]]
-```
+<!-- waso_plots[[4]] -->
+<!-- ``` -->
 
-:::
+<!-- ::: -->
 
-## Bland-Altman Analysis {transition="convex" .H2}
+<!-- ## Bland-Altman Analysis {transition="convex" .H2} -->
 
-```{r}
-source(here::here("presentation/code/table_ba.R"))
+<!-- ```{r} -->
+<!-- source(here::here("presentation/code/table_ba.R")) -->
 
-tab_ba_crude
-```
+<!-- tab_ba_crude -->
+<!-- ``` -->
 
-## In-bed classification flow {transition="convex" .H2}
+<!-- ## Descriptives of Sleep Quality Statistics Across Methods {transition="convex" .H2 .smaller} -->
 
-```{r}
-read_rds(here::here("/media/esbenlykke/My Passport/crude/fitted_models/axed_models/decision_tree_in_bed_AXED.rds")) %>%
-  extract_fit_engine() |>
-  visTree(
-    height = "550px",
-    nodesPopSize = TRUE,
-    minNodeSize = 10,
-    maxNodeSize = 30,
-    fallenLeaves = TRUE,
-    legend = FALSE,
-    digits = 3,
-    direction = "UD",
-    colorVar = c(
-      "#CC5500", "#8B0000", "#FFD700",
-      "#228B22", "#654321", "#8B0000", "#FFDB58"
-    ),
-    export = FALSE
-  ) |>
-  visNodes(font = list(
-    color = "#EEE8D5",
-    size = 20,
-    strokeWidth = 1
-  )) |>
-  visEdges(font = list(
-    color = "#EEE8D5",
-    size = 20,
-    strokeWidth = 1
-  )) |>
-  visInteraction(
-    dragNodes = FALSE,
-    dragView = FALSE,
-    zoomView = FALSE
-  )
-```
+<!-- ```{r} -->
+<!-- source(here::here("code/evaluate_models/table_descriptives_of_sleep_stats.R")) -->
 
----
+<!-- tab_descriptives -->
+<!-- ``` -->
 
-## Sleep classification flow {transition="convex" .H2}
+<!-- ## In-bed classification flow {transition="convex" .H2} -->
 
-```{r}
-read_rds(here::here("/media/esbenlykke/My Passport/crude/fitted_models/axed_models/decision_tree_sleep_AXED.rds")) %>% 
-  extract_fit_engine() |>
-  visTree(
-    height = "550px",
-    nodesPopSize = TRUE,
-    minNodeSize = 10,
-    maxNodeSize = 30,
-    fallenLeaves = TRUE,
-    legend = FALSE,
-    digits = 3,
-    direction = "UD",
-    colorVar = c(
-      "#CC5500", "#8B0000", "#FFD700",
-      "#228B22", "#654321", "#8B0000", "#FFDB58"
-    ),
-    export = FALSE
-  ) |>
-  visNodes(font = list(
-    color = "#EEE8D5",
-    strokeWidth = 1
-  )) |>
-  visEdges(font = list(
-    color = "#EEE8D5",
-    strokeWidth = 1
-  )) |>
-  visInteraction(
-    dragNodes = FALSE,
-    dragView = FALSE,
-    zoomView = FALSE
-  )
-```
+<!-- ```{r} -->
+<!-- read_rds(here::here("/media/esbenlykke/My Passport/crude/fitted_models/axed_models/decision_tree_in_bed_AXED.rds")) %>% -->
+<!--   extract_fit_engine() |> -->
+<!--   visTree( -->
+<!--     height = "550px", -->
+<!--     nodesPopSize = TRUE, -->
+<!--     minNodeSize = 10, -->
+<!--     maxNodeSize = 30, -->
+<!--     fallenLeaves = TRUE, -->
+<!--     legend = FALSE, -->
+<!--     digits = 3, -->
+<!--     direction = "UD", -->
+<!--     colorVar = c( -->
+<!--       "#CC5500", "#8B0000", "#FFD700", -->
+<!--       "#228B22", "#654321", "#8B0000", "#FFDB58" -->
+<!--     ), -->
+<!--     export = FALSE -->
+<!--   ) |> -->
+<!--   visNodes(font = list( -->
+<!--     color = "#EEE8D5", -->
+<!--     size = 20, -->
+<!--     strokeWidth = 1 -->
+<!--   )) |> -->
+<!--   visEdges(font = list( -->
+<!--     color = "#EEE8D5", -->
+<!--     size = 20, -->
+<!--     strokeWidth = 1 -->
+<!--   )) |> -->
+<!--   visInteraction( -->
+<!--     dragNodes = FALSE, -->
+<!--     dragView = FALSE, -->
+<!--     zoomView = FALSE -->
+<!--   ) -->
+<!-- ``` -->
 
-# Discussion {transition="convex"}
+<!-- --- -->
 
-## Discussion {transition="convex" .smaller}
+<!-- ## Sleep classification flow {transition="convex" .H2} -->
 
-- heteroscedasticity
-- Cheung 2018 table 4: actigraphy provides a sufficiently narrow range of possible mean differences (CI 95%) clinical significant thresholds
-- could be interesting to build models on thigh and hip ocmbined.
-- multiclass vs multilabel classification
-- in-bed awake/sleep is highly imbalanced -> maybe train a new classifier accounting for imbalanced data (SMOTE)
-- model combined preds instead?
+<!-- ```{r} -->
+<!-- read_rds(here::here("/media/esbenlykke/My Passport/crude/fitted_models/axed_models/decision_tree_sleep_AXED.rds")) %>%  -->
+<!--   extract_fit_engine() |> -->
+<!--   visTree( -->
+<!--     height = "550px", -->
+<!--     nodesPopSize = TRUE, -->
+<!--     minNodeSize = 10, -->
+<!--     maxNodeSize = 30, -->
+<!--     fallenLeaves = TRUE, -->
+<!--     legend = FALSE, -->
+<!--     digits = 3, -->
+<!--     direction = "UD", -->
+<!--     colorVar = c( -->
+<!--       "#CC5500", "#8B0000", "#FFD700", -->
+<!--       "#228B22", "#654321", "#8B0000", "#FFDB58" -->
+<!--     ), -->
+<!--     export = FALSE -->
+<!--   ) |> -->
+<!--   visNodes(font = list( -->
+<!--     color = "#EEE8D5", -->
+<!--     strokeWidth = 1 -->
+<!--   )) |> -->
+<!--   visEdges(font = list( -->
+<!--     color = "#EEE8D5", -->
+<!--     strokeWidth = 1 -->
+<!--   )) |> -->
+<!--   visInteraction( -->
+<!--     dragNodes = FALSE, -->
+<!--     dragView = FALSE, -->
+<!--     zoomView = FALSE -->
+<!--   ) -->
+<!-- ``` -->
 
-## References {transition="convex"}
+<!-- # Discussion {transition="convex"} -->
+
+<!-- ## Discussion {transition="convex" .smaller} -->
+
+<!-- - heteroscedasticity -->
+<!-- - Cheung 2018 table 4: actigraphy provides a sufficiently narrow range of possible mean differences (CI 95%) clinical significant thresholds -->
+<!-- - could be interesting to build models on thigh and hip ocmbined. -->
+<!-- - multiclass vs multilabel classification -->
+<!-- - in-bed awake/sleep is highly imbalanced -> maybe train a new classifier accounting for imbalanced data (SMOTE) -->
+<!-- - model combined preds instead? -->
+
+<!-- ## References {transition="convex"} -->

### 2023-04-28 code/create_models/chained_models/finalize_sleep_workflows.R

@@ -52,11 +52,17 @@ finalize_and_fit_best_wf <- function(grid_results_fnames, training_data) {
 }
 
 # List all grid result files in the specified directory
-grid_results_fnames <- 
-  list.files("/media/esbenlykke/My Passport/chained_models/grid_results/", full.names = TRUE) %>% 
-  str_subset(".rds")
+grid_results_fnames_30 <- 
+  list.files("/media/esbenlykke/My Passport/chained_models/grid_results", full.names = TRUE) %>% 
+  str_subset("30.rds")
 
 # Apply the finalize_and_fit_best_wf function to each grid result file and the training data
-walk(grid_results_fnames, ~ finalize_and_fit_best_wf(.x, train_30), 
+walk(grid_results_fnames_30, ~ finalize_and_fit_best_wf(.x, train_30), 
      .progress = TRUE)
 
+grid_results_fnames_10 <- 
+  list.files("/media/esbenlykke/My Passport/chained_models/grid_results", full.names = TRUE) %>% 
+  str_subset("10.rds")
+
+walk(grid_results_fnames_10, ~ finalize_and_fit_best_wf(.x, train_10), 
+     .progress = TRUE)

### 2023-04-28 code/evaluate_models/create_roc_data_all_models.R

@@ -16,31 +16,33 @@ get_roc <-
 # CRUDE -------------------------------------------------------------------
 
 
-in_bed_fits_files <- list.files("/media/esbenlykke/My Passport/crude/fitted_models/axed_models",
-  full.names = TRUE
-) %>%
-  str_subset("in_bed")
+fits_10_paths <- 
+  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows",
+  full.names = TRUE, recursive = TRUE
+) %>% 
+  str_subset("10_sec")
+
+fits_30_paths <- 
+  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows",
+             full.names = TRUE, recursive = TRUE
+  ) %>% 
+  str_subset("30_sec")
 
 
-sleep_fits_files <- list.files("/media/esbenlykke/My Passport/crude/fitted_models/axed_models",
-  full.names = TRUE
-) %>%
-  str_subset("sleep")
-
-in_bed_fits <- map(in_bed_fits_files, read_rds) |>
+fits_10 <- map(fits_10_paths, read_rds) |>
   set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
 
-sleep_fits <- map(sleep_fits_files, read_rds) |>
-  set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
+fits_30 <- map(fits_30_paths, read_rds) 
 
-test <- read_parquet("data/data_for_modelling/crude_testing_data.parquet")
+test <- read_parquet("data/data_for_modelling/chained_classifiers/testing_30_sec_data.parquet") %>% 
+  mutate(across(contains("sleep"), as_factor))
 
 
-map_dfr(in_bed_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
+map_dfr(fits_10, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
   write_parquet("data/processed/crude_roc_data_in_bed.parquet")
 
-map_dfr(sleep_fits, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
-  write_parquet("data/processed/crude_roc_data_sleep.parquet")
+map_dfr(fits_30, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
+  write_parquet("data/processed/")
 
 
 # Binary relevance --------------------------------------------------------
[…]

### 2023-04-28 presentation/code/plot_hyperparameter_tuning.R

@@ -1,11 +1,20 @@
+#!/usr/bin/env Rscript
+
 library(tidyverse)
 library(tidymodels)
+library(arrow)
+library(patchwork)
 
-data <- 
-  read_rds(here::here("presentation/data/wfset_example.rds"))
 
-hyper_plot <- 
-  data |> 
+in_bed_10 <- 
+  read_rds("/media/esbenlykke/My Passport/chained_models/grid_results/in_bed/in_bed_workflowsets_results_10_sec.rds")
+
+in_bed_30 <- 
+  read_rds("/media/esbenlykke/My Passport/chained_models/grid_results/in_bed/in_bed_workflowsets_results_30_sec.rds")
+
+
+hyper_plot_10 <- 
+  in_bed_10 |> 
   autoplot(
     rank_metric = "f_meas",  # <- how to order models
     metric = "f_meas",       # <- which metric to visualize
@@ -15,14 +24,14 @@ hyper_plot <-
   # lims(y = c(.96, 1)) +
   labs(
     title = "Hyperparameter Tuning",
-    subtitle = "5-fold cross validation using 10-level grid search",
+    subtitle = "5-fold cross validation using 5-level grid search",
     color = "Model Type",
     shape = NULL,
     y = "F1 Metric",
-    caption = "Note: this is based on simulated data and is not representative of the actual training data from the study"
+    caption = "Note: the decision tree model was chosen to extract in-bed time."
   ) +
   scale_color_brewer(palette = "Dark2",
-                     label = c("XGboost", "Neural Net", "Multinomial Logistic Regression", "Support Vector Machine")) +
+                     label = c("XGBoost", "Neural Net", "Decision Tree", "Logistic Regression")) +
   theme(
     text = element_text(color = "#EEE8D5", family = "ibm"),
     axis.title = element_text(family = "ibm", size = 20, color = "#EEE8D5"),
@@ -41,3 +50,46 @@ hyper_plot <-
     legend.position = "bottom"
   ) +
   guides(shape = "none")
+
+ggsave("presentation/visuals/plot_hyper_10.png", width = 14, height = 6)
+
+
+hyper_plot_30 <- 
+  in_bed_30 |> 
+  autoplot(
+    rank_metric = "f_meas",  # <- how to order models
+    metric = "f_meas",       # <- which metric to visualize
+    select_best = FALSE     # <- one point per workflow
+  ) +
+  # geom_text(aes(y = mean - .003, label = wflow_id), angle = 90, hjust = 1) +
[…]

### 2023-05-02 code/create_models/chained_models/finalize_sleep_workflows.R
create performance metrics for sequential models

@@ -57,8 +57,8 @@ grid_results_fnames_30 <-
   str_subset("30.rds")
 
 # Apply the finalize_and_fit_best_wf function to each grid result file and the training data
-walk(grid_results_fnames_30, ~ finalize_and_fit_best_wf(.x, train_30), 
-     .progress = TRUE)
+# walk(grid_results_fnames_30, ~ finalize_and_fit_best_wf(.x, train_30), 
+#      .progress = TRUE)
 
 grid_results_fnames_10 <- 
   list.files("/media/esbenlykke/My Passport/chained_models/grid_results", full.names = TRUE) %>% 

### 2023-05-02 code/evaluate_models/get_metrics_and_create_table_from_in_bed_models.R

@@ -1,0 +1,95 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+library(gt)
+library(showtext)
+
+# Define a set of metrics to be used for evaluation
+my_metrics <-
+  metric_set(f_meas, accuracy, sensitivity, precision, specificity)
+
+# Function to compute the specified performance metrics using the provided model and test data
+get_metrics <-
+  function(fit, test_data, truth, estimate, estimator = "binary") {
+    fit |>
+      augment(test_data) |>
+      my_metrics(
+        truth = {{ truth }}, estimate = {{ estimate }},
+        event_level = "second",
+        estimator = estimator
+      )
+  }
+
+# Load pre-trained models for 10-second and 30-second epochs
+in_bed_fit_10 <-
+  read_rds("/media/esbenlykke/My Passport/chained_models/fitted_workflows/in_bed/in_bed_simple_tree_fit_10_sec_AXED.rds")
+
+in_bed_fit_30 <-
+  read_rds("/media/esbenlykke/My Passport/chained_models/fitted_workflows/in_bed/in_bed_simple_tree_fit_30_sec_AXED.rds")
+
+# Load test data for 10-second and 30-second epochs and convert 'in_bed' column to factors
+test_10 <-
+  read_parquet("data/data_for_modelling/chained_classifiers/testing_10_sec_data.parquet") %>%
+  mutate(
+    in_bed = as_factor(in_bed)
+  )
+
+test_30 <-
+  read_parquet("data/data_for_modelling/chained_classifiers/testing_30_sec_data.parquet") %>%
+  mutate(
+    in_bed = as_factor(in_bed)
+  )
+
+# Calculate performance metrics for both models using the test datasets
+metrics <- 
+  get_metrics(in_bed_fit_10, test_10, in_bed, .pred_class) %>%
+  mutate(epoch_length = "10 sec") %>%
+  bind_rows(
+    get_metrics(in_bed_fit_30, test_30, in_bed, .pred_class) %>%
+      mutate(
+        epoch_length = "30 sec"
+      )
+  ) %>% 
+  select(-.estimator)
+
+# Save metrics as CSV
+# metrics %>% 
+#   write_csv("presentation/data/in_bed_metrics.csv")
+
+# Create a formatted table to display the metrics, grouped by epoch length
+tab_in_bed <- 
+  metrics %>% 
+  mutate(
+    .metric = factor(.metric,
[…]

### 2023-05-02 code/evaluate_models/get_metrics_from_in_bed_models.R

@@ -1,114 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(tidymodels)
-library(arrow)
-library(gt)
-library(showtext)
-
-my_metrics <-
-  metric_set(f_meas, accuracy, sensitivity, precision, specificity)
-
-get_metrics <-
-  function(fit, test_data, truth, estimate, estimator = "binary") {
-    fit |>
-      augment(test_data) |>
-      my_metrics(
-        truth = {{ truth }}, estimate = {{ estimate }},
-        event_level = "second",
-        estimator = estimator
-      )
-  }
-
-
-
-in_bed_fit_10 <-
-  read_rds("/media/esbenlykke/My Passport/chained_models/fitted_workflows/in_bed/in_bed_simple_tree_fit_10_sec_AXED.rds")
-
-in_bed_fit_30 <-
-  read_rds("/media/esbenlykke/My Passport/chained_models/fitted_workflows/in_bed/in_bed_simple_tree_fit_30_sec_AXED.rds")
-
-test_10 <-
-  read_parquet("data/data_for_modelling/chained_classifiers/testing_10_sec_data.parquet") %>%
-  mutate(
-    in_bed = as_factor(in_bed)
-  )
-
-test_30 <-
-  read_parquet("data/data_for_modelling/chained_classifiers/testing_30_sec_data.parquet") %>%
-  mutate(
-    in_bed = as_factor(in_bed)
-  )
-
-metrics <- 
-  get_metrics(in_bed_fit_10, test_10, in_bed, .pred_class) %>%
-  mutate(epoch_length = "10 sec") %>%
-  bind_rows(
-    get_metrics(in_bed_fit_30, test_30, in_bed, .pred_class) %>%
-      mutate(
-        epoch_length = "30 sec"
-      )
-  ) %>% 
-  select(-.estimator)
-
-metrics %>% 
-  write_csv("presentation/data/in_bed_metrics.csv")
-
-tab_in_bed <- 
-  metrics %>% 
-  mutate(
-    .metric = factor(.metric,
-                     levels = c("f_meas", "accuracy", "sensitivity", "precision", "specificity"),
-                     labels = c("F1 Score", "Accuracy", "Sensitivity", "Precision", "Specificity")
-    )
-  ) |> 
-  pivot_wider(names_from = epoch_length, values_from = .estimate) %>% 
-  gt() |>
-  fmt_percent(
-    columns = 2:3,
[…]

### 2023-05-02 code/evaluate_models/create_roc_data_all_models.R
prepare roc auc data

@@ -6,104 +6,168 @@ library(arrow)
 
 
 get_roc <-
-  function(fit, truth, estimate) {
+  function(fit, test, truth, estimate) {
     fit |>
       augment(test) |>
       roc_curve(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
   }
 
+get_roc_auc <-
+  function(fit, test, truth, estimate) {
+    fit |>
+      augment(test) |>
+      roc_auc(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
+  }
 
-# CRUDE -------------------------------------------------------------------
-
-
-fits_10_paths <- 
-  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows",
-  full.names = TRUE, recursive = TRUE
-) %>% 
-  str_subset("10_sec")
-
-fits_30_paths <- 
-  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows",
-             full.names = TRUE, recursive = TRUE
-  ) %>% 
+## Get the file names of the fitted models for the 10-second epochs, grouped by preprocessing type (raw, median5, and median10)
+fits_10_filenames_raw <- 
+  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/sleep",
+             full.names = TRUE) %>% 
+  str_subset("raw") %>% 
   str_subset("30_sec")
 
 
-fits_10 <- map(fits_10_paths, read_rds) |>
-  set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
-
-fits_30 <- map(fits_30_paths, read_rds) 
-
-test <- read_parquet("data/data_for_modelling/chained_classifiers/testing_30_sec_data.parquet") %>% 
-  mutate(across(contains("sleep"), as_factor))
+fits_10_filenames_median5 <- 
+  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/sleep",
+             full.names = TRUE) %>% 
+  str_subset("5_min") %>% 
+  str_subset("30_sec")
 
 
-map_dfr(fits_10, ~ get_roc(.x, in_bed, .pred_1), .id = "model") |>
-  write_parquet("data/processed/crude_roc_data_in_bed.parquet")
-
[…]

### 2023-05-02 code/evaluate_models/plot_performance_metrics.R

@@ -1,7 +1,0 @@
-#!/usr/bin/env RscriptsuppressMessages(library(tidyverse))
-
-library(tidyverse)
-
-
-metric_data <- read_csv("data/processed/performance_metrics.csv")
-

### 2023-05-02 code/plot_sleep_ROC_data.R

@@ -1,0 +1,73 @@
+#!/usr/bin/env
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+library(patchwork)
+
+
+sleep_roc_data_10 <- 
+  read_parquet("presentation/data/all_roc_sleep_10.parquet") %>% 
+  group_by(model, group) %>% 
+  slice_sample(n = 1000)
+
+sleep_roc_data_30 <- 
+  read_parquet("presentation/data/all_roc_sleep_30.parquet") %>% 
+  group_by(model, group) %>% 
+  slice_sample(n = 1000)
+
+
+
+roc_theme <- 
+  theme(
+    text = element_text(color = "#EEE8D5", family = "ibm"),
+    axis.title = element_text(family = "ibm", size = 16, color = "#EEE8D5"),
+    axis.text = element_text(size = 14, color = "#EEE8D5"),
+    plot.title = element_text(family = "ibm", size = 30, face = "bold", hjust = .5, 
+                              margin = margin(b = 40)),
+    plot.subtitle = element_text(size = 20, hjust = .5),
+    plot.caption = element_text(size = 12),
+    panel.grid.minor = element_blank(),
+    panel.grid = element_line(color = "#293D42"),
+    plot.background = element_rect(fill = "#002B36", color = "#002B36"),
+    panel.background = element_rect(color = "#002B36", fill = "#002B36"),
+    legend.background = element_rect(fill = "#002B36"),
+    legend.text = element_text(size = 16),
+    legend.title = element_text(size = 14, face = "bold"),
+    legend.key = element_rect(fill = "#002B36"),
+    legend.position = "bottom",
+    strip.background = element_rect(fill = "#002B36"),
+    strip.text = element_text(color = "#EEE8D5", face = "bold")
+  )
+
+plot_roc <- function(roc_data, title) {
+  roc_data %>%
+    ggplot(
+      aes(1 - specificity, sensitivity, color = model)
+    ) +
+    geom_line(linewidth = .8, alpha = .7) +
+    geom_abline(slope = 1, intercept = 0, linewidth = 0.4, lty = 2, color = "lightblue") +
+    scale_color_brewer(palette = "Dark2") +
+    scale_x_continuous(breaks = c(0, .5, 1)) +
+    scale_y_continuous(breaks = c(0, .5, 1)) +
+    facet_wrap(~ group) +
[…]

### 2023-05-03 presentation/index.qmd
create performance metrics for sequential models
prepare roc auc data
trying to figure out what to do with sensor problems. Calculate sleep stats accurately

@@
 
-tab_in_bed
+table_in_bed_metrics
 ```
@@
 
-## Lots of Metrics {transition="convex" .H2}
+## Roc curves for sleep models {transition="convex}
+:::: {.columns}
 
-## {transition="convex"}
-::: {layout="[[-1], [1], [-1]]"}
+::: {.column width="60%"}
+![](visuals/plot_sleep_roc.png)
+:::
+
+::: {.column width="40%"}
 ```{r}
-source(here::here("presentation/code/tables_performance.R"))
+source(here::here("presentation/code/table_roc_auc_sleep.R"))
 
-tab_multi_metrics
+gt::as_raw_html(table_sleep_roc_auc)
 ```
@@
 
-## {transition="convex" .H2}
+::::
 
-::: {layout="[[-1], [1], [-1]]"}
-```{r}
-tab_crude_metrics
-```
-:::
-
-## {transition="convex" .H2}
+## {transition="convex"}
 
 ```{r}
-tab_br_metrics
+table_sleep_metrics
 ```

### 2023-05-03 code/create_models/chained_models/predict_and_extract_in_bed_data.R

@@ -3,39 +3,53 @@ library(tidymodels)
 library(arrow)
 library(slider)
 
-
+# Define a function to filter and extract in-bed data based on the fitted model and test data
 extract_in_bed_data <- function(fit, test, epoch_length) {
   fit %>%
     augment(test) %>%
     mutate(
+      # Compute median-filtered in_bed_filtered values for the specified epoch_length
       in_bed_filtered = slider::slide_dbl(as.numeric(.pred_class) - 1, median,
-        .after = (7.5 * 60) / epoch_length, .before = (7.5 * 60) / epoch_length
+                                          .after = (7.5 * 60) / epoch_length, .before = (7.5 * 60) / epoch_length
       ),
     ) %>%
-    group_by(id, noon_day) %>%
+    group_by(id, noon_day, month) %>%
     group_modify(~ .x %>%
-      filter(row_number() > min(row_number()[in_bed_filtered == 1]) &
-        row_number() < max(row_number()[in_bed_filtered == 1]))) %>%
+                   # Filter rows where row_number is within the in_bed_filtered range
+                   filter(row_number() > min(row_number()[in_bed_filtered == 1]) &
+                            row_number() < max(row_number()[in_bed_filtered == 1]))) %>%
     ungroup()
 }
 
-
+# Read the 10-second epoch dataset
 data_10 <-
   read_parquet("data/data_for_modelling/all_data_incl_sensor_independent_features_10_sec_epochs.parquet") 
 
+# Read the 30-second epoch dataset
 data_30 <-
   read_parquet("data/data_for_modelling/all_data_incl_sensor_independent_features_30_sec_epochs.parquet") 
 
+# Read the fitted decision tree model for the 10-second epoch data
 in_bed_CART_fit_10 <-
   read_rds("/media/esbenlykke/My Passport/crude/fitted_models/axed_models/in_bed_simple_tree_fit_10_AXED.rds")
 
+# Read the fitted decision tree model for the 30-second epoch data
 in_bed_CART_fit_30 <-
   read_rds("/media/esbenlykke/My Passport/crude/fitted_models/axed_models/in_bed_simple_tree_fit_30_AXED.rds")
 
-### 10 sec data ###
-extract_in_bed_data(in_bed_CART_fit_10, data_10, 10) %>% 
-  write_parquet("data/data_for_modelling/only_in_bed_data_10.parquet")
+### Process 10-second epoch data ###
 
-### 30 sec data ###
[…]

### 2023-05-03 code/create_models/create_testing_training_data.R

@@ -1,31 +1,0 @@
-library(tidyverse)
-library(tidymodels)
-library(arrow)
-
-
-set.seed(132)
-data <- 
-  read_parquet("data/data_for_modelling/all_data_incl_sensor_independent_features.parquet")
-
-spl <- 
-  group_initial_split(data, group = id, prop = .5)
-
-training(spl) %>% 
-  write_parquet("data/data_for_modelling/crude_training_data.parquet")
-
-testing(spl) %>% 
-  write_parquet("data/data_for_modelling/crude_testing_data.parquet")
-
-# 30 sec data
-data30 <- 
-  read_parquet("data/data_for_modelling/all_data_incl_sensor_independent_features_30_sec_epochs.parquet")
-
-spl30 <- 
-  group_initial_split(data30, group = id, prop = .5)
-
-training(spl30) %>% 
-  write_parquet("data/data_for_modelling/crude_training_30_sec_data.parquet")
-
-testing(spl30) %>% 
-  write_parquet("data/data_for_modelling/crude_testing_30_sec_data.parquet")
-

### 2023-05-03 code/evaluate_models/compare_sleep_stats_zm_raw_vs_filtered.R

@@ -1,0 +1,67 @@
+library(tidyverse)
+library(furrr)
+library(slider)
+library(arrow)
+library(lubridate)
+
+
+# Read in the raw sleep data
+zm_raw <- read_tsv("data/processed/zm_scores.tsv")
+
+# Read in the summary statistics for all participants
+zm_stats <- read_csv("data/processed/all_zm_stats.csv")
+
+# Plot the raw and filtered scores over time
+zm_raw %>%
+  filter(id == 8505 & !score == -5) %>%
+  mutate(
+    score_filter_5 = slide_dbl(score, median, .after = 10),
+    score_filter_10 = slide_dbl(score, median, .after = 20),
+    noon_day = day(datetime - hours(12))
+  ) %>%
+  ggplot() +
+  geom_step(aes(datetime, score), alpha = .5) +
+  geom_step(aes(datetime, score_filter_5 - .2), color = "red") +
+  geom_step(aes(datetime, score_filter_10 - .4), color = "darkgreen") +
+  scale_x_datetime(breaks = "1 hours", date_labels = "%H:%M") +
+  facet_wrap(~noon_day, scales = "free") +
+  theme_light()
+
+# Filter the raw sleep data for ID 8505 and create a new dataset with sleep metrics
+zm_sleep <- zm_raw %>%
+  filter(id == 8505 & !score == -5) %>%
+  mutate(
+    noon_day = day(datetime - hours(12)),
+    month = month(datetime - hours(12)),
+    score_filtered = slide_dbl(score, median, .after = 10), # 5 min window
+    sleep_raw = if_else(score != 0, 1, 0),
+    sleep_filter_5 = if_else(score_filtered != 0, 1, 0),
+    sleep_12_cumsum_raw = slide_dbl(sleep_raw, sum, .after = 24),
+    sleep_12_cumsum_filter_5 = slide_dbl(sleep_filter_5, sum, .after = 24)
+  )
+
+# Define a function to count zeros in runs of at least three consecutive zeros 
+count_each_zero_in_consecutive_zeros <-
+  function(x, n = 3) {
+    rle_x <- rle(x == 0)
+    sum(rle_x$lengths[rle_x$values & rle_x$lengths >= n])
+  }
+
+# Calculate sleep metrics using both filtered scores and raw scores
+zm_sleep %>%
[…]

### 2023-05-03 code/evaluate_models/plot_sleep_ROC_data.R

@@ -1,0 +1,71 @@
+#!/usr/bin/env
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+library(patchwork)
+
+
+sleep_roc_data_10 <- 
+  read_parquet("presentation/data/all_roc_sleep_10.parquet") %>% 
+  group_by(model, group) %>% 
+  slice_sample(n = 1000)
+
+sleep_roc_data_30 <- 
+  read_parquet("presentation/data/all_roc_sleep_30.parquet") %>% 
+  group_by(model, group) %>% 
+  slice_sample(n = 1000)
+
+roc_theme <- 
+  theme(
+    text = element_text(color = "#EEE8D5", family = "ibm"),
+    axis.title = element_text(family = "ibm", size = 16, color = "#EEE8D5"),
+    axis.text = element_text(size = 14, color = "#EEE8D5"),
+    plot.title = element_text(family = "ibm", size = 30, face = "bold", hjust = .5, 
+                              margin = margin(b = 40)),
+    plot.subtitle = element_text(size = 20, hjust = .5),
+    plot.caption = element_text(size = 12),
+    panel.grid.minor = element_blank(),
+    panel.grid = element_line(color = "#293D42"),
+    plot.background = element_rect(fill = "#002B36", color = "#002B36"),
+    panel.background = element_rect(color = "#002B36", fill = "#002B36"),
+    legend.background = element_rect(fill = "#002B36"),
+    legend.text = element_text(size = 16),
+    legend.title = element_text(size = 14, face = "bold"),
+    legend.key = element_rect(fill = "#002B36"),
+    legend.position = "bottom",
+    strip.background = element_rect(fill = "#002B36"),
+    strip.text = element_text(color = "#EEE8D5", face = "bold")
+  )
+
+plot_roc <- function(roc_data, title) {
+  roc_data %>%
+    ggplot(
+      aes(1 - specificity, sensitivity, color = model)
+    ) +
+    geom_line(linewidth = .8, alpha = .7) +
+    geom_abline(slope = 1, intercept = 0, linewidth = 0.4, lty = 2, color = "lightblue") +
+    scale_color_brewer(palette = "Dark2") +
+    scale_x_continuous(breaks = c(0, .5, 1)) +
+    scale_y_continuous(breaks = c(0, .5, 1)) +
+    facet_wrap(~ group) +
[…]

### 2023-05-03 code/plot_sleep_ROC_data.R

@@ -1,73 +1,0 @@
-#!/usr/bin/env
-
-library(tidyverse)
-library(tidymodels)
-library(arrow)
-library(patchwork)
-
-
-sleep_roc_data_10 <- 
-  read_parquet("presentation/data/all_roc_sleep_10.parquet") %>% 
-  group_by(model, group) %>% 
-  slice_sample(n = 1000)
-
-sleep_roc_data_30 <- 
-  read_parquet("presentation/data/all_roc_sleep_30.parquet") %>% 
-  group_by(model, group) %>% 
-  slice_sample(n = 1000)
-
-
-
-roc_theme <- 
-  theme(
-    text = element_text(color = "#EEE8D5", family = "ibm"),
-    axis.title = element_text(family = "ibm", size = 16, color = "#EEE8D5"),
-    axis.text = element_text(size = 14, color = "#EEE8D5"),
-    plot.title = element_text(family = "ibm", size = 30, face = "bold", hjust = .5, 
-                              margin = margin(b = 40)),
-    plot.subtitle = element_text(size = 20, hjust = .5),
-    plot.caption = element_text(size = 12),
-    panel.grid.minor = element_blank(),
-    panel.grid = element_line(color = "#293D42"),
-    plot.background = element_rect(fill = "#002B36", color = "#002B36"),
-    panel.background = element_rect(color = "#002B36", fill = "#002B36"),
-    legend.background = element_rect(fill = "#002B36"),
-    legend.text = element_text(size = 16),
-    legend.title = element_text(size = 14, face = "bold"),
-    legend.key = element_rect(fill = "#002B36"),
-    legend.position = "bottom",
-    strip.background = element_rect(fill = "#002B36"),
-    strip.text = element_text(color = "#EEE8D5", face = "bold")
-  )
-
-plot_roc <- function(roc_data, title) {
-  roc_data %>%
-    ggplot(
-      aes(1 - specificity, sensitivity, color = model)
-    ) +
-    geom_line(linewidth = .8, alpha = .7) +
-    geom_abline(slope = 1, intercept = 0, linewidth = 0.4, lty = 2, color = "lightblue") +
-    scale_color_brewer(palette = "Dark2") +
-    scale_x_continuous(breaks = c(0, .5, 1)) +
-    scale_y_continuous(breaks = c(0, .5, 1)) +
-    facet_wrap(~ group) +
[…]

### 2023-05-03 code/prepare_data/resample_and_extract_features_screens.R

@@ -1,13 +1,13 @@
 #!/usr/bin/env Rscript
 
-suppressMessages(library(tidyverse))
-suppressMessages(library(furrr))
-suppressMessages(library(slider))
-suppressMessages(library(arrow))
-suppressMessages(library(read.cwa))
-suppressMessages(library(glue))
-suppressMessages(library(lubridate))
-suppressMessages(library(signal))
+library(tidyverse)
+library(furrr)
+library(slider)
+library(arrow)
+library(read.cwa)
+library(glue)
+library(lubridate)
+library(signal)
 
 args <- R.utils::commandArgs(trailingOnly = TRUE)
 

### 2023-05-03 code/prepare_data/zm_join_sceens_acc.R

@@ -5,70 +5,83 @@ library(lubridate)
 library(hms)
 library(arrow)
 
-path <- "data/processed/screens_baseline.parquet"
-
+# Read in command line arguments
 args <- R.utils::commandArgs(trailingOnly = TRUE)
-
 path <- as.character(args[1])
 dest <- as.character(args[2])
 
+# Read in accelerometer data
 acc_all <-
   read_parquet(path) |>
   relocate(datetime = time, .after = 1) |>
+  # Create columns for time intervals rounded to nearest multiple of 5
   mutate(
     time_int = as.integer(as_hms(datetime)),
     time_int_round5 = as.integer(time_int %/% 5 * 5),
     .after = 1
   ) |>
+  # Drop unnecessary columns and sort by id
   select(-c(time_int, sensor_code, datetime)) |>
   arrange(id)
 
+# Get list of child ids from accelerometer data
 child_ids_acc <- acc_all |>
   distinct(id) |>
   pull()
 
+# Read in Z-scores data and filter by child ids
 zm <-
   read_tsv("data/processed/zm_scores.tsv") |>
   filter(id %in% child_ids_acc) |>
+  # Repeat rows to get 5-minute intervals
   slice(rep(1:n(), each = 30 / 5)) %>%
+  # Create columns for time intervals rounded to nearest multiple of 5
   mutate(
     time_int = as.integer(as_hms(datetime)),
     time_int = as.integer(time_int + rep_len(seq(0, 25, 5), length.out = nrow(.))),
     time_int_round5 = as.integer(time_int %/% 5 * 5),
     .after = 1
   ) |>
+  # Drop unnecessary columns
   select(-time_int)
 
+# Get list of child ids from Z-scores data
 child_ids_zm <-
   zm |>
   distinct(id) |>
   pull()
 
+# Filter accelerometer data by child ids
 acc <-
   acc_all |>
   filter(id %in% child_ids_zm)
 
+# Remove original accelerometer data from memory
 rm(acc_all)
 gc()
 
-# TESTING
-
+# Create temporary directory
 dir.create(temp <- tempfile())
 
+# Write out accelerometer data to parquet files by child id
 acc |>
   group_by(id) |>
   group_walk(~ write_parquet(.x, file.path(temp, paste0(.y$id, ".parquet"))))
 
+# Get list of parquet files in temporary directory
 list.files(temp, "parquet$", full.names = TRUE)
 
+# Delete temporary directory and its contents
[…]

### 2023-05-03 presentation/code/table_roc_auc_sleep.R

@@ -1,0 +1,51 @@
+#!/usr/bin/env
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+library(gt)
+
+roc_auc_10 <-
+  read_csv(here::here("presentation/data/all_roc_auc_sleep_10_sec_epochs.csv"))
+
+roc_auc_30 <-
+  read_csv(here::here("presentation/data/all_roc_auc_sleep_30_sec_epochs.csv"))
+
+table_sleep_roc_auc <-
+  bind_rows(epoch_10 = roc_auc_10, epoch_30 = roc_auc_30, .id = "epoch_length") %>%
+  select(-.metric, -.estimator) %>%
+  pivot_wider(names_from = epoch_length, values_from = .estimate) %>%
+  gt(groupname_col = "group") %>%
+  fmt_number(3:4, decimals = 2) %>%
+  tab_spanner(label = "ROC AUC", columns = 3:4) %>% 
+  cols_label(epoch_10 = "10 sec epochs",
+             epoch_30 = "30 sec epochs",
+             model = "") %>% 
+  tab_options(
+    table.font.names = "ibm",
+    table.font.color.light = "#EEE8D5",
+    table.background.color = "#002B36",
+    heading.border.bottom.style = "none",
+    heading.align = "center",
+    heading.title.font.size = 26,
+    table.border.top.width = px(3),
+    table.border.top.style = "none", # transparent
+    table.border.bottom.style = "none",
+    source_notes.border.bottom.color = "#002B36",
+    column_labels.font.weight = "normal",
+    column_labels.border.top.style = "none",
+    column_labels.border.bottom.width = px(2),
+    column_labels.border.bottom.color = "#2C4E57",
+    column_labels.padding = px(1),
+    row_group.border.top.style = "none",
+    row_group.border.top.color = "#2C4E57",
+    row_group.border.bottom.width = px(0),
+    row_group.border.bottom.color = "#2C4E57",
+    stub.border.color = "#2C4E57",
+    stub.border.width = px(0),
+    stub.font.size = 14,
+    table_body.border.bottom.color = "#002B36",
+    table_body.hlines.color = "#2C4E57",
+    table_body.vlines.style = "none",
+    data_row.padding = px(1)
+  )

### 2023-05-03 presentation/code/tables_performance.R

@@ -103,7 +103,6 @@ table_sleep_metrics <-
   tab_header(
     title = md("Performance Metrics")
   ) |>
-  tab_options(table.font.names = "ibm") |>
   cols_width(everything() ~ px(120)) |>
   cols_align(align = "center") %>%
   cols_label(.metric = "") %>% 

### 2023-06-26 manuscript/manuscript.qmd
intro and methods 2nd draft. Begin results section.

---
title: Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality metrics
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction

An extensive array of research underlines the importance of sleep for
both mental and physical health[@ma2017; @meyer2022; @kpavlova2019;
@difrancesco2019]. Consequently, accurate sleep assessment methods are
essential for tracking sleep patterns, thereby enhancing our
comprehension of the sleep-health relationship. Furthermore, ensuring
high user-acceptability for these methods is essential in order to
conduct large-scale studies over prolonged periods.

While laboratory-based polysomnography is considered the gold standard
for objectively measuring sleep, its practicality in large-scale
epidemiological studies is limited due to its high cost, the need for
professional administration, and the substantial resources required for
specialized equipment[@vandewater2011]. As an alternative, diaries is
commonly used as low-cost and low-tech methods for sleep assessment in
population research. However, relying solely on diary-based methods may
introduce recall bias and other limitations[@moore2015]. A more feasible
approach in large-scale epidemiological studies is the use of
device-based measurement methods that can estimate sleep duration. This
method offers the advantage of being less burdensome for participants
and avoids potential biases associated with recall.

In this context, the Zmachine®️ Insight+ (ZM) emerges as a valuable
tool. Validated against polysomnography with favorable
results[@kaplan2014; @wang2015], the ZM provides comparable data without
the high costs or need for professional monitoring associated with
polysomnography. Furthermore, the ease of use of the ZM device makes it
compliant for free-living use[@pedersen2021]. This allows for the
analysis of multiple consecutive nights, as compared to single-night
recordings from PSG, thereby capturing the important variations in sleep
across multiple nights.

The introduction of body-worn accelerometers has provided an effective
and affordable alternative for objectively assessing sleep patterns in a
home environment over extended periods. These accelerometers collect
continuous, high-resolution data for several weeks without requiring
recharging, thus minimizing participant burden. Initial applications of
accelerometry for sleep and wake stage classification were based on
wrist movements. The original algorithm, developed in 1982 using simple
linear regression and validated with PSG [@webster1982], was later
refined in 1992 [@cole1992], leading to the widely used Cole-Kripke
model. Subsequent research on wrist-worn accelerometer data has employed
heuristic algorithms, advanced machine learning models, as well as
regression and deep learning techniques[@palotti2019; @cole1992;
@sazonov2004; @sadeh1994; @hees2015; @sundararajan2021].

Despite the well-developed field of accelerometer data analysis for
sleep detection from wrist and hip-worn devices, the same level of
advancement is not mirrored in studies utilizing thigh-worn
accelerometers. Methods for assessing sleep using wrist and hip-worn
accelerometers have greatly evolved over the years, employing an
extensive range of techniques. These include heuristic algorithms,
machine learning models, regression and deep learning techniques, all
tailored to the specific signal characteristics of wrist and hip-worn
devices [@palotti2019; @cole1992; @sazonov2004; @sadeh1994; @hees2015;
@sundararajan2021; @patterson_40_2023]. However, for thigh-worn
accelerometers, the landscape appears less mature, with only a handful
of studies investigating sleep detection algorithms. The majority of the
efforts are focused on delineating wakefulness from sleep, with
particular emphasis on the definition of 'waking time' and 'bedtime'
[@carlson2021; @inan-eroglu2021; @vanderberg2016; @winkler2016].
Furthermore, while strides have been made recently in estimating sleep
duration with these devices, with the introduction of a promising
algorithm and its comparison against PSG[@johansson_development_2023],
the field is still in its infancy when it comes to employing machine
learning techniques. Given the potential for accurate physical behavior
assessment that thigh-worn accelerometers
provide[@skotte_detection_2014; @arvidsson2019], a significant research
gap exists. Therefore, there is a pressing need for future studies to
develop techniques similar to those used for wrist and hip-worn
accelerometers, with the ultimate goal of establishing a more holistic,
accurate, and user-friendly method of sleep and physical activity
tracking.

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we
considered the gold standard for measuring sleep. Furthermore, our
secondary goal was to assess the developed models' performance in
evaluating important sleep quality metrics, including sleep period time
(SPT), total sleep time (TST), sleep efficiency (SE), latency until
persistent sleep (LPS), and wake after sleep onset (WASO). ~~By
analyzing these additional metrics, we aimed to provide a comprehensive
evaluation of the model's capability in assessing various aspects of
sleep quality, allowing us to determine the overall effectiveness of our
models in accurately estimating different parameters related to sleep
duration and quality.~~

# Methods

## Dataset and participants

The current study leverages data from the SCREENS
project[@rasmussen2020], a study conducted from October 2018 to March
2019 in Middelfart, Southern Denmark, that evaluated the impact of
screen media usage on Danish families. For our analysis, we isolated
data from child participants between the ages of 6 to 10 years within
the SCREENS cohort. Our main sources of data were accelerometer readings
from Axivity AX3 devices attached to the children's thighs, and
electroencephalography (EEG) data derived from the ZM device. The
Axivity AX3, an unobtrusive 3-axis accelerometer, was positioned midway
between the hip and knee on the right anterior thigh, recording
participant movement data.

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, utilizes advanced EEG hardware and signal
processing algorithms, employs three self-adhesive, disposable sensors
placed outside the hairline for reliable EEG signal acquisition. The ZM
uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is utilized
for accurate sleep detection, showcasing its aptness for in-home
monitoring[@kaplan2014], while the Z-PLUS effectively differentiates
sleep stages, as evidenced by its alignment with expert evaluations
using PSG data[@wang2015].

Finally, we affirm that the SCREENS study received approval from the
Regional Scientific Committee of Southern Denmark, and all data handling
processes complied with the General Data Protection Regulation (GDPR),
guaranteeing the ethical and secure management of participant
information.

## Data Preprocessing and Feature Extraction

In this study, raw data processing began with a low-pass filtration step
using a 4th order Butterworth filter with a 5 Hz cut-off frequency to
eliminate high-frequency noise. Following filtration, data were
partitioned into overlapping 2-second intervals, each successive
interval sharing a 50% overlap with the previous one similar to methods
described by Skotte et al.[@skotte_detection_2014]. Any non-wear data
was remove using previously described methods[@skovgaard2023] and data
was resampled to 30-second epochs so every sample classified by the
algorithms corresponds to a 30-second epoch scored during the ZM
recordings. Subsequently, we performed a feature extraction process that
yielded a set of 88 features, providing a robust characterization of the
data. Extracted from accelerometer and temperature signals, these
features include temporal elements that use both lag and lead values,
capturing dynamic data trends by incorporating measurements from
preceding and upcoming intervals. Furthermore, inspired by Walch et
al.[@walch2019], we developed sensor-independent features to encapsulate
circadian rhythms. These features offer unique insights not directly
discernible from sensor outputs and are meant to approximate the
changing drive of the circadian clock to sleep over the course of the
night (see @fig-sensor-independent). Furthermore, the feature set was
enriched by including signal characteristics, which encompass vector
magnitude, mean crossing rate, skewness, and kurtosis for each of the x,
y, and z dimensions. These features provide a comprehensive
understanding of the signal's attributes, facilitating a more nuanced
analysis of the underlying patterns. All features are summarized in
table ???. (overvej tabel til supp. material)

![Sensor-independent features of circadian rhythms across two
consecutive nights. A) cosinus feature, B) linear
feature.](visuals/sensor_independent.pdf){#fig-sensor-independent}

In addition to the engineered features, we chose to incorporate the
median-filtered raw predictions from the ZM device into our modeling
process. This decision stemmed from the understanding that children
typically undergo around five to eight sleep cycles per night, with
awakenings commonly occurring at the end of each
cycle[@galland_normal_2012]. Examining the raw ZM predictions, we noted
a significant overestimation in the number of awakenings per night for
the children in our study, exceeding what would be expected based on
typical sleep cycle patterns. Consequently, we elected to train and
evaluate our models using not only the raw ZM output, but also versions
that were subjected to 5-minute and 10-minute median filters. This
approach resulted in an anticipated, more age-appropriate count of
awakenings per night, providing a more accurate depiction of children's
sleep patterns (see @fig-zm-median).

![The difference in number of awakenings between the raw ZM predictions
vs. 5-minute, and 10-minute median filtered predictions for a random
night. Grey line is the raw predictions, black line is the median
filtered predictions. A: 5-minute median filter on raw ZM predictions,
B: 10-minute median filter on raw ZM
predictions.](visuals/zm_raw_vs_filtered.pdf){#fig-zm-median}

## Algorithms, Training and Validation

In our study, we utilize two distinct modeling strategies to predict
sleep patterns derived from thigh-mounted accelerometer data: a
sequential ensemble and a multiclass bidirectional Long Short-Term
Memory (biLSTM)[@hochreiter1997] neural network. The sequential ensemble
involves using four pairs of models, each pair consisting of the same
algorithm, that predict in-bed and sleep time successively. Conversely,
the biLSTM is a comprehensive model that predicts multiple classes
simultaneously, effectively handling complex temporal patterns. By
applying and comparing both, we aim to evaluate their relative
effectiveness.

### Models in Sequence

To predict in-bed time and sleep time accurately, we employed an
ensemble learning strategy based on sequential binary classification
models. This approach involved constructing a sequence of models using
multiple machine learning algorithms to improve predictive accuracy. The
process began with an initial model predicting in-bed time, followed by
a second model that utilized the output of the initial model to predict
sleep time. This sequential approach was applied across all four
algorithms detailed below, with each subsequent model leveraging the
outputs of the previous models for improved predictions.

1.  Logistic Regression (LREG): Logistic regression served as a simple
    and fast baseline model. However, due to its linear nature, it may
    struggle with capturing complex relationships and non-linear
    patterns present in the accelerometer data.

2.  Decision Tree (TREE): Decision trees are capable of handling
    non-linear patterns and are easily interpretable. However, they are
    prone to overfitting, particularly when dealing with complex
    patterns that require simultaneous consideration of multiple
    features.

3.  Single-layer Feed-forward Neural Network (SNN): Single-layer
    feed-forward neural networks can effectively capture non-linear
    relationships, even with their relatively simple structure. However,
    they tend to be more challenging to interpret compared to simpler
    models. Additionally, careful tuning of the network's architecture
    and training process is required to mitigate the risk of
    overfitting.

4.  XGBoost (XGB): XGBoost is a powerful algorithm known for its ability
    to provide highly accurate predictions and handle complex,
    non-linear patterns in the data. It also incorporates built-in
    methods to prevent overfitting. However, training XGBoost models can
    be computationally intensive, and interpreting the predictions it
    generates can pose challenges.

### Multiclass Model

The biLSTM is a multiclass classifier capable of predicting the classes
of out-bed-awake, in-bed-awake, and in-bed-asleep. The architecture of
the biLSTM consisted of four layers, each with 128 hidden units. The
choice of four layers and 128 hidden units in our model balanced
complexity and efficiency: it was deep enough to learn intricate
patterns yet feasible to train timely. Additionally, the bidirectional
nature of the LSTM, doubling the hidden units at each time step,
improved data comprehension and avoided overfitting.

The biLSTM model was chosen for sleep classification using accelerometer
data from a thigh-mounted device due to its advantages in modeling
temporal context, handling a large temporal scope, and enabling temporal
inference over any feature without the need for hand-designed temporal
features.

Previous studies have explored the use of LSTM models for sleep
detection, showing promising results in capturing complex temporal
patterns. Notable works by Sano et al. [@sano2019] and Chen et al.
[@chen2021] have demonstrated the potential of LSTM models in improving
sleep detection based on accelerometer data. However, further research
is necessary to fully comprehend the strengths and limitations of this
approach.

### Model Training

For the models in sequence, we trained four pairs of classification
models, each pair distinguishing between in-bed/out-of-bed and
asleep/awake states. The dataset was randomly split into a training set
(approximately 50% of the subjects) and a testing set (also
approximately 50%), ensuring that samples from the same subject were
never simultaneously used in both sets. Hyper-parameter optimization was
performed using 10-fold Monte Carlo cross-validation, with the F1 score
as the performance metric. The best-performing set of hyperparameters
was then used to fit the models to the full training dataset, maximizing
accuracy by leveraging all available data. Furthermore, the dataset for
predicting sleep was highly imbalanced which can pose challenges when
training machine learning models, as models may favor predicting the
majority class. To address this issue, we employed oversampling
techniques to ensure that each class had a roughly equal number of
training samples. Specifically, we utilized the Synthetic Minority
Over-sampling Technique (SMOTE)[@chawla2002], which generates new
samples by interpolating random samples with their nearest neighbors. In
our study, we implemented SMOTE using the themis R package[@themis],
resampling all classes to achieve a balanced distribution of training
samples.

The biLSTM model was trained using the Adam optimizer, which is
computationally efficient and adapts the learning rate during training.
Cross-entropy loss function was employed for its suitability in
multiclass classification with mutually exclusive classes. The softmax
activation function was selected for the output layer to obtain a
probability distribution over the classes. Data for training the biLSTM
were randomly divided into a training, validation, and test tensors
based on a 50/25/25 split. The model was evaluated using the F1 score on
both the training and validation sets. Early stopping was implemented
with a patience of 3 epochs, meaning training would stop if there was no
improvement in the validation loss for 3 consecutive epochs.

### Model Validation \[evt bare "statistics"\]

In our study, we employed standard evaluation metrics to assess the
epoch-to-epoch performance of each model. These metrics include accuracy
as $accuracy = \frac{TP+TN}{TP+TN+FP+FN}$, sensitivity as
$sensitivity = \frac{TP}{TP+FN}$, specificity as
$specificity = \frac{TN}{TN+FP}$, precision as
$precision = \frac{TP}{TP+FP}$, F1 score as
$F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}$, and
area under the precision-recall curves were also calculated. The Average
Precision, represented by the area under the precision-recall curve,
provides a quantitative measure of performance. A value of 1 for Average
Precision indicates the best performance. For the multiclass biLSTM
model, we calculated performance metrics as macro-averages, to treat all
classes equally. Consistent with previous studies in the
field[@hjorth2012; @kushida2001], we treated the in-bed/out-of-bed and
awake/asleep scoring tasks as a binary classification problems, where
in-bed and asleep was considered the positive label and out-of-bed and
awake the negative label.

To evaluate our developed models' performance in assessing sleep quality
metrics, we employed Bland-Altman plots and Pearson correlations. These
methods were used to compare the sleep quality summaries calculated by
our models with those obtained from EEG-based ZM sleep quality
summaries. The sleep quality summaries included several metrics:

1.  Sleep Period Time (SPT) - This refers to the total duration of the
    sleep period, which is defined as the time from the start to the end
    of the ZM recording.

2.  Total Sleep Time (TST) - This is the time spent asleep within the
    SPT.

3.  Sleep Efficiency (SE) - This is the ratio between TST and SPT,
    representing the proportion of the sleep period that was actually
    spent asleep.

4.  Latency Until Persistent Sleep (LPS) - This metric represents the
    time it takes to transition from wakefulness to sustained sleep. It
    is calculated as the time from the beginning of the ZM recording
    until a period when 10 out of 12 minutes are scored as sleep.

5.  Wake After Sleep Onset (WASO) - This refers to the time spent awake
    after initially falling asleep and before the final awakening. In
    our analysis, a period is counted as 'awake' only if it consists of
    3 or more contiguous 30-second epochs.

R version 4.3.0 (2023-04-21)[@R-lang] and the Tidymodels[@tidymodels]
and Tidyverse[@tidyverse] suite of packages were used as the core tools
for model development and analyses. Python version
3.10.6[@10.5555/1593511] and PyTorch[@NEURIPS2019_9015] were used to
implement the biLSTM model. All code used to perform the analysis and
generate the figures in this paper are available in [this
repository](https://github.com/esbenlykke/sleep_study).

# Results

![Flowchart of eligible nights included in the study - revise text
boxes!](visuals/flowchart_of_elligible_nights.pdf)

The children children icluded in the analyses were 9.4 years old (SD = 2.1).

As reported in @tbl-summary_zm the sleep quality metrics derived from ZMachine predictions were modified by the implementation of 5-minute and 10-minute median filters. The Sleep Period Time (SPT) remained consistent across raw and filtered datasets (mean: 9.2 ± 2.1 hours). Total Sleep Time (TST) and Sleep Efficiency (SE) increased in the filtered data, implying the filters categorize some wakefulness as sleep. Specifically, TST increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours (5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. Latency to Persistent Sleep (LPS) also elevated, suggesting the filter smooths out brief awakenings at sleep onset, leading to a prolonged time to persistent sleep. The most significant change was seen in Wake After Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and 10-minute filtered data, respectively. The results underscore the influential role of median filters on sleep metrics.

|||Raw| |Median 5-Minute| |Median 10-Minute| |
|-|-|-|-|-|-|-|-|
|||Mean|SD|Mean|SD|Mean|SD|
||SPT|9.2|2.1|9.2|2.1|9.2|2.1|
||TST|7.7|1.9|8.1|2.0|8.2|2.1|
||SE|82.6|12.0|86.4|12.7|87.5|12.9|
||LPS|34.5|27.9|36.3|39.8|38.0|48.7|
||WASO|39.0|33.6|30.6|46.8|22.3|55.4|

: Summary of sleep quality metrics as calculated based on the raw, the 5-minute, and 10-minute median filtered raw ZM prediction. {#tbl-summary_zm}

\newpage

# References

### 2023-06-26 manuscript/manuscript.tex
intro and methods 2nd draft. Begin results section.

--- 2023-06-26 manuscript/manuscript.qmd
@@
----
-title: Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device
-author:
-  - name: Esben Høegholm Lykke
-    email: eskovgaard@health.sdu.dk
-    affiliations: 
-        - id: some-tech
-          name: University of Southern Denmark
-          department: Department of Sports Science and Clinical Biomechanics
-          address: Campusvej 55
-          city: Odense
-          state: Denmark
-          postal-code: 5230
-    attributes:
-        corresponding: true
-  - name: Jan Christian Brønd
-    email: jbrond@health.sdu.dk
-    affiliations:
-        - id: another-u
-          name: University of Southern Denmark
-          department: Department of Sports Science and Clinical Biomechanics
-          address: Campusvej 55
-          city: Odense
-          state: Denmark
-          postal-code: 5230
-abstract: "LÆS IKKE!
-Det er volapyk på latin.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse vitae dictum eros, ullamcorper elementum orci.
-Sed laoreet nulla neque, pulvinar fermentum mi iaculis at.
-Nulla ultricies nibh sit amet vestibulum rutrum.
-Nam pharetra nisl sed ipsum maximus suscipit.
-Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante.
-Nunc vitae lectus nisi.
-Aliquam efficitur ut eros ut pellentesque.
-Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus.
-Curabitur sollicitudin lectus sit amet velit pellentesque laoreet.
-Ut posuere diam lobortis nisi eleifend tincidunt.
-Ut at euismod sem, sed dignissim ligula.
-Aliquam lacinia massa libero, id eleifend velit pulvinar ac.
-Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
-keywords: 
-  - Sleep
-  - Accelerometry
-  - EEG
-  - Machine learning
-  - Sleep quality metrics
-date: last-modified
-bibliography: bibliography.bib
-format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: npj Digital Medicine
-      formatting: preprint
-      model: 3p
-      cite-style: super
-    include-in-header: 
-      text: '\usepackage{lineno}\linenumbers'
-    fig-pos: 'b'
-  docx:
-    number-sections: true
-editor: 
-  markdown: 
-    wrap: 72
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  super,
+  preprint,
+  3p]{elsarticle}
 
-# Introduction
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\usepackage{soul}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\usepackage{lineno}\linenumbers
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\journal{npj Digital Medicine}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-num}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device},
+  pdfauthor={Esben Høegholm Lykke; Jan Christian Brønd},
+  pdfkeywords={Sleep, Accelerometry, EEG, Machine learning, Sleep
+quality metrics},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\setlength{\parindent}{6pt}
+\begin{document}
+
+\begin{frontmatter}
+\title{Estimating Sleep Quality Metrics Using Free-Living Accelerometer
+Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep
+Tracking Device}
+\author[1]{Esben Høegholm Lykke%
+\corref{cor1}%
+}
+ \ead{eskovgaard@health.sdu.dk} 
+\author[1]{Jan Christian Brønd%
+%
+}
+ \ead{jbrond@health.sdu.dk}
+
+\affiliation[1]{organization={University of Southern Denmark, Department
+of Sports Science and Clinical Biomechanics},addressline={Campusvej
+55},city={Odense},postcode={5230},postcodesep={}}
+
+\cortext[cor1]{Corresponding author}
+
+\begin{abstract}
+LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet,
+consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper
+elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis
+at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl
+sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum,
+tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros
+ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum
+fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin
+lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis
+nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula.
+Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce
+volutpat elit eu nulla viverra, nec tempus orci pellentesque.
+\end{abstract}
+
+\begin{keyword}
+    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
+    Sleep quality metrics
+\end{keyword}
+\end{frontmatter}
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
 An extensive array of research underlines the importance of sleep for
-both mental and physical health[@ma2017; @meyer2022; @kpavlova2019;
-@difrancesco2019]. Consequently, accurate sleep assessment methods are
-essential for tracking sleep patterns, thereby enhancing our
-comprehension of the sleep-health relationship. Furthermore, ensuring
-high user-acceptability for these methods is essential in order to
-conduct large-scale studies over prolonged periods.
+both mental and physical
+health\citep{ma2017, meyer2022, kpavlova2019, difrancesco2019}.
+Consequently, accurate sleep assessment methods are essential for
+tracking sleep patterns, thereby enhancing our comprehension of the
+sleep-health relationship. Furthermore, ensuring high user-acceptability
+for these methods is essential in order to conduct large-scale studies
+over prolonged periods.
 
@@
 professional administration, and the substantial resources required for
-specialized equipment[@vandewater2011]. As an alternative, diaries is
-commonly used as low-cost and low-tech methods for sleep assessment in
-population research. However, relying solely on diary-based methods may
-introduce recall bias and other limitations[@moore2015]. A more feasible
-approach in large-scale epidemiological studies is the use of
+specialized equipment\citep{vandewater2011}. As an alternative, diaries
+is commonly used as low-cost and low-tech methods for sleep assessment
+in population research. However, relying solely on diary-based methods
+may introduce recall bias and other limitations\citep{moore2015}. A more
+feasible approach in large-scale epidemiological studies is the use of
 device-based measurement methods that can estimate sleep duration. This
@@
 tool. Validated against polysomnography with favorable
-results[@kaplan2014; @wang2015], the ZM provides comparable data without
-the high costs or need for professional monitoring associated with
-polysomnography. Furthermore, the ease of use of the ZM device makes it
-compliant for free-living use[@pedersen2021]. This allows for the
-analysis of multiple consecutive nights, as compared to single-night
-recordings from PSG, thereby capturing the important variations in sleep
-across multiple nights.
+results\citep{kaplan2014, wang2015}, the ZM provides comparable data
+without the high costs or need for professional monitoring associated
+with polysomnography. Furthermore, the ease of use of the ZM device
+makes it compliant for free-living use\citep{pedersen2021}. This allows
+for the analysis of multiple consecutive nights, as compared to
+single-night recordings from PSG, thereby capturing the important
+variations in sleep across multiple nights.
 
@@
 wrist movements. The original algorithm, developed in 1982 using simple
-linear regression and validated with PSG [@webster1982], was later
-refined in 1992 [@cole1992], leading to the widely used Cole-Kripke
+linear regression and validated with PSG \citep{webster1982}, was later
+refined in 1992 \citep{cole1992}, leading to the widely used Cole-Kripke
 model. Subsequent research on wrist-worn accelerometer data has employed
 heuristic algorithms, advanced machine learning models, as well as
-regression and deep learning techniques[@palotti2019; @cole1992;
-@sazonov2004; @sadeh1994; @hees2015; @sundararajan2021].
+regression and deep learning
+techniques\citep{palotti2019, cole1992, sazonov2004, sadeh1994, hees2015, sundararajan2021}.
 
@@
 tailored to the specific signal characteristics of wrist and hip-worn
-devices [@palotti2019; @cole1992; @sazonov2004; @sadeh1994; @hees2015;
-@sundararajan2021; @patterson_40_2023]. However, for thigh-worn
-accelerometers, the landscape appears less mature, with only a handful
-of studies investigating sleep detection algorithms. The majority of the
-efforts are focused on delineating wakefulness from sleep, with
-particular emphasis on the definition of 'waking time' and 'bedtime'
-[@carlson2021; @inan-eroglu2021; @vanderberg2016; @winkler2016].
+devices
+\citep{palotti2019, cole1992, sazonov2004, sadeh1994, hees2015, sundararajan2021, patterson_40_2023}.
+However, for thigh-worn accelerometers, the landscape appears less
+mature, with only a handful of studies investigating sleep detection
+algorithms. The majority of the efforts are focused on delineating
+wakefulness from sleep, with particular emphasis on the definition of
+`waking time' and `bedtime'
+\citep{carlson2021, inan-eroglu2021, vanderberg2016, winkler2016}.
 Furthermore, while strides have been made recently in estimating sleep
 duration with these devices, with the introduction of a promising
-algorithm and its comparison against PSG[@johansson_development_2023],
-the field is still in its infancy when it comes to employing machine
-learning techniques. Given the potential for accurate physical behavior
-assessment that thigh-worn accelerometers
-provide[@skotte_detection_2014; @arvidsson2019], a significant research
-gap exists. Therefore, there is a pressing need for future studies to
-develop techniques similar to those used for wrist and hip-worn
-accelerometers, with the ultimate goal of establishing a more holistic,
-accurate, and user-friendly method of sleep and physical activity
-tracking.
+algorithm and its comparison against
+PSG\citep{johansson_development_2023}, the field is still in its infancy
+when it comes to employing machine learning techniques. Given the
+potential for accurate physical behavior assessment that thigh-worn
+accelerometers provide\citep{skotte_detection_2014, arvidsson2019}, a
+significant research gap exists. Therefore, there is a pressing need for
+future studies to develop techniques similar to those used for wrist and
+hip-worn accelerometers, with the ultimate goal of establishing a more
+holistic, accurate, and user-friendly method of sleep and physical
+activity tracking.
 
@@
 (SPT), total sleep time (TST), sleep efficiency (SE), latency until
-persistent sleep (LPS), and wake after sleep onset (WASO). ~~By
+persistent sleep (LPS), and wake after sleep onset (WASO). \st{By
 analyzing these additional metrics, we aimed to provide a comprehensive
@@
 models in accurately estimating different parameters related to sleep
-duration and quality.~~
+duration and quality.}
 
-# Methods
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-## Dataset and participants
+\hypertarget{dataset-and-participants}{%
+\subsection{Dataset and participants}\label{dataset-and-participants}}
 
 The current study leverages data from the SCREENS
-project[@rasmussen2020], a study conducted from October 2018 to March
-2019 in Middelfart, Southern Denmark, that evaluated the impact of
+project\citep{rasmussen2020}, a study conducted from October 2018 to
+March 2019 in Middelfart, Southern Denmark, that evaluated the impact of
 screen media usage on Danish families. For our analysis, we isolated
@@
 for accurate sleep detection, showcasing its aptness for in-home
-monitoring[@kaplan2014], while the Z-PLUS effectively differentiates
-sleep stages, as evidenced by its alignment with expert evaluations
-using PSG data[@wang2015].
+monitoring\citep{kaplan2014}, while the Z-PLUS effectively
+differentiates sleep stages, as evidenced by its alignment with expert
+evaluations using PSG data\citep{wang2015}.
 
@@
 
-## Data Preprocessing and Feature Extraction
+\hypertarget{data-preprocessing-and-feature-extraction}{%
+\subsection{Data Preprocessing and Feature
+Extraction}\label{data-preprocessing-and-feature-extraction}}
 
@@
 partitioned into overlapping 2-second intervals, each successive
-interval sharing a 50% overlap with the previous one similar to methods
-described by Skotte et al.[@skotte_detection_2014]. Any non-wear data
-was remove using previously described methods[@skovgaard2023] and data
-was resampled to 30-second epochs so every sample classified by the
-algorithms corresponds to a 30-second epoch scored during the ZM
+interval sharing a 50\% overlap with the previous one similar to methods
+described by Skotte et al.\citep{skotte_detection_2014}. Any non-wear
+data was remove using previously described methods\citep{skovgaard2023}
+and data was resampled to 30-second epochs so every sample classified by
+the algorithms corresponds to a 30-second epoch scored during the ZM
 recordings. Subsequently, we performed a feature extraction process that
@@
 preceding and upcoming intervals. Furthermore, inspired by Walch et
-al.[@walch2019], we developed sensor-independent features to encapsulate
-circadian rhythms. These features offer unique insights not directly
-discernible from sensor outputs and are meant to approximate the
-changing drive of the circadian clock to sleep over the course of the
-night (see @fig-sensor-independent). Furthermore, the feature set was
-enriched by including signal characteristics, which encompass vector
-magnitude, mean crossing rate, skewness, and kurtosis for each of the x,
-y, and z dimensions. These features provide a comprehensive
-understanding of the signal's attributes, facilitating a more nuanced
-analysis of the underlying patterns. All features are summarized in
-table ???. (overvej tabel til supp. material)
+al.\citep{walch2019}, we developed sensor-independent features to
+encapsulate circadian rhythms. These features offer unique insights not
+directly discernible from sensor outputs and are meant to approximate
+the changing drive of the circadian clock to sleep over the course of
+the night (see Figure~\ref{fig-sensor-independent}). Furthermore, the
+feature set was enriched by including signal characteristics, which
+encompass vector magnitude, mean crossing rate, skewness, and kurtosis
+for each of the x, y, and z dimensions. These features provide a
+comprehensive understanding of the signal's attributes, facilitating a
+more nuanced analysis of the underlying patterns. All features are
+summarized in table ???. (overvej tabel til supp. material)
 
-![Sensor-independent features of circadian rhythms across two
-consecutive nights. A) cosinus feature, B) linear
-feature.](visuals/sensor_independent.pdf){#fig-sensor-independent}
+\begin{figure}[b]
+
+{\centering \includegraphics{visuals/sensor_independent.pdf}
+
+}
+
+\caption{\label{fig-sensor-independent}Sensor-independent features of
+circadian rhythms across two consecutive nights. A) cosinus feature, B)
+linear feature.}
+
+\end{figure}
 
@@
 awakenings commonly occurring at the end of each
-cycle[@galland_normal_2012]. Examining the raw ZM predictions, we noted
-a significant overestimation in the number of awakenings per night for
-the children in our study, exceeding what would be expected based on
+cycle\citep{galland_normal_2012}. Examining the raw ZM predictions, we
+noted a significant overestimation in the number of awakenings per night
+for the children in our study, exceeding what would be expected based on
 typical sleep cycle patterns. Consequently, we elected to train and
@@
 awakenings per night, providing a more accurate depiction of children's
-sleep patterns (see @fig-zm-median).
+sleep patterns (see Figure~\ref{fig-zm-median}).
 
-![The difference in number of awakenings between the raw ZM predictions
-vs. 5-minute, and 10-minute median filtered predictions for a random
-night. Grey line is the raw predictions, black line is the median
-filtered predictions. A: 5-minute median filter on raw ZM predictions,
-B: 10-minute median filter on raw ZM
-predictions.](visuals/zm_raw_vs_filtered.pdf){#fig-zm-median}
+\begin{figure}[b]
 
-## Algorithms, Training and Validation
+{\centering \includegraphics{visuals/zm_raw_vs_filtered.pdf}
+
+}
+
+\caption{\label{fig-zm-median}The difference in number of awakenings
+between the raw ZM predictions vs.~5-minute, and 10-minute median
+filtered predictions for a random night. Grey line is the raw
+predictions, black line is the median filtered predictions. A: 5-minute
+median filter on raw ZM predictions, B: 10-minute median filter on raw
+ZM predictions.}
+
+\end{figure}
+
+\hypertarget{algorithms-training-and-validation}{%
+\subsection{Algorithms, Training and
+Validation}\label{algorithms-training-and-validation}}
 
@@
 sequential ensemble and a multiclass bidirectional Long Short-Term
-Memory (biLSTM)[@hochreiter1997] neural network. The sequential ensemble
-involves using four pairs of models, each pair consisting of the same
-algorithm, that predict in-bed and sleep time successively. Conversely,
-the biLSTM is a comprehensive model that predicts multiple classes
-simultaneously, effectively handling complex temporal patterns. By
-applying and comparing both, we aim to evaluate their relative
+Memory (biLSTM)\citep{hochreiter1997} neural network. The sequential
+ensemble involves using four pairs of models, each pair consisting of
+the same algorithm, that predict in-bed and sleep time successively.
+Conversely, the biLSTM is a comprehensive model that predicts multiple
+classes simultaneously, effectively handling complex temporal patterns.
+By applying and comparing both, we aim to evaluate their relative
 effectiveness.
 
-### Models in Sequence
+\hypertarget{models-in-sequence}{%
+\subsubsection{Models in Sequence}\label{models-in-sequence}}
 
@@
 
-1.  Logistic Regression (LREG): Logistic regression served as a simple
-    and fast baseline model. However, due to its linear nature, it may
-    struggle with capturing complex relationships and non-linear
-    patterns present in the accelerometer data.
-
-2.  Decision Tree (TREE): Decision trees are capable of handling
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Logistic Regression (LREG): Logistic regression served as a simple and
+  fast baseline model. However, due to its linear nature, it may
+  struggle with capturing complex relationships and non-linear patterns
+  present in the accelerometer data.
+\item
+  Decision Tree (TREE): Decision trees are capable of handling
   non-linear patterns and are easily interpretable. However, they are
-    prone to overfitting, particularly when dealing with complex
-    patterns that require simultaneous consideration of multiple
-    features.
-
-3.  Single-layer Feed-forward Neural Network (SNN): Single-layer
+  prone to overfitting, particularly when dealing with complex patterns
+  that require simultaneous consideration of multiple features.
+\item
+  Single-layer Feed-forward Neural Network (SNN): Single-layer
   feed-forward neural networks can effectively capture non-linear
@@
   they tend to be more challenging to interpret compared to simpler
-    models. Additionally, careful tuning of the network's architecture
-    and training process is required to mitigate the risk of
-    overfitting.
+  models. Additionally, careful tuning of the network's architecture and
+  training process is required to mitigate the risk of overfitting.
+\item
+  XGBoost (XGB): XGBoost is a powerful algorithm known for its ability
+  to provide highly accurate predictions and handle complex, non-linear
+  patterns in the data. It also incorporates built-in methods to prevent
+  overfitting. However, training XGBoost models can be computationally
+  intensive, and interpreting the predictions it generates can pose
+  challenges.
+\end{enumerate}
 
-4.  XGBoost (XGB): XGBoost is a powerful algorithm known for its ability
-    to provide highly accurate predictions and handle complex,
-    non-linear patterns in the data. It also incorporates built-in
-    methods to prevent overfitting. However, training XGBoost models can
-    be computationally intensive, and interpreting the predictions it
-    generates can pose challenges.
-
-### Multiclass Model
+\hypertarget{multiclass-model}{%
+\subsubsection{Multiclass Model}\label{multiclass-model}}
 
@@
 detection, showing promising results in capturing complex temporal
-patterns. Notable works by Sano et al. [@sano2019] and Chen et al.
-[@chen2021] have demonstrated the potential of LSTM models in improving
-sleep detection based on accelerometer data. However, further research
-is necessary to fully comprehend the strengths and limitations of this
-approach.
+patterns. Notable works by Sano et al. \citep{sano2019} and Chen et al.
+\citep{chen2021} have demonstrated the potential of LSTM models in
+improving sleep detection based on accelerometer data. However, further
+research is necessary to fully comprehend the strengths and limitations
+of this approach.
 
-### Model Training
+\hypertarget{model-training}{%
+\subsubsection{Model Training}\label{model-training}}
 
@@
 asleep/awake states. The dataset was randomly split into a training set
-(approximately 50% of the subjects) and a testing set (also
-approximately 50%), ensuring that samples from the same subject were
+(approximately 50\% of the subjects) and a testing set (also
+approximately 50\%), ensuring that samples from the same subject were
 never simultaneously used in both sets. Hyper-parameter optimization was
@@
 training samples. Specifically, we utilized the Synthetic Minority
-Over-sampling Technique (SMOTE)[@chawla2002], which generates new
+Over-sampling Technique (SMOTE)\citep{chawla2002}, which generates new
 samples by interpolating random samples with their nearest neighbors. In
-our study, we implemented SMOTE using the themis R package[@themis],
-resampling all classes to achieve a balanced distribution of training
-samples.
+our study, we implemented SMOTE using the themis R
+package\citep{themis}, resampling all classes to achieve a balanced
+distribution of training samples.
 
@@
 
-### Model Validation \[evt bare "statistics"\]
+\hypertarget{model-validation-evt-bare-statistics}{%
+\subsubsection{Model Validation {[}evt bare
+``statistics''{]}}\label{model-validation-evt-bare-statistics}}
 
@@
 epoch-to-epoch performance of each model. These metrics include accuracy
-as $accuracy = \frac{TP+TN}{TP+TN+FP+FN}$, sensitivity as
-$sensitivity = \frac{TP}{TP+FN}$, specificity as
-$specificity = \frac{TN}{TN+FP}$, precision as
-$precision = \frac{TP}{TP+FP}$, F1 score as
-$F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}$, and
-area under the precision-recall curves were also calculated. The Average
-Precision, represented by the area under the precision-recall curve,
-provides a quantitative measure of performance. A value of 1 for Average
-Precision indicates the best performance. For the multiclass biLSTM
-model, we calculated performance metrics as macro-averages, to treat all
-classes equally. Consistent with previous studies in the
-field[@hjorth2012; @kushida2001], we treated the in-bed/out-of-bed and
-awake/asleep scoring tasks as a binary classification problems, where
-in-bed and asleep was considered the positive label and out-of-bed and
-awake the negative label.
+as \(accuracy = \frac{TP+TN}{TP+TN+FP+FN}\), sensitivity as
+\(sensitivity = \frac{TP}{TP+FN}\), specificity as
+\(specificity = \frac{TN}{TN+FP}\), precision as
+\(precision = \frac{TP}{TP+FP}\), F1 score as
+\(F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}\),
+and area under the precision-recall curves were also calculated. The
+Average Precision, represented by the area under the precision-recall
+curve, provides a quantitative measure of performance. A value of 1 for
+Average Precision indicates the best performance. For the multiclass
+biLSTM model, we calculated performance metrics as macro-averages, to
+treat all classes equally. Consistent with previous studies in the
+field\citep{hjorth2012, kushida2001}, we treated the in-bed/out-of-bed
+and awake/asleep scoring tasks as a binary classification problems,
+where in-bed and asleep was considered the positive label and out-of-bed
+and awake the negative label.
 
@@
 
-1.  Sleep Period Time (SPT) - This refers to the total duration of the
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Sleep Period Time (SPT) - This refers to the total duration of the
   sleep period, which is defined as the time from the start to the end
   of the ZM recording.
-
-2.  Total Sleep Time (TST) - This is the time spent asleep within the
-    SPT.
-
-3.  Sleep Efficiency (SE) - This is the ratio between TST and SPT,
+\item
+  Total Sleep Time (TST) - This is the time spent asleep within the SPT.
+\item
+  Sleep Efficiency (SE) - This is the ratio between TST and SPT,
   representing the proportion of the sleep period that was actually
   spent asleep.
+\item
+  Latency Until Persistent Sleep (LPS) - This metric represents the time
+  it takes to transition from wakefulness to sustained sleep. It is
+  calculated as the time from the beginning of the ZM recording until a
+  period when 10 out of 12 minutes are scored as sleep.
+\item
+  Wake After Sleep Onset (WASO) - This refers to the time spent awake
+  after initially falling asleep and before the final awakening. In our
+  analysis, a period is counted as `awake' only if it consists of 3 or
+  more contiguous 30-second epochs.
+\end{enumerate}
 
-4.  Latency Until Persistent Sleep (LPS) - This metric represents the
-    time it takes to transition from wakefulness to sustained sleep. It
-    is calculated as the time from the beginning of the ZM recording
-    until a period when 10 out of 12 minutes are scored as sleep.
+R version 4.3.0 (2023-04-21)\citep{R-lang} and the
+Tidymodels\citep{tidymodels} and Tidyverse\citep{tidyverse} suite of
+packages were used as the core tools for model development and analyses.
+Python version 3.10.6\citep{10.5555/1593511} and
+PyTorch\citep{NEURIPS2019_9015} were used to implement the biLSTM model.
+All code used to perform the analysis and generate the figures in this
+paper are available in
+\href{https://github.com/esbenlykke/sleep_study}{this repository}.
 
-5.  Wake After Sleep Onset (WASO) - This refers to the time spent awake
-    after initially falling asleep and before the final awakening. In
-    our analysis, a period is counted as 'awake' only if it consists of
-    3 or more contiguous 30-second epochs.
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-R version 4.3.0 (2023-04-21)[@R-lang] and the Tidymodels[@tidymodels]
-and Tidyverse[@tidyverse] suite of packages were used as the core tools
-for model development and analyses. Python version
-3.10.6[@10.5555/1593511] and PyTorch[@NEURIPS2019_9015] were used to
-implement the biLSTM model. All code used to perform the analysis and
-generate the figures in this paper are available in [this
-repository](https://github.com/esbenlykke/sleep_study).
+\begin{figure}[b]
 
-# Results
+{\centering \includegraphics{visuals/flowchart_of_elligible_nights.pdf}
 
-![Flowchart of eligible nights included in the study - revise text
-boxes!](visuals/flowchart_of_elligible_nights.pdf)
+}
 
-The children children icluded in the analyses were 9.4 years old (SD = 2.1).
+\caption{Flowchart of eligible nights included in the study - revise
+text boxes!}
 
-As reported in @tbl-summary_zm the sleep quality metrics derived from ZMachine predictions were modified by the implementation of 5-minute and 10-minute median filters.
-The Sleep Period Time (SPT) remained consistent across raw and filtered datasets (mean: 9.2 ± 2.1 hours).
-Total Sleep Time (TST) and Sleep Efficiency (SE) increased in the filtered data, implying the filters categorize some wakefulness as sleep.
-Specifically, TST increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours (5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively.
-Latency to Persistent Sleep (LPS) also elevated, suggesting the filter smooths out brief awakenings at sleep onset, leading to a prolonged time to persistent sleep.
-The most significant change was seen in Wake After Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and 10-minute filtered data, respectively.
-The results underscore the influential role of median filters on sleep metrics.
+\end{figure}
 
-|||Raw| |Median 5-Minute| |Median 10-Minute| |
-|-|-|-|-|-|-|-|-|
-|||Mean|SD|Mean|SD|Mean|SD|
-||SPT|9.2|2.1|9.2|2.1|9.2|2.1|
-||TST|7.7|1.9|8.1|2.0|8.2|2.1|
-||SE|82.6|12.0|86.4|12.7|87.5|12.9|
-||LPS|34.5|27.9|36.3|39.8|38.0|48.7|
-||WASO|39.0|33.6|30.6|46.8|22.3|55.4|
+The children children icluded in the analyses were 9.4 years old (SD =
+2.1).
 
-: Summary of sleep quality metrics as calculated based on the raw, the 5-minute, and 10-minute median filtered raw ZM prediction. {#tbl-summary_zm}
+As reported in Table~\ref{tbl-summary_zm} the sleep quality metrics
+derived from ZMachine predictions were modified by the implementation of
+5-minute and 10-minute median filters. The Sleep Period Time (SPT)
+remained consistent across raw and filtered datasets (mean: 9.2 ± 2.1
+hours). Total Sleep Time (TST) and Sleep Efficiency (SE) increased in
+the filtered data, implying the filters categorize some wakefulness as
+sleep. Specifically, TST increased from a raw mean of 7.7 ± 1.9 hours to
+8.1 ± 2.0 hours (5-minute filter) and 8.2 ± 2.1 hours (10-minute
+filter), while SE rose from 82.6 ± 12.0\% to 86.4 ± 12.7\% and 87.5 ±
+12.9\% respectively. Latency to Persistent Sleep (LPS) also elevated,
+suggesting the filter smooths out brief awakenings at sleep onset,
+leading to a prolonged time to persistent sleep. The most significant
+change was seen in Wake After Sleep Onset (WASO), which dropped from
+39.0 ± 33.6 minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4
+minutes in the 5-minute and 10-minute filtered data, respectively. The
+results underscore the influential role of median filters on sleep
+metrics.
+
+\hypertarget{tbl-summary_zm}{}
+\begin{longtable}[]{@{}llllllll@{}}
+\caption{\label{tbl-summary_zm}Summary of sleep quality metrics as
+calculated based on the raw, the 5-minute, and 10-minute median filtered
+raw ZM prediction.}\tabularnewline
+\toprule\noalign{}
+& & Raw & & Median 5-Minute & & Median 10-Minute & \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+& & Raw & & Median 5-Minute & & Median 10-Minute & \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+& & Mean & SD & Mean & SD & Mean & SD \\
+& SPT & 9.2 & 2.1 & 9.2 & 2.1 & 9.2 & 2.1 \\
+& TST & 7.7 & 1.9 & 8.1 & 2.0 & 8.2 & 2.1 \\
+& SE & 82.6 & 12.0 & 86.4 & 12.7 & 87.5 & 12.9 \\
+& LPS & 34.5 & 27.9 & 36.3 & 39.8 & 38.0 & 48.7 \\
+& WASO & 39.0 & 33.6 & 30.6 & 46.8 & 22.3 & 55.4 \\
+\end{longtable}
 
@@
 
-# References
+\renewcommand\refname{References}
+  \bibliography{bibliography.bib}
+
+\end{document}

### 2023-06-26 manuscript/code/flowchart_elligible_nights.R

@@ -1,0 +1,195 @@
+library(tidyverse)
+library(ggtext)
+library(showtext)
+
+
+font_add_google("Mukta", family = "mukta")
+font_add_google("IBM Plex Serif", family = "ibm")
+showtext_auto()
+
+data <-
+  tibble(x = 1:100, y = 1:100)
+
+annotate_text_size <- 2
+annotate_lineheight <- 1
+
+flow <-
+  data |>
+  ggplot(aes(x, y)) +
+  scale_x_continuous(minor_breaks = seq(10, 100, 10)) +
+  scale_y_continuous(minor_breaks = seq(10, 130, 10)) +
+  # main boxes
+  geom_rect(
+    xmin = 10, xmax = 50, ymin = 117, ymax = 127, color = "#93A1A1",
+    fill = "grey80", linewidth = 0.25
+  ) +
+  annotate("text",
+           x = 30, y = 122,
+           label = "2278 nights recorded with\nZmachine during SCREENS trial",
+           size = annotate_text_size, color = "grey15", lineheight = annotate_lineheight
+  ) +
+  geom_rect(
+    xmin = 10, xmax = 50, ymin = 92, ymax = 102, color = "#93A1A1",
+    fill = "grey80", linewidth = 0.25
+  ) +
+  annotate("text",
+           x = 30, y = 97,
+           label = "1174 night from children\nwith accelerometer recordings",
+           size = annotate_text_size, color = "grey15", lineheight = annotate_lineheight
+  ) +
+  geom_rect(
+    xmin = 10, xmax = 50, ymin = 67, ymax = 77, color = "#93A1A1",
+    fill = "grey80", linewidth = 0.25
+  ) +
+  annotate("text",
+           x = 30, y = 72, label = "1032 night with measured sleep",
+           size = annotate_text_size, color = "grey15", lineheight = annotate_lineheight
+  ) +
+  geom_rect(
+    xmin = 10, xmax = 50, ymin = 42, ymax = 52, color = "#93A1A1",
+    fill = "grey80", linewidth = 0.25
+  ) +
+  annotate("text",
+           x = 30, y = 47, label = "779 nights with a duration\nbetween 7 and 14 hours",
+           size = annotate_text_size, color = "grey15", lineheight = annotate_lineheight
+  ) +
+  geom_rect(
+    xmin = 10, xmax = 50, ymin = 17, ymax = 27, color = "#93A1A1",
[…]

### 2023-06-26 manuscript/code/table_age_and_summary_sleep_stats_from_zm.R

@@ -3,61 +3,24 @@ library(tidyverse)
 library(slider)
 library(arrow)
 library(stringr)
+library(gt)
 
-# Function to count zeros in runs of at least three consecutive zeros (WASO)
+
 count_each_zero_in_consecutive_zeros <- function(x, n = 3) {
   rle_x <- rle(x == 0)
   sum(rle_x$lengths[rle_x$values & rle_x$lengths >= n])
 }
 
-# File path
-all_preds_filenames <- list.files("data/data_for_modelling/chained_classifiers/sleep_predictions", full.names = TRUE)
 
-# Extract model names
-model_names <- str_extract(all_preds_filenames, "(?<=/)[^/]+(?=_sleep_prediction_complete.parquet)")
+acc_ids <-
+  read_parquet("data/data_for_modelling/no_edge_sp_incl_features_30_sec_epochs.parquet") %>%
+  pull(id) %>%
+  unique()
 
-# Different measures of sleep
-sleep_types <- c("raw", "median_5", "median_10")
-
-# Use map from purrr package to apply function over sleep measure
-all_sleep_stats <- all_preds_filenames %>%
-  set_names(model_names) %>%
-  map(function(file) {
-    test_dc <- read_parquet(file) %>%
-      select(id, noon_day, month, datetime, matches("in_bed|sleep")) %>%
-      # filter(id == 8504) %>%
-      group_by(id, noon_day, month) %>%
-      mutate(
-        across(where(is.factor), as.numeric) - 1,
-        across(matches("sleep|in_bed"), ~ slide_sum(.x, before = 12, after = 12), .names = "{.col}_slide_sum_12")
-      )
-
-    map_dfr(sleep_types, function(st) {
-      test_dc %>%
-        summarise(
-          spt_hrs = if (any(in_bed_class_slide_sum_12 >= 20)) {
-            (max(row_number()[in_bed_class_slide_sum_12 >= 20]) - min(row_number()[in_bed_class_slide_sum_12 >= 20])) * 30 / 60 / 60
-          } else {
-            NA
-          },
-          tst_hrs = sum(get(paste0("sleep_", st, "_pred_class"))[get(paste0("sleep_", st, "_pred_class_slide_sum_12")) >= 20]) * 30 / 60 / 60,
[…]

### 2023-06-27 manuscript/manuscript.tex
pr and roc curves

@@
 
+\usepackage{booktabs}
+\usepackage{caption}
+\usepackage{longtable}
 \usepackage{lineno}\linenumbers
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \hypertarget{tbl-summary_zm}{}
-\begin{longtable}[]{@{}llllllll@{}}
-\caption{\label{tbl-summary_zm}Summary of sleep quality metrics as
-calculated based on the raw, the 5-minute, and 10-minute median filtered
-raw ZM prediction.}\tabularnewline
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2400}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1500}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1500}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1500}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1500}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1500}}@{}}
+\caption{\label{tbl-summary_zm}Summary of Sleep Quality Metrics based on
+Raw ZM Predictions and Median Filtering.}\tabularnewline
 \toprule\noalign{}
-& & Raw & & Median 5-Minute & & Median 10-Minute & \\
+\begin{minipage}[b]{\linewidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+SPT
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+TST
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+SE
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+LPS
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+WASO
+\end{minipage} \\
 \midrule\noalign{}
@@
 \toprule\noalign{}
-& & Raw & & Median 5-Minute & & Median 10-Minute & \\
+\begin{minipage}[b]{\linewidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+SPT
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+TST
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+SE
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+LPS
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+WASO
+\end{minipage} \\
 \midrule\noalign{}
@@
 \endlastfoot
-& & Mean & SD & Mean & SD & Mean & SD \\
-& SPT & 9.2 & 2.1 & 9.2 & 2.1 & 9.2 & 2.1 \\
-& TST & 7.7 & 1.9 & 8.1 & 2.0 & 8.2 & 2.1 \\
-& SE & 82.6 & 12.0 & 86.4 & 12.7 & 87.5 & 12.9 \\
-& LPS & 34.5 & 27.9 & 36.3 & 39.8 & 38.0 & 48.7 \\
-& WASO & 39.0 & 33.6 & 30.6 & 46.8 & 22.3 & 55.4 \\
+Raw & 9.2 (2.1) & 7.7 (1.9) & 82.6 (12.0) & 34.5 (27.9) & 39.0 (33.6) \\
+5-Min Median & 9.2 (2.1) & 8.1 (2.0) & 86.4 (12.7) & 36.3 (39.8) & 30.6
+(46.8) \\
+10-Min Median & 9.2 (2.1) & 8.2 (2.1) & 87.5 (12.9) & 38.0 (48.7) & 22.3
+(55.4) \\
 \end{longtable}
+
+\hypertarget{performance-on-epoch-to-epoch-basis}{%
+\subsection{Performance on Epoch-To-Epoch
+Basis}\label{performance-on-epoch-to-epoch-basis}}
+
+The results from the epoch-to-epoch evaluation of in-bed performance
+metrics for our included models are presented in
+Table~\ref{tbl-in_bed_performance}. The epoch-to-epoch evaluation of
+in-bed performance metrics, outlined in @tbl-in\_bed\_performance,
+demonstrates practically equivalent performance across all model types.
+While the Decision Tree model posted an F1 score of 94.4\% and accuracy
+of 95.3\%, the Logistic Regression and Feed-Forward Neural Network
+models each exhibited an F1 score of 95.0\% and similar accuracies of
+95.7\% and 95.8\%, respectively. The XGBoost model, despite recording
+the highest metrics with an F1 score of 95.4\% and accuracy of 96.1\%,
+outpaced the others only marginally. This underscores the consistency of
+performance among these models in monitoring in-bed conditions.
+
+\hypertarget{tbl-in_bed_performance}{}
+\begin{longtable}{lrrrrr}
+\caption{\label{tbl-in_bed_performance}In-Bed Performance Metrics }\tabularnewline
+
+\toprule
+model & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
+\midrule
+Decision Tree & $9,436.3\%$ & $9,527.2\%$ & $9,312.3\%$ & $9,563.6\%$ & $9,686.0\%$ \\ 
+Logistic Regression & $9,498.9\%$ & $9,573.9\%$ & $9,503.5\%$ & $9,494.3\%$ & $9,626.0\%$ \\ 
+Feed-Forward Neural Network & $9,502.9\%$ & $9,577.4\%$ & $9,507.1\%$ & $9,498.7\%$ & $9,629.2\%$ \\ 
+XGBoost & $9,538.4\%$ & $9,605.8\%$ & $9,582.7\%$ & $9,494.4\%$ & $9,622.9\%$ \\ 
+\bottomrule
+\end{longtable}
+
+Table~\ref{tbl-sleep_performance} details the performance metrics of the
+four machine learning models---Decision Tree, Logistic Regression,
+Neural Network, and XGBoost---applied to raw ZM predictions and 5- and
+10-minute median filtered predictions. For raw ZM predictions, all
+models demonstrated comparable performance, with F1 Scores between
+93.27\% and 93.58\%. The XGBoost model performed marginally better.
+However, it's crucial to highlight that Specificity was notably low
+across models, with the Neural Network model achieving the highest at
+48.31\%. Performance improved on 5-minute median data. XGBoost again
+outperformed other models in F1 Score and Accuracy, yet Specificity
+remained a concern, with scores ranging from 37.47\% to 50.90\%. On
+10-minute median data, all models showed further improvement, and
+XGBoost continued to lead in F1 Score and Accuracy. It is significant
+that both Neural Network and XGBoost models achieved a Sensitivity of
+99.10\%. However, low Specificity scores persisted, with the highest
+only at 50.98\% for the XGBoost model.
+
+\hypertarget{tbl-sleep_performance}{}
+\begin{longtable}{cccccc}
+\caption{\label{tbl-sleep_performance}Sleep Performance Metrics }\tabularnewline
+
+\toprule
+ & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
+\midrule
+\multicolumn{6}{l}{Raw ZM Predictions} \\ 
+\midrule
+Decision Tree & $93.27$ & $88.25$ & $96.92$ & $89.88$ & $42.73$ \\ 
+Logistic Regression & $93.30$ & $88.14$ & $98.33$ & $88.76$ & $34.65$ \\ 
+Neural Network & $93.54$ & $88.80$ & $96.52$ & $90.74$ & $48.31$ \\ 
+XGBoost & $93.58$ & $88.84$ & $96.92$ & $90.47$ & $46.41$ \\ 
+\midrule
+\multicolumn{6}{l}{5-Min Median} \\ 
+\midrule
+Decision Tree & $95.48$ & $91.87$ & $98.02$ & $93.07$ & $48.01$ \\ 
+Logistic Regression & $95.37$ & $91.55$ & $99.14$ & $91.87$ & $37.47$ \\ 
+Neural Network & $95.75$ & $92.33$ & $98.39$ & $93.24$ & $49.14$ \\ 
+XGBoost & $95.78$ & $92.41$ & $98.23$ & $93.45$ & $50.90$ \\ 
+\midrule
+\multicolumn{6}{l}{10-Min Median} \\ 
+\midrule
+Decision Tree & $96.02$ & $92.75$ & $98.30$ & $93.84$ & $47.95$ \\ 
+Logistic Regression & $96.04$ & $92.73$ & $99.20$ & $93.09$ & $40.56$ \\ 
+Neural Network & $96.18$ & $93.06$ & $98.29$ & $94.17$ & $50.90$ \\ 
+XGBoost & $96.60$ & $93.79$ & $99.10$ & $94.22$ & $50.98$ \\ 
+\bottomrule
+\end{longtable}
+
+Analyzing the area under the curve for precision-recall (AUC-PR) for all
+models across raw ZM predictions and 5- and 10-minute median filtered
+predictions yielded distinct results. Across all types of ZM
+predictions, the decision tree consistently yielded the highest AUC-PR,
+whereas the Neural Network generally performed least effectively. This
+highlights the superior predictive accuracy of the decision tree model
+in this context (see Figure~\ref{fig-pr_curves}). The Area Under the
+Receiver Operating Characteristics Curve (ROC-AUC) results indicate a
+varying performance of the four models across the ZM predictions types.
+The XGBoost model consistently outperforms the other models across all
+data types, suggesting its superior capacity for distinguishing between
+classes. In contrast, the Neural Network model consistently exhibits the
+lowest ROC-AUC values, suggesting a weaker performance in this specific
+context (see Figure~\ref{fig-roc_curves}).
+
+\begin{figure}[b]
+
+{\centering \includegraphics{visuals/plot_sleep_pr.pdf}
+
+}
+
+\caption{\label{fig-pr_curves}Precision-Recall curves of the models
+evaluated across the different ZM predictions, including raw ZM
+predictions, as well as 5-minute and 10-minute median smoothing of the
+ZM raw predictions. The x-axis of the plot represents the proportion of
+true wake epochs that were correctly classified as wake, while the
+y-axis represents the proportion of all epochs labeled as wake by the
+classifier that were classified correctly. The AUC values displayed in
+the plot indicate the area under the Precision-Recall curve for each
+model and condition.}
+
+\end{figure}
+
+\begin{figure}[b]
+
+{\centering \includegraphics{visuals/plot_sleep_roc.pdf}
+
+}
+
+\caption{\label{fig-roc_curves}Text bla bla}
+
+\end{figure}
 

### 2023-06-27 code/evaluate_models/create_pr_and_roc_data_all_models.R

@@ -1,0 +1,44 @@
+#!/usr/bin/env
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+
+# File processing function
+process_files <- function(suffix){
+  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/sleep",
+             full.names = TRUE
+  ) %>%
+    str_subset(suffix) %>%
+    str_subset("30_sec") %>%
+    map(read_rds) %>%
+    set_names(c("Decision Tree", "Logistic Regression", "Neural Network", "XGBoost"))
+}
+
+fits_30_raw <- process_files("raw")
+fits_30_median5 <- process_files("5_min")
+fits_30_median10 <- process_files("10_min")
+
+test_30 <-
+  read_parquet("data/data_for_modelling/chained_classifiers/30_sec_testing_data.parquet") %>%
+  mutate(
+    across(contains("sleep"), as_factor)
+  )
+
+# Metric calculation function
+calc_metrics <- function(fits, target, file_prefix){
+  pr <- fits %>% map_dfr(., ~ augment(.x, test_30) %>% pr_curve(., target, .pred_1), .id = "model")
+  pr_auc <- fits %>% map_dfr(., ~ .x %>% augment(test_30) %>% pr_auc(., target, .pred_1), .id = "model")
+  roc <- fits %>% map_dfr(., ~ augment(.x, test_30) %>% roc_curve(., target, .pred_1), .id = "model")
+  roc_auc <- fits %>% map_dfr(., ~ .x %>% augment(test_30) %>% roc_auc(., target, .pred_1), .id = "model")
+  
+  # Writing files
+  bind_rows(pr, pr_auc, roc, roc_auc, .id = "group") %>%
+    write_parquet(paste0("data/processed/all_", file_prefix, "_sleep_30.parquet"))
+  bind_rows(pr_auc, roc_auc, .id = "group") %>%
+    write_csv(paste0("data/processed/all_", file_prefix, "_auc_sleep_30_sec_epochs.csv"))
+}
+
+calc_metrics(fits_30_raw, "sleep", "raw")
+calc_metrics(fits_30_median5, "sleep_median5", "median_5")
+calc_metrics(fits_30_median10, "sleep_median10", "median_10")

### 2023-06-27 code/evaluate_models/create_roc_data_all_models.R

@@ -1,173 +1,0 @@
-#!/usr/bin/env
-
-library(tidyverse)
-library(tidymodels)
-library(arrow)
-
-
-get_roc <-
-  function(fit, test, truth, estimate) {
-    fit |>
-      augment(test) |>
-      roc_curve(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
-  }
-
-get_roc_auc <-
-  function(fit, test, truth, estimate) {
-    fit |>
-      augment(test) |>
-      roc_auc(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second")
-  }
-
-## Get the file names of the fitted models for the 10-second epochs, grouped by preprocessing type (raw, median5, and median10)
-fits_10_filenames_raw <- 
-  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/sleep",
-             full.names = TRUE) %>% 
-  str_subset("raw") %>% 
-  str_subset("30_sec")
-
-
-fits_10_filenames_median5 <- 
-  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/sleep",
-             full.names = TRUE) %>% 
-  str_subset("5_min") %>% 
-  str_subset("30_sec")
-
-
-fits_10_filenames_median10 <- 
-  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/sleep",
-             full.names = TRUE) %>% 
-  str_subset("10_min") %>% 
-  str_subset("30_sec")
-
-
-# Read the models from the file names and store them in separate lists
-fits_10_raw <- map(fits_10_filenames_raw, read_rds)
-fits_10_median5 <- map(fits_10_filenames_median5, read_rds)
-fits_10_median10 <- map(fits_10_filenames_median10, read_rds)
-
-# Repeat the same process for the 30-second epochs
-fits_30_filenames_raw <- 
-  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/sleep",
-             full.names = TRUE) %>% 
-  str_subset("raw") %>% 
-  str_subset("30_sec")
-
-fits_30_filenames_median5 <- 
[…]

### 2023-06-27 code/evaluate_models/get_metrics_for_in_bed_models.R

@@ -1,0 +1,84 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+library(gt)
+library(showtext)
+
+# # Define a set of metrics to be used for evaluation
+# my_metrics <-
+#   metric_set(f_meas, accuracy, sensitivity, precision, specificity)
+#
+# # Function to compute the specified performance metrics using the provided model and test data
+# get_metrics <-
+#   function(fit, test_data, truth, estimate, estimator = "binary") {
+#     fit %>%
+#       augment(test_data) %>%
+#       my_metrics(
+#         truth = {{ truth }}, estimate = {{ estimate }},
+#         event_level = "second",
+#         estimator = estimator
+#       )
+#   }
+#
+# in_bed_filenames <-
+#   list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/in_bed/", full.names = TRUE)
+#
+# in_bed_fits <-
+#   map(in_bed_filenames, read_rds) %>%
+#   set_names(c("decision_tree", "log_reg", "mlp", "xgboost"))
+#
+#
+# test_data_30 <-
+#   read_parquet("data/data_for_modelling/chained_classifiers/30_sec_testing_data.parquet") %>%
+#   mutate(
+#     in_bed = as_factor(in_bed)
+#   )
+#
+# # Calculate performance metrics for both models using the test datasets
+# metrics <-
+#   map_df(in_bed_fits, ~ get_metrics(.x, test_data_30, in_bed, .pred_class), .id = "model")
+
+
+# Save metrics as CSV
+# metrics %>%
+#   write_csv("data/processed/in_bed_metrics.csv")
+
+metrics <-
+  read_csv("data/processed/in_bed_metrics.csv")
+
+# Create a formatted table to display the metrics, grouped by epoch length
+tab_in_bed <-
+  metrics %>%
+  mutate(
+    .metric = factor(.metric,
+      levels = c("f_meas", "accuracy", "sensitivity", "precision", "specificity"),
+      labels = c("F1 Score", "Accuracy", "Sensitivity", "Precision", "Specificity")
+    )
+  ) %>%
+  select(-.estimator) %>%
+  pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  gt() %>%
+  fmt_percent(
+    columns = 2:6,
+    decimals = 2
+  ) %>%
+  tab_style(
[…]

### 2023-06-27 manuscript/code/table_in_bed_performance.R

@@ -1,0 +1,32 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+library(gt)
+library(showtext)
+
+metrics <-
+  read_csv(here::here("data/processed/in_bed_metrics.csv"))
+
+# Create a formatted table to display the metrics, grouped by epoch length
+tbl_in_bed <-
+  metrics %>%
+  mutate(
+    .estimate = .estimate * 100,
+    .metric = factor(.metric,
+                     levels = c("f_meas", "accuracy", "sensitivity", "precision", "specificity"),
+                     labels = c("F1 Score (%)", "Accuracy (%)", "Sensitivity (%)", "Precision (%)", "Specificity (%)")
+    ),
+    model = case_when(model == "decision_tree" ~ "Decision Tree",
+                      model == "log_reg" ~ "Logistic Regression",
+                      model == "mlp" ~ "Feed-Forward Neural Network",
+                      model == "xgboost" ~ "XGBoost")
+  ) %>%
+  select(-.estimator) %>%
+  pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  gt() %>%
+  fmt_percent(
+    columns = 2:6,
+    decimals = 1
+  ) 

### 2023-06-29 manuscript/manuscript.qmd
fix metrics and calculate n awakenings

@@
 analysis of the underlying patterns. All features are summarized in
-table ???. (overvej tabel til supp. material)
+table ??? (overvej tabel til supp. material)
 
@@
 
-### Model Validation \[evt bare "statistics"\]
+### Model Validation \[evt. bare "statistics"\]
 
-In our study, we employed standard evaluation metrics to assess the
-epoch-to-epoch performance of each model. These metrics include accuracy
-as $accuracy = \frac{TP+TN}{TP+TN+FP+FN}$, sensitivity as
-$sensitivity = \frac{TP}{TP+FN}$, specificity as
-$specificity = \frac{TN}{TN+FP}$, precision as
-$precision = \frac{TP}{TP+FP}$, F1 score as
-$F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}$, and
-area under the precision-recall curves were also calculated. The Average
-Precision, represented by the area under the precision-recall curve,
-provides a quantitative measure of performance. A value of 1 for Average
-Precision indicates the best performance. For the multiclass biLSTM
-model, we calculated performance metrics as macro-averages, to treat all
-classes equally. Consistent with previous studies in the
-field[@hjorth2012; @kushida2001], we treated the in-bed/out-of-bed and
-awake/asleep scoring tasks as a binary classification problems, where
-in-bed and asleep was considered the positive label and out-of-bed and
-awake the negative label.
+In our study, we utilized standard evaluation metrics to assess the
+performance of each model on an epoch-to-epoch basis. These include
+accuracy ($accuracy = \frac{TP+TN}{TP+TN+FP+FN}$), sensitivity
+($sensitivity = \frac{TP}{TP+FN}$), specificity
+($specificity = \frac{TN}{TN+FP}$), precision
+($precision = \frac{TP}{TP+FP}$), negative predictive value (NPV,
+$NPV = \frac{TN}{TN + FN}$), and F1 score
+($F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}$).
+
+In the context of our sequential learning strategy, the initial models
+were tasked with the binary classification of in-bed vs. out-of-bed. For
+this task, we assessed performance using the F1-score, accuracy,
+sensitivity, specificity, and precision metrics. The second models in
+our sequential learning strategy focused on the binary classification of
+asleep vs. awake. For these models, we considered the same metrics, in
+addition to the negative predictive rate. The class imbalance in this
+case led us to compute the F1 score as an unweighted macro-average.
+Additionally, we evaluated the multiclass classifier, biLSTM, using
+macro-averaged F1-score, sensitivity, and precision. Furthermore, we
+present precision-recall curves and receiver operating characteristic
+curves, including the area under the curve, for each model and for each
+type of ZM prediction.We considered both the in-bed/out-of-bed and
+awake/asleep scoring tasks as binary classification problems,
+designating in-bed and asleep as the positive labels and out-of-bed and
+awake as the negative labels in accordance with previous
+research[@hjorth2012; @kushida2001].
 
@@
 
-The children children icluded in the analyses were 9.4 years old (SD = 2.1).
+The children children included in the analyses were 9.4 years old (SD =
+2.1). The number of sleep epochs recorded in the study was 2,035,261,
+which accounted for approximately 86% of the total sleep duration. This
+metric indicates the granularity of sleep measurement, capturing the
+discrete moments of sleep throughout the sleep period. Sleep epochs are
+essential for analyzing sleep patterns and stages, as well as for
+assessing sleep quality and efficiency.
 
-As reported in @tbl-summary_zm the sleep quality metrics derived from ZMachine predictions were modified by the implementation of 5-minute and 10-minute median filters.
-The Sleep Period Time (SPT) remained consistent across raw and filtered datasets (mean: 9.2 ± 2.1 hours).
-Total Sleep Time (TST) and Sleep Efficiency (SE) increased in the filtered data, implying the filters categorize some wakefulness as sleep.
-Specifically, TST increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours (5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively.
-Latency to Persistent Sleep (LPS) also elevated, suggesting the filter smooths out brief awakenings at sleep onset, leading to a prolonged time to persistent sleep.
-The most significant change was seen in Wake After Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and 10-minute filtered data, respectively.
-The results underscore the influential role of median filters on sleep metrics.
+As reported in @tbl-zm_overview the sleep quality metrics derived from
+ZM predictions were modified by the implementation of 5-minute and
+10-minute median filters. The Sleep Period Time (SPT) were consistent
+across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as this
+corresponds to the length of the ZM recording. Total Sleep Time (TST)
+and Sleep Efficiency (SE) increased in the filtered data, implying the
+filters categorize some wakefulness as sleep. Specifically, TST
+increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
+(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
+from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. Latency
+to Persistent Sleep (LPS) also elevated, suggesting the filter smooths
+out brief awakenings at sleep onset, leading to a prolonged time to
+persistent sleep. The most significant change was seen in Wake After
+Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
+to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
+10-minute filtered data, respectively. The results underscore the role
+of median filters on sleep metrics.
 
-|||Raw| |Median 5-Minute| |Median 10-Minute| |
-|-|-|-|-|-|-|-|-|
-|||Mean|SD|Mean|SD|Mean|SD|
-||SPT|9.2|2.1|9.2|2.1|9.2|2.1|
-||TST|7.7|1.9|8.1|2.0|8.2|2.1|
-||SE|82.6|12.0|86.4|12.7|87.5|12.9|
-||LPS|34.5|27.9|36.3|39.8|38.0|48.7|
-||WASO|39.0|33.6|30.6|46.8|22.3|55.4|
+```{r}
+#| echo: false
+#| message: false
+#| label: tbl-zm_overview
+#| tbl-cap: Overview of characteristics of the ZM sleep quality summaries. Values are represented as mean (SD).
 
-: Summary of sleep quality metrics as calculated based on the raw, the 5-minute, and 10-minute median filtered raw ZM prediction. {#tbl-summary_zm}
+source("code/table_overview_ZM_stats.R")
+
+tbl_overview
+```
+
+## Performance on Epoch-To-Epoch Basis
+
+The epoch-to-epoch evaluation of in-bed performance metrics, outlined in
+@tbl-in_bed_performance, demonstrates practically equivalent performance
+across all model types. While the Decision Tree model posted an F1 score
+of 94.4% and accuracy of 95.3%, the Logistic Regression and Feed-Forward
+Neural Network models each exhibited an F1 score of 95.0% and similar
+accuracies of 95.7% and 95.8%, respectively. The XGBoost model, despite
+recording the highest metrics with an F1 score of 95.4% and accuracy of
+96.1%, outpaced the others only marginally. This underscores the
+consistency of performance among these models in classifying in-bed
+conditions.
+
+```{r}
+#| echo: false
+#| message: false
+#| label: tbl-in_bed_performance
+#| tbl-cap: In-Bed Performance Metrics
+
+source("code/table_in_bed_performance.R")
+
+tbl_in_bed
+```
+
+@tbl-sleep_performance details the performance of the sequential models
+on raw and median-filtered (5 and 10 minute) ZM predictions. The F1
+Scores, which are unweighted macro averages, for raw ZM predictions
+range from 71.05% to 76.18%. The models perform comparably, but the low
+Specificity values (62.84% to 70.93%) suggest difficulty in correctly
+classifying negative instances. Applying 5-minute median filtering
+improves the performance metrics. The XGBoost model tops the charts with
+an F1 Score of 79.22% and NPV of 74.00%. However, Specificity still
+remains low, with values between 54.68% and 74.84% across all models.
+With 10-minute median filtering, the metrics improve further. The
+XGBoost model still leads with an F1 Score of 80.87% and an NPV of
+75.76%. But, Specificity remains a concern, ranging from 57.47% to
+76.35% across all models.
+
+```{r}
+#| echo: false
+#| message: false
+#| tbl-colwidths: auto
+#| label: tbl-sleep_performance
+#| tbl-cap: Sleep Performance Metrics
+
+source("code/table_sleep_performance.R")
+
+tbl_sleep_performance
+```
+
+The analysis of precision-recall and ROC curves across different models
+and ZM prediction types shows varying performance. In terms of
+precision-recall AUC, the Decision Tree model consistently outperforms
+others, indicating its superior predictive accuracy (see
+@fig-pr_curves). Conversely, the Neural Network model generally shows
+weaker performance. However, for ROC AUC, the XGBoost model consistently
+excels across all data types, indicating a strong ability to
+differentiate between classes, while the Neural Network model tends to
+underperform (see @fig-roc_curves). The F-measure (F1 score) shows
+variable performance across different configurations but generally, the
+Decision Tree model yields higher scores.
+
+![Precision-recall curves of the models evaluated across the different
+ZM predictions, including raw ZM predictions, as well as 5-minute and
+10-minute median smoothing of the ZM raw predictions. The x-axis of the
+plot represents the proportion of true wake epochs that were correctly
+classified as wake, while the y-axis represents the proportion of all
+epochs labeled as wake by the classifier that were classified correctly.
+The area under the curve values are displayed as color-coded text in the
+plot indicate the area under the Precision-Recall curve for each model
+and condition.](visuals/plot_sleep_pr.pdf){#fig-pr_curves}
+
+![Receiver operating characteristic curves of the models evaluated
+across the different ZM predictions, including raw ZM predictions, as
+well as 5-minute and 10-minute median smoothing of the ZM raw
+predictions. The x-axis of the plot represents the proportion of true
+asleep epochs that were incorrectly classified as awake, while the
+y-axis represents the proportion of all epochs labeled as awake by the
+classifier that were correctly classified. The area under the curve
+values displayed are displayed as color-coded text in the plot to
+indicate the area under the receiver operating characteristic curve for
+each model and condition.](visuals/plot_sleep_roc.pdf){#fig-roc_curves}
+
+The performance metrics of the biLSTM were...
 

### 2023-06-29 manuscript/manuscript.tex
fix metrics and calculate n awakenings

--- 2023-06-26 manuscript/manuscript.tex
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 more nuanced analysis of the underlying patterns. All features are
-summarized in table ???. (overvej tabel til supp. material)
+summarized in table ??? (overvej tabel til supp. material)
 
@@
 
-\hypertarget{model-validation-evt-bare-statistics}{%
-\subsubsection{Model Validation {[}evt bare
-``statistics''{]}}\label{model-validation-evt-bare-statistics}}
+\hypertarget{model-validation-evt.-bare-statistics}{%
+\subsubsection{Model Validation {[}evt. bare
+``statistics''{]}}\label{model-validation-evt.-bare-statistics}}
 
-In our study, we employed standard evaluation metrics to assess the
-epoch-to-epoch performance of each model. These metrics include accuracy
-as \(accuracy = \frac{TP+TN}{TP+TN+FP+FN}\), sensitivity as
-\(sensitivity = \frac{TP}{TP+FN}\), specificity as
-\(specificity = \frac{TN}{TN+FP}\), precision as
-\(precision = \frac{TP}{TP+FP}\), F1 score as
-\(F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}\),
-and area under the precision-recall curves were also calculated. The
-Average Precision, represented by the area under the precision-recall
-curve, provides a quantitative measure of performance. A value of 1 for
-Average Precision indicates the best performance. For the multiclass
-biLSTM model, we calculated performance metrics as macro-averages, to
-treat all classes equally. Consistent with previous studies in the
-field\citep{hjorth2012, kushida2001}, we treated the in-bed/out-of-bed
-and awake/asleep scoring tasks as a binary classification problems,
-where in-bed and asleep was considered the positive label and out-of-bed
-and awake the negative label.
+In our study, we utilized standard evaluation metrics to assess the
+performance of each model on an epoch-to-epoch basis. These include
+accuracy (\(accuracy = \frac{TP+TN}{TP+TN+FP+FN}\)), sensitivity
+(\(sensitivity = \frac{TP}{TP+FN}\)), specificity
+(\(specificity = \frac{TN}{TN+FP}\)), precision
+(\(precision = \frac{TP}{TP+FP}\)), negative predictive value (NPV,
+\(NPV = \frac{TN}{TN + FN}\)), and F1 score
+(\(F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}\)).
+
+In the context of our sequential learning strategy, the initial models
+were tasked with the binary classification of in-bed vs.~out-of-bed. For
+this task, we assessed performance using the F1-score, accuracy,
+sensitivity, specificity, and precision metrics. The second models in
+our sequential learning strategy focused on the binary classification of
+asleep vs.~awake. For these models, we considered the same metrics, in
+addition to the negative predictive rate. The class imbalance in this
+case led us to compute the F1 score as an unweighted macro-average.
+Additionally, we evaluated the multiclass classifier, biLSTM, using
+macro-averaged F1-score, sensitivity, and precision. Furthermore, we
+present precision-recall curves and receiver operating characteristic
+curves, including the area under the curve, for each model and for each
+type of ZM prediction.We considered both the in-bed/out-of-bed and
+awake/asleep scoring tasks as binary classification problems,
+designating in-bed and asleep as the positive labels and out-of-bed and
+awake as the negative labels in accordance with previous
+research\citep{hjorth2012, kushida2001}.
 
@@
 
-The children children icluded in the analyses were 9.4 years old (SD =
-2.1).
+The children children included in the analyses were 9.4 years old (SD =
+2.1). The number of sleep epochs recorded in the study was 2,035,261,
+which accounted for approximately 86\% of the total sleep duration. This
+metric indicates the granularity of sleep measurement, capturing the
+discrete moments of sleep throughout the sleep period. Sleep epochs are
+essential for analyzing sleep patterns and stages, as well as for
+assessing sleep quality and efficiency.
 
-As reported in Table~\ref{tbl-summary_zm} the sleep quality metrics
-derived from ZMachine predictions were modified by the implementation of
-5-minute and 10-minute median filters. The Sleep Period Time (SPT)
-remained consistent across raw and filtered datasets (mean: 9.2 ± 2.1
-hours). Total Sleep Time (TST) and Sleep Efficiency (SE) increased in
-the filtered data, implying the filters categorize some wakefulness as
-sleep. Specifically, TST increased from a raw mean of 7.7 ± 1.9 hours to
-8.1 ± 2.0 hours (5-minute filter) and 8.2 ± 2.1 hours (10-minute
-filter), while SE rose from 82.6 ± 12.0\% to 86.4 ± 12.7\% and 87.5 ±
-12.9\% respectively. Latency to Persistent Sleep (LPS) also elevated,
-suggesting the filter smooths out brief awakenings at sleep onset,
-leading to a prolonged time to persistent sleep. The most significant
-change was seen in Wake After Sleep Onset (WASO), which dropped from
-39.0 ± 33.6 minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4
-minutes in the 5-minute and 10-minute filtered data, respectively. The
-results underscore the influential role of median filters on sleep
-metrics.
+As reported in Table~\ref{tbl-zm_overview} the sleep quality metrics
+derived from ZM predictions were modified by the implementation of
+5-minute and 10-minute median filters. The Sleep Period Time (SPT) were
+consistent across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as
+this corresponds to the length of the ZM recording. Total Sleep Time
+(TST) and Sleep Efficiency (SE) increased in the filtered data, implying
+the filters categorize some wakefulness as sleep. Specifically, TST
+increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
+(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
+from 82.6 ± 12.0\% to 86.4 ± 12.7\% and 87.5 ± 12.9\% respectively.
+Latency to Persistent Sleep (LPS) also elevated, suggesting the filter
+smooths out brief awakenings at sleep onset, leading to a prolonged time
+to persistent sleep. The most significant change was seen in Wake After
+Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
+to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
+10-minute filtered data, respectively. The results underscore the role
+of median filters on sleep metrics.
 
-\hypertarget{tbl-summary_zm}{}
-\begin{longtable}[]{@{}llllllll@{}}
-\caption{\label{tbl-summary_zm}Summary of sleep quality metrics as
-calculated based on the raw, the 5-minute, and 10-minute median filtered
-raw ZM prediction.}\tabularnewline
-\toprule\noalign{}
-& & Raw & & Median 5-Minute & & Median 10-Minute & \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-& & Raw & & Median 5-Minute & & Median 10-Minute & \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-& & Mean & SD & Mean & SD & Mean & SD \\
-& SPT & 9.2 & 2.1 & 9.2 & 2.1 & 9.2 & 2.1 \\
-& TST & 7.7 & 1.9 & 8.1 & 2.0 & 8.2 & 2.1 \\
-& SE & 82.6 & 12.0 & 86.4 & 12.7 & 87.5 & 12.9 \\
-& LPS & 34.5 & 27.9 & 36.3 & 39.8 & 38.0 & 48.7 \\
-& WASO & 39.0 & 33.6 & 30.6 & 46.8 & 22.3 & 55.4 \\
+\hypertarget{tbl-zm_overview}{}
+\begin{longtable}{lrrrrrr}
+\caption{\label{tbl-zm_overview}Overview of characteristics of the ZM sleep quality summaries. Values
+are represented as mean (SD). }\tabularnewline
+
+\toprule
+ & SPT (hrs) & TST (hrs) & SE (\%) & LPS (min) & WASO (min) & Awakenings (N) \\ 
+\midrule
+Raw ZM Predictions & 9.21 (2.15) & 7.74 (1.93) & 82.61 (11.97) & 34.53 (27.87) & 39 (33.56) & 34.46 (11.33) \\ 
+5-Min Median & 9.21 (2.15) & 8.1 (2.03) & 86.35 (12.72) & 36.26 (39.8) & 30.63 (46.84) & 4.43 (3.26) \\ 
+10-Min Median & 9.21 (2.15) & 8.21 (2.06) & 87.54 (12.89) & 37.99 (48.7) & 22.27 (55.38) & 1.95 (2.01) \\ 
+\bottomrule
 \end{longtable}
+
+\hypertarget{performance-on-epoch-to-epoch-basis}{%
+\subsection{Performance on Epoch-To-Epoch
+Basis}\label{performance-on-epoch-to-epoch-basis}}
+
+The epoch-to-epoch evaluation of in-bed performance metrics, outlined in
+Table~\ref{tbl-in_bed_performance}, demonstrates practically equivalent
+performance across all model types. While the Decision Tree model posted
+an F1 score of 94.4\% and accuracy of 95.3\%, the Logistic Regression
+and Feed-Forward Neural Network models each exhibited an F1 score of
+95.0\% and similar accuracies of 95.7\% and 95.8\%, respectively. The
+XGBoost model, despite recording the highest metrics with an F1 score of
+95.4\% and accuracy of 96.1\%, outpaced the others only marginally. This
+underscores the consistency of performance among these models in
+classifying in-bed conditions.
+
+\hypertarget{tbl-in_bed_performance}{}
+\begin{longtable}{lrrrrr}
+\caption{\label{tbl-in_bed_performance}In-Bed Performance Metrics }\tabularnewline
+
+\toprule
+model & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
+\midrule
+Decision Tree & $94.36$ & $95.27$ & $93.12$ & $95.64$ & $96.86$ \\ 
+Logistic Regression & $94.99$ & $95.74$ & $95.03$ & $94.94$ & $96.26$ \\ 
+Feed-Forward Neural Net & $95.03$ & $95.77$ & $95.07$ & $94.99$ & $96.29$ \\ 
+XGBoost & $95.38$ & $96.06$ & $95.83$ & $94.94$ & $96.23$ \\ 
+\bottomrule
+\end{longtable}
+
+Table~\ref{tbl-sleep_performance} details the performance of the
+sequential models on raw and median-filtered (5 and 10 minute) ZM
+predictions. The F1 Scores, which are unweighted macro averages, for raw
+ZM predictions range from 71.05\% to 76.18\%. The models perform
+comparably, but the low Specificity values (62.84\% to 70.93\%) suggest
+difficulty in correctly classifying negative instances. Applying
+5-minute median filtering improves the performance metrics. The XGBoost
+model tops the charts with an F1 Score of 79.22\% and NPV of 74.00\%.
+However, Specificity still remains low, with values between 54.68\% and
+74.84\% across all models. With 10-minute median filtering, the metrics
+improve further. The XGBoost model still leads with an F1 Score of
+80.87\% and an NPV of 75.76\%. But, Specificity remains a concern,
+ranging from 57.47\% to 76.35\% across all models.
+
+\hypertarget{tbl-sleep_performance}{}
+\begin{longtable}{cccccc}
+\caption{\label{tbl-sleep_performance}Sleep Performance Metrics }\tabularnewline
+
+\toprule
+ & F1 Score (\%) & Precision (\%) & NPV (\%) & Sensitivity (\%) & Specificity (\%) \\ 
+\midrule
+\multicolumn{6}{l}{Raw ZM Predictions} \\ 
+\midrule
+Decision Tree & $72.94$ & $93.24$ & $48.36$ & $86.34$ & $67.15$ \\ 
+Logistic Regression & $71.05$ & $93.72$ & $43.88$ & $82.72$ & $70.93$ \\ 
+Neural Network & $71.76$ & $93.77$ & $45.13$ & $83.59$ & $70.83$ \\ 
+XGBoost & $76.18$ & $92.80$ & $57.98$ & $91.32$ & $62.84$ \\ 
+\midrule
+\multicolumn{6}{l}{5-Min Median} \\ 
+\midrule
+Decision Tree & $75.48$ & $94.20$ & $55.46$ & $93.35$ & $59.01$ \\ 
+Logistic Regression & $68.32$ & $95.84$ & $36.03$ & $81.36$ & $74.84$ \\ 
+Neural Network & $71.74$ & $95.78$ & $41.64$ & $85.62$ & $73.11$ \\ 
+XGBoost & $79.22$ & $93.87$ & $74.00$ & $97.30$ & $54.68$ \\ 
+\midrule
+\multicolumn{6}{l}{10-Min Median} \\ 
+\midrule
+Decision Tree & $76.28$ & $94.74$ & $58.06$ & $94.86$ & $57.47$ \\ 
+Logistic Regression & $67.96$ & $96.54$ & $34.29$ & $81.87$ & $76.35$ \\ 
+Neural Network & $70.95$ & $96.06$ & $39.54$ & $86.48$ & $71.37$ \\ 
+XGBoost & $80.87$ & $94.90$ & $75.76$ & $97.72$ & $57.60$ \\ 
+\bottomrule
+\end{longtable}
+
+The analysis of precision-recall and ROC curves across different models
+and ZM prediction types shows varying performance. In terms of
+precision-recall AUC, the Decision Tree model consistently outperforms
+others, indicating its superior predictive accuracy (see
+Figure~\ref{fig-pr_curves}). Conversely, the Neural Network model
+generally shows weaker performance. However, for ROC AUC, the XGBoost
+model consistently excels across all data types, indicating a strong
+ability to differentiate between classes, while the Neural Network model
+tends to underperform (see Figure~\ref{fig-roc_curves}). The F-measure
+(F1 score) shows variable performance across different configurations
+but generally, the Decision Tree model yields higher scores.
+
+\begin{figure}[b]
+
+{\centering \includegraphics{visuals/plot_sleep_pr.pdf}
+
+}
+
+\caption{\label{fig-pr_curves}Precision-recall curves of the models
+evaluated across the different ZM predictions, including raw ZM
+predictions, as well as 5-minute and 10-minute median smoothing of the
+ZM raw predictions. The x-axis of the plot represents the proportion of
+true wake epochs that were correctly classified as wake, while the
+y-axis represents the proportion of all epochs labeled as wake by the
+classifier that were classified correctly. The area under the curve
+values are displayed as color-coded text in the plot indicate the area
+under the Precision-Recall curve for each model and condition.}
+
+\end{figure}
+
+\begin{figure}[b]
+
+{\centering \includegraphics{visuals/plot_sleep_roc.pdf}
+
+}
+
+\caption{\label{fig-roc_curves}Receiver operating characteristic curves
+of the models evaluated across the different ZM predictions, including
+raw ZM predictions, as well as 5-minute and 10-minute median smoothing
+of the ZM raw predictions. The x-axis of the plot represents the
+proportion of true asleep epochs that were incorrectly classified as
+awake, while the y-axis represents the proportion of all epochs labeled
+as awake by the classifier that were correctly classified. The area
+under the curve values displayed are displayed as color-coded text in
+the plot to indicate the area under the receiver operating
+characteristic curve for each model and condition.}
+
+\end{figure}
+
+The performance metrics of the biLSTM were\ldots{}
 

### 2023-06-29 code/control_jobs.sh

@@ -8,14 +8,16 @@
 # 
 # ./code/create_models/chained_models/in_bed_workflows.R &&
 
-./code/create_models/chained_models/sleep_workflows.R &&
+# ./code/create_models/chained_models/sleep_workflows.R &&
 
 # ./code/create_models/chained_models/finalize_in_bed_workflows.R &&
 
-./code/create_models/chained_models/finalize_sleep_workflows.R &&
+# ./code/create_models/chained_models/finalize_sleep_workflows.R &&
 
 ./code/evaluate_models/create_pr_and_roc_data_all_models.R &&
 
 ./manuscript/code/plot_sleep_PR_and_ROC_data.R &&
 
 echo "Job's done!"
+
+sleep 5 && shutdown now

### 2023-06-29 code/create_models/chained_models/sleep_workflows.R

@@ -171,7 +171,8 @@ cat("Creating workflow sets...\n")
 #       decision_tree = CART_spec,
 #       # MARS = mars_spec, # works in parallel
 #       # random_forest = rf_spec, # works in parallel
-#       xgboost = xgb_spec # works in parallel
+#       xgboost = xgtest_30 <- load_data("data/data_for_modelling/chained_classifiers/30_sec_only_in_bed_testing_data.parquet")
+b_spec # works in parallel
 #     ),
 #     cross = TRUE
 #   ) %>%
@@ -281,7 +282,8 @@ grid_ctrl <-
     allow_par = TRUE,
     parallel_over = "everything",
     save_pred = FALSE,
-    save_workflow = FALSE
+    save_workflow = FALSE,
+    
   )
 
 tune_wf_and_write <- function(wfs, fname, folds) {

### 2023-06-29 code/evaluate_models/create_pr_and_roc_data_all_models.R

@@ -4,41 +4,59 @@ library(tidyverse)
 library(tidymodels)
 library(arrow)
 
-# File processing function
-process_files <- function(suffix){
-  list.files("/media/esbenlykke/My Passport/chained_models/fitted_workflows/sleep",
-             full.names = TRUE
-  ) %>%
-    str_subset(suffix) %>%
-    str_subset("30_sec") %>%
-    map(read_rds) %>%
-    set_names(c("Decision Tree", "Logistic Regression", "Neural Network", "XGBoost"))
-}
-
-fits_30_raw <- process_files("raw")
-fits_30_median5 <- process_files("5_min")
-fits_30_median10 <- process_files("10_min")
-
-test_30 <-
-  read_parquet("data/data_for_modelling/chained_classifiers/30_sec_testing_data.parquet") %>%
-  mutate(
-    across(contains("sleep"), as_factor)
-  )
-
-# Metric calculation function
-calc_metrics <- function(fits, target, file_prefix){
-  pr <- fits %>% map_dfr(., ~ augment(.x, test_30) %>% pr_curve(., target, .pred_1), .id = "model")
-  pr_auc <- fits %>% map_dfr(., ~ .x %>% augment(test_30) %>% pr_auc(., target, .pred_1), .id = "model")
-  roc <- fits %>% map_dfr(., ~ augment(.x, test_30) %>% roc_curve(., target, .pred_1), .id = "model")
-  roc_auc <- fits %>% map_dfr(., ~ .x %>% augment(test_30) %>% roc_auc(., target, .pred_1), .id = "model")
+# Define a function to read the models and calculate the PR and ROC curves
+process_models <- function(filenames, test_data, sleep_col) {
+  models <- map(filenames, read_rds) %>% set_names(model_names)
   
-  # Writing files
-  bind_rows(pr, pr_auc, roc, roc_auc, .id = "group") %>%
-    write_parquet(paste0("data/processed/all_", file_prefix, "_sleep_30.parquet"))
-  bind_rows(pr_auc, roc_auc, .id = "group") %>%
-    write_csv(paste0("data/processed/all_", file_prefix, "_auc_sleep_30_sec_epochs.csv"))
+  pr <- models %>% map_dfr(~ augment(.x, test_data) %>% pr_curve(., {{sleep_col}}, .pred_0), .id = "model")
[…]

### 2023-06-29 code/evaluate_models/get_metrics_for_sleep_models.R

@@ -8,13 +8,22 @@ library(gt)
 library(showtext)
 
 # Define a set of metrics to be used for evaluation
-my_metrics <- metric_set(f_meas, accuracy, sensitivity, precision, specificity)
+my_metrics <- metric_set(precision, npv, sensitivity, specificity)
 
 # Function to compute the specified performance metrics using the provided model and test data
-get_metrics <- function(fit, test_data, truth, estimate, estimator = "binary") {
-  fit %>%
-    augment(test_data) %>%
-    my_metrics(truth = {{ truth }}, estimate = {{ estimate }}, event_level = "second", estimator = estimator)
+get_metrics <- function(fit, test_data, truth, estimate) {
+  preds <- fit %>%
+    augment(test_data)
+  
+  f1_score <- preds %>%
+    f_meas(truth = {{ truth }}, estimate = {{ estimate }}, 
+           event_level = "second", estimator = "macro")
+  
+  other_metrics <- preds %>%
+    my_metrics(truth = {{ truth }}, estimate = {{ estimate }},
+               event_level = "second")
+  
+  bind_rows(f1_score, other_metrics)
 }
 
 # Define file paths for the model

### 2023-06-29 code/evaluate_models/get_summary_sleep_stats_from_zm.R

@@ -1,15 +1,69 @@
 #!/usr/bin/env Rscript
+
 library(tidyverse)
-library(slider)
+library(tidymodels)
 library(arrow)
-library(stringr)
-library(gt)
 
 
 count_each_zero_in_consecutive_zeros <- function(x, n = 3) {
   rle_x <- rle(x == 0)
   sum(rle_x$lengths[rle_x$values & rle_x$lengths >= n])
 }
+
+count_awakenings_min_duration <- function(x, min_length = 1) {
+  # Run-length encoding (RLE) is a simple form of data compression in which 
+  # runs of data (sequences in which the same data value occurs in many consecutive 
+  # data elements) are stored as a single data value and count.
+  rle_x <- rle(x == 0)
+  
+  # Create a sequence of positions where each 'awake' period starts
+  awakening_starts <- cumsum(rle_x$lengths) - rle_x$lengths + 1
+  
+  # Create a sequence of positions where each 'awake' period ends
+  awakening_ends <- cumsum(rle_x$lengths)
+  
+  # Record the lengths of each 'awake' period
+  awakening_lengths <- rle_x$lengths
+  
+  # Find positions where the 'awake' periods have lengths greater or equal to 
+  # the specified minimum length and are actually 'awake' (values are TRUE in rle_x)
+  valid_awakening_start_positions <- which(rle_x$values & awakening_lengths >= min_length)
+  
+  # If there are no valid awakenings, return 0
+  if (length(valid_awakening_start_positions) == 0) {
+    return(0)
+  }
+  
+  # Get start and end positions of valid awakenings
+  valid_awakening_starts <- awakening_starts[valid_awakening_start_positions]
+  valid_awakening_ends <- awakening_ends[valid_awakening_start_positions]
+  
+  # Initiate the count of valid awakenings
+  number_of_valid_awakenings <- sum(sapply(1:length(valid_awakening_starts), function(i) {
+    start_pos <- valid_awakening_starts[i]
+    end_pos <- valid_awakening_ends[i]
+    
+    # Check if awakening is surrounded by sleep epochs
+    # If there are no sleep epochs, then every awakening is considered valid
+    if (all(x == 0)) {
[…]

### 2023-06-29 manuscript/code/plot_sleep_PR_and_ROC_data.R

@@ -20,15 +20,13 @@ sleep_pr_auc <-
 pr_auc_labels <-
   sleep_pr_auc %>% 
   select(-.metric, -.estimator) %>% 
-    mutate(recall = .2, precision = rep(c(.25, .23, .21, .19), 3),
+    mutate(recall = .2, precision = rep(.5 + c(.25, .23, .21, .19), 3),
            label = paste(model, round(.estimate, digits = 2))) %>% 
   mutate(
     group = case_when(group == "raw" ~ "Raw ZM Predictions",
                       group == "median_5" ~ "5-Min Median",
                       group == "median_10" ~ "10-Min Median")
   ) 
-
-levels(sleep_pr_data$group)
 
 sleep_pr_data %>%
   mutate(
@@ -73,7 +71,7 @@ sleep_roc_auc <-
 roc_auc_labels <-
   sleep_pr_auc %>% 
   select(-.metric, -.estimator) %>% 
-  mutate(specificity = .75, sensitivity = rep(c(.25, .23, .21, .19), 3),
+  mutate(specificity = .6, sensitivity = rep(c(.25, .21, .17, .13), 3),
          label = paste(model, round(.estimate, digits = 2))) %>% 
   mutate(
     group = case_when(group == "raw" ~ "Raw ZM Predictions",

### 2023-06-29 manuscript/code/table_overview_ZM_stats.R

@@ -1,0 +1,32 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(glue)
+library(gt)
+library(showtext)
+
+stats <-
+  read_csv(here::here("data/processed/tbl_overview_sleep_summaries.csv"))
+
+# Create a formatted table to display the metrics, grouped by epoch length
+tbl_overview <-
+  stats %>% 
+  # pivot_wider(names_from = row, values_from = raw_stat:median_10_stat2) %>%
+  # mutate(raw_stat2_sleep_count_and_ratio = raw_stat2_sleep_count_and_ratio * 100,
+  #        median_5_stat2_sleep_count_and_ratio = median_5_stat2_sleep_count_and_ratio * 100,
+  #        median_10_stat2_sleep_count_and_ratio = median_5_stat2_sleep_count_and_ratio * 100)
+  transmute(
+    row = row,
+    "Raw ZM Predictions" = glue("{round(raw_stat, 2)} ({round(raw_stat2, 2)})"),
+    "5-Min Median" = glue("{round(median_5_stat, 2)} ({round(median_5_stat2, 2)})"),
+    "10-Min Median" = glue("{round(median_10_stat, 2)} ({round(median_10_stat2, 2)})")
+  ) %>%
+  pivot_longer(-row) %>%
+  pivot_wider(names_from = row, values_from = value) %>%
+  gt() %>%
+  cols_label(
+    name = "", spt = "SPT (hrs)", tst = "TST (hrs)", se = "SE (%)", 
+    lps = "LPS (min)", waso = "WASO (min)", no = "Awakenings (N)"
+  ) %>%
+  fmt_number(decimals = 1) %>% 
+  cols_align(align = "right", columns = -name)

### 2023-07-03 manuscript/manuscript.qmd
all stats for ML models calculated. Now stats for biLSTM.

@@
 
+something on sequence length and step size
+
 The biLSTM model was chosen for sleep classification using accelerometer
@@
     our analysis, a period is counted as 'awake' only if it consists of
-    3 or more contiguous 30-second epochs.
+    3 or more contiguous 30-second epochs which is also how the ZM
+    summarizes WASO.
 
@@
 
-The children children included in the analyses were 9.4 years old (SD =
-2.1). The number of sleep epochs recorded in the study was 2,035,261,
-which accounted for approximately 86% of the total sleep duration. This
-metric indicates the granularity of sleep measurement, capturing the
-discrete moments of sleep throughout the sleep period. Sleep epochs are
-essential for analyzing sleep patterns and stages, as well as for
-assessing sleep quality and efficiency.
+The analysis included children with an average age of 9.4 years (SD =
+2.1). The raw ZM predictions covered 2,035,261 sleep epochs, or about
+86% of the total recording duration.
 
@@
 to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
-10-minute filtered data, respectively. The results underscore the role
-of median filters on sleep metrics.
+10-minute filtered data, respectively. The number of awakenings was also
+considerably reduced with the application of filters. In the raw data,
+the average number of awakenings was 34.46 ± 11.33 per night, which
+reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the 5-minute and 10-minute
+filtered data sets respectively. These results underscore the role of
+median filters on sleep metrics.
 
@@
 #| label: tbl-zm_overview
-#| tbl-cap: Overview of characteristics of the ZM sleep quality summaries. Values are represented as mean (SD).
+#| tbl-cap: Overview of characteristics of the ZM sleep quality summaries per night. Values are represented as mean (SD).
 
@@
 
-The performance metrics of the biLSTM were...
+@tbl-biLSTM_performance presents the performance of the three-class
+biLSTM multiclassifier on raw and median-filtered (5-minute and
+10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores ranging
+from 71.36% to 76.04%, indicating overall good performance. Applying
+5-minute median filtering improves the metrics further, resulting in F1
+Scores ranging from 75.99% to 78.53%, demonstrating enhanced precision
+and sensitivity. However, Specificity values are not available for this
+configuration. With 10-minute median filtering, F1 Scores range from
+73.45% to 73.93%, maintaining good performance but with limited
+information on Specificity. Further analysis is required to assess
+performance across all classes.
+
+```{r}
+#| echo: false
+#| message: false
+#| tbl-colwidths: auto
+#| label: tbl-biLSTM_performance
+#| tbl-cap: Performance of the three-class biLSTM multicclassifier.
+source("code/table_biLSTM_performance.R")
+
+tbl_biLSTM_performance
+```
+
+## Evaluation of Sleep Quality Summaries
+
+BA plots
+
+pearson regression
+
+scatterplots with identity line and best linear fit
 

### 2023-07-03 manuscript/manuscript.tex
all stats for ML models calculated. Now stats for biLSTM.

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
   analysis, a period is counted as `awake' only if it consists of 3 or
-  more contiguous 30-second epochs.
+  more contiguous 30-second epochs which is also how the ZM summarizes
+  WASO.
 \end{enumerate}
@@
 
-The children children included in the analyses were 9.4 years old (SD =
-2.1). The number of sleep epochs recorded in the study was 2,035,261,
-which accounted for approximately 86\% of the total sleep duration. This
-metric indicates the granularity of sleep measurement, capturing the
-discrete moments of sleep throughout the sleep period. Sleep epochs are
-essential for analyzing sleep patterns and stages, as well as for
-assessing sleep quality and efficiency.
+The analysis included children with an average age of 9.4 years (SD =
+2.1). The raw ZM predictions covered 2,035,261 sleep epochs, or about
+86\% of the total recording duration.
 
@@
 \begin{longtable}{lrrrrrr}
-\caption{\label{tbl-zm_overview}Overview of characteristics of the ZM sleep quality summaries. Values
-are represented as mean (SD). }\tabularnewline
+\caption{\label{tbl-zm_overview}Overview of characteristics of the ZM sleep quality summaries per night.
+Values are represented as mean (SD). }\tabularnewline
 
@@
 \toprule
-model & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
+ & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
 \midrule
@@
 
-The performance metrics of the biLSTM were\ldots{}
+Table~\ref{tbl-biLSTM_performance} presents the performance of the
+three-class biLSTM multiclassifier on raw and median-filtered (5-minute
+and 10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores
+ranging from 71.36\% to 76.04\%, indicating overall good performance.
+Applying 5-minute median filtering improves the metrics further,
+resulting in F1 Scores ranging from 75.99\% to 78.53\%, demonstrating
+enhanced precision and sensitivity. However, Specificity values are not
+available for this configuration. With 10-minute median filtering, F1
+Scores range from 73.45\% to 73.93\%, maintaining good performance but
+with limited information on Specificity. Further analysis is required to
+assess performance across all classes.
+
+\hypertarget{tbl-biLSTM_performance}{}
+\begin{longtable}{cccc}
+\caption{\label{tbl-biLSTM_performance}Performance of the three-class biLSTM multicclassifier. }\tabularnewline
+
+\toprule
+ & F1 Score & Sensitivity & Precision \\ 
+\midrule
+Raw ZM Predictions & $71.36$ & $70.42$ & $76.04$ \\ 
+5-Min Median & $75.99$ & $74.62$ & $78.53$ \\ 
+10-Min Median & $73.45$ & $73.93$ & $73.07$ \\ 
+\bottomrule
+\end{longtable}
+
+\hypertarget{evaluation-of-sleep-quality-summaries}{%
+\subsection{Evaluation of Sleep Quality
+Summaries}\label{evaluation-of-sleep-quality-summaries}}
+
+BA plots
+
+pearson regression
+
+scatterplots with identity line and best linear fit
 

### 2023-07-03 code/create_models/lstm/create_tensors_pytorch.py

@@ -4,7 +4,7 @@ from torch.utils.data import Dataset
 from sklearn.model_selection import train_test_split
 
 # Define the helper function
-def prepare_data(data_path, sequence_length, step_size):
+def prepare_data(data_path, sequence_length, step_size, target_var):
     # Load the data using pandas
     data = pd.read_parquet(data_path)
     data = data[['age', 'weekday', 'incl', 'theta', 'x_mean', 'y_mean', 'z_mean',
@@ -25,16 +25,16 @@ def prepare_data(data_path, sequence_length, step_size):
               'vector_magnitude',
               'x_crossing_rate', 'y_crossing_rate', 'z_crossing_rate',
               'x_skewness', 'y_skewness', 'z_skewness',
-              'x_kurtosis', 'y_kurtosis', 'z_kurtosis', 'score_simple_filtered']]
+              'x_kurtosis', 'y_kurtosis', 'z_kurtosis', target_var]]
 
-    # Separate out 'score_simple_filtered'
-    score_simple_filtered = data.pop('score_simple_filtered')
+    # Separate out the target variable
+    target = data.pop(target_var)
 
     # Normalize the data
     data = (data - data.mean()) / data.std()
 
-    # Add 'score_simple_filtered' back to the dataframe
-    data['score_simple_filtered'] = score_simple_filtered
+    # Add the target variable back to the dataframe
+    data[target_var] = target
 
     # Convert to PyTorch tensors
     data_tensor = torch.tensor(data.values, dtype=torch.float32)
@@ -50,27 +50,30 @@ def prepare_data(data_path, sequence_length, step_size):
     sequences_tensor = torch.stack(sequences)
     labels_tensor = torch.stack(labels)
 
-    return sequences_tensor, labels_tensor
+    # Generate file names
+    base_name = data_path.split("/")[-1].replace(".parquet", "")
+    sequences_file = f"data/data_for_modelling/lstm/pytorch_{base_name}_sequences_{target_var}.pt"
+    labels_file = f"data/data_for_modelling/lstm/pytorch_{base_name}_labels_{target_var}.pt"
 
+    # Save tensors to files
+    torch.save(sequences_tensor, sequences_file)
+    torch.save(labels_tensor, labels_file)
 
+    return sequences_file, labels_file
 
 # Sequence length and step size
 sequence_length = 20  # Corresponding to 10 minutes in 30 sec epoch data
[…]

### 2023-07-03 code/create_models/lstm/inference.py

@@ -6,67 +6,146 @@ from torch.utils.data import TensorDataset
 from torch.utils.data import DataLoader
 from tqdm import tqdm
 
-# Load the trained model
-model = torch.load("/home/esbenlykke/projects/sleep_study/models/lstm_model.pt")
+# Hyperparameters
+input_size = 64
+sequence_length = 20  # or n, depending on your data
+step = 1
+hidden_size = 128
+num_layers = 4
+num_classes = 3
+learning_rate = 3e-4
+batch_size = 64
+num_epochs = 10
+
+# Load the trained models
+model_raw = biLSTM(input_size, hidden_size, num_layers, num_classes)
+model_raw.load_state_dict(torch.load("/home/esbenlykke/projects/sleep_study/models/score_simple_lstm_model.pt"))
+
+model_median_5 = biLSTM(input_size, hidden_size, num_layers, num_classes)
+model_median_5.load_state_dict(torch.load("/home/esbenlykke/projects/sleep_study/models/score_simple_median_5_lstm_model.pt"))
+
+model_median_10 = biLSTM(input_size, hidden_size, num_layers, num_classes)
+model_median_10.load_state_dict(torch.load("/home/esbenlykke/projects/sleep_study/models/score_simple_median_10_lstm_model.pt"))
 
 # Set device
 device = "cuda" if torch.cuda.is_available() else "cpu"
 
-model.to(device)
-model.eval()
+model_raw.to(device)
+model_raw.eval()
+model_median_5.to(device)
+model_median_5.eval()
+model_median_10.to(device)
+model_median_10.eval()
 
 # Load test tensors
-test_predictors = torch.load("/home/esbenlykke/projects/sleep_study/data/data_for_modelling/lstm/pytorch_test_sequences.pt")
-test_labels = torch.load("/home/esbenlykke/projects/sleep_study/data/data_for_modelling/lstm/pytorch_test_labels.pt")
+test_predictors_raw = torch.load("/home/esbenlykke/projects/sleep_study/data/data_for_modelling/lstm/pytorch_30_sec_testing_sequences_score_simple.pt")
+test_labels_raw = torch.load("/home/esbenlykke/projects/sleep_study/data/data_for_modelling/lstm/pytorch_30_sec_testing_labels_score_simple.pt")
+test_predictors_median_5 = torch.load("/home/esbenlykke/projects/sleep_study/data/data_for_modelling/lstm/pytorch_30_sec_testing_sequences_score_simple_median_5.pt")
[…]

### 2023-07-03 code/create_models/lstm/train_lstm_pytorch.py

@@ -1,11 +1,11 @@
 # Imports
 import torch
-import torch.nn as nn  
-import torch.optim as optim  
-import torch.nn.functional as F  
+import torch.nn as nn
+import torch.optim as optim
+import torch.nn.functional as F
 from torch.utils.data import DataLoader
 from torch.utils.data import TensorDataset
-from tqdm import tqdm  
+from tqdm import tqdm
 from sklearn.metrics import f1_score
 import pandas as pd
 
@@ -15,7 +15,7 @@ torch.cuda.manual_seed_all(42)
 # Set device
 device = "cuda" if torch.cuda.is_available() else "cpu"
 
-print("The device being used is: {} \n".format(device))
+print("The device being used is: {}\n".format(device))
 
 # Hyperparameters
 input_size = 64
@@ -45,13 +45,12 @@ class biLSTM(nn.Module):
         c0 = torch.zeros(self.num_layers * 2, x.size(0), self.hidden_size).to(device)
 
         out, _ = self.lstm(x, (h0, c0))
-        out = out[:, -1, :] # remove this line to get predictions for the entire input sequence (does not work currently)
+        out = out[:, -1, :]
         out = self.fc1(out)  # Apply linear transformation
         out = self.fc2(out)  # Output layer
         out = F.softmax(out, dim=1)  # Apply softmax
 
         return out
-
 
 
 # define check_accuracy function
@@ -61,7 +60,7 @@ def check_accuracy(loader, model, dataset_name):
     num_correct = 0
     num_samples = 0
     model.eval()
-    
+
     all_predictions = []
     all_targets = []
 
@@ -78,10 +77,10 @@ def check_accuracy(loader, model, dataset_name):
             # Store all predictions and targets for F1 score calculation
             all_predictions.extend(predictions.cpu().numpy())
             all_targets.extend(y.cpu().numpy())
-            
+
         accuracy = float(num_correct) / float(num_samples) * 100
         print(f"Got {num_correct} / {num_samples} with accuracy {accuracy:.2f}")
-        
+
         # Calculate and print F1 score
         f1 = f1_score(all_targets, all_predictions, average='weighted')
         print(f"F1 score on {dataset_name}: {f1:.2f}")
@@ -91,109 +90,114 @@ def check_accuracy(loader, model, dataset_name):
     return accuracy, f1
 
 
-# Load train tensors
-train_predictors = torch.load("/home/esbenlykke/projects/sleep_study/data/data_for_modelling/lstm/pytorch_train_sequences.pt")
[…]

### 2023-07-03 code/evaluate_models/create_performance_metrics_all_models.R

@@ -1,144 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(tidymodels)
-library(arrow)
-
-in_bed_filenames <-
-  list.files("/media/esbenlykke/My Passport/crude/fitted_models/axed_models",
-    full.names = TRUE
-  ) |>
-  str_subset("in_bed")
-
-sleep_filenames <-
-  list.files("/media/esbenlykke/My Passport/crude/fitted_models/axed_models",
-    full.names = TRUE
-  ) |>
-  str_subset("sleep")
-
-in_bed_fits <- map(in_bed_filenames, read_rds) |>
-  set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
-
-sleep_fits <- map(sleep_filenames, read_rds) |>
-  set_names(c("decision_tree", "logistic_regression", "neural_net", "xgboost"))
-
-my_metrics <-
-  metric_set(f_meas, accuracy, sensitivity, precision, specificity)
-
-get_metrics <-
-  function(fit, truth, estimate, estimator = "binary") {
-    fit |>
-      augment(test) |>
-      my_metrics(
-        truth = {{ truth }}, estimate = {{ estimate }},
-        event_level = "second",
-        estimator = estimator
-      )
-  }
-
-# calculate metrics -------------------------------------------------------
-
-### crude
-test <- read_parquet("data/data_for_modelling/crude_testing_data.parquet")
-
-in_bed_metrics <-
-  map_dfr(in_bed_fits, ~ get_metrics(.x, in_bed, .pred_class), .id = "model")
-
-sleep_metrics <-
-  map_dfr(sleep_fits, ~ get_metrics(.x, sleep, .pred_class), .id = "model")
-
-in_bed_metrics |>
-  select(-.estimator) |>
-  pivot_wider(names_from = model, values_from = .estimate) |>
-  mutate(event = "In-bed Prediction") |>
-  bind_rows(
-    sleep_metrics |>
-      select(-.estimator) |>
-      pivot_wider(names_from = model, values_from = .estimate) |>
-      mutate(event = "Sleep Prediction")
-  ) |>
-  write_csv("data/processed/crude_performance_metrics.csv")
-
-### multiclass
-test <-
-  read_parquet("data/data_for_modelling/multiclass_testing_data.parquet") %>%
-  mutate(
[…]

### 2023-07-03 code/evaluate_models/get_biLSTM_performance.R

@@ -1,0 +1,68 @@
+library(tidyverse)
+library(tidymodels)
+library(arrow)
+
+# Define the renaming function
+rename_cols <- function(df, suffix) {
+  df %>%
+    rename_with(.fn = ~ paste0(.x, "_", suffix))
+}
+
+preds_filenames <-
+  list.files("data/processed/predictions/", full.names = TRUE)
+
+preds <-
+  map(preds_filenames, ~ read_csv(.x) %>%
+    janitor::clean_names()) %>%
+  set_names(c("raw", "median_5", "median_10")) %>%
+  imap(~ rename_cols(.x, .y)) %>%
+  reduce(bind_cols)
+
+test <-
+  read_parquet("data/data_for_modelling/lstm/30_sec_testing.parquet")
+
+test_preds <-
+  tibble(
+    predicted_class_raw = rep(1, 20),
+    probability_class_0_raw = rep(0, 20),
+    probability_class_1_raw = rep(0, 20),
+    predicted_class_median_5 = rep(1, 20),
+    probability_class_0_median_5 = rep(0, 20),
+    probability_class_1_median_5 = rep(0, 20),
+    predicted_class_median_10 = rep(1, 20),
+    probability_class_0_median_10 = rep(0, 20),
+    probability_class_1_median_10 = rep(0, 20)
+  ) %>%
+  bind_rows(preds) %>%
+  bind_cols(test) %>%
+  mutate(
+    across(c(
+      score_simple, score_simple_median_5, score_simple_median_10,
+      matches("predicted|probability")
+    ), as_factor)
+  )
+
+
+# write_parquet(test_preds, "data/data_for_modelling/lstm/predictions/lstm_multiclass_preds.parquet")
+
+
+my_metrics <- metric_set(f_meas, sensitivity, precision)
+
+get_metrics <- function(data, truth, estimate) {
+  data %>%
+    my_metrics(truth = {{ truth }}, estimate = {{ estimate }}, 
+               event_level = "second", estimator = "macro")
+}
+
+truths <- c("score_simple", "score_simple_median_5", "score_simple_median_10")
+estimates <- c("predicted_class_raw", "predicted_class_median_5", "predicted_class_median_10")
+
+results <- 
+  map2_dfr(truths, estimates, ~ get_metrics(test_preds, .x, .y), .id = "type") %>% 
+  mutate(type = case_when(type == 1 ~ "raw",
+                          type == 2 ~ "median_5", 
[…]

### 2023-07-03 code/prepare_data/create_features.R

@@ -29,12 +29,13 @@ mean_crossing_rate <- function(signal) {
 # Process the data for 10-second epochs
 data_10 <-
   data %>% 
-  # filter(id == 3404) %>% 
+  # filter(id == 3404) %>%
   # Create new features for sleep, sleep_median5, sleep_median10
   mutate(
     score = if_else(is.na(score), 1, score), # 1 = out-bed awake, otherwise see ZM guide doc
     score_simple = if_else(score %in% c(2, 3, 5, -5), 2, score), # 0 = in-bed awake, 1 = out-bed awake, 2 = in-bed asleep
-    score_simple_filtered = slide_dbl(score_simple, median, .before = 15, .after = 15), # these are 5 minute windows
+    score_simple_median_5 = slide_dbl(score_simple, median, .before = 15, .after = 15), # these are 5 minute windows
+    score_simple_median_10 = slide_dbl(score_simple, median, .before = 30, .after = 30),
     in_bed = if_else(score %in% c(0L, 2L, 3L, 5L, -5L), 1L, 0L),
     in_bed_median5 = slide_dbl(in_bed, median, .before = 15, .after = 15),
     sleep = if_else(score %in% c(2L, 3L, 5L), 1L, 0L),

### 2023-07-03 manuscript/manuscript.qmd
bias, loa, ci95% calculated for ml models. Next is biLSTM.

@@
 To evaluate our developed models' performance in assessing sleep quality
-metrics, we employed Bland-Altman plots and Pearson correlations. These
-methods were used to compare the sleep quality summaries calculated by
-our models with those obtained from EEG-based ZM sleep quality
-summaries. The sleep quality summaries included several metrics:
+metrics, we employed Bland-Altman plots and Pearson correlations.
+Additionally, we used the Bland-Altman method to assess the agreement
+between two measurement techniques. Given the multiple observations per
+subject in our dataset, we incorporated a bootstrap procedure to account
+for this additional source of variability. We computed the mean
+difference (bias) and the limits of agreement as the mean difference
+±1.96 times the standard deviation of the differences. The 95%
+confidence intervals for the bias and the limits of agreement were
+estimated using the bias-corrected and accelerated bootstrap method with
+10,000 bootstrap replicates, to ensure our measurements are robust and
+account for intra-subject variability. These methods were used to
+compare the sleep quality summaries calculated by our models with those
+obtained from EEG-based ZM sleep quality summaries. The sleep quality
+summaries included the following metrics:
 
@@
     summarizes WASO.
+
+Our models were trained to predict sleep quality metrics using the full
+span of accelerometer recordings, beyond the ZM recording duration. This
+approach, applied to all models and ZM prediction types, allows for
+comprehensive sleep pattern analysis on longitudinal accelerometer data.
 
@@
 
+# Discussion
+
+The present study embarked on an in-depth exploration of the utility and
+comparative performance of different machine learning models in the
+analysis of pediatric sleep data. The results offer promising insights
+into the role of machine learning in enhancing the precision of sleep
+quality metrics.
+
+A significant finding was the impact of median filters on the sleep
+metrics. Implementing 5-minute and 10-minute median filters led to
+notable changes in key sleep parameters. In particular, the Total Sleep
+Time (TST), Sleep Efficiency (SE), and Latency to Persistent Sleep (LPS)
+increased, suggesting that the filters may categorize some instances of
+wakefulness as sleep and smooth out brief awakenings. This underscores
+the potential of median filters in refining sleep data and mitigating
+the potential misclassification of sleep states.
+
+Interestingly, the most marked change was observed in Wake After Sleep
+Onset (WASO), with a substantial reduction in the filtered data. This
+reduction, coupled with the decrease in the number of awakenings,
+implies a significant role of median filters in sleep-wake cycle
+characterization, potentially enhancing the reliability of sleep
+assessment.
+
+The machine learning models demonstrated high overall performance on an
+epoch-to-epoch basis, with the XGBoost model marginally outperforming
+the others. However, the models had more varied performance in
+classifying sleep conditions, as evidenced by the sleep performance
+metrics. The relatively lower Specificity values across all models
+suggest a common challenge in correctly classifying negative instances.
+
+In the precision-recall and ROC curve analyses, the Decision Tree model
+showed superior precision-recall AUC, indicating strong predictive
+accuracy. In contrast, the XGBoost model consistently demonstrated a
+high ROC AUC, indicating a robust ability to differentiate between
+classes. However, it's worth noting that the Neural Network model
+generally showed weaker performance in these areas.
+
+The three-class biLSTM multiclassifier showed overall good performance,
+further supporting the potential of machine learning in sleep study.
+However, the absence of Specificity values limits a complete assessment
+of its performance, calling for further exploration in future studies.
+
+In conclusion, this study provides compelling evidence of the utility of
+machine learning models and median filters in enhancing the precision
+and reliability of pediatric sleep assessment. However, more research is
+needed to overcome the challenge of correctly classifying negative
+instances and to fully understand the performance of these models across
+all classes and conditions. The findings underscore the potential of
+machine learning in advancing sleep medicine, paving the way for more
+accurate, reliable, and personalized sleep assessment in pediatric
+populations.
+
 \newpage

### 2023-07-03 manuscript/manuscript.tex
bias, loa, ci95% calculated for ml models. Next is biLSTM.

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced]}{\end{tcolorbox}}\fi
 
@@
 To evaluate our developed models' performance in assessing sleep quality
-metrics, we employed Bland-Altman plots and Pearson correlations. These
-methods were used to compare the sleep quality summaries calculated by
-our models with those obtained from EEG-based ZM sleep quality
-summaries. The sleep quality summaries included several metrics:
+metrics, we employed Bland-Altman plots and Pearson correlations.
+Additionally, we used the Bland-Altman method to assess the agreement
+between two measurement techniques. Given the multiple observations per
+subject in our dataset, we incorporated a bootstrap procedure to account
+for this additional source of variability. We computed the mean
+difference (bias) and the limits of agreement as the mean difference
+±1.96 times the standard deviation of the differences. The 95\%
+confidence intervals for the bias and the limits of agreement were
+estimated using the bias-corrected and accelerated bootstrap method with
+10,000 bootstrap replicates, to ensure our measurements are robust and
+account for intra-subject variability. These methods were used to
+compare the sleep quality summaries calculated by our models with those
+obtained from EEG-based ZM sleep quality summaries. The sleep quality
+summaries included the following metrics:
 
@@
 \end{enumerate}
+
+Our models were trained to predict sleep quality metrics using the full
+span of accelerometer recordings, beyond the ZM recording duration. This
+approach, applied to all models and ZM prediction types, allows for
+comprehensive sleep pattern analysis on longitudinal accelerometer data.
 
@@
 
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The present study embarked on an in-depth exploration of the utility and
+comparative performance of different machine learning models in the
+analysis of pediatric sleep data. The results offer promising insights
+into the role of machine learning in enhancing the precision of sleep
+quality metrics.
+
+A significant finding was the impact of median filters on the sleep
+metrics. Implementing 5-minute and 10-minute median filters led to
+notable changes in key sleep parameters. In particular, the Total Sleep
+Time (TST), Sleep Efficiency (SE), and Latency to Persistent Sleep (LPS)
+increased, suggesting that the filters may categorize some instances of
+wakefulness as sleep and smooth out brief awakenings. This underscores
+the potential of median filters in refining sleep data and mitigating
+the potential misclassification of sleep states.
+
+Interestingly, the most marked change was observed in Wake After Sleep
+Onset (WASO), with a substantial reduction in the filtered data. This
+reduction, coupled with the decrease in the number of awakenings,
+implies a significant role of median filters in sleep-wake cycle
+characterization, potentially enhancing the reliability of sleep
+assessment.
+
+The machine learning models demonstrated high overall performance on an
+epoch-to-epoch basis, with the XGBoost model marginally outperforming
+the others. However, the models had more varied performance in
+classifying sleep conditions, as evidenced by the sleep performance
+metrics. The relatively lower Specificity values across all models
+suggest a common challenge in correctly classifying negative instances.
+
+In the precision-recall and ROC curve analyses, the Decision Tree model
+showed superior precision-recall AUC, indicating strong predictive
+accuracy. In contrast, the XGBoost model consistently demonstrated a
+high ROC AUC, indicating a robust ability to differentiate between
+classes. However, it's worth noting that the Neural Network model
+generally showed weaker performance in these areas.
+
+The three-class biLSTM multiclassifier showed overall good performance,
+further supporting the potential of machine learning in sleep study.
+However, the absence of Specificity values limits a complete assessment
+of its performance, calling for further exploration in future studies.
+
+In conclusion, this study provides compelling evidence of the utility of
+machine learning models and median filters in enhancing the precision
+and reliability of pediatric sleep assessment. However, more research is
+needed to overcome the challenge of correctly classifying negative
+instances and to fully understand the performance of these models across
+all classes and conditions. The findings underscore the potential of
+machine learning in advancing sleep medicine, paving the way for more
+accurate, reliable, and personalized sleep assessment in pediatric
+populations.
+
 \newpage

### 2023-07-03 code/evaluate_models/get_bootstrap_ba_cis.R

@@ -1,0 +1,75 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(boot)
+library(furrr)
+
+# Set up a parallel backend using the future package
+# Adjust the number of workers as necessary
+plan(multisession, workers = 5)
+
+# Define a function to calculate the mean difference and LoA
+get_ba <- function(d, i) {
+  d <- d[i] # resample with replacement
+  mean_diff <- mean(d)
+  sd_diff <- sd(d)
+  loa_upper <- mean_diff + 1.96 * sd_diff
+  loa_lower <- mean_diff - 1.96 * sd_diff
+  return(c(mean_diff, loa_upper, loa_lower))
+}
+
+# Define a function to calculate bootstrap confidence intervals
+get_bootstrap_ci <- function(data, var) {
+  # Run the bootstrap
+  results <- boot(data[[var]], get_ba, R = 10000)
+  
+  # Calculate the 95% confidence intervals
+  bias_ci <- boot.ci(results, type = "bca", index = 1)$bca
+  loa_upper_ci <- boot.ci(results, type = "bca", index = 2)$bca
+  loa_lower_ci <- boot.ci(results, type = "bca", index = 3)$bca
+  
+  # Return a tibble with the results
+  return(tibble(
+    variable = var,
+    bias = results$t0[1],
+    loa_upper = results$t0[2],
+    loa_lower = results$t0[3],
+    bias_ci_lower = bias_ci[4],
+    bias_ci_upper = bias_ci[5],
+    loa_upper_ci_lower = loa_upper_ci[4],
+    loa_upper_ci_upper = loa_upper_ci[5],
+    loa_lower_ci_lower = loa_lower_ci[4],
+    loa_lower_ci_upper = loa_lower_ci[5]
+  ))
+}
+
+# List of variables to bootstrap
+vars_to_bootstrap <- c("diff_spt", "diff_tst", "diff_se", "diff_lps", "diff_waso")
+
+# Read in the data
+stats <-
+  list.files("data/data_for_modelling/chained_classifiers/sleep_stats", full.names = TRUE) %>%
+  read_csv(id = "model") %>%
+  mutate(
+    model = str_extract(model, "(?<=sleep_stats\\/).*(?=_stats.csv)")
+  ) %>%
+  mutate(
+    diff_spt = (spt_hrs - zm_spt) * 60,
+    diff_tst = (tst_hrs - zm_tst) * 60,
+    diff_se = se_percent - zm_se,
[…]

### 2023-07-03 code/evaluate_models/get_stats_ml_models.R

@@ -155,14 +155,15 @@ calculated_zm_stats <-
   
 
 all_sleep_stats_incl_zm <-
-  all_sleep_stats %>% 
+  all_sleep_stats %>%
     map(~ .x %>% mutate(
       sleep_type = case_when(sleep_type == "median_5" ~ "median5",
                              sleep_type == "median_10" ~ "median10",
                              TRUE ~ sleep_type)
     ) %>% 
     left_join(calculated_zm_stats, by = c("id", "noon_day", "month", "sleep_type" = "type")) %>%
-    drop_na())
+    drop_na()) %>%
+  map(~ .x %>% filter(!id %in% c(255704, 649105)))
 
 walk2(
   all_sleep_stats_incl_zm, model_names,

### 2023-07-03 code/evaluate_models/mixed_effect_LOA.R

@@ -2,50 +2,65 @@
 
 library(tidyverse)
 library(arrow)
-library(SimplyAgree)
-library(furrr)
+library(boot)
+
+stats <-
+  list.files("data/data_for_modelling/chained_classifiers/sleep_stats", full.names = TRUE) %>% 
+  read_csv(id = "model") %>% 
+  mutate(
+    model = str_extract(model, "(?<=sleep_stats\\/).*(?=_stats.csv)")
+  ) %>% 
+  mutate(
+    diff_spt = spt_hrs - zm_spt,
+    diff_tst = tst_hrs - zm_tst,
+    diff_se = se_percent - zm_se,
+    diff_lps = lps_min - zm_lps,
+    diff_waso = waso_min - zm_waso
+  )
+
+stats %>% 
+  group_by(model) %>% 
+  summarise(
+    bias_spt = mean(diff_spt) * 60,
+    upper_spt = bias_spt + (1.96 * sd(diff_spt)),
+    lower_spt = bias_spt - (1.96 * sd(diff_spt)),
+    bias_tst = mean(diff_tst) * 60,
+    upper_tst = bias_tst + (1.96 * sd(diff_tst)),
+    lower_tst = bias_tst - (1.96 * sd(diff_tst)),
+    bias_se = mean(diff_se),
+    upper_se = bias_se + (1.96 * sd(diff_se)),
+    lower_se = bias_se - (1.96 * sd(diff_se)),
+    bias_lps = mean(diff_lps),
+    upper_lps = bias_lps + (1.96 * sd(diff_lps)),
+    lower_lps = bias_lps - (1.96 * sd(diff_lps)),
+    bias_waso = mean(diff_waso),
+    upper_waso = bias_waso + (1.96 * sd(diff_waso)),
+    lower_waso = bias_waso - (1.96 * sd(diff_waso)),
+  )
 
 
-crude_stats <-
-  read_parquet("data/processed/crude_stats.parquet")
+# Define a function to calculate the mean difference and LoA
+get_ba <- function(d, i) {
+  d <- d[i]  # resample with replacement
+  mean_diff <- mean(d)
+  sd_diff <- sd(d)
+  loa_upper <- mean_diff + 1.96 * sd_diff
+  loa_lower <- mean_diff - 1.96 * sd_diff
+  return(c(mean_diff, loa_upper, loa_lower))
+}
 
-multi_stats <- 
-  read_parquet("data/processed/multiclass_stats.parquet")
+# Run the bootstrap
[…]

### 2023-07-03 code/evaluate_models/plot_ba_preds_vs_zm.R

@@ -5,48 +5,30 @@ library(showtext)
 
 # TODO Fix theming for journal publication
 
-ba_metrics <-
-  read_csv(here::here("data/processed/mixed_effect_ba.csv"))
+stats <-
+  list.files("data/data_for_modelling/chained_classifiers/sleep_stats", full.names = TRUE) %>% 
+  read_csv(id = "model") %>% 
+  mutate(
+    model = str_extract(model, "(?<=sleep_stats\\/).*(?=_stats.csv)")
+  )
 
-stats_files <-
-  list.files("~/sleep_study/data/processed/stats_predictions", full.names = TRUE)
-
-all_stats <-
-  map(stats_files, read_parquet) |>
-  set_names(c("logistic_regression", "neural_net", "decision_tree", "xgboost"))
-# |>
-#   map(
-#     ~ filter(.x, (id != 1742705 | noon_day != 24) &
-#         (id != 1044005 | noon_day != 16) &
-#         (id != 1750904 | noon_day != 21) &
-#         (id != 1718904 | noon_day != 8)
-#     )
-#   ) # TODO look into these lps_min outliers
-
-get_diff_stats <-
-  function(tbl) {
-    tbl |>
-      mutate(
-        diff_spt_hrs = spt_hrs - zm_spt_hrs,
-        diff_tst_hrs = tst_hrs - zm_tst_hrs,
-        diff_se_percent = se_percent - zm_se_percent,
-        diff_lps_min = lps_min - zm_lps_min,
-        diff_waso_min = waso_min - zm_waso_min
-      ) |>
-      rowwise() |>
-      mutate(
-        avg_spt_hrs = mean(c(spt_hrs, zm_spt_hrs)),
-        avg_tst_hrs = mean(c(tst_hrs, zm_tst_hrs)),
-        avg_se_percent = mean(c(se_percent, zm_se_percent)),
-        avg_lps_min = mean(c(lps_min, zm_lps_min)),
-        avg_waso_min = mean(c(waso_min, zm_waso_min))
-      ) |>
-      ungroup()
-  }
-
-diff_avg <-
-  all_stats |>
-  map(get_diff_stats) |>
+# diff_avg <-
+  stats %>% 
+    mutate(
+    diff_spt_hrs = spt_hrs - zm_spt,
+    diff_tst_hrs = tst_hrs - zm_tst,
+    diff_se_percent = se_percent - zm_se,
[…]

### 2023-07-07 manuscript/manuscript.qmd
results almost done. XXgboost plots are ok.

@@
   - Machine learning
-  - Sleep quality metrics
+  - Sleep quality
 date: last-modified
@@
     wrap: 72
+editor_options: 
+  chunk_output_type: inline
 ---
@@
 secondary goal was to assess the developed models' performance in
-evaluating important sleep quality metrics, including sleep period time
-(SPT), total sleep time (TST), sleep efficiency (SE), latency until
-persistent sleep (LPS), and wake after sleep onset (WASO). ~~By
-analyzing these additional metrics, we aimed to provide a comprehensive
-evaluation of the model's capability in assessing various aspects of
-sleep quality, allowing us to determine the overall effectiveness of our
-models in accurately estimating different parameters related to sleep
-duration and quality.~~
+evaluating important sleep quality parameters, including sleep period
+time (SPT), total sleep time (TST), sleep efficiency (SE), latency until
+persistent sleep (LPS), and wake after sleep onset (WASO).
 
@@
 2019 in Middelfart, Southern Denmark, that evaluated the impact of
-screen media usage on Danish families. For our analysis, we isolated
-data from child participants between the ages of 6 to 10 years within
-the SCREENS cohort. Our main sources of data were accelerometer readings
+screen media usage on Danish families. For our analysis, we focused on
+data from child participants aged between 6 and 10 years within the
+SCREENS cohort. Our primary sources of data were accelerometer readings
 from Axivity AX3 devices attached to the children's thighs, and
@@
 Sleep state information was extracted using the ZM, a product of General
-Sleep Corporation. The ZM, utilizes advanced EEG hardware and signal
-processing algorithms, employs three self-adhesive, disposable sensors
-placed outside the hairline for reliable EEG signal acquisition. The ZM
-uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is utilized
-for accurate sleep detection, showcasing its aptness for in-home
-monitoring[@kaplan2014], while the Z-PLUS effectively differentiates
-sleep stages, as evidenced by its alignment with expert evaluations
-using PSG data[@wang2015].
+Sleep Corporation. The ZM, which utilizes advanced EEG hardware and
+signal processing algorithms, employs three self-adhesive, disposable
+sensors placed outside the hairline for reliable EEG signal acquisition.
+The ZM uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is
+utilized for accurate sleep detection, showcasing its suitability for
+in-home monitoring[@kaplan2014], while the Z-PLUS effectively
+differentiates sleep stages, as evidenced by its alignment with expert
+evaluations using PSG data[@wang2015]. In the current study, we treated
+all sleep stages as a single category effectively deducing the output of
+the ZM to "awake" and "asleep" as the ability to distinguish sleep
+stages are not a necessity to derive sleep quality parameters of
+interest and to simplify the learning process of the models.
+
+@fig-flow illustrates the selection criteria applied to the children's
+recordings from the SCREENS study. We included only ZM recordings that
+were accompanied by complete accelerometer data and lasted between 7 and
+14 hours. Any night when the ZM reported sensor issues was excluded. The
+children whose recordings were considered had an average age of 9.4
+years, with a standard deviation of 2.1. In their raw form, the ZM
+predictions encompassed 696,779 epochs, each 30 seconds long. Notably,
+approximately 84% of the total ZM recording duration was classified as
+sleep, resulting in an imbalance of the dataset.
+
+![Flowchart of eligible ZM recording nights included in the
+study](visuals/flowchart_of_elligible_nights.pdf){#fig-flow}
 
@@
 processes complied with the General Data Protection Regulation (GDPR),
-guaranteeing the ethical and secure management of participant
-information.
+ensuring the ethical and secure management of participant information.
 
@@
 
-In this study, raw data processing began with a low-pass filtration step
-using a 4th order Butterworth filter with a 5 Hz cut-off frequency to
-eliminate high-frequency noise. Following filtration, data were
-partitioned into overlapping 2-second intervals, each successive
-interval sharing a 50% overlap with the previous one similar to methods
-described by Skotte et al.[@skotte_detection_2014]. Any non-wear data
-was remove using previously described methods[@skovgaard2023] and data
-was resampled to 30-second epochs so every sample classified by the
-algorithms corresponds to a 30-second epoch scored during the ZM
-recordings. Subsequently, we performed a feature extraction process that
-yielded a set of 88 features, providing a robust characterization of the
-data. Extracted from accelerometer and temperature signals, these
-features include temporal elements that use both lag and lead values,
-capturing dynamic data trends by incorporating measurements from
-preceding and upcoming intervals. Furthermore, inspired by Walch et
-al.[@walch2019], we developed sensor-independent features to encapsulate
-circadian rhythms. These features offer unique insights not directly
-discernible from sensor outputs and are meant to approximate the
-changing drive of the circadian clock to sleep over the course of the
-night (see @fig-sensor-independent). Furthermore, the feature set was
-enriched by including signal characteristics, which encompass vector
-magnitude, mean crossing rate, skewness, and kurtosis for each of the x,
-y, and z dimensions. These features provide a comprehensive
-understanding of the signal's attributes, facilitating a more nuanced
-analysis of the underlying patterns. All features are summarized in
-table ??? (overvej tabel til supp. material)
+In this study, data processing of the raw accelerometer data began with
+a low-pass filtration step using a 4th order Butterworth filter with a 5
+Hz cut-off frequency to eliminate high-frequency noise. Following
+filtration, data were partitioned into overlapping 2-second intervals,
+each successive interval sharing a 50% overlap with the previous one
+similar to methods described by Skotte et al.[@skotte_detection_2014].
+Any non-wear data was remove using previously described
+methods[@skovgaard2023] and data was resampled to 30-second epochs so
+every sample classified by the algorithms corresponds to a 30-second
+epoch scored during the ZM recordings. Subsequently, we performed a
+feature extraction process that yielded a set of 88 features, providing
+a robust characterization of the data. Extracted from accelerometer and
+temperature signals, these features include temporal elements that use
+both lag and lead values, capturing dynamic data trends by incorporating
+measurements from preceding and upcoming intervals. Furthermore,
+inspired by Walch et al.[@walch2019], we incorporated sensor-independent
+features to encapsulate circadian rhythms. These features offer unique
+insights not directly discernible from sensor outputs and are meant to
+approximate the changing drive of the circadian clock to sleep over the
+course of the night (see @fig-sensor-independent). Furthermore, the
+feature set was enriched by including signal characteristics, which
+encompass vector magnitude, mean crossing rate, skewness, and kurtosis
+for each of the x, y, and z dimensions. All features are summarized in
+table ??? **in the supplementary matrials**. Subsequently, we merged the
+ZM and corresponding accelerometer recordings. Any overlapping time
+between the ZM and accelerometer data was treated as 'in-bed' time, with
+the remaining time considered 'out-of-bed'. This process yielded a
+dataset providing a around the clock temporal view of each participant's
+activity and sleep patterns.
 
@@
 typically undergo around five to eight sleep cycles per night, with
-awakenings commonly occurring at the end of each
-cycle[@galland_normal_2012]. Examining the raw ZM predictions, we noted
-a significant overestimation in the number of awakenings per night for
-the children in our study, exceeding what would be expected based on
-typical sleep cycle patterns. Consequently, we elected to train and
-evaluate our models using not only the raw ZM output, but also versions
-that were subjected to 5-minute and 10-minute median filters. This
-approach resulted in an anticipated, more age-appropriate count of
-awakenings per night, providing a more accurate depiction of children's
-sleep patterns (see @fig-zm-median).
+awakenings most likely occurring at the end of each
+cycle[@galland_normal_2012]. Upon examining the raw ZM predictions, we
+noted a significant overestimation in the number of awakenings per night
+for the children in our study, exceeding what would be expected based on
+typical sleep cycle patterns. In particular, many of these brief
+awakenings could be considered as noise, which when present in the data,
+can potentially hinder the learning process of machine learning models
+by obscuring the underlying patterns that the models are trying to
+learn, leading to less accurate predictions. Consequently, we elected to
+train and evaluate our models using not only the raw ZM output, but also
+versions that were subjected to 5-minute and 10-minute median filters.
+This approach, by mitigating this noise, resulted in an anticipated,
+more age-appropriate count of awakenings per night, providing a more
+accurate depiction of children's sleep patterns (see @fig-zm-median).
 
@@
 
-In our study, we utilize two distinct modeling strategies to predict
-sleep patterns derived from thigh-mounted accelerometer data: a
-sequential ensemble and a multiclass bidirectional Long Short-Term
-Memory (biLSTM)[@hochreiter1997] neural network. The sequential ensemble
-involves using four pairs of models, each pair consisting of the same
-algorithm, that predict in-bed and sleep time successively. Conversely,
-the biLSTM is a comprehensive model that predicts multiple classes
-simultaneously, effectively handling complex temporal patterns. By
-applying and comparing both, we aim to evaluate their relative
-effectiveness.
+In this study, we employed two distinct modeling strategies to analyze
+sleep patterns from thigh-mounted accelerometer data. We used a
+sequential strategy, comprising an ensemble of four pairs of models,
+each pair featuring the same algorithm. This strategy aimed to make the
+prediction task more manageable for the algorithms by breaking it down
+into a sequence of two binary classifications: first predicting 'in-bed'
+time, then 'sleep' time. Simultaneously, we also used a multiclass
+approach utilizing a bidirectional Long Short-Term Memory
+(biLSTM)[@hochreiter1997] neural network.
 
@@
 
-The biLSTM is a multiclass classifier capable of predicting the classes
-of out-bed-awake, in-bed-awake, and in-bed-asleep. The architecture of
-the biLSTM consisted of four layers, each with 128 hidden units. The
-choice of four layers and 128 hidden units in our model balanced
-complexity and efficiency: it was deep enough to learn intricate
-patterns yet feasible to train timely. Additionally, the bidirectional
-nature of the LSTM, doubling the hidden units at each time step,
-improved data comprehension and avoided overfitting.
-
-something on sequence length and step size
-
-The biLSTM model was chosen for sleep classification using accelerometer
-data from a thigh-mounted device due to its advantages in modeling
-temporal context, handling a large temporal scope, and enabling temporal
-inference over any feature without the need for hand-designed temporal
-features.
-
-Previous studies have explored the use of LSTM models for sleep
-detection, showing promising results in capturing complex temporal
-patterns. Notable works by Sano et al. [@sano2019] and Chen et al.
-[@chen2021] have demonstrated the potential of LSTM models in improving
-sleep detection based on accelerometer data. However, further research
-is necessary to fully comprehend the strengths and limitations of this
-approach.
+In this study, we employed a biLSTM, a multiclass classifier, to predict
+three distinct states: out-of-bed-awake, in-bed-awake, and
+in-bed-asleep. The architecture of the biLSTM was set up with four
+layers, each equipped with 128 hidden units. This configuration was
+intentionally chosen to balance between model complexity and training
+efficiency: it provided the depth necessary for learning intricate
+patterns while remaining feasible for timely training. The bidirectional
+design of the LSTM served to enhance data interpretation and mitigate
+overfitting by doubling the hidden units at each time step. For input,
+we used tensors shaped as sequences, with each sequence spanning 10
+minutes and a step size of one.This approach follows in the footsteps of
+previous studies that utilized LSTM models for sleep detection. These
+studies showcased the promising potential of LSTMs in capturing complex
+temporal patterns. Particularly, the works of Sano et al. [@sano2019]
+and Chen et al. [@chen2021] demonstrated the effectiveness of LSTM
+models in improving sleep detection using accelerometer data,
+underscoring the value of this modeling approach.
 
@@
 For the models in sequence, we trained four pairs of classification
-models, each pair distinguishing between in-bed/out-of-bed and
-asleep/awake states. The dataset was randomly split into a training set
-(approximately 50% of the subjects) and a testing set (also
-approximately 50%), ensuring that samples from the same subject were
-never simultaneously used in both sets. Hyper-parameter optimization was
-performed using 10-fold Monte Carlo cross-validation, with the F1 score
-as the performance metric. The best-performing set of hyperparameters
-was then used to fit the models to the full training dataset, maximizing
-accuracy by leveraging all available data. Furthermore, the dataset for
-predicting sleep was highly imbalanced which can pose challenges when
-training machine learning models, as models may favor predicting the
-majority class. To address this issue, we employed oversampling
-techniques to ensure that each class had a roughly equal number of
-training samples. Specifically, we utilized the Synthetic Minority
-Over-sampling Technique (SMOTE)[@chawla2002], which generates new
-samples by interpolating random samples with their nearest neighbors. In
-our study, we implemented SMOTE using the themis R package[@themis],
-resampling all classes to achieve a balanced distribution of training
-samples.
+models. Each pair was designed to distinguish between in-bed/out-of-bed
+and asleep/awake states, respectively. The dataset was randomly split
+into a training set and a testing set, each containing approximately 50%
+of the subjects. This division ensured that samples from the same night
+were never simultaneously present in both sets. To optimize
+hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
+a regular grid, comprising 20 different combinations of hyperparameters.
+The F1 score served as the optimization metric. The best-performing set
+of hyperparameters was then used to fit the models to the full training
+dataset. This approach allowed us to maximize performance by leveraging
+all available data. Moreover, after extracting the in-bed time from the
+initial sequential models, the imbalance on the resulting dataset could
+cause biases during model training, as models may favor predicting the
+majority class. To rectify this, we employed the Synthetic Minority
+Over-sampling Technique (SMOTE)[@chawla2002]. SMOTE generates new
+samples by interpolating random samples with their nearest neighbors. We
+utilized the themis R package[@themis] to implement SMOTE, resulting in
+a balanced distribution of training samples across both classes.
 
-The biLSTM model was trained using the Adam optimizer, which is
-computationally efficient and adapts the learning rate during training.
-Cross-entropy loss function was employed for its suitability in
-multiclass classification with mutually exclusive classes. The softmax
-activation function was selected for the output layer to obtain a
-probability distribution over the classes. Data for training the biLSTM
-were randomly divided into a training, validation, and test tensors
-based on a 50/25/25 split. The model was evaluated using the F1 score on
-both the training and validation sets. Early stopping was implemented
-with a patience of 3 epochs, meaning training would stop if there was no
-improvement in the validation loss for 3 consecutive epochs.
+The biLSTM model was trained to differentiate between three states:
+out-of-bed-awake, in-bed-awake, and in-bed-asleep. The data used for
+training the biLSTM was randomly divided into training, validation, and
+test sets, based on a 50/25/25 split. We ensured that data from the same
+night was not present across different sets. The model was trained using
+the Adam optimizer, selected for its computational efficiency and
+adaptability of the learning rate during training. Given the multiclass
+classification task with mutually exclusive classes, we employed the
+cross-entropy loss function. To obtain a probability distribution over
+the classes, the softmax activation function was applied at the output
+layer. We evaluated the model's performance using the F1 score on both
+the training and validation sets. We implemented early stopping with a
+patience of 3 epochs, halting the training process if there was no
+improvement in the validation loss over three consecutive epochs.
 
-### Model Validation \[evt. bare "statistics"\]
+### Model Validation
 
@@
 Additionally, we evaluated the multiclass classifier, biLSTM, using
-macro-averaged F1-score, sensitivity, and precision. Furthermore, we
-present precision-recall curves and receiver operating characteristic
-curves, including the area under the curve, for each model and for each
-type of ZM prediction.We considered both the in-bed/out-of-bed and
-awake/asleep scoring tasks as binary classification problems,
-designating in-bed and asleep as the positive labels and out-of-bed and
-awake as the negative labels in accordance with previous
-research[@hjorth2012; @kushida2001].
+macro-averaged F1-score, sensitivity, and precision. To further
+illustrate model performance, we provide confusion matrices for the full
+dataset, encompassing both in-bed and out-of-bed data. These matrices
+report relative counts, column percentages (the proportion of the true
+class accurately predicted), and row percentages (the proportion of
+predictions correctly classified). We considered both the
+in-bed/out-of-bed and awake/asleep scoring tasks as binary
+classification problems, designating in-bed and asleep as the positive
+labels and out-of-bed and awake as the negative labels in accordance
+with previous research[@hjorth2012; @kushida2001].
 
-To evaluate our developed models' performance in assessing sleep quality
-metrics, we employed Bland-Altman plots and Pearson correlations.
-Additionally, we used the Bland-Altman method to assess the agreement
-between two measurement techniques. Given the multiple observations per
-subject in our dataset, we incorporated a bootstrap procedure to account
-for this additional source of variability. We computed the mean
-difference (bias) and the limits of agreement as the mean difference
-±1.96 times the standard deviation of the differences. The 95%
-confidence intervals for the bias and the limits of agreement were
-estimated using the bias-corrected and accelerated bootstrap method with
-10,000 bootstrap replicates, to ensure our measurements are robust and
-account for intra-subject variability. These methods were used to
-compare the sleep quality summaries calculated by our models with those
-obtained from EEG-based ZM sleep quality summaries. The sleep quality
-summaries included the following metrics:
+To assess the performance of our models in deriving sleep quality
+parameters, we utilized Bland-Altman plots and Pearson correlations. The
+Bland-Altman method was employed specifically to determine the level of
+agreement between two measurement techniques. Considering our dataset
+contained multiple observations per subject, we integrated a bootstrap
+procedure to address this extra source of variability. We calculated the
+mean difference (bias) and defined the limits of agreement as the mean
+difference plus or minus 1.96 times the standard deviation of these
+differences. To ensure our measurements were robust and accounted for
+intra-subject variability, we estimated the 95% confidence intervals for
+both the bias and the limits of agreement using a bias-corrected and
+accelerated bootstrap method, utilizing 10,000 bootstrap replicates. The
+sleep quality parameteres included the following:
 
@@
 
-Our models were trained to predict sleep quality metrics using the full
-span of accelerometer recordings, beyond the ZM recording duration. This
-approach, applied to all models and ZM prediction types, allows for
-comprehensive sleep pattern analysis on longitudinal accelerometer data.
-
 R version 4.3.0 (2023-04-21)[@R-lang] and the Tidymodels[@tidymodels]
@@
 
-![Flowchart of eligible nights included in the study - revise text
-boxes!](visuals/flowchart_of_elligible_nights.pdf)
-
-The analysis included children with an average age of 9.4 years (SD =
-2.1). The raw ZM predictions covered 2,035,261 sleep epochs, or about
-86% of the total recording duration.
-
-As reported in @tbl-zm_overview the sleep quality metrics derived from
-ZM predictions were modified by the implementation of 5-minute and
-10-minute median filters. The Sleep Period Time (SPT) were consistent
-across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as this
-corresponds to the length of the ZM recording. Total Sleep Time (TST)
-and Sleep Efficiency (SE) increased in the filtered data, implying the
+As reported in @tbl-zm_overview the sleep quality parameters derived
+from ZM predictions were modified by the implementation of 5-minute and
+10-minute median filters. SPT were consistent across raw and filtered
+datasets (mean: 9.2 ± 2.1 hours), corresponding to the length of the ZM
+recording. TST and SE increased in the filtered data, implying the
 filters categorize some wakefulness as sleep. Specifically, TST
@@
 (5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
-from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. Latency
-to Persistent Sleep (LPS) also elevated, suggesting the filter smooths
-out brief awakenings at sleep onset, leading to a prolonged time to
-persistent sleep. The most significant change was seen in Wake After
-Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
-to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
-10-minute filtered data, respectively. The number of awakenings was also
-considerably reduced with the application of filters. In the raw data,
-the average number of awakenings was 34.46 ± 11.33 per night, which
-reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the 5-minute and 10-minute
-filtered data sets respectively. These results underscore the role of
-median filters on sleep metrics.
+from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. LPS
+also elevated, suggesting the filter smooths out brief awakenings at
+sleep onset, leading to a prolonged time to persistent sleep. A
+significant change was seen in WASO, which dropped from 39.0 ± 33.6
+minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in
+the 5-minute and 10-minute filtered data, respectively. The number of
+awakenings was also considerably reduced with the application of
+filters. In the raw data, the average number of awakenings was 34.46 ±
+11.33 per night, which reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the
+5-minute and 10-minute filtered data sets respectively.
 
@@
 
-## Performance on Epoch-To-Epoch Basis
+## Performance on Epoch-to-Epoch Basis
 
-The epoch-to-epoch evaluation of in-bed performance metrics, outlined in
-@tbl-in_bed_performance, demonstrates practically equivalent performance
-across all model types. While the Decision Tree model posted an F1 score
-of 94.4% and accuracy of 95.3%, the Logistic Regression and Feed-Forward
-Neural Network models each exhibited an F1 score of 95.0% and similar
-accuracies of 95.7% and 95.8%, respectively. The XGBoost model, despite
-recording the highest metrics with an F1 score of 95.4% and accuracy of
-96.1%, outpaced the others only marginally. This underscores the
-consistency of performance among these models in classifying in-bed
-conditions.
+The epoch-to-epoch evaluation of predicting in-bed time is outlined in
+@tbl-in_bed_performance, and demonstrates practically equivalent
+performance across all model types. The F1 score ranges from 94.4%
+(Decision Tree) to 95.4% (XGBoost), while accuracy ranges from 95.3%
+(Decision Tree) to 96.1% (XGBoost). Sensitivity, Precision, and
+Specificity also demonstrate consistent results across the models. The
+XGBoost model, despite recording the highest metrics with an F1 score of
+95.4% and accuracy of 96.1%, outpaced the others only marginally.
 
@@
 
-@tbl-sleep_performance details the performance of the sequential models
-on raw and median-filtered (5 and 10 minute) ZM predictions. The F1
-Scores, which are unweighted macro averages, for raw ZM predictions
-range from 71.05% to 76.18%. The models perform comparably, but the low
-Specificity values (62.84% to 70.93%) suggest difficulty in correctly
-classifying negative instances. Applying 5-minute median filtering
-improves the performance metrics. The XGBoost model tops the charts with
-an F1 Score of 79.22% and NPV of 74.00%. However, Specificity still
-remains low, with values between 54.68% and 74.84% across all models.
-With 10-minute median filtering, the metrics improve further. The
-XGBoost model still leads with an F1 Score of 80.87% and an NPV of
-75.76%. But, Specificity remains a concern, ranging from 57.47% to
-76.35% across all models.
+@tbl-sleep_performance details the performance of the second sequential
+models on raw and median-filtered (5 and 10 minute) ZM predictions.
+These are the predictions of the sequential models' ability to predict
+asleep/awake on only the extracted in-bed time. The F1 scores, which are
+unweighted macro averages, for raw ZM predictions range from 71.05% to
+76.18%. The models perform comparably, but the low specificity values
+(62.84% to 70.93%) suggest difficulty in correctly classifying awake
+epochs. Applying 5-minute median filtering improves the performance
+metrics. The XGBoost model tops the charts with an F1 score of 79.22%
+and NPV of 74.00%. However, specificity still remains low, with values
+between 54.68% and 74.84% across all models. With 10-minute median
+filtering, the metrics improve further. The XGBoost model still leads
+with an F1 score of 80.87% and an NPV of 75.76%. But, specificity
+remains low, ranging from 57.47% to 76.35% across all models.
 
@@
 
-The analysis of precision-recall and ROC curves across different models
-and ZM prediction types shows varying performance. In terms of
-precision-recall AUC, the Decision Tree model consistently outperforms
-others, indicating its superior predictive accuracy (see
-@fig-pr_curves). Conversely, the Neural Network model generally shows
-weaker performance. However, for ROC AUC, the XGBoost model consistently
-excels across all data types, indicating a strong ability to
-differentiate between classes, while the Neural Network model tends to
-underperform (see @fig-roc_curves). The F-measure (F1 score) shows
-variable performance across different configurations but generally, the
-Decision Tree model yields higher scores.
+The complete set of confusion matrices generated from data both
+containing the out-of-bed and in-bed time are presented in
+@fig-conf_mat. The figure shows favorable epoch-to-epoch performance
+across across all sequential models, however, it is evident that the
+biLSTM is less successful in classifying the in-bed-awake class which
+cannot be deduced from the confusion matrices from the sequential
+models. **\[fejl i biLSTM confmat. Alle mats er ens\]**
 
-![Precision-recall curves of the models evaluated across the different
-ZM predictions, including raw ZM predictions, as well as 5-minute and
-10-minute median smoothing of the ZM raw predictions. The x-axis of the
-plot represents the proportion of true wake epochs that were correctly
-classified as wake, while the y-axis represents the proportion of all
-epochs labeled as wake by the classifier that were classified correctly.
-The area under the curve values are displayed as color-coded text in the
-plot indicate the area under the Precision-Recall curve for each model
-and condition.](visuals/plot_sleep_pr.pdf){#fig-pr_curves}
-
-![Receiver operating characteristic curves of the models evaluated
-across the different ZM predictions, including raw ZM predictions, as
-well as 5-minute and 10-minute median smoothing of the ZM raw
-predictions. The x-axis of the plot represents the proportion of true
-asleep epochs that were incorrectly classified as awake, while the
-y-axis represents the proportion of all epochs labeled as awake by the
-classifier that were correctly classified. The area under the curve
-values displayed are displayed as color-coded text in the plot to
-indicate the area under the receiver operating characteristic curve for
-each model and condition.](visuals/plot_sleep_roc.pdf){#fig-roc_curves}
+![Confusion matrices for binary sleep prediction. The middle of each
+tile is the normalized count (overall percentage) and, beneath it, the
+count. The bottom number is the column percentage (target). At the right
+side of each tile is the row percentage (prediction). i) decision tree,
+ii) logistic regression, iii) feed-forward neural net, iv) xgboost, and
+v) biLSTM.](visuals/all_conf_mats.pdf){#fig-conf_mat}
 
@@
 Scores ranging from 75.99% to 78.53%, demonstrating enhanced precision
-and sensitivity. However, Specificity values are not available for this
-configuration. With 10-minute median filtering, F1 Scores range from
-73.45% to 73.93%, maintaining good performance but with limited
-information on Specificity. Further analysis is required to assess
-performance across all classes.
+and sensitivity. **\[Overvej flere metrics. Evt. per class... pt er det
+lidt svært at sammenligne på tværs af lstm og andre modeller ud over
+F1.\]** With 10-minute median filtering, F1 Scores range from 73.45% to
+73.93%, maintaining good performance but with limited information on
+Specificity. Further analysis is required to assess performance across
+all classes.
 
@@
 
-## Evaluation of Sleep Quality Summaries
+## Evaluation of Sleep Quality Parameters
 
-BA plots
+@tbl-ba_cor presents a comparative analysis of the included models used
+to predict various sleep quality parameters (SPT, TST, SE, LPS, WASO)
+using the 5-minute median filtered ZM predictions. To see the full table
+including models developed from raw ZM predictions and 10-minute median
+filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
+decision tree model consistently underestimated SPT, TST, and SE, and
+overestimated LPS and WASO in comparison to ZM. The logistic regression
+model had similar trends, with more pronounced underestimation in TST
+and overestimation in LPS. The eed-forward neural network also exhibited
+similar bias as the decision tree and the logistic regression models,
+but with a higher overestimation in WASO. On the other hand, the XGboost
+model showed least bias among all, especially in its 5-minute median
+predictions. Considering LOA, the decision tree had higher variability
+across different sleep quality parameters and filtering techniques,
+particularly for LPS and WASO, which indicates lower agreement with ZM.
+Other models had comparable LOA but with notable exceptions. For
+example, TST LOA for the logistic regression model was particularly wide
+in the 5-minute median predictions. Correlation-wise, the pearson
+coefficient, revealed that the XGboost model consistently had the
+highest correlation with ZM across all sleep qualityparameters and
+filtering methods Notably, the XGboost's 5-minute median predictions
+showed the strongest correlation (0.66) for TST among all models and
+filtering techniques.
 
-pearson regression
+```{r}
+#| echo: false
+#| message: false
+#| label: tbl-ba_cor
+#| tbl-cap: Summary of Bias, Limits of Agreement (LOA), and Pearson Correlation for various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using different Machine Learning Models (Decision Tree, Logistic Regression, Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and 10-Min Median as predictors.
+Each value is provided with its 95% Confidence Interval (CI).
 
-scatterplots with identity line and best linear fit
+source("code/table_ba_cor_statistics.R")
+
+short_tbl_ba_cor
+
+```
+
+@fig-xgb_ba_cor shows the agreement of the XGBoost model trained on the
+5-minute median filtered ZM predictions in estimating sleep quality
+parameters as evaluated 5-minute median-smoothed ZM derived sleep
+quality parameter. As can be seen on from the Bland-Altman plots, a
+large portion of the nights are in almost perfect agreement with the ZM
+and also a portion that can be considered outliers. This is true for all
+sleep quality parameters.
+
+**\[Overvej at smide samtlige 75 plots i supp.mat.\]**
+
+![Comparison of sleep quality parameters derived from the XGboost model
+trained on the 5-minute smoothed ZM predictions. The left column
+displays Bland-Altman plots. Dashed lines represent the bias (the
+average difference between the two measurements) and limits of
+agreement, with the 95% confidence intervals represented as the grayed
+areas. The right column displays scatter plots of XGboost-derived vs
+ZM-derived sleep quality parameters. The dashed line represents the
+identity line, while the full-drawn line represents the best linear fit.
+Pearson's correlations are annotated in the upper left
+corner.](visuals/median_5_xgboost_ba_cor.pdf){#fig-xgb_ba_cor}
 
@@
 
+It is unclear whether Johansson et al.[@johansson_development_2023]\]
+provide a method for detecting the in-bed time...
+
 In conclusion, this study provides compelling evidence of the utility of

### 2023-07-07 manuscript/manuscript.tex
results almost done. XXgboost plots are ok.

@@
 \usepackage{xcolor}
-\usepackage{soul}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   pdfkeywords={Sleep, Accelerometry, EEG, Machine learning, Sleep
-quality metrics},
+quality},
   colorlinks=true,
@@
     Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
-    Sleep quality metrics
+    Sleep quality
 \end{keyword}
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 
@@
 secondary goal was to assess the developed models' performance in
-evaluating important sleep quality metrics, including sleep period time
-(SPT), total sleep time (TST), sleep efficiency (SE), latency until
-persistent sleep (LPS), and wake after sleep onset (WASO). \st{By
-analyzing these additional metrics, we aimed to provide a comprehensive
-evaluation of the model's capability in assessing various aspects of
-sleep quality, allowing us to determine the overall effectiveness of our
-models in accurately estimating different parameters related to sleep
-duration and quality.}
+evaluating important sleep quality parameters, including sleep period
+time (SPT), total sleep time (TST), sleep efficiency (SE), latency until
+persistent sleep (LPS), and wake after sleep onset (WASO).
 
@@
 March 2019 in Middelfart, Southern Denmark, that evaluated the impact of
-screen media usage on Danish families. For our analysis, we isolated
-data from child participants between the ages of 6 to 10 years within
-the SCREENS cohort. Our main sources of data were accelerometer readings
+screen media usage on Danish families. For our analysis, we focused on
+data from child participants aged between 6 and 10 years within the
+SCREENS cohort. Our primary sources of data were accelerometer readings
 from Axivity AX3 devices attached to the children's thighs, and
@@
 Sleep state information was extracted using the ZM, a product of General
-Sleep Corporation. The ZM, utilizes advanced EEG hardware and signal
-processing algorithms, employs three self-adhesive, disposable sensors
-placed outside the hairline for reliable EEG signal acquisition. The ZM
-uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is utilized
-for accurate sleep detection, showcasing its aptness for in-home
-monitoring\citep{kaplan2014}, while the Z-PLUS effectively
+Sleep Corporation. The ZM, which utilizes advanced EEG hardware and
+signal processing algorithms, employs three self-adhesive, disposable
+sensors placed outside the hairline for reliable EEG signal acquisition.
+The ZM uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is
+utilized for accurate sleep detection, showcasing its suitability for
+in-home monitoring\citep{kaplan2014}, while the Z-PLUS effectively
 differentiates sleep stages, as evidenced by its alignment with expert
-evaluations using PSG data\citep{wang2015}.
+evaluations using PSG data\citep{wang2015}. In the current study, we
+treated all sleep stages as a single category effectively deducing the
+output of the ZM to ``awake'' and ``asleep'' as the ability to
+distinguish sleep stages are not a necessity to derive sleep quality
+parameters of interest and to simplify the learning process of the
+models.
+
+Figure~\ref{fig-flow} illustrates the selection criteria applied to the
+children's recordings from the SCREENS study. We included only ZM
+recordings that were accompanied by complete accelerometer data and
+lasted between 7 and 14 hours. Any night when the ZM reported sensor
+issues was excluded. The children whose recordings were considered had
+an average age of 9.4 years, with a standard deviation of 2.1. In their
+raw form, the ZM predictions encompassed 696,779 epochs, each 30 seconds
+long. Notably, approximately 84\% of the total ZM recording duration was
+classified as sleep, resulting in an imbalance of the dataset.
+
+\begin{figure}[b]
+
+{\centering \includegraphics{visuals/flowchart_of_elligible_nights.pdf}
+
+}
+
+\caption{\label{fig-flow}Flowchart of eligible ZM recording nights
+included in the study}
+
+\end{figure}
 
@@
 
-In this study, raw data processing began with a low-pass filtration step
-using a 4th order Butterworth filter with a 5 Hz cut-off frequency to
-eliminate high-frequency noise. Following filtration, data were
-partitioned into overlapping 2-second intervals, each successive
-interval sharing a 50\% overlap with the previous one similar to methods
-described by Skotte et al.\citep{skotte_detection_2014}. Any non-wear
-data was remove using previously described methods\citep{skovgaard2023}
-and data was resampled to 30-second epochs so every sample classified by
-the algorithms corresponds to a 30-second epoch scored during the ZM
-recordings. Subsequently, we performed a feature extraction process that
-yielded a set of 88 features, providing a robust characterization of the
-data. Extracted from accelerometer and temperature signals, these
-features include temporal elements that use both lag and lead values,
-capturing dynamic data trends by incorporating measurements from
-preceding and upcoming intervals. Furthermore, inspired by Walch et
-al.\citep{walch2019}, we developed sensor-independent features to
+In this study, data processing of the raw accelerometer data began with
+a low-pass filtration step using a 4th order Butterworth filter with a 5
+Hz cut-off frequency to eliminate high-frequency noise. Following
+filtration, data were partitioned into overlapping 2-second intervals,
+each successive interval sharing a 50\% overlap with the previous one
+similar to methods described by Skotte et
+al.\citep{skotte_detection_2014}. Any non-wear data was remove using
+previously described methods\citep{skovgaard2023} and data was resampled
+to 30-second epochs so every sample classified by the algorithms
+corresponds to a 30-second epoch scored during the ZM recordings.
+Subsequently, we performed a feature extraction process that yielded a
+set of 88 features, providing a robust characterization of the data.
+Extracted from accelerometer and temperature signals, these features
+include temporal elements that use both lag and lead values, capturing
+dynamic data trends by incorporating measurements from preceding and
+upcoming intervals. Furthermore, inspired by Walch et
+al.\citep{walch2019}, we incorporated sensor-independent features to
 encapsulate circadian rhythms. These features offer unique insights not
@@
 encompass vector magnitude, mean crossing rate, skewness, and kurtosis
-for each of the x, y, and z dimensions. These features provide a
-comprehensive understanding of the signal's attributes, facilitating a
-more nuanced analysis of the underlying patterns. All features are
-summarized in table ??? (overvej tabel til supp. material)
+for each of the x, y, and z dimensions. All features are summarized in
+table ??? \textbf{in the supplementary matrials}. Subsequently, we
+merged the ZM and corresponding accelerometer recordings. Any
+overlapping time between the ZM and accelerometer data was treated as
+`in-bed' time, with the remaining time considered `out-of-bed'. This
+process yielded a dataset providing a around the clock temporal view of
+each participant's activity and sleep patterns.
 
@@
 typically undergo around five to eight sleep cycles per night, with
-awakenings commonly occurring at the end of each
-cycle\citep{galland_normal_2012}. Examining the raw ZM predictions, we
-noted a significant overestimation in the number of awakenings per night
-for the children in our study, exceeding what would be expected based on
-typical sleep cycle patterns. Consequently, we elected to train and
-evaluate our models using not only the raw ZM output, but also versions
-that were subjected to 5-minute and 10-minute median filters. This
-approach resulted in an anticipated, more age-appropriate count of
-awakenings per night, providing a more accurate depiction of children's
-sleep patterns (see Figure~\ref{fig-zm-median}).
+awakenings most likely occurring at the end of each
+cycle\citep{galland_normal_2012}. Upon examining the raw ZM predictions,
+we noted a significant overestimation in the number of awakenings per
+night for the children in our study, exceeding what would be expected
+based on typical sleep cycle patterns. In particular, many of these
+brief awakenings could be considered as noise, which when present in the
+data, can potentially hinder the learning process of machine learning
+models by obscuring the underlying patterns that the models are trying
+to learn, leading to less accurate predictions. Consequently, we elected
+to train and evaluate our models using not only the raw ZM output, but
+also versions that were subjected to 5-minute and 10-minute median
+filters. This approach, by mitigating this noise, resulted in an
+anticipated, more age-appropriate count of awakenings per night,
+providing a more accurate depiction of children's sleep patterns (see
+Figure~\ref{fig-zm-median}).
 
@@
 
-In our study, we utilize two distinct modeling strategies to predict
-sleep patterns derived from thigh-mounted accelerometer data: a
-sequential ensemble and a multiclass bidirectional Long Short-Term
-Memory (biLSTM)\citep{hochreiter1997} neural network. The sequential
-ensemble involves using four pairs of models, each pair consisting of
-the same algorithm, that predict in-bed and sleep time successively.
-Conversely, the biLSTM is a comprehensive model that predicts multiple
-classes simultaneously, effectively handling complex temporal patterns.
-By applying and comparing both, we aim to evaluate their relative
-effectiveness.
+In this study, we employed two distinct modeling strategies to analyze
+sleep patterns from thigh-mounted accelerometer data. We used a
+sequential strategy, comprising an ensemble of four pairs of models,
+each pair featuring the same algorithm. This strategy aimed to make the
+prediction task more manageable for the algorithms by breaking it down
+into a sequence of two binary classifications: first predicting `in-bed'
+time, then `sleep' time. Simultaneously, we also used a multiclass
+approach utilizing a bidirectional Long Short-Term Memory
+(biLSTM)\citep{hochreiter1997} neural network.
 
@@
 
-The biLSTM is a multiclass classifier capable of predicting the classes
-of out-bed-awake, in-bed-awake, and in-bed-asleep. The architecture of
-the biLSTM consisted of four layers, each with 128 hidden units. The
-choice of four layers and 128 hidden units in our model balanced
-complexity and efficiency: it was deep enough to learn intricate
-patterns yet feasible to train timely. Additionally, the bidirectional
-nature of the LSTM, doubling the hidden units at each time step,
-improved data comprehension and avoided overfitting.
-
-something on sequence length and step size
-
-The biLSTM model was chosen for sleep classification using accelerometer
-data from a thigh-mounted device due to its advantages in modeling
-temporal context, handling a large temporal scope, and enabling temporal
-inference over any feature without the need for hand-designed temporal
-features.
-
-Previous studies have explored the use of LSTM models for sleep
-detection, showing promising results in capturing complex temporal
-patterns. Notable works by Sano et al. \citep{sano2019} and Chen et al.
-\citep{chen2021} have demonstrated the potential of LSTM models in
-improving sleep detection based on accelerometer data. However, further
-research is necessary to fully comprehend the strengths and limitations
-of this approach.
+In this study, we employed a biLSTM, a multiclass classifier, to predict
+three distinct states: out-of-bed-awake, in-bed-awake, and
+in-bed-asleep. The architecture of the biLSTM was set up with four
+layers, each equipped with 128 hidden units. This configuration was
+intentionally chosen to balance between model complexity and training
+efficiency: it provided the depth necessary for learning intricate
+patterns while remaining feasible for timely training. The bidirectional
+design of the LSTM served to enhance data interpretation and mitigate
+overfitting by doubling the hidden units at each time step. For input,
+we used tensors shaped as sequences, with each sequence spanning 10
+minutes and a step size of one.This approach follows in the footsteps of
+previous studies that utilized LSTM models for sleep detection. These
+studies showcased the promising potential of LSTMs in capturing complex
+temporal patterns. Particularly, the works of Sano et al.
+\citep{sano2019} and Chen et al. \citep{chen2021} demonstrated the
+effectiveness of LSTM models in improving sleep detection using
+accelerometer data, underscoring the value of this modeling approach.
 
@@
 For the models in sequence, we trained four pairs of classification
-models, each pair distinguishing between in-bed/out-of-bed and
-asleep/awake states. The dataset was randomly split into a training set
-(approximately 50\% of the subjects) and a testing set (also
-approximately 50\%), ensuring that samples from the same subject were
-never simultaneously used in both sets. Hyper-parameter optimization was
-performed using 10-fold Monte Carlo cross-validation, with the F1 score
-as the performance metric. The best-performing set of hyperparameters
-was then used to fit the models to the full training dataset, maximizing
-accuracy by leveraging all available data. Furthermore, the dataset for
-predicting sleep was highly imbalanced which can pose challenges when
-training machine learning models, as models may favor predicting the
-majority class. To address this issue, we employed oversampling
-techniques to ensure that each class had a roughly equal number of
-training samples. Specifically, we utilized the Synthetic Minority
-Over-sampling Technique (SMOTE)\citep{chawla2002}, which generates new
-samples by interpolating random samples with their nearest neighbors. In
-our study, we implemented SMOTE using the themis R
-package\citep{themis}, resampling all classes to achieve a balanced
-distribution of training samples.
+models. Each pair was designed to distinguish between in-bed/out-of-bed
+and asleep/awake states, respectively. The dataset was randomly split
+into a training set and a testing set, each containing approximately
+50\% of the subjects. This division ensured that samples from the same
+night were never simultaneously present in both sets. To optimize
+hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
+a regular grid, comprising 20 different combinations of hyperparameters.
+The F1 score served as the optimization metric. The best-performing set
+of hyperparameters was then used to fit the models to the full training
+dataset. This approach allowed us to maximize performance by leveraging
+all available data. Moreover, after extracting the in-bed time from the
+initial sequential models, the imbalance on the resulting dataset could
+cause biases during model training, as models may favor predicting the
+majority class. To rectify this, we employed the Synthetic Minority
+Over-sampling Technique (SMOTE)\citep{chawla2002}. SMOTE generates new
+samples by interpolating random samples with their nearest neighbors. We
+utilized the themis R package\citep{themis} to implement SMOTE,
+resulting in a balanced distribution of training samples across both
+classes.
 
-The biLSTM model was trained using the Adam optimizer, which is
-computationally efficient and adapts the learning rate during training.
-Cross-entropy loss function was employed for its suitability in
-multiclass classification with mutually exclusive classes. The softmax
-activation function was selected for the output layer to obtain a
-probability distribution over the classes. Data for training the biLSTM
-were randomly divided into a training, validation, and test tensors
-based on a 50/25/25 split. The model was evaluated using the F1 score on
-both the training and validation sets. Early stopping was implemented
-with a patience of 3 epochs, meaning training would stop if there was no
-improvement in the validation loss for 3 consecutive epochs.
+The biLSTM model was trained to differentiate between three states:
+out-of-bed-awake, in-bed-awake, and in-bed-asleep. The data used for
+training the biLSTM was randomly divided into training, validation, and
+test sets, based on a 50/25/25 split. We ensured that data from the same
+night was not present across different sets. The model was trained using
+the Adam optimizer, selected for its computational efficiency and
+adaptability of the learning rate during training. Given the multiclass
+classification task with mutually exclusive classes, we employed the
+cross-entropy loss function. To obtain a probability distribution over
+the classes, the softmax activation function was applied at the output
+layer. We evaluated the model's performance using the F1 score on both
+the training and validation sets. We implemented early stopping with a
+patience of 3 epochs, halting the training process if there was no
+improvement in the validation loss over three consecutive epochs.
 
-\hypertarget{model-validation-evt.-bare-statistics}{%
-\subsubsection{Model Validation {[}evt. bare
-``statistics''{]}}\label{model-validation-evt.-bare-statistics}}
+\hypertarget{model-validation}{%
+\subsubsection{Model Validation}\label{model-validation}}
 
@@
 Additionally, we evaluated the multiclass classifier, biLSTM, using
-macro-averaged F1-score, sensitivity, and precision. Furthermore, we
-present precision-recall curves and receiver operating characteristic
-curves, including the area under the curve, for each model and for each
-type of ZM prediction.We considered both the in-bed/out-of-bed and
-awake/asleep scoring tasks as binary classification problems,
-designating in-bed and asleep as the positive labels and out-of-bed and
-awake as the negative labels in accordance with previous
-research\citep{hjorth2012, kushida2001}.
+macro-averaged F1-score, sensitivity, and precision. To further
+illustrate model performance, we provide confusion matrices for the full
+dataset, encompassing both in-bed and out-of-bed data. These matrices
+report relative counts, column percentages (the proportion of the true
+class accurately predicted), and row percentages (the proportion of
+predictions correctly classified). We considered both the
+in-bed/out-of-bed and awake/asleep scoring tasks as binary
+classification problems, designating in-bed and asleep as the positive
+labels and out-of-bed and awake as the negative labels in accordance
+with previous research\citep{hjorth2012, kushida2001}.
 
-To evaluate our developed models' performance in assessing sleep quality
-metrics, we employed Bland-Altman plots and Pearson correlations.
-Additionally, we used the Bland-Altman method to assess the agreement
-between two measurement techniques. Given the multiple observations per
-subject in our dataset, we incorporated a bootstrap procedure to account
-for this additional source of variability. We computed the mean
-difference (bias) and the limits of agreement as the mean difference
-±1.96 times the standard deviation of the differences. The 95\%
-confidence intervals for the bias and the limits of agreement were
-estimated using the bias-corrected and accelerated bootstrap method with
-10,000 bootstrap replicates, to ensure our measurements are robust and
-account for intra-subject variability. These methods were used to
-compare the sleep quality summaries calculated by our models with those
-obtained from EEG-based ZM sleep quality summaries. The sleep quality
-summaries included the following metrics:
+To assess the performance of our models in deriving sleep quality
+parameters, we utilized Bland-Altman plots and Pearson correlations. The
+Bland-Altman method was employed specifically to determine the level of
+agreement between two measurement techniques. Considering our dataset
+contained multiple observations per subject, we integrated a bootstrap
+procedure to address this extra source of variability. We calculated the
+mean difference (bias) and defined the limits of agreement as the mean
+difference plus or minus 1.96 times the standard deviation of these
+differences. To ensure our measurements were robust and accounted for
+intra-subject variability, we estimated the 95\% confidence intervals
+for both the bias and the limits of agreement using a bias-corrected and
+accelerated bootstrap method, utilizing 10,000 bootstrap replicates. The
+sleep quality parameteres included the following:
 
@@
 
-Our models were trained to predict sleep quality metrics using the full
-span of accelerometer recordings, beyond the ZM recording duration. This
-approach, applied to all models and ZM prediction types, allows for
-comprehensive sleep pattern analysis on longitudinal accelerometer data.
-
 R version 4.3.0 (2023-04-21)\citep{R-lang} and the
@@
 
-\begin{figure}[b]
-
-{\centering \includegraphics{visuals/flowchart_of_elligible_nights.pdf}
-
-}
-
-\caption{Flowchart of eligible nights included in the study - revise
-text boxes!}
-
-\end{figure}
-
-The analysis included children with an average age of 9.4 years (SD =
-2.1). The raw ZM predictions covered 2,035,261 sleep epochs, or about
-86\% of the total recording duration.
-
-As reported in Table~\ref{tbl-zm_overview} the sleep quality metrics
+As reported in Table~\ref{tbl-zm_overview} the sleep quality parameters
 derived from ZM predictions were modified by the implementation of
-5-minute and 10-minute median filters. The Sleep Period Time (SPT) were
-consistent across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as
-this corresponds to the length of the ZM recording. Total Sleep Time
-(TST) and Sleep Efficiency (SE) increased in the filtered data, implying
-the filters categorize some wakefulness as sleep. Specifically, TST
-increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
+5-minute and 10-minute median filters. SPT were consistent across raw
+and filtered datasets (mean: 9.2 ± 2.1 hours), corresponding to the
+length of the ZM recording. TST and SE increased in the filtered data,
+implying the filters categorize some wakefulness as sleep. Specifically,
+TST increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
 (5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
-from 82.6 ± 12.0\% to 86.4 ± 12.7\% and 87.5 ± 12.9\% respectively.
-Latency to Persistent Sleep (LPS) also elevated, suggesting the filter
-smooths out brief awakenings at sleep onset, leading to a prolonged time
-to persistent sleep. The most significant change was seen in Wake After
-Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
-to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
-10-minute filtered data, respectively. The number of awakenings was also
-considerably reduced with the application of filters. In the raw data,
-the average number of awakenings was 34.46 ± 11.33 per night, which
-reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the 5-minute and 10-minute
-filtered data sets respectively. These results underscore the role of
-median filters on sleep metrics.
+from 82.6 ± 12.0\% to 86.4 ± 12.7\% and 87.5 ± 12.9\% respectively. LPS
+also elevated, suggesting the filter smooths out brief awakenings at
+sleep onset, leading to a prolonged time to persistent sleep. A
+significant change was seen in WASO, which dropped from 39.0 ± 33.6
+minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in
+the 5-minute and 10-minute filtered data, respectively. The number of
+awakenings was also considerably reduced with the application of
+filters. In the raw data, the average number of awakenings was 34.46 ±
+11.33 per night, which reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the
+5-minute and 10-minute filtered data sets respectively.
 
@@
 \midrule
-Raw ZM Predictions & 9.21 (2.15) & 7.74 (1.93) & 82.61 (11.97) & 34.53 (27.87) & 39 (33.56) & 34.46 (11.33) \\ 
-5-Min Median & 9.21 (2.15) & 8.1 (2.03) & 86.35 (12.72) & 36.26 (39.8) & 30.63 (46.84) & 4.43 (3.26) \\ 
-10-Min Median & 9.21 (2.15) & 8.21 (2.06) & 87.54 (12.89) & 37.99 (48.7) & 22.27 (55.38) & 1.95 (2.01) \\ 
+Raw ZM Predictions & 9.2 (2.1) & 7.7 (1.9) & 82.6 (12) & 34.5 (27.9) & 39 (33.6) & 34.5 (11.3) \\ 
+5-Min Median & 9.2 (2.1) & 8.1 (2) & 86.4 (12.7) & 36.3 (39.8) & 30.6 (46.8) & 4.4 (3.3) \\ 
+10-Min Median & 9.2 (2.1) & 8.2 (2.1) & 87.5 (12.9) & 38 (48.7) & 22.3 (55.4) & 1.9 (2) \\ 
 \bottomrule
@@
 \hypertarget{performance-on-epoch-to-epoch-basis}{%
-\subsection{Performance on Epoch-To-Epoch
+\subsection{Performance on Epoch-to-Epoch
 Basis}\label{performance-on-epoch-to-epoch-basis}}
 
-The epoch-to-epoch evaluation of in-bed performance metrics, outlined in
-Table~\ref{tbl-in_bed_performance}, demonstrates practically equivalent
-performance across all model types. While the Decision Tree model posted
-an F1 score of 94.4\% and accuracy of 95.3\%, the Logistic Regression
-and Feed-Forward Neural Network models each exhibited an F1 score of
-95.0\% and similar accuracies of 95.7\% and 95.8\%, respectively. The
+The epoch-to-epoch evaluation of predicting in-bed time is outlined in
+Table~\ref{tbl-in_bed_performance}, and demonstrates practically
+equivalent performance across all model types. The F1 score ranges from
+94.4\% (Decision Tree) to 95.4\% (XGBoost), while accuracy ranges from
+95.3\% (Decision Tree) to 96.1\% (XGBoost). Sensitivity, Precision, and
+Specificity also demonstrate consistent results across the models. The
 XGBoost model, despite recording the highest metrics with an F1 score of
-95.4\% and accuracy of 96.1\%, outpaced the others only marginally. This
-underscores the consistency of performance among these models in
-classifying in-bed conditions.
+95.4\% and accuracy of 96.1\%, outpaced the others only marginally.
 
@@
 \midrule
-Decision Tree & $94.36$ & $95.27$ & $93.12$ & $95.64$ & $96.86$ \\ 
-Logistic Regression & $94.99$ & $95.74$ & $95.03$ & $94.94$ & $96.26$ \\ 
-Feed-Forward Neural Net & $95.03$ & $95.77$ & $95.07$ & $94.99$ & $96.29$ \\ 
-XGBoost & $95.38$ & $96.06$ & $95.83$ & $94.94$ & $96.23$ \\ 
+Decision Tree & $94.4$ & $95.3$ & $93.1$ & $95.6$ & $96.9$ \\ 
+Logistic Regression & $95.0$ & $95.7$ & $95.0$ & $94.9$ & $96.3$ \\ 
+Feed-Forward Neural Net & $95.0$ & $95.8$ & $95.1$ & $95.0$ & $96.3$ \\ 
+XGBoost & $95.4$ & $96.1$ & $95.8$ & $94.9$ & $96.2$ \\ 
 \bottomrule
@@
 
-Table~\ref{tbl-sleep_performance} details the performance of the
+Table~\ref{tbl-sleep_performance} details the performance of the second
 sequential models on raw and median-filtered (5 and 10 minute) ZM
-predictions. The F1 Scores, which are unweighted macro averages, for raw
-ZM predictions range from 71.05\% to 76.18\%. The models perform
-comparably, but the low Specificity values (62.84\% to 70.93\%) suggest
-difficulty in correctly classifying negative instances. Applying
-5-minute median filtering improves the performance metrics. The XGBoost
-model tops the charts with an F1 Score of 79.22\% and NPV of 74.00\%.
-However, Specificity still remains low, with values between 54.68\% and
-74.84\% across all models. With 10-minute median filtering, the metrics
-improve further. The XGBoost model still leads with an F1 Score of
-80.87\% and an NPV of 75.76\%. But, Specificity remains a concern,
-ranging from 57.47\% to 76.35\% across all models.
+predictions. These are the predictions of the sequential models' ability
+to predict asleep/awake on only the extracted in-bed time. The F1
+scores, which are unweighted macro averages, for raw ZM predictions
+range from 71.05\% to 76.18\%. The models perform comparably, but the
+low specificity values (62.84\% to 70.93\%) suggest difficulty in
+correctly classifying awake epochs. Applying 5-minute median filtering
+improves the performance metrics. The XGBoost model tops the charts with
+an F1 score of 79.22\% and NPV of 74.00\%. However, specificity still
+remains low, with values between 54.68\% and 74.84\% across all models.
+With 10-minute median filtering, the metrics improve further. The
+XGBoost model still leads with an F1 score of 80.87\% and an NPV of
+75.76\%. But, specificity remains low, ranging from 57.47\% to 76.35\%
+across all models.
 
 \hypertarget{tbl-sleep_performance}{}
-\begin{longtable}{cccccc}
+\begin{longtable}{lrrrrr}
 \caption{\label{tbl-sleep_performance}Sleep Performance Metrics }\tabularnewline
@@
 \midrule
-Decision Tree & $72.94$ & $93.24$ & $48.36$ & $86.34$ & $67.15$ \\ 
-Logistic Regression & $71.05$ & $93.72$ & $43.88$ & $82.72$ & $70.93$ \\ 
-Neural Network & $71.76$ & $93.77$ & $45.13$ & $83.59$ & $70.83$ \\ 
-XGBoost & $76.18$ & $92.80$ & $57.98$ & $91.32$ & $62.84$ \\ 
+Decision Tree & $72.9$ & $93.2$ & $48.4$ & $86.3$ & $67.1$ \\ 
+Logistic Regression & $71.0$ & $93.7$ & $43.9$ & $82.7$ & $70.9$ \\ 
+Neural Network & $71.8$ & $93.8$ & $45.1$ & $83.6$ & $70.8$ \\ 
+XGBoost & $76.2$ & $92.8$ & $58.0$ & $91.3$ & $62.8$ \\ 
 \midrule
@@
 \midrule
-Decision Tree & $75.48$ & $94.20$ & $55.46$ & $93.35$ & $59.01$ \\ 
-Logistic Regression & $68.32$ & $95.84$ & $36.03$ & $81.36$ & $74.84$ \\ 
-Neural Network & $71.74$ & $95.78$ & $41.64$ & $85.62$ & $73.11$ \\ 
-XGBoost & $79.22$ & $93.87$ & $74.00$ & $97.30$ & $54.68$ \\ 
+Decision Tree & $75.5$ & $94.2$ & $55.5$ & $93.4$ & $59.0$ \\ 
+Logistic Regression & $68.3$ & $95.8$ & $36.0$ & $81.4$ & $74.8$ \\ 
+Neural Network & $71.7$ & $95.8$ & $41.6$ & $85.6$ & $73.1$ \\ 
+XGBoost & $79.2$ & $93.9$ & $74.0$ & $97.3$ & $54.7$ \\ 
 \midrule
@@
 \midrule
-Decision Tree & $76.28$ & $94.74$ & $58.06$ & $94.86$ & $57.47$ \\ 
-Logistic Regression & $67.96$ & $96.54$ & $34.29$ & $81.87$ & $76.35$ \\ 
-Neural Network & $70.95$ & $96.06$ & $39.54$ & $86.48$ & $71.37$ \\ 
-XGBoost & $80.87$ & $94.90$ & $75.76$ & $97.72$ & $57.60$ \\ 
+Decision Tree & $76.3$ & $94.7$ & $58.1$ & $94.9$ & $57.5$ \\ 
+Logistic Regression & $68.0$ & $96.5$ & $34.3$ & $81.9$ & $76.4$ \\ 
+Neural Network & $71.0$ & $96.1$ & $39.5$ & $86.5$ & $71.4$ \\ 
+XGBoost & $80.9$ & $94.9$ & $75.8$ & $97.7$ & $57.6$ \\ 
 \bottomrule
@@
 
-The analysis of precision-recall and ROC curves across different models
-and ZM prediction types shows varying performance. In terms of
-precision-recall AUC, the Decision Tree model consistently outperforms
-others, indicating its superior predictive accuracy (see
-Figure~\ref{fig-pr_curves}). Conversely, the Neural Network model
-generally shows weaker performance. However, for ROC AUC, the XGBoost
-model consistently excels across all data types, indicating a strong
-ability to differentiate between classes, while the Neural Network model
-tends to underperform (see Figure~\ref{fig-roc_curves}). The F-measure
-(F1 score) shows variable performance across different configurations
-but generally, the Decision Tree model yields higher scores.
+The complete set of confusion matrices generated from data both
+containing the out-of-bed and in-bed time are presented in
+Figure~\ref{fig-conf_mat}. The figure shows favorable epoch-to-epoch
+performance across across all sequential models, however, it is evident
+that the biLSTM is less successful in classifying the in-bed-awake class
+which cannot be deduced from the confusion matrices from the sequential
+models. \textbf{{[}fejl i biLSTM confmat. Alle mats er ens{]}}
 
@@
 
-{\centering \includegraphics{visuals/plot_sleep_pr.pdf}
+{\centering \includegraphics{visuals/all_conf_mats.pdf}
 
@@
 
-\caption{\label{fig-pr_curves}Precision-recall curves of the models
-evaluated across the different ZM predictions, including raw ZM
-predictions, as well as 5-minute and 10-minute median smoothing of the
-ZM raw predictions. The x-axis of the plot represents the proportion of
-true wake epochs that were correctly classified as wake, while the
-y-axis represents the proportion of all epochs labeled as wake by the
-classifier that were classified correctly. The area under the curve
-values are displayed as color-coded text in the plot indicate the area
-under the Precision-Recall curve for each model and condition.}
-
-\end{figure}
-
-\begin{figure}[b]
-
-{\centering \includegraphics{visuals/plot_sleep_roc.pdf}
-
-}
-
-\caption{\label{fig-roc_curves}Receiver operating characteristic curves
-of the models evaluated across the different ZM predictions, including
-raw ZM predictions, as well as 5-minute and 10-minute median smoothing
-of the ZM raw predictions. The x-axis of the plot represents the
-proportion of true asleep epochs that were incorrectly classified as
-awake, while the y-axis represents the proportion of all epochs labeled
-as awake by the classifier that were correctly classified. The area
-under the curve values displayed are displayed as color-coded text in
-the plot to indicate the area under the receiver operating
-characteristic curve for each model and condition.}
+\caption{\label{fig-conf_mat}Confusion matrices for binary sleep
+prediction. The middle of each tile is the normalized count (overall
+percentage) and, beneath it, the count. The bottom number is the column
+percentage (target). At the right side of each tile is the row
+percentage (prediction). i) decision tree, ii) logistic regression, iii)
+feed-forward neural net, iv) xgboost, and v) biLSTM.}
 
@@
 resulting in F1 Scores ranging from 75.99\% to 78.53\%, demonstrating
-enhanced precision and sensitivity. However, Specificity values are not
-available for this configuration. With 10-minute median filtering, F1
-Scores range from 73.45\% to 73.93\%, maintaining good performance but
-with limited information on Specificity. Further analysis is required to
-assess performance across all classes.
+enhanced precision and sensitivity. \textbf{{[}Overvej flere metrics.
+Evt. per class\ldots{} pt er det lidt svært at sammenligne på tværs af
+lstm og andre modeller ud over F1.{]}} With 10-minute median filtering,
+F1 Scores range from 73.45\% to 73.93\%, maintaining good performance
+but with limited information on Specificity. Further analysis is
+required to assess performance across all classes.
 
 \hypertarget{tbl-biLSTM_performance}{}
-\begin{longtable}{cccc}
+\begin{longtable}{cccccc}
 \caption{\label{tbl-biLSTM_performance}Performance of the three-class biLSTM multicclassifier. }\tabularnewline
@@
 \toprule
- & F1 Score & Sensitivity & Precision \\ 
+ & F1 Score & Precision & NPV & Sensitivity & Specificity \\ 
 \midrule
-Raw ZM Predictions & $71.36$ & $70.42$ & $76.04$ \\ 
-5-Min Median & $75.99$ & $74.62$ & $78.53$ \\ 
-10-Min Median & $73.45$ & $73.93$ & $73.07$ \\ 
+Raw ZM Predictions & $71.4$ & $76.0$ & $95.5$ & $70.4$ & $95.0$ \\ 
+5-Min Median & $76.0$ & $78.5$ & $96.4$ & $74.6$ & $96.2$ \\ 
+10-Min Median & $73.4$ & $73.1$ & $95.6$ & $73.9$ & $95.7$ \\ 
 \bottomrule
@@
 
-\hypertarget{evaluation-of-sleep-quality-summaries}{%
+\hypertarget{evaluation-of-sleep-quality-parameters}{%
 \subsection{Evaluation of Sleep Quality
-Summaries}\label{evaluation-of-sleep-quality-summaries}}
+Parameters}\label{evaluation-of-sleep-quality-parameters}}
 
-BA plots
+Table~\ref{tbl-ba_cor} presents a comparative analysis of the included
+models used to predict various sleep quality parameters (SPT, TST, SE,
+LPS, WASO) using the 5-minute median filtered ZM predictions. To see the
+full table including models developed from raw ZM predictions and
+10-minute median filtered ZM predictions, see \textbf{{[}SUPP. MAT.{]}}.
+In terms of bias, the decision tree model consistently underestimated
+SPT, TST, and SE, and overestimated LPS and WASO in comparison to ZM.
+The logistic regression model had similar trends, with more pronounced
+underestimation in TST and overestimation in LPS. The eed-forward neural
+network also exhibited similar bias as the decision tree and the
+logistic regression models, but with a higher overestimation in WASO. On
+the other hand, the XGboost model showed least bias among all,
+especially in its 5-minute median predictions. Considering LOA, the
+decision tree had higher variability across different sleep quality
+parameters and filtering techniques, particularly for LPS and WASO,
+which indicates lower agreement with ZM. Other models had comparable LOA
+but with notable exceptions. For example, TST LOA for the logistic
+regression model was particularly wide in the 5-minute median
+predictions. Correlation-wise, the pearson coefficient, revealed that
+the XGboost model consistently had the highest correlation with ZM
+across all sleep qualityparameters and filtering methods Notably, the
+XGboost's 5-minute median predictions showed the strongest correlation
+(0.66) for TST among all models and filtering techniques.
 
-pearson regression
+\hypertarget{tbl-ba_cor}{}
+\begin{longtable}{lrrrr}
+\caption{\label{tbl-ba_cor}Summary of Bias, Limits of Agreement (LOA), and Pearson Correlation for
+various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using
+different Machine Learning Models (Decision Tree, Logistic Regression,
+Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and
+10-Min Median as predictors. Each value is provided with its 95\%
+Confidence Interval (CI). }\tabularnewline
 
-scatterplots with identity line and best linear fit
+\toprule
+ & Bias (95\% CI) & LOA (95\% CI) & LOA (95\% CI) & Pearson, \emph{r} (95\% CI) \\ 
+\midrule
+\multicolumn{5}{l}{5-Min Median - Decision Tree} \\ 
+\midrule
+SPT (min) & -21.6 (-25.6;-17.6) & -117.5 (-125.6;-110.7) & 74.2 (63.9;85.9) & 0.54 (0.48;0.6) \\ 
+TST (min) & -50.5 (-55.2;-46) & -161.4 (-175.8;-151.3) & 60.4 (51.5;71.7) & 0.48 (0.42;0.54) \\ 
+SE (\%) & -5.5 (-6.3;-4.7) & -23.9 (-26.4;-22.2) & 12.9 (11.6;14.6) & 0.22 (0.14;0.29) \\ 
+LPS (min) & 24.6 (19.7;29.1) & -88.8 (-115;-77.3) & 138 (126.2;156.7) & 0.06 (-0.02;0.14) \\ 
+WASO (min) & 9.9 (6.5;14) & -79.4 (-109;-63.1) & 99.2 (80;136.1) & 0.15 (0.07;0.22) \\ 
+\midrule
+\multicolumn{5}{l}{5-Min Median - Logistic Regression} \\ 
+\midrule
+SPT (min) & -3.7 (-8;1) & -112.2 (-120.9;-105.2) & 104.8 (94;117.4) & 0.38 (0.3;0.44) \\ 
+TST (min) & -139.7 (-146.9;-133) & -305.6 (-323.6;-291.8) & 26.2 (16.1;38.6) & 0.09 (0.01;0.17) \\ 
+SE (\%) & -23.2 (-24.3;-22.2) & -48.1 (-50.9;-46.1) & 1.7 (0.1;3.8) & 0.13 (0.05;0.21) \\ 
+LPS (min) & 58.1 (53.4;62.6) & -52.3 (-75;-40.1) & 168.6 (155.9;187.7) & 0.05 (-0.03;0.13) \\ 
+WASO (min) & 45.4 (41.7;49.7) & -50.7 (-74.4;-38.4) & 141.5 (126.8;173) & 0.19 (0.11;0.27) \\ 
+\midrule
+\multicolumn{5}{l}{5-Min Median - Feed-Forward Neural Net} \\ 
+\midrule
+SPT (min) & -3.9 (-8.1;0.9) & -112.7 (-122;-105.2) & 104.9 (94.1;118.4) & 0.38 (0.3;0.44) \\ 
+TST (min) & -126.5 (-132.8;-120.3) & -276.8 (-291.3;-264.7) & 23.9 (14.8;33.9) & 0.25 (0.17;0.32) \\ 
+SE (\%) & -20.9 (-21.9;-19.9) & -44.3 (-46.3;-42.5) & 2.5 (1.1;4) & 0.21 (0.13;0.29) \\ 
+LPS (min) & 35.3 (30.7;39.8) & -75.8 (-102.3;-63.4) & 146.5 (134.4;166.9) & 0.07 (-0.01;0.15) \\ 
+WASO (min) & 45 (41.2;49.2) & -51.8 (-76.4;-39.1) & 141.7 (125.8;174.1) & 0.21 (0.14;0.29) \\ 
+\midrule
+\multicolumn{5}{l}{5-Min Median - XGboost} \\ 
+\midrule
+SPT (min) & 0.2 (-3.7;4.5) & -97.4 (-106.2;-90.3) & 97.8 (86.6;111) & 0.56 (0.5;0.61) \\ 
+TST (min) & -7 (-10.8;-3.3) & -95.5 (-105.2;-88) & 81.4 (72.4;92.5) & 0.66 (0.61;0.7) \\ 
+SE (\%) & -1.1 (-1.7;-0.5) & -15.6 (-17;-14.4) & 13.3 (12.2;14.7) & 0.44 (0.38;0.51) \\ 
+LPS (min) & 28.5 (23.9;32.6) & -76.4 (-104.2;-63.3) & 133.4 (120.4;154.2) & 0.12 (0.04;0.2) \\ 
+WASO (min) & -0.9 (-3.9;3) & -83.4 (-113.1;-66) & 81.7 (62;119.6) & 0.26 (0.18;0.33) \\ 
+\midrule
+\multicolumn{5}{l}{5-Min Median - biLSTM} \\ 
+\midrule
+SPT (min) & -36.1 (-41.7;-30) & -136.1 (-146.3;-126.9) & 64 (51.1;78.6) & 0.54 (0.45;0.62) \\ 
+TST (min) & 12.8 (7.4;18.3) & -80.1 (-89.8;-72.3) & 105.8 (94.3;118.8) & 0.63 (0.55;0.69) \\ 
+SE (\%) & 8 (7.2;8.8) & -5.1 (-6.8;-3.8) & 21.1 (19.5;23.1) & 0.16 (0.04;0.27) \\ 
+LPS (min) & -15.7 (-25.9;-7.5) & -169 (-230.7;-127.9) & 137.6 (101.1;184.9) & 0.09 (-0.02;0.2) \\ 
+WASO (min) & -3 (-9.9;7.7) & -144.1 (-197.2;-107.2) & 138.1 (90.8;211.4) & 0.02 (-0.1;0.13) \\ 
+\bottomrule
+\end{longtable}
+
+Figure~\ref{fig-xgb_ba_cor} shows the agreement of the XGBoost model
+trained on the 5-minute median filtered ZM predictions in estimating
+sleep quality parameters as evaluated 5-minute median-smoothed ZM
+derived sleep quality parameter. As can be seen on from the Bland-Altman
+plots, a large portion of the nights are in almost perfect agreement
+with the ZM and also a portion that can be considered outliers. This is
+true for all sleep quality parameters.
+
+\textbf{{[}Overvej at smide samtlige 75 plots i supp.mat.{]}}
+
+\begin{figure}[b]
+
+{\centering \includegraphics{visuals/median_5_xgboost_ba_cor.pdf}
+
+}
+
+\caption{\label{fig-xgb_ba_cor}Comparison of sleep quality parameters
+derived from the XGboost model trained on the 5-minute smoothed ZM
+predictions. The left column displays Bland-Altman plots. Dashed lines
+represent the bias (the average difference between the two measurements)
+and limits of agreement, with the 95\% confidence intervals represented
+as the grayed areas. The right column displays scatter plots of
+XGboost-derived vs ZM-derived sleep quality parameters. The dashed line
+represents the identity line, while the full-drawn line represents the
+best linear fit. Pearson's correlations are annotated in the upper left
+corner.}
+
+\end{figure}
 
@@
 
+It is unclear whether Johansson et
+al.\citep{johansson_development_2023}{]} provide a method for detecting
+the in-bed time\ldots{}
+
 In conclusion, this study provides compelling evidence of the utility of

### 2023-07-07 manuscript/supp_mat/supp.qmd
results almost done. XXgboost plots are ok.

---
title: "Supplmentary Material for DOI:"
date: last-modified
format:
  pdf
---

```{r}
#| echo: false
#| message: false
#| label: tbl-ba_cor
#| tbl-cap: Summary of Bias, Limits of Agreement (LOA), and Pearson Correlation for various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using different Machine Learning Models (Decision Tree, Logistic Regression, Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and 10-Min Median as predictors. Each value is provided with its 95% Confidence Interval (CI).

source(here::here("manuscript/code/table_ba_cor_statistics.R"))

tbl_ba_cor
```

The analysis of precision-recall and ROC curves across different models
and ZM prediction types shows varying performance. In terms of
precision-recall AUC, the Decision Tree model consistently outperforms
others, indicating its superior predictive accuracy (see
@fig-pr_curves). Conversely, the Neural Network model generally shows
weaker performance. However, for ROC AUC, the XGBoost model consistently
excels across all data types, indicating a strong ability to
differentiate between classes, while the Neural Network model tends to
underperform (see @fig-roc_curves). The F-measure (F1 score) shows
variable performance across different configurations but generally, the
Decision Tree model yields higher scores.

![Precision-recall curves of the models evaluated across the different ZM predictions, including raw ZM predictions, as well as 5-minute and 10-minute median smoothing of the ZM raw predictions. The x-axis of the plot represents the proportion of true wake epochs that were correctly classified as wake, while the y-axis represents the proportion of all epochs labeled as wake by the classifier that were classified correctly. The area under the curve values are displayed as color-coded text in the plot indicate the area under the Precision-Recall curve for each model and condition.](../visuals/plot_sleep_pr.pdf){#fig-pr_curves}

![Receiver operating characteristic curves of the models evaluated across the different ZM predictions, including raw ZM predictions, as well as 5-minute and 10-minute median smoothing of the ZM raw predictions. The x-axis of the plot represents the proportion of true asleep epochs that were incorrectly classified as awake, while the y-axis represents the proportion of all epochs labeled as awake by the classifier that were correctly classified. The area under the curve values displayed are displayed as color-coded text in the plot to indicate the area under the receiver operating characteristic curve for each model and condition.](../visuals/plot_sleep_roc.pdf){#fig-roc_curves}

![all ba and scatterplots](../visuals/all_ba_scatterplots.pdf)

### 2023-07-07 code/create_models/chained_models/in_bed_workflows.R

@@ -305,7 +305,7 @@ tune_wf_and_write <- function(wfs, fname, folds) {
       seed = 123,
       "tune_grid",
       resamples = folds,
-      grid = 5,
+      grid = 20,
       control = grid_ctrl,
       metrics = metric_set(f_meas),
       verbose = TRUE

### 2023-07-07 code/create_models/chained_models/sleep_workflows.R

@@ -293,7 +293,7 @@ tune_wf_and_write <- function(wfs, fname, folds) {
       seed = 123,
       "tune_grid",
       resamples = folds,
-      grid = 5,
+      grid = 20,
       control = grid_ctrl,
       metrics = metric_set(f_meas),
       verbose = TRUE

### 2023-07-07 code/evaluate_models/get_biLSTM_performance.R

@@ -46,7 +46,7 @@ test_preds <-
 # write_parquet(test_preds, "data/data_for_modelling/lstm/predictions/lstm_multiclass_preds.parquet")
 
 
-my_metrics <- metric_set(f_meas, sensitivity, precision)
+my_metrics <- metric_set(f_meas, precision, npv, sensitivity, specificity)
 
 get_metrics <- function(data, truth, estimate) {
   data %>%

### 2023-07-07 code/evaluate_models/get_bootstrap_ba_cis.R

@@ -22,12 +22,12 @@ get_ba <- function(d, i) {
 get_bootstrap_ci <- function(data, var) {
   # Run the bootstrap
   results <- boot(data[[var]], get_ba, R = 10000)
-  
+
   # Calculate the 95% confidence intervals
   bias_ci <- boot.ci(results, type = "bca", index = 1)$bca
   loa_upper_ci <- boot.ci(results, type = "bca", index = 2)$bca
   loa_lower_ci <- boot.ci(results, type = "bca", index = 3)$bca
-  
+
   # Return a tibble with the results
   return(tibble(
     variable = var,
@@ -47,7 +47,7 @@ get_bootstrap_ci <- function(data, var) {
 vars_to_bootstrap <- c("diff_spt", "diff_tst", "diff_se", "diff_lps", "diff_waso")
 
 # Read in the data
-stats <-
+ml_stats <-
   list.files("data/data_for_modelling/chained_classifiers/sleep_stats", full.names = TRUE) %>%
   read_csv(id = "model") %>%
   mutate(
@@ -59,17 +59,83 @@ stats <-
     diff_se = se_percent - zm_se,
     diff_lps = lps_min - zm_lps,
     diff_waso = waso_min - zm_waso
-  ) %>% 
-  relocate(sleep_type)
+  ) %>%
+  relocate(type = sleep_type)
 
 # Apply get_bootstrap_ci to each variable within each model group and sleep type, in parallel
-bootstrap_results <- stats %>%
-  group_by(sleep_type, model) %>%
-  do(bind_rows(future_map(vars_to_bootstrap, get_bootstrap_ci, data = ., 
-                          .options = furrr_options(seed = 123), .progress = TRUE))) %>%
-  ungroup()
+ml_ba_res <- ml_stats %>%
+  group_by(type, model) %>%
+  group_modify(~ future_map_dfr(vars_to_bootstrap, get_bootstrap_ci,
+    data = .x,
+    .options = furrr_options(seed = 123)
+  ), .keep = TRUE)
 
 # Save the results
-write_csv(bootstrap_results, "data/processed/bootstrap_ba_results_ml.csv")
+write_csv(ml_ba_res, "data/processed/bootstrap_ba_results_ml.csv")
+
+lstm_stats <- 
+  read_csv("data/data_for_modelling/lstm/stats/biLSTM_stats.csv") %>%
+  mutate(
[…]

### 2023-07-07 manuscript/code/plot_ba_and_scattercors.R

@@ -1,0 +1,428 @@
+library(tidyverse)
+library(arrow)
+library(patchwork)
+library(showtext)
+library(glue)
+library(ggtext)
+
+# model <- c("decision_tree", "logistic_regression", "neural_network", "xgboost", "biLSTM")
+# sleep_parameter <- c("spt", "tst", "se", "lps", "waso")
+
+# Read in the data
+ml_stats <-
+  list.files("data/data_for_modelling/chained_classifiers/sleep_stats", full.names = TRUE) %>%
+  read_csv(id = "model") %>%
+  mutate(
+    model = str_extract(model, "(?<=sleep_stats\\/).*(?=_stats.csv)")
+  ) %>%
+  select(-waso_min) %>%
+  rename_with(.fn = ~ str_remove(.x, pattern = "_.*"), .cols = spt_hrs:waso_3) %>%
+  mutate(
+    diff_spt = (spt - zm_spt) * 60,
+    diff_tst = (tst - zm_tst) * 60,
+    diff_se = se - zm_se,
+    diff_lps = lps - zm_lps,
+    diff_waso = waso - zm_waso) %>% 
+  rowwise() %>% 
+  mutate(
+    avg_spt = mean(c(spt, zm_spt)),
+    avg_tst = mean(c(tst, zm_tst)),
+    avg_se = mean(c(se, zm_se)),
+    avg_lps = mean(c(lps, zm_lps)),
+    avg_lps = mean(c(waso, zm_waso))
+  ) %>%
+  ungroup() %>% 
+  relocate(type = sleep_type)
+
+lstm_stats <-
+  read_csv("data/data_for_modelling/lstm/stats/biLSTM_stats.csv") %>%
+  drop_na() %>%
+  mutate(
+    diff_spt = (spt - zm_spt) * 60,
+    diff_tst = (tst - zm_tst) * 60,
+    diff_se = se - zm_se,
+    diff_lps = lps - zm_lps,
+    diff_waso = waso - zm_waso
+  ) %>%
+  rowwise() %>%
+  mutate(
+    avg_spt = mean(c(spt, zm_spt)),
+    avg_tst = mean(c(tst, zm_tst)),
+    avg_se = mean(c(se, zm_se)),
+    avg_lps = mean(c(lps, zm_lps)),
+    avg_lps = mean(c(waso, zm_waso))
+  ) %>%
+  ungroup() %>%
+  relocate(type, model, .after = 1) %>%
[…]

### 2023-07-07 manuscript/code/plot_ba_and_scatterplots.R

@@ -1,0 +1,191 @@
+library(tidyverse)
+library(arrow)
+library(patchwork)
+library(showtext)
+library(glue)
+library(ggtext)
+
+# Read in the data
+ml_stats <-
+  list.files("data/data_for_modelling/chained_classifiers/sleep_stats", full.names = TRUE) %>%
+  read_csv(id = "model") %>%
+  mutate(
+    model = str_extract(model, "(?<=sleep_stats\\/).*(?=_stats.csv)")
+  ) %>%
+  select(-waso_min) %>%
+  rename_with(.fn = ~ str_remove(.x, pattern = "_.*"), .cols = spt_hrs:waso_3) %>%
+  mutate(
+    diff_spt = (spt - zm_spt) * 60,
+    diff_tst = (tst - zm_tst) * 60,
+    diff_se = se - zm_se,
+    diff_lps = lps - zm_lps,
+    diff_waso = waso - zm_waso
+  ) %>%
+  rowwise() %>%
+  mutate(
+    avg_spt = mean(c(spt, zm_spt)),
+    avg_tst = mean(c(tst, zm_tst)),
+    avg_se = mean(c(se, zm_se)),
+    avg_lps = mean(c(lps, zm_lps)),
+    avg_waso = mean(c(waso, zm_waso))
+  ) %>%
+  ungroup() %>%
+  relocate(type = sleep_type)
+
+lstm_stats <-
+  read_csv("data/data_for_modelling/lstm/stats/biLSTM_stats.csv") %>%
+  drop_na() %>%
+  mutate(
+    diff_spt = (spt - zm_spt) * 60,
+    diff_tst = (tst - zm_tst) * 60,
+    diff_se = se - zm_se,
+    diff_lps = lps - zm_lps,
+    diff_waso = waso - zm_waso
+  ) %>%
+  rowwise() %>%
+  mutate(
+    avg_spt = mean(c(spt, zm_spt)),
+    avg_tst = mean(c(tst, zm_tst)),
+    avg_se = mean(c(se, zm_se)),
+    avg_lps = mean(c(lps, zm_lps)),
+    avg_waso = mean(c(waso, zm_waso))
+  ) %>%
+  ungroup() %>%
+  relocate(type, model, .after = 1) %>%
+  filter(!id %in% c(255704, 649105))
+
+all_cors <-
+  read_csv(here::here("data/processed/all_cors.csv"))
+
+ba_data <-
+  read_csv(here::here("data/processed/all_boostrap_ba_cis.csv"))
+
[…]

### 2023-07-07 manuscript/code/table_ba_cor_statistics.R

@@ -1,0 +1,121 @@
+library(tidyverse)
+library(glue)
+library(gt)
+library(showtext)
+
+
+stats <-
+  read_csv(here::here("data/processed/all_boostrap_ba_cis.csv")) %>%
+  mutate(
+    variable = str_remove(variable, "diff_")
+  ) %>%
+  left_join(read_csv(here::here("data/processed/all_cors.csv")),
+    by = c("type", "model", "variable" = "correlation")
+  )
+
+tbl_ba_cor <-
+  stats %>%
+  mutate(
+    model = factor(model, levels = c(
+      "decision_tree", "logistic_regression",
+      "neural_network", "xgboost", "biLSTM"
+    ))
+  ) %>%
+  arrange(model, desc(type)) %>%
+  mutate(
+    across(bias:loa_lower_ci_upper, ~ round(.x, 1)),
+    across(estimate:upper.ci, ~ round(.x, 2))
+  ) %>%
+  transmute(
+    type = type, model = model, variable = variable,
+    bias = glue("{bias} ({bias_ci_lower};{bias_ci_upper})"),
+    lower_loa = glue("{loa_lower} ({loa_lower_ci_lower};{loa_lower_ci_upper})"),
+    upper_loa = glue("{loa_upper} ({loa_upper_ci_lower};{loa_upper_ci_upper})"),
+    pearson = glue("{estimate} ({lower.ci};{upper.ci})")
+  ) %>%
+  mutate(
+    type = case_when(
+      type == "raw" ~ "Raw ZM Predictions",
+      type == "median5" ~ "5-Min Median",
+      type == "median10" ~ "10-Min Median"
+    ),
+    model = case_when(
+      model == "decision_tree" ~ "Decision Tree",
+      model == "logistic_regression" ~ "Logistic Regression",
+      model == "neural_network" ~ "Feed-Forward Neural Net",
+      model == "xgboost" ~ "XGboost",
+      TRUE ~ model
+    ),
+    variable = str_to_upper(variable),
+    variable = case_when(
+      variable == "SPT" ~ "SPT (min)",
+      variable == "TST" ~ "TST (min)",
+      variable == "SE" ~ "SE (%)",
+      variable == "LPS" ~ "LPS (min)",
+      variable == "WASO" ~ "WASO (min)"
+    )
+  ) %>%
+  gt(groupname_col = c("type", "model")) %>%
+  cols_label(
[…]

### 2023-07-07 manuscript/code/table_overview_ZM_stats.R

@@ -17,9 +17,9 @@ tbl_overview <-
   #        median_10_stat2_sleep_count_and_ratio = median_5_stat2_sleep_count_and_ratio * 100)
   transmute(
     row = row,
-    "Raw ZM Predictions" = glue("{round(raw_stat, 2)} ({round(raw_stat2, 2)})"),
-    "5-Min Median" = glue("{round(median_5_stat, 2)} ({round(median_5_stat2, 2)})"),
-    "10-Min Median" = glue("{round(median_10_stat, 2)} ({round(median_10_stat2, 2)})")
+    "Raw ZM Predictions" = glue("{round(raw_stat, 1)} ({round(raw_stat2, 1)})"),
+    "5-Min Median" = glue("{round(median_5_stat, 1)} ({round(median_5_stat2, 1)})"),
+    "10-Min Median" = glue("{round(median_10_stat, 1)} ({round(median_10_stat2, 1)})")
   ) %>%
   pivot_longer(-row) %>%
   pivot_wider(names_from = row, values_from = value) %>%
@@ -28,5 +28,4 @@ tbl_overview <-
     name = "", spt = "SPT (hrs)", tst = "TST (hrs)", se = "SE (%)", 
     lps = "LPS (min)", waso = "WASO (min)", no = "Awakenings (N)"
   ) %>%
-  fmt_number(decimals = 1) %>% 
   cols_align(align = "right", columns = -name)

### 2023-07-07 manuscript/code/table_sleep_performance.R

@@ -27,10 +27,11 @@ tbl_sleep_performance <-
   pivot_wider(names_from = .metric, values_from = .estimate) %>%
   rename_with(.cols = "F1 Score":"Specificity", ~ paste(.x, "(%)")) %>% 
   gt(groupname_col = "group") |>
-  fmt_number(everything(), decimals = 2) %>% 
+  fmt_number(everything(), decimals = 1) %>% 
   tab_options(table.font.names = "ibm") |>
   cols_width(everything() ~ px(140)) |>
-  cols_align(align = "center") %>%
+  cols_align(align = "left", columns = 1:2) %>%
+  cols_align(align = "right", columns = 3:7) %>% 
   cols_label(model = "") %>%
   tab_options(
     # Configure table appearance, such as font, colors, borders, padding, and alignment

### 2023-07-08 manuscript/manuscript.qmd
introduction makes more sense now. ZM is presented in the end of the intro.

@@
 
-An extensive array of research underlines the importance of sleep for
-both mental and physical health[@ma2017; @meyer2022; @kpavlova2019;
-@difrancesco2019]. Consequently, accurate sleep assessment methods are
-essential for tracking sleep patterns, thereby enhancing our
-comprehension of the sleep-health relationship. Furthermore, ensuring
-high user-acceptability for these methods is essential in order to
-conduct large-scale studies over prolonged periods.
+A vast body of research highlights the critical role of sleep in
+maintaining both mental and physical health[@ma2017; @meyer2022;
+@kpavlova2019; @difrancesco2019]. Consequently, accurate sleep
+assessment methods are crucial for tracking sleep patterns and improving
+our understanding of the sleep-health relationship. Furthermore, the
+ease of use and high acceptability of these methods are essential to
+facilitate large-scale, longitudinal studies.
 
-While laboratory-based polysomnography is considered the gold standard
-for objectively measuring sleep, its practicality in large-scale
-epidemiological studies is limited due to its high cost, the need for
-professional administration, and the substantial resources required for
-specialized equipment[@vandewater2011]. As an alternative, diaries is
-commonly used as low-cost and low-tech methods for sleep assessment in
-population research. However, relying solely on diary-based methods may
-introduce recall bias and other limitations[@moore2015]. A more feasible
-approach in large-scale epidemiological studies is the use of
-device-based measurement methods that can estimate sleep duration. This
-method offers the advantage of being less burdensome for participants
-and avoids potential biases associated with recall.
+While laboratory-based polysomnography (PSG) is typically considered the
+gold standard for objective sleep measurement, its practicality in
+large-scale epidemiological studies is hindered due to high costs, the
+necessity for professional administration, and it is also subject to
+potential rater bias[@vandewater2011; @lee2022]. As an alternative,
+diaries are commonly employed as cost-effective and low-tech methods for
+sleep assessment in population research. However, reliance on
+diary-based methods may lead to recall bias and other
+limitations[@moore2015]. A more feasible approach in large-scale
+epidemiological studies is to use device-based measurement methods that
+can estimate sleep duration. This approach offers the advantage of being
+less burdensome for participants and eliminates potential biases
+associated with recall.
 
-In this context, the Zmachine®️ Insight+ (ZM) emerges as a valuable
-tool. Validated against polysomnography with favorable
-results[@kaplan2014; @wang2015], the ZM provides comparable data without
-the high costs or need for professional monitoring associated with
-polysomnography. Furthermore, the ease of use of the ZM device makes it
-compliant for free-living use[@pedersen2021]. This allows for the
-analysis of multiple consecutive nights, as compared to single-night
-recordings from PSG, thereby capturing the important variations in sleep
-across multiple nights.
+Body-worn accelerometers have emerged as an effective and affordable
+alternative for objectively assessing sleep patterns at home over
+extended periods. These devices gather continuous, high-resolution data
+for several weeks without the need for recharging, thus reducing
+participant burden. Initial applications of accelerometry for sleep and
+wake stage classification were based on wrist movements, starting with
+an algorithm developed in 1982 and validated with PSG [@webster1982].
+This model was later refined in 1992 [@cole1992], giving rise to the
+widely used Cole-Kripke model. As the field advanced, an array of
+techniques, including heuristic algorithms, machine learning models,
+regression, and deep learning, were employed to analyze wrist-worn
+accelerometer data[@palotti2019; @cole1992; @sazonov2004; @sadeh1994;
+@hees2015; @sundararajan2021].
 
-The introduction of body-worn accelerometers has provided an effective
-and affordable alternative for objectively assessing sleep patterns in a
-home environment over extended periods. These accelerometers collect
-continuous, high-resolution data for several weeks without requiring
-recharging, thus minimizing participant burden. Initial applications of
-accelerometry for sleep and wake stage classification were based on
-wrist movements. The original algorithm, developed in 1982 using simple
-linear regression and validated with PSG [@webster1982], was later
-refined in 1992 [@cole1992], leading to the widely used Cole-Kripke
-model. Subsequent research on wrist-worn accelerometer data has employed
-heuristic algorithms, advanced machine learning models, as well as
-regression and deep learning techniques[@palotti2019; @cole1992;
-@sazonov2004; @sadeh1994; @hees2015; @sundararajan2021].
+While wrist and hip-worn devices have benefited from extensive
+methodological development, thigh-worn accelerometers have not seen the
+same level of advancement. Existing studies mainly focus on
+distinguishing sleep from wakefulness, with emphasis on defining 'waking
+time' and 'bedtime' [@carlson2021; @inan-eroglu2021; @vanderberg2016;
+@winkler2016]. Recent strides in estimating sleep duration using these
+devices have been made, including the introduction of a promising
+algorithm and its comparison against PSG[@johansson_development_2023].
+Despite these advancements, the application of machine learning
+techniques in this area is still relatively unexplored. Considering the
+potential of thigh-worn accelerometers for accurate physical behavior
+assessment[@skotte_detection_2014; @arvidsson2019], there is a
+significant research gap. Therefore, future studies need to develop
+techniques similar to those used for wrist and hip-worn accelerometers,
+with the ultimate goal of establishing a more holistic, accurate, and
+user-friendly method of sleep and physical activity tracking.
 
-Despite the well-developed field of accelerometer data analysis for
-sleep detection from wrist and hip-worn devices, the same level of
-advancement is not mirrored in studies utilizing thigh-worn
-accelerometers. Methods for assessing sleep using wrist and hip-worn
-accelerometers have greatly evolved over the years, employing an
-extensive range of techniques. These include heuristic algorithms,
-machine learning models, regression and deep learning techniques, all
-tailored to the specific signal characteristics of wrist and hip-worn
-devices [@palotti2019; @cole1992; @sazonov2004; @sadeh1994; @hees2015;
-@sundararajan2021; @patterson_40_2023]. However, for thigh-worn
-accelerometers, the landscape appears less mature, with only a handful
-of studies investigating sleep detection algorithms. The majority of the
-efforts are focused on delineating wakefulness from sleep, with
-particular emphasis on the definition of 'waking time' and 'bedtime'
-[@carlson2021; @inan-eroglu2021; @vanderberg2016; @winkler2016].
-Furthermore, while strides have been made recently in estimating sleep
-duration with these devices, with the introduction of a promising
-algorithm and its comparison against PSG[@johansson_development_2023],
-the field is still in its infancy when it comes to employing machine
-learning techniques. Given the potential for accurate physical behavior
-assessment that thigh-worn accelerometers
-provide[@skotte_detection_2014; @arvidsson2019], a significant research
-gap exists. Therefore, there is a pressing need for future studies to
-develop techniques similar to those used for wrist and hip-worn
-accelerometers, with the ultimate goal of establishing a more holistic,
-accurate, and user-friendly method of sleep and physical activity
-tracking.
+The Zmachine®️ Insight+ (ZM) emerges as a valuable tool within this
+landscape. Favorably validated against PSG[@kaplan2014; @wang2015], the
+ZM provides comparable data without the high costs or the need for
+professional monitoring typically associated with PSG. Crucially, the ZM
+facilitates multi-night analysis in free-living conditions due to its
+ease of use[@pedersen2021], capturing the natural variations in sleep
+patterns. This makes it advantageous over single-night PSG, particularly
+as a gold standard data source in machine learning tasks, as it provides
+multiple nights of measurements without inter-rater bias. Despite these
+benefits, the ZM, like PSG, still poses a significant participant burden
+and cost, reinforcing the need for more accessible alternatives like
+accelerometers.
 

### 2023-07-08 manuscript/manuscript.tex
introduction makes more sense now. ZM is presented in the end of the intro.

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, interior hidden, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-An extensive array of research underlines the importance of sleep for
-both mental and physical
+A vast body of research highlights the critical role of sleep in
+maintaining both mental and physical
 health\citep{ma2017, meyer2022, kpavlova2019, difrancesco2019}.
-Consequently, accurate sleep assessment methods are essential for
-tracking sleep patterns, thereby enhancing our comprehension of the
-sleep-health relationship. Furthermore, ensuring high user-acceptability
-for these methods is essential in order to conduct large-scale studies
-over prolonged periods.
+Consequently, accurate sleep assessment methods are crucial for tracking
+sleep patterns and improving our understanding of the sleep-health
+relationship. Furthermore, the ease of use and high acceptability of
+these methods are essential to facilitate large-scale, longitudinal
+studies.
 
-While laboratory-based polysomnography is considered the gold standard
-for objectively measuring sleep, its practicality in large-scale
-epidemiological studies is limited due to its high cost, the need for
-professional administration, and the substantial resources required for
-specialized equipment\citep{vandewater2011}. As an alternative, diaries
-is commonly used as low-cost and low-tech methods for sleep assessment
-in population research. However, relying solely on diary-based methods
-may introduce recall bias and other limitations\citep{moore2015}. A more
-feasible approach in large-scale epidemiological studies is the use of
-device-based measurement methods that can estimate sleep duration. This
-method offers the advantage of being less burdensome for participants
-and avoids potential biases associated with recall.
+While laboratory-based polysomnography (PSG) is typically considered the
+gold standard for objective sleep measurement, its practicality in
+large-scale epidemiological studies is hindered due to high costs, the
+necessity for professional administration, and it is also subject to
+potential rater bias\citep{vandewater2011, lee2022}. As an alternative,
+diaries are commonly employed as cost-effective and low-tech methods for
+sleep assessment in population research. However, reliance on
+diary-based methods may lead to recall bias and other
+limitations\citep{moore2015}. A more feasible approach in large-scale
+epidemiological studies is to use device-based measurement methods that
+can estimate sleep duration. This approach offers the advantage of being
+less burdensome for participants and eliminates potential biases
+associated with recall.
 
-In this context, the Zmachine®️ Insight+ (ZM) emerges as a valuable
-tool. Validated against polysomnography with favorable
-results\citep{kaplan2014, wang2015}, the ZM provides comparable data
-without the high costs or need for professional monitoring associated
-with polysomnography. Furthermore, the ease of use of the ZM device
-makes it compliant for free-living use\citep{pedersen2021}. This allows
-for the analysis of multiple consecutive nights, as compared to
-single-night recordings from PSG, thereby capturing the important
-variations in sleep across multiple nights.
+Body-worn accelerometers have emerged as an effective and affordable
+alternative for objectively assessing sleep patterns at home over
+extended periods. These devices gather continuous, high-resolution data
+for several weeks without the need for recharging, thus reducing
+participant burden. Initial applications of accelerometry for sleep and
+wake stage classification were based on wrist movements, starting with
+an algorithm developed in 1982 and validated with PSG
+\citep{webster1982}. This model was later refined in 1992
+\citep{cole1992}, giving rise to the widely used Cole-Kripke model. As
+the field advanced, an array of techniques, including heuristic
+algorithms, machine learning models, regression, and deep learning, were
+employed to analyze wrist-worn accelerometer
+data\citep{palotti2019, cole1992, sazonov2004, sadeh1994, hees2015, sundararajan2021}.
 
-The introduction of body-worn accelerometers has provided an effective
-and affordable alternative for objectively assessing sleep patterns in a
-home environment over extended periods. These accelerometers collect
-continuous, high-resolution data for several weeks without requiring
-recharging, thus minimizing participant burden. Initial applications of
-accelerometry for sleep and wake stage classification were based on
-wrist movements. The original algorithm, developed in 1982 using simple
-linear regression and validated with PSG \citep{webster1982}, was later
-refined in 1992 \citep{cole1992}, leading to the widely used Cole-Kripke
-model. Subsequent research on wrist-worn accelerometer data has employed
-heuristic algorithms, advanced machine learning models, as well as
-regression and deep learning
-techniques\citep{palotti2019, cole1992, sazonov2004, sadeh1994, hees2015, sundararajan2021}.
+While wrist and hip-worn devices have benefited from extensive
+methodological development, thigh-worn accelerometers have not seen the
+same level of advancement. Existing studies mainly focus on
+distinguishing sleep from wakefulness, with emphasis on defining `waking
+time' and `bedtime'
+\citep{carlson2021, inan-eroglu2021, vanderberg2016, winkler2016}.
+Recent strides in estimating sleep duration using these devices have
+been made, including the introduction of a promising algorithm and its
+comparison against PSG\citep{johansson_development_2023}. Despite these
+advancements, the application of machine learning techniques in this
+area is still relatively unexplored. Considering the potential of
+thigh-worn accelerometers for accurate physical behavior
+assessment\citep{skotte_detection_2014, arvidsson2019}, there is a
+significant research gap. Therefore, future studies need to develop
+techniques similar to those used for wrist and hip-worn accelerometers,
+with the ultimate goal of establishing a more holistic, accurate, and
+user-friendly method of sleep and physical activity tracking.
 
-Despite the well-developed field of accelerometer data analysis for
-sleep detection from wrist and hip-worn devices, the same level of
-advancement is not mirrored in studies utilizing thigh-worn
-accelerometers. Methods for assessing sleep using wrist and hip-worn
-accelerometers have greatly evolved over the years, employing an
-extensive range of techniques. These include heuristic algorithms,
-machine learning models, regression and deep learning techniques, all
-tailored to the specific signal characteristics of wrist and hip-worn
-devices
-\citep{palotti2019, cole1992, sazonov2004, sadeh1994, hees2015, sundararajan2021, patterson_40_2023}.
-However, for thigh-worn accelerometers, the landscape appears less
-mature, with only a handful of studies investigating sleep detection
-algorithms. The majority of the efforts are focused on delineating
-wakefulness from sleep, with particular emphasis on the definition of
-`waking time' and `bedtime'
-\citep{carlson2021, inan-eroglu2021, vanderberg2016, winkler2016}.
-Furthermore, while strides have been made recently in estimating sleep
-duration with these devices, with the introduction of a promising
-algorithm and its comparison against
-PSG\citep{johansson_development_2023}, the field is still in its infancy
-when it comes to employing machine learning techniques. Given the
-potential for accurate physical behavior assessment that thigh-worn
-accelerometers provide\citep{skotte_detection_2014, arvidsson2019}, a
-significant research gap exists. Therefore, there is a pressing need for
-future studies to develop techniques similar to those used for wrist and
-hip-worn accelerometers, with the ultimate goal of establishing a more
-holistic, accurate, and user-friendly method of sleep and physical
-activity tracking.
+The Zmachine®️ Insight+ (ZM) emerges as a valuable tool within this
+landscape. Favorably validated against PSG\citep{kaplan2014, wang2015},
+the ZM provides comparable data without the high costs or the need for
+professional monitoring typically associated with PSG. Crucially, the ZM
+facilitates multi-night analysis in free-living conditions due to its
+ease of use\citep{pedersen2021}, capturing the natural variations in
+sleep patterns. This makes it advantageous over single-night PSG,
+particularly as a gold standard data source in machine learning tasks,
+as it provides multiple nights of measurements without inter-rater bias.
+Despite these benefits, the ZM, like PSG, still poses a significant
+participant burden and cost, reinforcing the need for more accessible
+alternatives like accelerometers.
 

### 2023-07-11 manuscript/manuscript.qmd
biLSTM performance as two binary.

@@
 ---
-title: Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device
+title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
 author:
@@
   elsevier-pdf:
-    keep-tex: true
+    keep-tex: false
     journal:
@@
 
-In this study, we employed a biLSTM, a multiclass classifier, to predict
-three distinct states: out-of-bed-awake, in-bed-awake, and
-in-bed-asleep. The architecture of the biLSTM was set up with four
-layers, each equipped with 128 hidden units. This configuration was
-intentionally chosen to balance between model complexity and training
-efficiency: it provided the depth necessary for learning intricate
-patterns while remaining feasible for timely training. The bidirectional
-design of the LSTM served to enhance data interpretation and mitigate
-overfitting by doubling the hidden units at each time step. For input,
-we used tensors shaped as sequences, with each sequence spanning 10
-minutes and a step size of one.This approach follows in the footsteps of
-previous studies that utilized LSTM models for sleep detection. These
-studies showcased the promising potential of LSTMs in capturing complex
-temporal patterns. Particularly, the works of Sano et al. [@sano2019]
-and Chen et al. [@chen2021] demonstrated the effectiveness of LSTM
-models in improving sleep detection using accelerometer data,
-underscoring the value of this modeling approach.
+We also employed a biLSTM, a multiclass classifier, to predict three
+distinct states: out-of-bed-awake, in-bed-awake, and in-bed-asleep. The
+architecture of the biLSTM was set up with four layers, each equipped
+with 128 hidden units. This configuration was intentionally chosen to
+balance between model complexity and training efficiency: it provided
+the depth necessary for learning intricate patterns while remaining
+feasible for timely training. The bidirectional design of the LSTM
+served to enhance data interpretation and mitigate overfitting by
+doubling the hidden units at each time step. For input, we used tensors
+shaped as sequences, with each sequence spanning 10 minutes and a step
+size of one.This approach follows in the footsteps of previous studies
+that utilized LSTM models for sleep detection. These studies showcased
+the promising potential of LSTMs in capturing complex temporal patterns.
+Particularly, the works of Sano et al. [@sano2019] and Chen et al.
+[@chen2021] demonstrated the effectiveness of LSTM models in improving
+sleep detection using accelerometer data, underscoring the value of this
+modeling approach.
 
@@
 case led us to compute the F1 score as an unweighted macro-average.
-Additionally, we evaluated the multiclass classifier, biLSTM, using
-macro-averaged F1-score, sensitivity, and precision. To further
-illustrate model performance, we provide confusion matrices for the full
-dataset, encompassing both in-bed and out-of-bed data. These matrices
-report relative counts, column percentages (the proportion of the true
-class accurately predicted), and row percentages (the proportion of
-predictions correctly classified). We considered both the
+Additionally, we evaluated the multiclass classifier, biLSTM, using the
+same metrics. To do this, we considered the multiclass output as to
+binary classifications, where the first was out-of-bed vs the rest and
+the second binary classification as in-bed-awake vs in-bed-asleep. To
+further illustrate model performance, we provide confusion matrices for
+the full dataset, encompassing both in-bed and out-of-bed data. These
+matrices report relative counts, column percentages (the proportion of
+the true class accurately predicted), and row percentages (the
+proportion of predictions correctly classified). We considered both the
 in-bed/out-of-bed and awake/asleep scoring tasks as binary
@@
 procedure to address this extra source of variability. We calculated the
-mean difference (bias) and defined the limits of agreement as the mean
-difference plus or minus 1.96 times the standard deviation of these
-differences. To ensure our measurements were robust and accounted for
-intra-subject variability, we estimated the 95% confidence intervals for
-both the bias and the limits of agreement using a bias-corrected and
-accelerated bootstrap method, utilizing 10,000 bootstrap replicates. The
-sleep quality parameteres included the following:
+mean difference (bias) and defined the LOA as the mean difference plus
+or minus 1.96 times the standard deviation of these differences. To
+ensure our measurements were robust and accounted for intra-subject
+variability, we estimated the 95% confidence intervals for both the bias
+and the LOA using a bias-corrected and accelerated bootstrap method,
+utilizing 10,000 bootstrap replicates. The sleep quality parameteres
+included are defined as follews in accordance with the ZM definitions:
 
@@
     is calculated as the time from the beginning of the ZM recording
-    until a period when 10 out of 12 minutes are scored as sleep.
+    until the first period when 10 out of 12 minutes are scored as
+    sleep.
 
@@
 #| label: tbl-in_bed_performance
-#| tbl-cap: In-Bed Performance Metrics
+#| tbl-cap: In-bed performance metrics
 
@@
 
-@tbl-sleep_performance details the performance of the second sequential
-models on raw and median-filtered (5 and 10 minute) ZM predictions.
-These are the predictions of the sequential models' ability to predict
-asleep/awake on only the extracted in-bed time. The F1 scores, which are
-unweighted macro averages, for raw ZM predictions range from 71.05% to
-76.18%. The models perform comparably, but the low specificity values
-(62.84% to 70.93%) suggest difficulty in correctly classifying awake
-epochs. Applying 5-minute median filtering improves the performance
-metrics. The XGBoost model tops the charts with an F1 score of 79.22%
-and NPV of 74.00%. However, specificity still remains low, with values
-between 54.68% and 74.84% across all models. With 10-minute median
-filtering, the metrics improve further. The XGBoost model still leads
-with an F1 score of 80.87% and an NPV of 75.76%. But, specificity
-remains low, ranging from 57.47% to 76.35% across all models.
+@tbl-sleep_performance details the performance of all sequential model
+types on raw and median-filtered (5 and 10 minute) ZM predictions for
+sleep/wake classification. For raw ZM predictions, the F1 scores, which
+are unweighted macro averages, range from 65.6% (biLSTM) to 76.2%
+(XGBoost). All models perform comparably, but the low specificity values
+(62.5% to 70.9%) suggest difficulty in correctly classifying awake
+epochs. Applying a 5-minute median filter improves the performance
+metrics. The XGBoost model tops the charts with an F1 score of 79.2% and
+NPV of 74.0%. However, specificity still remains low, with values
+between 54.7% (XGBoost) and 74.8% (Logistic Regression) across all
+models. With a 10-minute median filter, the metrics improve further. The
+XGBoost model still leads with an F1 score of 80.9% and an NPV of 75.9%.
+But, specificity remains low, ranging from 57.5% (Decision Tree) to
+76.4% (Logistic Regression) across all models.
 
@@
 #| label: tbl-sleep_performance
-#| tbl-cap: Sleep Performance Metrics
+#| tbl-cap: Performance of the sleep/wake classification of the sequential models.
 
@@
 
-The complete set of confusion matrices generated from data both
-containing the out-of-bed and in-bed time are presented in
-@fig-conf_mat. The figure shows favorable epoch-to-epoch performance
-across across all sequential models, however, it is evident that the
-biLSTM is less successful in classifying the in-bed-awake class which
-cannot be deduced from the confusion matrices from the sequential
-models. **\[fejl i biLSTM confmat. Alle mats er ens\]**
+A complete set of confusion matrices generated from data both containing
+the out-of-bed and in-bed time are presented in @fig-conf_mat. The
+figure shows favorable epoch-to-epoch performance across across all
+sequential models, however, it is evident that the biLSTM is less
+successful in classifying the in-bed-awake class which cannot be deduced
+from the confusion matrices from the sequential models.
 
@@
 side of each tile is the row percentage (prediction). i) decision tree,
-ii) logistic regression, iii) feed-forward neural net, iv) xgboost, and
+ii) logistic regression, iii) feed-forward neural net, iv) XGBoost, and
 v) biLSTM.](visuals/all_conf_mats.pdf){#fig-conf_mat}
-
-@tbl-biLSTM_performance presents the performance of the three-class
-biLSTM multiclassifier on raw and median-filtered (5-minute and
-10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores ranging
-from 71.36% to 76.04%, indicating overall good performance. Applying
-5-minute median filtering improves the metrics further, resulting in F1
-Scores ranging from 75.99% to 78.53%, demonstrating enhanced precision
-and sensitivity. **\[Overvej flere metrics. Evt. per class... pt er det
-lidt svært at sammenligne på tværs af lstm og andre modeller ud over
-F1.\]** With 10-minute median filtering, F1 Scores range from 73.45% to
-73.93%, maintaining good performance but with limited information on
-Specificity. Further analysis is required to assess performance across
-all classes.
-
-```{r}
-#| echo: false
-#| message: false
-#| tbl-colwidths: auto
-#| label: tbl-biLSTM_performance
-#| tbl-cap: Performance of the three-class biLSTM multicclassifier.
-source("code/table_biLSTM_performance.R")
-
-tbl_biLSTM_performance
-```
 
@@
 similar bias as the decision tree and the logistic regression models,
-but with a higher overestimation in WASO. On the other hand, the XGboost
+but with a higher overestimation in WASO. On the other hand, the XGBoost
 model showed least bias among all, especially in its 5-minute median
@@
 in the 5-minute median predictions. Correlation-wise, the pearson
-coefficient, revealed that the XGboost model consistently had the
+coefficient, revealed that the XGBoost model consistently had the
 highest correlation with ZM across all sleep qualityparameters and
-filtering methods Notably, the XGboost's 5-minute median predictions
+filtering methods Notably, the XGBoost's 5-minute median predictions
 showed the strongest correlation (0.66) for TST among all models and
@@
 #| label: tbl-ba_cor
-#| tbl-cap: Summary of Bias, Limits of Agreement (LOA), and Pearson Correlation for various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using different Machine Learning Models (Decision Tree, Logistic Regression, Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and 10-Min Median as predictors.
+#| tbl-cap: Summary of Bias, Limits of Agreement, and Pearson Correlation for various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using different Machine Learning Models (Decision Tree, Logistic Regression, Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and 10-Min Median as predictors.
 Each value is provided with its 95% Confidence Interval (CI).
@@
 
-@fig-xgb_ba_cor shows the agreement of the XGBoost model trained on the
-5-minute median filtered ZM predictions in estimating sleep quality
-parameters as evaluated 5-minute median-smoothed ZM derived sleep
-quality parameter. As can be seen on from the Bland-Altman plots, a
-large portion of the nights are in almost perfect agreement with the ZM
-and also a portion that can be considered outliers. This is true for all
-sleep quality parameters.
+@fig-xgb_ba_cor shows the agreement between the XGBoost model, trained
+on 5-minute median filtered ZM predictions, and the 5-minute
+median-smoothed ZM-derived sleep quality parameters. The Bland-Altman
+plot for the SPT reveals a significant level of agreement with the ZM,
+as evidenced by a positive correlation. However, the presence of extreme
+outliers widens the limits of agreement (LOA). The scatterplot for SPT
+also demonstrates a positive trend, indicating a moderate linear
+correlation between the XGBoost model and the ZM-derived sleep quality
+parameters.In terms of TST, perfect agreement is not observed for a
+substantial number of nights, but there is a positive correlation
+between the XGBoost model and ZM-derived sleep quality parameters. The
+bias and LOA for TST are comparable to those observed for SPT,
+indicating a consistent level of agreement between the two methods. The
+scatterplot for TST also shows a slightly higher correlation, primarily
+driven by the absence of extreme outliers.Furthermore, the remaining
+three sleep quality parameters, SE, LPS, and WASO, exhibit
+heteroscedasticity in contrast to SPT and TST. This outcome is expected
+as achieving 100% sleep efficiency is relatively rare, resulting in less
+disagreement between the methods as values approach the upper limit.
+However, as sleep efficiency decreases, the potential for discrepancies
+and differing interpretations between the methods increases, leading to
+greater heteroscedasticity. A moderate linear correlation is observed
+between the XGBoost model and ZM-derived sleep quality parameters for
+SE, indicating a positive relationship. However, a poor correlation is
+observed for LPS and WASO, suggesting less agreement between the two
+methods for these parameters. Similar plots for all models are available
+in the supplementary materials.
 
-**\[Overvej at smide samtlige 75 plots i supp.mat.\]**
-
-![Comparison of sleep quality parameters derived from the XGboost model
+![Comparison of sleep quality parameters derived from the XGBoost model
 trained on the 5-minute smoothed ZM predictions. The left column
 displays Bland-Altman plots. Dashed lines represent the bias (the
-average difference between the two measurements) and limits of
-agreement, with the 95% confidence intervals represented as the grayed
-areas. The right column displays scatter plots of XGboost-derived vs
-ZM-derived sleep quality parameters. The dashed line represents the
-identity line, while the full-drawn line represents the best linear fit.
-Pearson's correlations are annotated in the upper left
+average difference between the two measurements) and LOA, with the 95%
+confidence intervals represented as the grayed areas. The right column
+displays scatter plots of XGBoost-derived vs ZM-derived sleep quality
+parameters. The dashed line represents the identity line, while the
+full-drawn line represents the best linear fit. Pearson's correlations
+are annotated in the upper left
 corner.](visuals/median_5_xgboost_ba_cor.pdf){#fig-xgb_ba_cor}
@@
 
-The present study embarked on an in-depth exploration of the utility and
-comparative performance of different machine learning models in the
-analysis of pediatric sleep data. The results offer promising insights
-into the role of machine learning in enhancing the precision of sleep
-quality metrics.
+In an effort to mature the methods for estimating sleep from thigh-worn
+accelerometers, we evaluated various models for predicting in-bed and
+sleep time and their derived sleep quality parameters. Furthermore, we
+trained and evaluated the models using raw and median-filtered
+gold-standard predictions from the ZM. In general, all sequential models
+performed well at predicting in-bed time. More challenging was it to
+distinguish wake from sleep on the extracted in-bed time, however, the
+performance of the sequential models were enhanced by the application of
+median filters. Moreover, even though the multiclass biLSTM showed good
+performance across F1 score, precision and NPV, the derived sleep
+quality parameters were not on par with the XGBoost model which
+demonstrated the highest performance metrics across all evaluations,
+including epoch-to-epoch prediction and sleep quality parameter
+derivatives. Despite this, all sequential models showed low specificity
+values, indicating difficulty in correctly classifying awake epochs
+during time in bed. The application of 5-minute and 10-minute median
+filters improved the performance metrics of all models, with the XGBoost
+model consistently leading. The filters increased total sleep time and
+sleep efficiency, while reducing wake after sleep onset and the number
+of awakenings. The XGBoost model also showed the least bias and highest
+correlation with ZM across sleep quality parameters.
+
+Limited research exists regarding the epoch-to-epoch effectiveness of
+classifying in-bed time based on data from thigh-worn accelerometers.
+Nevertheless, Carlson and colleagues provided compelling insights. They
+demonstrated that a third-party algorithm, "ProcessingPal," and a
+proprietary one, "CREA," achieved accuracies of 91% and 86%
+respectively. These algorithms, evaluated against self-reported
+measures[@carlson2021], produced F1 scores as high as 95% and 96%. These
+figures are consistent with the performance of our sequential models,
+which also achieved F1 scores and accuracy scores exceeding 95% in
+identifying in-bed time. In our study, in-bed time is equated with SPT.
+All models, with the exception of XGBoost, underestimated SPT. The
+biLSTM model showed the greatest underestimation, with a bias of -36
+minutes, reflecting trends observed in previous research. Winkler et al.
+developed an algorithm that, despite a strong correlation (Pearson
+correlation coefficient = .67) between their algorithmic results and
+diary-recorded waking times, overestimated waking wear time by more than
+30 minutes, resulting in an underestimation of in-bed
+time[@winkler2016]. This trend was further confirmed when Inan-Eroglu et
+al. examined Winkler et al.'s algorithm, revealing a underestimation of
+9.8 minutes in bed time compared to self-reported
+measures[@inan-eroglu2021]. In contrast, a study by van der Berg et al.
+reported a slight underestimation of in-bed time. They employed a unique
+approach with their algorithm, which relied on quantifying the number
+and duration of sedentary periods to determine time in bed, and active
+periods (standing or stepping) to identify wake times[@vanderberg2016].
+Finally. it is important to note that high predictive performance in
+determining in-bed time does not necessarily translate to accurate
+predictions of broader sleep quality parameters. The crucial task of
+detecting awake periods during in-bed time, a key factor in assessing
+sleep quality, may not be effectively captured by in-bed time
+predictions alone. Indeed, underestimating in-bed time could result in
+overestimating waking time during in-bed time. Furthermore, the
+distinction between actual sleep and time spent in bed, often overlooked
+but vital in sleep research, is critical for a comprehensive
+understanding of sleep quality.
+
+To the best of our knowledge, Johansson and
+colleagues[@johansson_development_2023] are the only researchers who
+have reported epoch-to-epoch performance metrics for sleep scoring using
+thigh-worn accelerometers, beyond just "waking time" and "in-bed time."
+They achieved a mean sensitivity of 0.84, specificity of 0.55, and
+accuracy of 0.80, using a single-night evaluation dataset of 71
+subjects. Despite our models achieving a sensitivity above 97%, they,
+like Johansson et al.'s algorithm, struggled with detecting in-bed awake
+epochs. This is reflected in the low specificity scores, ranging from
+54.7% to 76.4%, reported in our study. The challenge of low specificity
+is not unique to methods using thigh-worn devices. Conley et al.'s
+meta-analysis[@conley2019] reported similar findings when estimating
+sleep using wrist-worn accelerometers among healthy adults, with a mean
+sensitivity, accuracy, and specificity of 0.89, 0.88, and 0.53,
+respectively. These figures align with our results regarding in-bed
+time. Patterson and colleagues[@patterson_40_2023] recently summarized
+the performance of various heuristic algorithms, machine learning, and
+deep learning models used to predict sleep. They found the mean
+sensitivity and specificity to be 93% (SD = 2.8) and 60% (SD = 11.1)
+respectively. These findings underscore the challenge of automating the
+detection of in-bed awake periods. Interestingly, despite low
+specificity values for most of our models and configurations, we
+observed an overestimation of LPS and WASO, contrasting with most
+previous research[@conley2019; @palotti2019]. This overestimation of
+wake epochs is evident from the low NPV scores, indicating that only a
+small proportion of the wake predictions are actually correct. This
+discrepancy may be driven by the SMOTE process used to balance the
+dataset. If the synthetic "wake" examples created by SMOTE are not
+representative of the true "wake" data, the models might learn to
+incorrectly classify certain "sleep" epochs as "wake". This could lead
+to an overestimation of LPS and WASO, as the models are incorrectly
+identifying more periods of wakefulness during the sleep period.
+
+often times imbalanced datasets only containing night recordings
+
+The sleep quality parameter that our models consistently corresponded
+most with the ZM derived sleep quality parameters were SPT and TST
+
+something on outliers...maybe related to incl 180 degrees problem.
+
+single night vs multiple nights/days
+
+It is unclear whether Johansson et al.[@johansson_development_2023]\]
+provide a method for detecting the in-bed time which hinders its
+applicability for researchers that want to ...
 
@@
 
-It is unclear whether Johansson et al.[@johansson_development_2023]\]
-provide a method for detecting the in-bed time...
+strengths and limitations
+
+ZM as gold-standard? Without a gold standard, both methods can both
+contribute to disagreement.
+
+external validation dataset/generalization
+
+expand the feature set (Palotti)
 

### 2023-07-11 manuscript/supp_mat/supp.qmd
biLSTM performance as two binary.

@@
 
-![all ba and scatterplots](../visuals/all_ba_scatterplots.pdf)
+![Decision tree, raw](../visuals/raw_decision_tree_ba_cor.pdf){#tbl-raw_dc}
+![Decision tree, median_5](../visuals/median_5_decision_tree_ba_cor.pdf){#tbl-m5_dc}
+![Decision tree, median_10](../visuals/median_10_decision_tree_ba_cor.pdf){#tbl-m10_dc}
+
+![Logistic regression, raw](../visuals/raw_logistic_regression_ba_cor.pdf)
+![Logistic regression, median_5](../visuals/median_5_logistic_regression_ba_cor.pdf)
+![Logistic regression, median_10](../visuals/median_10_logistic_regression_ba_cor.pdf)
+
+![Neural_network, raw](../visuals/raw_neural_network_ba_cor.pdf)
+![Neural_network, median_5](../visuals/median_5_neural_network_ba_cor.pdf)
+![Neural_network, median_10](../visuals/median_10_neural_network_ba_cor.pdf)
+
+![XGBoost, raw](../visuals/raw_xgboost_ba_cor.pdf)
+![XGBoost, median_5](../visuals/median_5_xgboost_ba_cor.pdf)
+![XGBoost, median_10](../visuals/median_10_xgboost_ba_cor.pdf)
+
+![biLSTM, raw](../visuals/raw_biLSTM_ba_cor.pdf)
+![biLSTM, median_5](../visuals/median_5_biLSTM_ba_cor.pdf)
+![biLSTM, median_10](../visuals/median_10_biLSTM_ba_cor.pdf)

### 2023-07-11 code/evaluate_models/get_biLSTM_performance.R

@@ -39,30 +39,76 @@ test_preds <-
     across(c(
       score_simple, score_simple_median_5, score_simple_median_10,
       matches("predicted|probability")
-    ), as_factor)
+    ), as_factor),
+    in_bed_raw = if_else(score_simple %in% c(0, 2), 1, 0),
+    pred_in_bed_raw = if_else(predicted_class_raw %in% c(0, 2), 1, 0),
+    sleep_raw = if_else(score_simple == 2 & in_bed_raw == 1, 1, 0),
+    pred_sleep_raw = if_else(predicted_class_raw == 2 & pred_in_bed_raw == 1, 1, 0),
+    
+    in_bed_median_5 = if_else(score_simple %in% c(0, 2), 1, 0),
+    pred_in_bed_median_5 = if_else(predicted_class_median_5 %in% c(0, 2), 1, 0),
+    sleep_median_5 = if_else(score_simple == 2 & in_bed_median_5 == 1, 1, 0),
+    pred_sleep_median_5 = if_else(predicted_class_median_5 == 2 & pred_in_bed_median_5 == 1, 1, 0),
+    
+    in_bed_median_10 = if_else(score_simple %in% c(0, 2), 1, 0),
+    pred_in_bed_median_10 = if_else(predicted_class_median_10 %in% c(0, 2), 1, 0),
+    sleep_median_10 = if_else(score_simple == 2 & in_bed_median_10 == 1, 1, 0),
+    pred_sleep_median_10 = if_else(predicted_class_median_10 == 2 & pred_in_bed_median_10 == 1, 1, 0),
+    across(in_bed_raw:pred_sleep_median_10, as_factor)
   )
-
 
 # write_parquet(test_preds, "data/data_for_modelling/lstm/predictions/lstm_multiclass_preds.parquet")
 
 
-my_metrics <- metric_set(f_meas, precision, npv, sensitivity, specificity)
+my_metrics_in_bed <- metric_set(f_meas, precision, accuracy, sensitivity, specificity)
+my_metrics_sleep <- metric_set(f_meas, precision, npv, sensitivity, specificity)
 
-get_metrics <- function(data, truth, estimate) {
-  data %>%
-    my_metrics(truth = {{ truth }}, estimate = {{ estimate }}, 
-               event_level = "second", estimator = "macro")
+get_metrics <- function(data, truth, estimate, filter = FALSE, filter_var) {
+  if (filter) {
+    data %>%
+      filter(!!sym(filter_var) == 1) %>%
[…]

### 2023-07-11 manuscript/code/plot_ba_and_scatterplots.R

@@ -78,10 +78,11 @@ plot_ba_cor <- function(model, variable, type) {
     lstm_stats
   } else {
     ml_stats %>%
-      filter(model == {{ model }} & type == {{ type }} & 
-               !!sym(paste0("zm_", variable)) >= 0 & !!sym(variable) >= 0 
-             # & abs(!!sym(paste0("diff_", variable))) / max(abs(!!sym(paste0("diff_", variable))))  < .1
-             )
+      filter(
+        model == {{ model }} & type == {{ type }} &
+          !!sym(paste0("zm_", variable)) >= 0 & !!sym(variable) >= 0
+        # & abs(!!sym(paste0("diff_", variable))) / max(abs(!!sym(paste0("diff_", variable))))  < .1
+      )
   }
 
   x_coords <- stats_data %>% pull(!!sym(variable))
@@ -115,15 +116,29 @@ plot_ba_cor <- function(model, variable, type) {
     geom_hline(data = bias_data, aes(yintercept = loa_upper), color = "grey25", lty = 2) +
     geom_point(
       data = stats_data, aes(x = !!sym(paste0("avg_", variable)), y = !!sym(paste0("diff_", variable))),
-      color = "grey80", fill = "grey50", shape = 21, size = 2.5, stroke = .2, alpha = .7
+      color = "grey20", fill = "steelblue", shape = 21, size = 2.5, stroke = .2, alpha = .7
     ) +
     labs(
-      x = paste("Mean of Model and ZM", str_to_upper(variable), 
-                switch(variable, "spt" = "(hrs)", "tst" = "(hrs)", "se" = "(%)", 
-                "lps" = "(min)", "waso" = "(min)")),
-      y = paste("Difference between\nModel and ZM", str_to_upper(variable), 
-                switch(variable, "spt" = "(hrs)", "tst" = "(hrs)", "se" = "(%)", 
-                "lps" = "(min)", "waso" = "(min)"))
+      x = paste(
+        "Mean of Model and ZM", str_to_upper(variable),
+        switch(variable,
+          "spt" = "(hrs)",
+          "tst" = "(hrs)",
+          "se" = "(%)",
+          "lps" = "(min)",
+          "waso" = "(min)"
+        )
+      ),
+      y = paste(
[…]

### 2023-07-11 manuscript/code/plot_confusion_matrices.R

@@ -66,7 +66,7 @@ plots <-
   map(truth_estimate_pairs, ~ create_confusion_matrix_plot(data, .x$truth, .x$estimate))
 
 # Function to create confusion matrix and plot for the biLSTM
-create_confusion_matrix_plot <- function(target, preds) {
+create_confusion_matrix_plot_lstm <- function(target, preds) {
   table(target = target, preds = preds) %>%
     as_tibble() %>%
     mutate(
@@ -105,21 +105,21 @@ lstm_preds <-
 
 # Generate confusion matrix plots
 conf_mat_lstm_raw <-
-  create_confusion_matrix_plot(lstm_preds$score_simple, lstm_preds$predicted_class_raw)
+  create_confusion_matrix_plot_lstm(lstm_preds$score_simple, lstm_preds$predicted_class_raw)
 conf_mat_lstm_median5 <-
-  create_confusion_matrix_plot(lstm_preds$score_simple_median_5, lstm_preds$predicted_class_median_5)
+  create_confusion_matrix_plot_lstm(lstm_preds$score_simple_median_5, lstm_preds$predicted_class_median_5)
 conf_mat_lstm_median10 <-
-  create_confusion_matrix_plot(lstm_preds$score_simple_median_10, lstm_preds$predicted_class_median_10)
+  create_confusion_matrix_plot_lstm(lstm_preds$score_simple_median_10, lstm_preds$predicted_class_median_10)
 
 
 raw <-
   wrap_plots(plots[[1]][[1]], plots[[1]][[2]], plots[[1]][[3]], plots[[1]][[4]], conf_mat_lstm_raw, nrow = 1) + plot_layout(widths = c(1, 1, 1, 1, 2), heights = c(1, 1, 1, 1, 2)) & plot_annotation(tag_levels = "i", title = "Raw")
 
 median5 <-
-  wrap_plots(plots[[2]][[1]], plots[[2]][[2]], plots[[2]][[3]], plots[[2]][[4]], conf_mat_lstm_raw, nrow = 1) + plot_layout(widths = c(1, 1, 1, 1, 2), heights = c(1, 1, 1, 1, 2)) & plot_annotation(title = "5-Min Median")
[…]

### 2023-07-11 manuscript/code/table_sleep_performance.R

@@ -8,6 +8,9 @@ library(showtext)
 
 all_metrics_30 <-
   read_csv(here::here("data/processed/all_metrics_sleep_30.csv")) %>% 
+  bind_rows(
+  read_csv(here::here("data/processed/biLSTM_sleep_performance_metrics.csv")) %>% 
+  rename(group = type)) %>% 
   mutate(
     group = case_when(group == "raw" ~ "Raw ZM Predictions",
                       group == "median_5" ~ "5-Min Median",

### 2023-07-12 manuscript/manuscript.qmd
van hees spt chatter

@@
   elsevier-pdf:
+    pdf-engine: lualatex
     keep-tex: false
@@
     fig-pos: 'b'
+  simple-article-pdf:
+    indent: false
+    classoption: 
+     - 9pt
+     - lucida
+     - twocolumn
+     - twoside
   docx:
@@
 sleep time and their derived sleep quality parameters. Furthermore, we
-trained and evaluated the models using raw and median-filtered
-gold-standard predictions from the ZM. In general, all sequential models
+trained and evaluated the models using raw and median-filtered gold
+standard predictions from the ZM. In general, all sequential models
 performed well at predicting in-bed time. More challenging was it to
@@
 filters improved the performance metrics of all models, with the XGBoost
-model consistently leading. The filters increased total sleep time and
-sleep efficiency, while reducing wake after sleep onset and the number
-of awakenings. The XGBoost model also showed the least bias and highest
-correlation with ZM across sleep quality parameters.
+model consistently leading. The median filters increased total sleep
+time and sleep efficiency, while reducing wake after sleep onset and the
+number of awakenings. The XGBoost model also showed the least bias and
+highest correlation with ZM across sleep quality parameters.
 
@@
 54.7% to 76.4%, reported in our study. The challenge of low specificity
-is not unique to methods using thigh-worn devices. Conley et al.'s
-meta-analysis[@conley2019] reported similar findings when estimating
-sleep using wrist-worn accelerometers among healthy adults, with a mean
-sensitivity, accuracy, and specificity of 0.89, 0.88, and 0.53,
-respectively. These figures align with our results regarding in-bed
-time. Patterson and colleagues[@patterson_40_2023] recently summarized
-the performance of various heuristic algorithms, machine learning, and
-deep learning models used to predict sleep. They found the mean
-sensitivity and specificity to be 93% (SD = 2.8) and 60% (SD = 11.1)
-respectively. These findings underscore the challenge of automating the
-detection of in-bed awake periods. Interestingly, despite low
-specificity values for most of our models and configurations, we
-observed an overestimation of LPS and WASO, contrasting with most
-previous research[@conley2019; @palotti2019]. This overestimation of
-wake epochs is evident from the low NPV scores, indicating that only a
-small proportion of the wake predictions are actually correct. This
-discrepancy may be driven by the SMOTE process used to balance the
-dataset. If the synthetic "wake" examples created by SMOTE are not
-representative of the true "wake" data, the models might learn to
-incorrectly classify certain "sleep" epochs as "wake". This could lead
-to an overestimation of LPS and WASO, as the models are incorrectly
-identifying more periods of wakefulness during the sleep period.
+is not unique to methods using data collected from thigh-worn devices.
+Conley et al.'s meta-analysis[@conley2019] reported similar findings
+when estimating sleep using wrist-worn accelerometers among healthy
+adults, with a mean sensitivity, accuracy, and specificity of 0.89,
+0.88, and 0.53, respectively. Furthermore, Patterson and
+colleagues[@patterson_40_2023] recently summarized the performance of
+various heuristic algorithms, machine learning, and deep learning models
+used to predict sleep. They found the mean sensitivity and specificity
+to be 93% (SD = 2.8) and 60% (SD = 11.1) respectively. These findings
+underscore the challenge of automating the detection of in-bed awake
+periods. Interestingly, despite low specificity values for most of our
+models and configurations, we observed an overestimation of LPS and
+WASO, contrasting with most previous research[@conley2019;
+@palotti2019]. This overestimation of wake epochs is evident from the
+low NPV scores, indicating that only a small proportion of the wake
+predictions are actually correct. This discrepancy may be driven by the
+SMOTE process used to balance the dataset. If the synthetic "wake"
+samples created by SMOTE are not representative of the true "wake" data,
+the models might learn to incorrectly classify certain "sleep" epochs as
+"wake". This could lead to an overestimation of LPS and WASO, as the
+models are incorrectly identifying more periods of wakefulness during
+the sleep period.
 
-often times imbalanced datasets only containing night recordings
+The use of the SMOTE technique likely improved the performance of our
+models by addressing the class imbalance in our data. However, this
+technique also introduced synthetic"wake" samples that may not be fully
+representative of true wake data. This could potentially lead some
+models to overestimate the wake class. Interestingly, the biLSTM model,
+which was not trained on SMOTE-processed data, was the only one to
+overestimate TST and SE. On the other hand, the XGBoost model, which was
+trained on data subjected to the SMOTE process, was able to handle the
+synthetic "wake" samples better than the other models, and it did not
+overestimate TST to the same degree. The Bland-Altman statistics for the
+XGBoost model trained on the 5-minute median filtered ZM predictions
+showed a mean difference of -7 minutes for TST and -1.1% for SE, with
+limits of agreement ranging from -95.5 to 81.4 minutes and from -15.6%
+to 13.3% respectively. This suggests that the XGBoost model was able to
+maintain a balance between sensitivity and specificity, and it was not
+overly influenced by the synthetic "wake" samples. The XGBoost model's
+success with the SMOTE dataset may be due to its ability to handle
+non-representative synthetic samples. XGBoost's gradient boosting
+mechanism allows it to iteratively learn from the errors of previous
+models, which can help it to better distinguish between true wake data
+and synthetic wake samples created by SMOTE. This iterative learning
+process could make XGBoost more robust to the inaccuracies introduced by
+the synthetic samples, leading to better overall performance.
 
-The sleep quality parameter that our models consistently corresponded
-most with the ZM derived sleep quality parameters were SPT and TST
+In our evaluation of sleep quality parameters, we found that Latency to
+Persistent Sleep (LPS) had the largest mean error relative to absolute
+time allocated to LPS. This suggests that the initial epochs of Sleep
+Period Time (SPT) are particularly challenging to classify correctly.
+This is also supported by the poor Pearson correlations between LPS
+derived from model predictions and the ZM. The XGBoost model, which was
+the best performer among all models, overestimated LPS by an average of
+26.4 minutes for models trained on raw ZM predictions, 28.5 minutes for
+models trained on 5-minute filtered ZM predictions, and 34.5 minutes for
+models trained on 10-minute filtered ZM predictions. This level of
+discrepancy is comparable to the mean error of sleep latency of 23
+minutes reported by Johansson et al.[@johansson_development_2023].
+Johansson et al. suggest that the discrepancy with the gold standard is
+likely due to the multifaceted nature of the sleep state, which is a
+complex physiological process. Short awakenings or sleep episodes may
+not necessarily correspond to noticeable changes in thigh movement,
+making them difficult to detect and accurately classify. These results
+align with several methods for wrist-worn devices reviewed by Conley and
+colleagues[@conley2019]. They reported correlations between
+accelerometer and PSG sleep onset latency (equivalent to LPS) from 10
+studies with a mean correlation of 0.2 (ranging from -0.69 to 0.69),
+indicating the inherent difficulty in estimating this parameter using
+accelerometry alone.
 
-something on outliers...maybe related to incl 180 degrees problem.
+In this study, we used the ZM as the reference method, rather than PSG,
+which is considered the gold standard for sleep measurement. This choice
+may contribute to discrepancies between our models and the ZM, as
+without a true gold standard, it's difficult to determine the source of
+disagreement. However, we believe that the use of ZM, which allows for
+multiple consecutive nights of recording, is valuable. This approach
+captures intra-individual variances in sleep, which is impractical with
+PSG. It also enabled us to include more nights in our study typically
+compared to those relying on PSG. For instance, the widely used
+Newcastle dataset[@hees2015] only contains data from 28 participants.
+However, upon examining the ZM outputs, we found that the raw
+predictions were not optimal for developing machine learning models due
+to a seemingly low signal-to-noise ratio (see @fig-zm-median). The ZM
+itself mitigates this issue by applying certain filtering processes when
+generating sleep quality parameters. For example, epochs contributing to
+WASO must be in contiguous epochs of 3, and sleep only counts towards
+sleep quality parameters if 10 out of 12 minutes are scored as sleep. To
+improve the prospect of our machine learning algorithms, we applied
+median filters to the ZM raw predictions. This did in fact alter the
+derived sleep quality parameters. Notably, the mean WASO decreased from
+39 minutes in the raw predictions to 30.6 minutes in the 5-minute median
+filtered predictions, and further decreased to 22.3 minutes in the
+10-minute filtered predictions. The application of 5-minute and
+10-minute median filters also led to increases in TST, SE, and LPS. This
+suggests that the filters may categorize some instances of wakefulness
+as sleep and smooth out brief awakenings. Despite these changes, the
+overall sleep quality profile derived from the median-filtered
+predictions is still comparable to that from the raw predictions,
+justifying our approach.
 
-single night vs multiple nights/days
+Our study's XGBoost model demonstrated relatively narrower limits of
+agreements (LOAs) for TST, SE, and WASO, with ranges of -95.5 to 81.4
+min, -15.6 to 13.3%, and -83.4 to 81.7 min, respectively when compared
+with other models such as the Van Hees algorithm[@hees2015], Oakley rsc
+(rescored)[@palotti2019], and LSTM-50[@palotti2019] evaluated in the
+Patterson et al. study[@patterson_40_2023]. Furthermore, comparing the
+LOAs between our XGBoost model and the algorithm developed for
+thigh-worn devices by Johansson et al. study[@johansson2022], our
+XGBoost model showed narrower LOAs for TST , SE, LPS , and WASO, but not
+SPT. Generally, all methods, both from this study and from the reviewed
+literature, exibit wide LOAs sugesting that there is high variability in
+the derevide sleep quality parameters. These findings imply that the
+current methods, are only reasonbly reliable for assessing sleep quality
+parameters at a group level. However, caution should be exercised when
+applying the models and methods to individual-level sleep assessments.
+Therefore, further improvements and refinements are needed to enhance
+the precision and reliability of these models for individual sleep
+assessments.
 
-It is unclear whether Johansson et al.[@johansson_development_2023]\]
+Typically, sleep detection methods are applied in two contexts: either
+to night recordings or to 24-hour recordings. In night recordings, it is
+possible to derive sleep quality parameters like SE and LPS because the
+SPT is already known because it is inferred from the length of the
+recording[@conley2019; @patterson_40_2023]. On the other hand, when
+sleep detection methods are applied to 24-hour recordings, most methods
+do not have the ability to infer the SPT with sleep
+diaries[@girschik2012]. Consequently, these methods are unable to
+generate certain sleep quality parameters that rely on the
+SPT[@doherty2017; @anderson2014]. We have attempted to circumvent this
+issue by employing models able to distinguish between in-bed awake time
+and in-bed asleep time over a 24h recording making our models capable of
+estimating all commonly used sleep quality parameters. Van Hees et
+al.[@van_hees_estimating_2018] have developed an algorithm to identify
+SPT from wrist-worn devices which was recently validated by Plekhanova
+and colleagues[@plekhanova2023]. which in combination with other
+algorithms or models on said SPT can then be used infer further sleep
+quality parameters.
+
+Strengths and limitations It is not always clear the methods for
+detecting sleep from accelerometer data is
+
+It is unclear whether Johansson et al.[@johansson_development_2023]
 provide a method for detecting the in-bed time which hinders its
 applicability for researchers that want to ...
-
-A significant finding was the impact of median filters on the sleep
-metrics. Implementing 5-minute and 10-minute median filters led to
-notable changes in key sleep parameters. In particular, the Total Sleep
-Time (TST), Sleep Efficiency (SE), and Latency to Persistent Sleep (LPS)
-increased, suggesting that the filters may categorize some instances of
-wakefulness as sleep and smooth out brief awakenings. This underscores
-the potential of median filters in refining sleep data and mitigating
-the potential misclassification of sleep states.
-
-Interestingly, the most marked change was observed in Wake After Sleep
-Onset (WASO), with a substantial reduction in the filtered data. This
-reduction, coupled with the decrease in the number of awakenings,
-implies a significant role of median filters in sleep-wake cycle
-characterization, potentially enhancing the reliability of sleep
-assessment.
-
-The machine learning models demonstrated high overall performance on an
-epoch-to-epoch basis, with the XGBoost model marginally outperforming
-the others. However, the models had more varied performance in
-classifying sleep conditions, as evidenced by the sleep performance
-metrics. The relatively lower Specificity values across all models
-suggest a common challenge in correctly classifying negative instances.
-
-In the precision-recall and ROC curve analyses, the Decision Tree model
-showed superior precision-recall AUC, indicating strong predictive
-accuracy. In contrast, the XGBoost model consistently demonstrated a
-high ROC AUC, indicating a robust ability to differentiate between
-classes. However, it's worth noting that the Neural Network model
-generally showed weaker performance in these areas.
-
-The three-class biLSTM multiclassifier showed overall good performance,
-further supporting the potential of machine learning in sleep study.
-However, the absence of Specificity values limits a complete assessment
-of its performance, calling for further exploration in future studies.
-
-strengths and limitations
-
-ZM as gold-standard? Without a gold standard, both methods can both
-contribute to disagreement.
 
@@
 
-In conclusion, this study provides compelling evidence of the utility of
-machine learning models and median filters in enhancing the precision
-and reliability of pediatric sleep assessment. However, more research is
-needed to overcome the challenge of correctly classifying negative
-instances and to fully understand the performance of these models across
-all classes and conditions. The findings underscore the potential of
-machine learning in advancing sleep medicine, paving the way for more
-accurate, reliable, and personalized sleep assessment in pediatric
-populations.
+In conclusion, our study has advanced the methods for estimating sleep
+from thigh-worn accelerometers by evaluating various models for
+predicting in-bed and sleep time. Our results indicate that all
+sequential models and the multiclass model performed well at predicting
+in-bed time. However, all models showed difficulty in correctly
+classifying awake epochs during time in bed, highlighting the challenge
+of automating the detection of in-bed awake periods. The application of
+median filters improved the performance metrics of all models,
+suggesting that this approach may be beneficial in future studies.
+Despite these promising results, our findings also underscore the need
+for caution when applying these models to individual-level sleep
+assessments due to the high variability in derived sleep quality
+parameters. Future research should focus on refining these models to
+enhance their precision and reliability for individual sleep
+assessments.
+
+Our open-source algorithm and the inferred sleep parameters will open
+the door to future studies on sleep and sleep architecture using
+large-scale accelerometer databases.
 

### 2023-07-12 manuscript/manuscript.tex
van hees spt chatter

@@
 \usepackage{longtable}
+\usepackage{colortbl}
 \usepackage{lineno}\linenumbers
@@
 \hypersetup{
-  pdftitle={Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device},
+  pdftitle={Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking},
   pdfauthor={Esben Høegholm Lykke; Jan Christian Brønd},
@@
 \begin{frontmatter}
-\title{Estimating Sleep Quality Metrics Using Free-Living Accelerometer
-Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep
-Tracking Device}
+\title{Improving Sleep Quality Estimation: A Comparative Study of
+Machine Learning and Deep Learning Techniques Utilizing Free-Living
+Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking}
 \author[1]{Esben Høegholm Lykke%
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, interior hidden, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, enhanced, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-In this study, we employed a biLSTM, a multiclass classifier, to predict
-three distinct states: out-of-bed-awake, in-bed-awake, and
-in-bed-asleep. The architecture of the biLSTM was set up with four
-layers, each equipped with 128 hidden units. This configuration was
-intentionally chosen to balance between model complexity and training
-efficiency: it provided the depth necessary for learning intricate
-patterns while remaining feasible for timely training. The bidirectional
-design of the LSTM served to enhance data interpretation and mitigate
-overfitting by doubling the hidden units at each time step. For input,
-we used tensors shaped as sequences, with each sequence spanning 10
-minutes and a step size of one.This approach follows in the footsteps of
-previous studies that utilized LSTM models for sleep detection. These
-studies showcased the promising potential of LSTMs in capturing complex
-temporal patterns. Particularly, the works of Sano et al.
-\citep{sano2019} and Chen et al. \citep{chen2021} demonstrated the
-effectiveness of LSTM models in improving sleep detection using
-accelerometer data, underscoring the value of this modeling approach.
+We also employed a biLSTM, a multiclass classifier, to predict three
+distinct states: out-of-bed-awake, in-bed-awake, and in-bed-asleep. The
+architecture of the biLSTM was set up with four layers, each equipped
+with 128 hidden units. This configuration was intentionally chosen to
+balance between model complexity and training efficiency: it provided
+the depth necessary for learning intricate patterns while remaining
+feasible for timely training. The bidirectional design of the LSTM
+served to enhance data interpretation and mitigate overfitting by
+doubling the hidden units at each time step. For input, we used tensors
+shaped as sequences, with each sequence spanning 10 minutes and a step
+size of one.This approach follows in the footsteps of previous studies
+that utilized LSTM models for sleep detection. These studies showcased
+the promising potential of LSTMs in capturing complex temporal patterns.
+Particularly, the works of Sano et al. \citep{sano2019} and Chen et al.
+\citep{chen2021} demonstrated the effectiveness of LSTM models in
+improving sleep detection using accelerometer data, underscoring the
+value of this modeling approach.
 
@@
 procedure to address this extra source of variability. We calculated the
-mean difference (bias) and defined the limits of agreement as the mean
-difference plus or minus 1.96 times the standard deviation of these
-differences. To ensure our measurements were robust and accounted for
-intra-subject variability, we estimated the 95\% confidence intervals
-for both the bias and the limits of agreement using a bias-corrected and
-accelerated bootstrap method, utilizing 10,000 bootstrap replicates. The
-sleep quality parameteres included the following:
+mean difference (bias) and defined the LOA as the mean difference plus
+or minus 1.96 times the standard deviation of these differences. To
+ensure our measurements were robust and accounted for intra-subject
+variability, we estimated the 95\% confidence intervals for both the
+bias and the LOA using a bias-corrected and accelerated bootstrap
+method, utilizing 10,000 bootstrap replicates. The sleep quality
+parameteres included are defined as follews in accordance with the ZM
+definitions:
 
@@
   it takes to transition from wakefulness to sustained sleep. It is
-  calculated as the time from the beginning of the ZM recording until a
-  period when 10 out of 12 minutes are scored as sleep.
+  calculated as the time from the beginning of the ZM recording until
+  the first period when 10 out of 12 minutes are scored as sleep.
 \item
@@
 \begin{longtable}{lrrrrr}
-\caption{\label{tbl-in_bed_performance}In-Bed Performance Metrics }\tabularnewline
+\caption{\label{tbl-in_bed_performance}In-bed performance metrics }\tabularnewline
 
@@
 XGBoost & $95.4$ & $96.1$ & $95.8$ & $94.9$ & $96.2$ \\ 
+biLSTM & $95.2$ & $95.3$ & $95.3$ & $95.1$ & $95.3$ \\ 
 \bottomrule
@@
 
-Table~\ref{tbl-sleep_performance} details the performance of the second
-sequential models on raw and median-filtered (5 and 10 minute) ZM
-predictions. These are the predictions of the sequential models' ability
-to predict asleep/awake on only the extracted in-bed time. The F1
-scores, which are unweighted macro averages, for raw ZM predictions
-range from 71.05\% to 76.18\%. The models perform comparably, but the
-low specificity values (62.84\% to 70.93\%) suggest difficulty in
-correctly classifying awake epochs. Applying 5-minute median filtering
-improves the performance metrics. The XGBoost model tops the charts with
-an F1 score of 79.22\% and NPV of 74.00\%. However, specificity still
-remains low, with values between 54.68\% and 74.84\% across all models.
-With 10-minute median filtering, the metrics improve further. The
-XGBoost model still leads with an F1 score of 80.87\% and an NPV of
-75.76\%. But, specificity remains low, ranging from 57.47\% to 76.35\%
-across all models.
+Table~\ref{tbl-sleep_performance} details the performance of all
+sequential model types on raw and median-filtered (5 and 10 minute) ZM
+predictions for sleep/wake classification. For raw ZM predictions, the
+F1 scores, which are unweighted macro averages, range from 65.6\%
+(biLSTM) to 76.2\% (XGBoost). All models perform comparably, but the low
+specificity values (62.5\% to 70.9\%) suggest difficulty in correctly
+classifying awake epochs. Applying a 5-minute median filter improves the
+performance metrics. The XGBoost model tops the charts with an F1 score
+of 79.2\% and NPV of 74.0\%. However, specificity still remains low,
+with values between 54.7\% (XGBoost) and 74.8\% (Logistic Regression)
+across all models. With a 10-minute median filter, the metrics improve
+further. The XGBoost model still leads with an F1 score of 80.9\% and an
+NPV of 75.9\%. But, specificity remains low, ranging from 57.5\%
+(Decision Tree) to 76.4\% (Logistic Regression) across all models.
 
@@
 \begin{longtable}{lrrrrr}
-\caption{\label{tbl-sleep_performance}Sleep Performance Metrics }\tabularnewline
+\caption{\label{tbl-sleep_performance}Performance of the sleep/wake classification of the sequential models. }\tabularnewline
 
@@
 XGBoost & $76.2$ & $92.8$ & $58.0$ & $91.3$ & $62.8$ \\ 
+biLSTM & $65.6$ & $80.6$ & $80.6$ & $62.5$ & $62.5$ \\ 
 \midrule
@@
 XGBoost & $79.2$ & $93.9$ & $74.0$ & $97.3$ & $54.7$ \\ 
+biLSTM & $70.3$ & $84.6$ & $84.6$ & $66.2$ & $66.2$ \\ 
 \midrule
@@
 XGBoost & $80.9$ & $94.9$ & $75.8$ & $97.7$ & $57.6$ \\ 
+biLSTM & $70.9$ & $75.1$ & $75.1$ & $68.5$ & $68.5$ \\ 
 \bottomrule
@@
 
-The complete set of confusion matrices generated from data both
-containing the out-of-bed and in-bed time are presented in
+A complete set of confusion matrices generated from data both containing
+the out-of-bed and in-bed time are presented in
 Figure~\ref{fig-conf_mat}. The figure shows favorable epoch-to-epoch
@@
 which cannot be deduced from the confusion matrices from the sequential
-models. \textbf{{[}fejl i biLSTM confmat. Alle mats er ens{]}}
+models.
 
@@
 percentage (prediction). i) decision tree, ii) logistic regression, iii)
-feed-forward neural net, iv) xgboost, and v) biLSTM.}
+feed-forward neural net, iv) XGBoost, and v) biLSTM.}
 
 \end{figure}
-
-Table~\ref{tbl-biLSTM_performance} presents the performance of the
-three-class biLSTM multiclassifier on raw and median-filtered (5-minute
-and 10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores
-ranging from 71.36\% to 76.04\%, indicating overall good performance.
-Applying 5-minute median filtering improves the metrics further,
-resulting in F1 Scores ranging from 75.99\% to 78.53\%, demonstrating
-enhanced precision and sensitivity. \textbf{{[}Overvej flere metrics.
-Evt. per class\ldots{} pt er det lidt svært at sammenligne på tværs af
-lstm og andre modeller ud over F1.{]}} With 10-minute median filtering,
-F1 Scores range from 73.45\% to 73.93\%, maintaining good performance
-but with limited information on Specificity. Further analysis is
-required to assess performance across all classes.
-
-\hypertarget{tbl-biLSTM_performance}{}
-\begin{longtable}{cccccc}
-\caption{\label{tbl-biLSTM_performance}Performance of the three-class biLSTM multicclassifier. }\tabularnewline
-
-\toprule
- & F1 Score & Precision & NPV & Sensitivity & Specificity \\ 
-\midrule
-Raw ZM Predictions & $71.4$ & $76.0$ & $95.5$ & $70.4$ & $95.0$ \\ 
-5-Min Median & $76.0$ & $78.5$ & $96.4$ & $74.6$ & $96.2$ \\ 
-10-Min Median & $73.4$ & $73.1$ & $95.6$ & $73.9$ & $95.7$ \\ 
-\bottomrule
-\end{longtable}
 
@@
 logistic regression models, but with a higher overestimation in WASO. On
-the other hand, the XGboost model showed least bias among all,
+the other hand, the XGBoost model showed least bias among all,
 especially in its 5-minute median predictions. Considering LOA, the
@@
 predictions. Correlation-wise, the pearson coefficient, revealed that
-the XGboost model consistently had the highest correlation with ZM
+the XGBoost model consistently had the highest correlation with ZM
 across all sleep qualityparameters and filtering methods Notably, the
-XGboost's 5-minute median predictions showed the strongest correlation
+XGBoost's 5-minute median predictions showed the strongest correlation
 (0.66) for TST among all models and filtering techniques.
@@
 \begin{longtable}{lrrrr}
-\caption{\label{tbl-ba_cor}Summary of Bias, Limits of Agreement (LOA), and Pearson Correlation for
+\caption{\label{tbl-ba_cor}Summary of Bias, Limits of Agreement, and Pearson Correlation for
 various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using
@@
 
-Figure~\ref{fig-xgb_ba_cor} shows the agreement of the XGBoost model
-trained on the 5-minute median filtered ZM predictions in estimating
-sleep quality parameters as evaluated 5-minute median-smoothed ZM
-derived sleep quality parameter. As can be seen on from the Bland-Altman
-plots, a large portion of the nights are in almost perfect agreement
-with the ZM and also a portion that can be considered outliers. This is
-true for all sleep quality parameters.
-
-\textbf{{[}Overvej at smide samtlige 75 plots i supp.mat.{]}}
+Figure~\ref{fig-xgb_ba_cor} shows the agreement between the XGBoost
+model, trained on 5-minute median filtered ZM predictions, and the
+5-minute median-smoothed ZM-derived sleep quality parameters. The
+Bland-Altman plot for the SPT reveals a significant level of agreement
+with the ZM, as evidenced by a positive correlation. However, the
+presence of extreme outliers widens the limits of agreement (LOA). The
+scatterplot for SPT also demonstrates a positive trend, indicating a
+moderate linear correlation between the XGBoost model and the ZM-derived
+sleep quality parameters.In terms of TST, perfect agreement is not
+observed for a substantial number of nights, but there is a positive
+correlation between the XGBoost model and ZM-derived sleep quality
+parameters. The bias and LOA for TST are comparable to those observed
+for SPT, indicating a consistent level of agreement between the two
+methods. The scatterplot for TST also shows a slightly higher
+correlation, primarily driven by the absence of extreme
+outliers.Furthermore, the remaining three sleep quality parameters, SE,
+LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST.
+This outcome is expected as achieving 100\% sleep efficiency is
+relatively rare, resulting in less disagreement between the methods as
+values approach the upper limit. However, as sleep efficiency decreases,
+the potential for discrepancies and differing interpretations between
+the methods increases, leading to greater heteroscedasticity. A moderate
+linear correlation is observed between the XGBoost model and ZM-derived
+sleep quality parameters for SE, indicating a positive relationship.
+However, a poor correlation is observed for LPS and WASO, suggesting
+less agreement between the two methods for these parameters. Similar
+plots for all models are available in the supplementary materials.
 
@@
 \caption{\label{fig-xgb_ba_cor}Comparison of sleep quality parameters
-derived from the XGboost model trained on the 5-minute smoothed ZM
+derived from the XGBoost model trained on the 5-minute smoothed ZM
 predictions. The left column displays Bland-Altman plots. Dashed lines
 represent the bias (the average difference between the two measurements)
-and limits of agreement, with the 95\% confidence intervals represented
-as the grayed areas. The right column displays scatter plots of
-XGboost-derived vs ZM-derived sleep quality parameters. The dashed line
-represents the identity line, while the full-drawn line represents the
-best linear fit. Pearson's correlations are annotated in the upper left
-corner.}
+and LOA, with the 95\% confidence intervals represented as the grayed
+areas. The right column displays scatter plots of XGBoost-derived vs
+ZM-derived sleep quality parameters. The dashed line represents the
+identity line, while the full-drawn line represents the best linear fit.
+Pearson's correlations are annotated in the upper left corner.}
 
@@
 
-The present study embarked on an in-depth exploration of the utility and
-comparative performance of different machine learning models in the
-analysis of pediatric sleep data. The results offer promising insights
-into the role of machine learning in enhancing the precision of sleep
-quality metrics.
-
-A significant finding was the impact of median filters on the sleep
-metrics. Implementing 5-minute and 10-minute median filters led to
-notable changes in key sleep parameters. In particular, the Total Sleep
-Time (TST), Sleep Efficiency (SE), and Latency to Persistent Sleep (LPS)
-increased, suggesting that the filters may categorize some instances of
-wakefulness as sleep and smooth out brief awakenings. This underscores
-the potential of median filters in refining sleep data and mitigating
-the potential misclassification of sleep states.
-
-Interestingly, the most marked change was observed in Wake After Sleep
-Onset (WASO), with a substantial reduction in the filtered data. This
-reduction, coupled with the decrease in the number of awakenings,
-implies a significant role of median filters in sleep-wake cycle
-characterization, potentially enhancing the reliability of sleep
-assessment.
-
-The machine learning models demonstrated high overall performance on an
-epoch-to-epoch basis, with the XGBoost model marginally outperforming
-the others. However, the models had more varied performance in
-classifying sleep conditions, as evidenced by the sleep performance
-metrics. The relatively lower Specificity values across all models
-suggest a common challenge in correctly classifying negative instances.
-
-In the precision-recall and ROC curve analyses, the Decision Tree model
-showed superior precision-recall AUC, indicating strong predictive
-accuracy. In contrast, the XGBoost model consistently demonstrated a
-high ROC AUC, indicating a robust ability to differentiate between
-classes. However, it's worth noting that the Neural Network model
-generally showed weaker performance in these areas.
-
-The three-class biLSTM multiclassifier showed overall good performance,
-further supporting the potential of machine learning in sleep study.
-However, the absence of Specificity values limits a complete assessment
-of its performance, calling for further exploration in future studies.
-
-It is unclear whether Johansson et
-al.\citep{johansson_development_2023}{]} provide a method for detecting
-the in-bed time\ldots{}
-
-In conclusion, this study provides compelling evidence of the utility of
-machine learning models and median filters in enhancing the precision
-and reliability of pediatric sleep assessment. However, more research is
-needed to overcome the challenge of correctly classifying negative
-instances and to fully understand the performance of these models across
-all classes and conditions. The findings underscore the potential of
-machine learning in advancing sleep medicine, paving the way for more
-accurate, reliable, and personalized sleep assessment in pediatric
-populations.
+In an effort to mature the methods for estimating sleep from thigh-worn
+accelerometers, we evaluated various models for predicting in-bed and
+sleep time and their derived sleep quality parameters. Furthermore, we
+trained and evaluated the models using raw and median-filtered gold
+standard predictions from the ZM. In general, all sequential models
+performed well at predicting in-bed time. More challenging was it to
+distinguish wake from sleep on the extracted in-bed time, however, the
+performance of the sequential models were enhanced by the application of
+median filters. Moreover, even though the multiclass biLSTM showed good
+performance across F1 score, precision and NPV, the derived sleep
+quality parameters were not on par with the XGBoost model which
+demonstrated the highest performance metrics across all evaluations,
+including epoch-to-epoch prediction and sleep quality parameter
+derivatives. Despite this, all sequential models showed low specificity
+values, indicating difficulty in correctly classifying awake epochs
+during time in bed. The application of 5-minute and 10-minute median
+filters improved the performance metrics of all models, with the XGBoost
+model consistently leading. The median filters increased total sleep
+time and sleep efficiency, while reducing wake after sleep onset and the
+number of awakenings. The XGBoost model also showed the least bias and
+highest correlation with ZM across sleep quality parameters.
+
+Limited research exists regarding the epoch-to-epoch effectiveness of
+classifying in-bed time based on data from thigh-worn accelerometers.
+Nevertheless, Carlson and colleagues provided compelling insights. They
+demonstrated that a third-party algorithm, ``ProcessingPal,'' and a
+proprietary one, ``CREA,'' achieved accuracies of 91\% and 86\%
+respectively. These algorithms, evaluated against self-reported
+measures\citep{carlson2021}, produced F1 scores as high as 95\% and
+96\%. These figures are consistent with the performance of our
+sequential models, which also achieved F1 scores and accuracy scores
+exceeding 95\% in identifying in-bed time. In our study, in-bed time is
+equated with SPT. All models, with the exception of XGBoost,
+underestimated SPT. The biLSTM model showed the greatest
+underestimation, with a bias of -36 minutes, reflecting trends observed
+in previous research. Winkler et al. developed an algorithm that,
+despite a strong correlation (Pearson correlation coefficient = .67)
+between their algorithmic results and diary-recorded waking times,
+overestimated waking wear time by more than 30 minutes, resulting in an
+underestimation of in-bed time\citep{winkler2016}. This trend was
+further confirmed when Inan-Eroglu et al.~examined Winkler et al.'s
+algorithm, revealing a underestimation of 9.8 minutes in bed time
+compared to self-reported measures\citep{inan-eroglu2021}. In contrast,
+a study by van der Berg et al. reported a slight underestimation of
+in-bed time. They employed a unique approach with their algorithm, which
+relied on quantifying the number and duration of sedentary periods to
+determine time in bed, and active periods (standing or stepping) to
+identify wake times\citep{vanderberg2016}. Finally. it is important to
+note that high predictive performance in determining in-bed time does
+not necessarily translate to accurate predictions of broader sleep
+quality parameters. The crucial task of detecting awake periods during
+in-bed time, a key factor in assessing sleep quality, may not be
+effectively captured by in-bed time predictions alone. Indeed,
+underestimating in-bed time could result in overestimating waking time
+during in-bed time. Furthermore, the distinction between actual sleep
+and time spent in bed, often overlooked but vital in sleep research, is
+critical for a comprehensive understanding of sleep quality.
+
+To the best of our knowledge, Johansson and
+colleagues\citep{johansson_development_2023} are the only researchers
+who have reported epoch-to-epoch performance metrics for sleep scoring
+using thigh-worn accelerometers, beyond just ``waking time'' and
+``in-bed time.'' They achieved a mean sensitivity of 0.84, specificity
+of 0.55, and accuracy of 0.80, using a single-night evaluation dataset
+of 71 subjects. Despite our models achieving a sensitivity above 97\%,
+they, like Johansson et al.'s algorithm, struggled with detecting in-bed
+awake epochs. This is reflected in the low specificity scores, ranging
+from 54.7\% to 76.4\%, reported in our study. The challenge of low
+specificity is not unique to methods using data collected from
+thigh-worn devices. Conley et al.'s meta-analysis\citep{conley2019}
+reported similar findings when estimating sleep using wrist-worn
+accelerometers among healthy adults, with a mean sensitivity, accuracy,
+and specificity of 0.89, 0.88, and 0.53, respectively. Furthermore,
+Patterson and colleagues\citep{patterson_40_2023} recently summarized
+the performance of various heuristic algorithms, machine learning, and
+deep learning models used to predict sleep. They found the mean
+sensitivity and specificity to be 93\% (SD = 2.8) and 60\% (SD = 11.1)
+respectively. These findings underscore the challenge of automating the
+detection of in-bed awake periods. Interestingly, despite low
+specificity values for most of our models and configurations, we
+observed an overestimation of LPS and WASO, contrasting with most
+previous research\citep{conley2019, palotti2019}. This overestimation of
+wake epochs is evident from the low NPV scores, indicating that only a
+small proportion of the wake predictions are actually correct. This
+discrepancy may be driven by the SMOTE process used to balance the
+dataset. If the synthetic ``wake'' samples created by SMOTE are not
+representative of the true ``wake'' data, the models might learn to
+incorrectly classify certain ``sleep'' epochs as ``wake''. This could
+lead to an overestimation of LPS and WASO, as the models are incorrectly
+identifying more periods of wakefulness during the sleep period.
+
+The use of the SMOTE technique likely improved the performance of our
+models by addressing the class imbalance in our data. However, this
+technique also introduced synthetic''wake'' samples that may not be
+fully representative of true wake data. This could potentially lead some
+models to overestimate the wake class. Interestingly, the biLSTM model,
+which was not trained on SMOTE-processed data, was the only one to
+overestimate TST and SE. On the other hand, the XGBoost model, which was
+trained on data subjected to the SMOTE process, was able to handle the
+synthetic ``wake'' samples better than the other models, and it did not
+overestimate TST to the same degree. The Bland-Altman statistics for the
+XGBoost model trained on the 5-minute median filtered ZM predictions
+showed a mean difference of -7 minutes for TST and -1.1\% for SE, with
+limits of agreement ranging from -95.5 to 81.4 minutes and from -15.6\%
+to 13.3\% respectively. This suggests that the XGBoost model was able to
+maintain a balance between sensitivity and specificity, and it was not
+overly influenced by the synthetic ``wake'' samples. The XGBoost model's
+success with the SMOTE dataset may be due to its ability to handle
+non-representative synthetic samples. XGBoost's gradient boosting
+mechanism allows it to iteratively learn from the errors of previous
+models, which can help it to better distinguish between true wake data
+and synthetic wake samples created by SMOTE. This iterative learning
+process could make XGBoost more robust to the inaccuracies introduced by
+the synthetic samples, leading to better overall performance.
+
+In our evaluation of sleep quality parameters, we found that Latency to
+Persistent Sleep (LPS) had the largest mean error relative to absolute
+time allocated to LPS. This suggests that the initial epochs of Sleep
+Period Time (SPT) are particularly challenging to classify correctly.
+This is also supported by the poor Pearson correlations between LPS
+derived from model predictions and the ZM. The XGBoost model, which was
+the best performer among all models, overestimated LPS by an average of
+26.4 minutes for models trained on raw ZM predictions, 28.5 minutes for
+models trained on 5-minute filtered ZM predictions, and 34.5 minutes for
+models trained on 10-minute filtered ZM predictions. This level of
+discrepancy is comparable to the mean error of sleep latency of 23
+minutes reported by Johansson et al.\citep{johansson_development_2023}.
+Johansson et al.~suggest that the discrepancy with the gold standard is
+likely due to the multifaceted nature of the sleep state, which is a
+complex physiological process. Short awakenings or sleep episodes may
+not necessarily correspond to noticeable changes in thigh movement,
+making them difficult to detect and accurately classify. These results
+align with several methods for wrist-worn devices reviewed by Conley and
+colleagues\citep{conley2019}. They reported correlations between
+accelerometer and PSG sleep onset latency (equivalent to LPS) from 10
+studies with a mean correlation of 0.2 (ranging from -0.69 to 0.69),
+indicating the inherent difficulty in estimating this parameter using
+accelerometry alone.
+
+In this study, we used the ZM as the reference method, rather than PSG,
+which is considered the gold standard for sleep measurement. This choice
+may contribute to discrepancies between our models and the ZM, as
+without a true gold standard, it's difficult to determine the source of
+disagreement. However, we believe that the use of ZM, which allows for
+multiple consecutive nights of recording, is valuable. This approach
+captures intra-individual variances in sleep, which is impractical with
+PSG. It also enabled us to include more nights in our study typically
+compared to those relying on PSG. For instance, the widely used
+Newcastle dataset\citep{hees2015} only contains data from 28
+participants. However, upon examining the ZM outputs, we found that the
+raw predictions were not optimal for developing machine learning models
+due to a seemingly low signal-to-noise ratio (see
+Figure~\ref{fig-zm-median}). The ZM itself mitigates this issue by
+applying certain filtering processes when generating sleep quality
+parameters. For example, epochs contributing to WASO must be in
+contiguous epochs of 3, and sleep only counts towards sleep quality
+parameters if 10 out of 12 minutes are scored as sleep. To improve the
+prospect of our machine learning algorithms, we applied median filters
+to the ZM raw predictions. This did in fact alter the derived sleep
+quality parameters. Notably, the mean WASO decreased from 39 minutes in
+the raw predictions to 30.6 minutes in the 5-minute median filtered
+predictions, and further decreased to 22.3 minutes in the 10-minute
+filtered predictions. The application of 5-minute and 10-minute median
+filters also led to increases in TST, SE, and LPS. This suggests that
+the filters may categorize some instances of wakefulness as sleep and
+smooth out brief awakenings. Despite these changes, the overall sleep
+quality profile derived from the median-filtered predictions is still
+comparable to that from the raw predictions, justifying our approach.
+
+Our study's XGBoost model demonstrated relatively narrower limits of
+agreements (LOAs) for TST, SE, and WASO, with ranges of -95.5 to 81.4
+min, -15.6 to 13.3\%, and -83.4 to 81.7 min, respectively when compared
+with other models such as the Van Hees algorithm\citep{hees2015}, Oakley
+rsc (rescored)\citep{palotti2019}, and LSTM-50\citep{palotti2019}
+evaluated in the Patterson et al.~study\citep{patterson_40_2023}.
+Furthermore, comparing the LOAs between our XGBoost model and the
+algorithm developed for thigh-worn devices by Johansson et
+al.~study\citep{johansson2022}, our XGBoost model showed narrower LOAs
+for TST , SE, LPS , and WASO, but not SPT. Generally, all methods, both
+from this study and from the reviewed literature, exibit wide LOAs
+sugesting that there is high variability in the derevide sleep quality
+parameters. These findings imply that the current methods, are only
+reasonbly reliable for assessing sleep quality parameters at a group
+level. However, caution should be exercised when applying the models and
+methods to individual-level sleep assessments. Therefore, further
+improvements and refinements are needed to enhance the precision and
+reliability of these models for individual sleep assessments.
+
+Typically, sleep detection methods are applied in two contexts: either
+to night recordings or to 24-hour recordings. In night recordings, it is
+often possible to derive sleep quality parameters like SE and LPS
+because the SPT is already known. This knowledge usually comes either
+from sleep diaries\citep{girschik2012} or from the length of the
+recording itself. On the other hand, when sleep detection methods are
+applied to 24-hour recordings, most methods do not have the ability to
+infer the SPT. Consequently, these methods are unable to generate
+certain sleep quality parameters that rely on the
+SPT\citep{doherty2017, anderson2014}.
+
+Strengths and limitations It is not always clear the methods for
+detecting sleep from accelerometer data is
+
+It is unclear whether Johansson et al.\citep{johansson_development_2023}
+provide a method for detecting the in-bed time which hinders its
+applicability for researchers that want to \ldots{}
+
+external validation dataset/generalization
+
+expand the feature set (Palotti)
+
+In conclusion, our study has advanced the methods for estimating sleep
+from thigh-worn accelerometers by evaluating various models for
+predicting in-bed and sleep time. Our results indicate that all
+sequential models and the multiclass model performed well at predicting
+in-bed time. However, all models showed difficulty in correctly
+classifying awake epochs during time in bed, highlighting the challenge
+of automating the detection of in-bed awake periods. The application of
+median filters improved the performance metrics of all models,
+suggesting that this approach may be beneficial in future studies.
+Despite these promising results, our findings also underscore the need
+for caution when applying these models to individual-level sleep
+assessments due to the high variability in derived sleep quality
+parameters. Future research should focus on refining these models to
+enhance their precision and reliability for individual sleep
+assessments.
+
+Our open-source algorithm and the inferred sleep parameters will open
+the door to future studies on sleep and sleep architecture using
+large-scale accelerometer databases.
 

### 2023-07-12 manuscript/manuscript.qmd
all sections of the discussion is ok

@@
 generate certain sleep quality parameters that rely on the
-SPT[@doherty2017; @anderson2014]. We have attempted to circumvent this
-issue by employing models able to distinguish between in-bed awake time
-and in-bed asleep time over a 24h recording making our models capable of
-estimating all commonly used sleep quality parameters. Van Hees et
-al.[@van_hees_estimating_2018] have developed an algorithm to identify
-SPT from wrist-worn devices which was recently validated by Plekhanova
-and colleagues[@plekhanova2023]. which in combination with other
-algorithms or models on said SPT can then be used infer further sleep
-quality parameters.
+SPT[@doherty2017; @anderson2014]. To overcome this limitation, we have
+incorporated models that can differentiate between in-bed awake time and
+in-bed asleep from out-bed awake time over a 24-hour recording. This
+approach allows our models to estimate all commonly used sleep quality
+parameters. Van Hees et al. [@van_hees_estimating_2018] have proposed an
+algorithm to determine SPT from data collected by wrist-worn devices.
+This algorithm was recently validated by Plekhanova and her team
+[@plekhanova2023]. By combining this algorithm with other methods,
+further sleep quality parameters can be inferred based on the identified
+SPT. Van Hees et al.[@van_hees_estimating_2018] reported good agreements
+and low mean differences compared to self-report and PSG on SPT,
+findings later confirmed by Plekhanova and colleagues. However, they
+also observed poor agreement with LPS and Wake After Sleep Onset (WASO).
+They found low reliability with PSG, indicating difficulties in
+detecting wakefulness during in-bed time. These challenges parallel
+those we experienced in our study.
 
-Strengths and limitations It is not always clear the methods for
-detecting sleep from accelerometer data is
+The study boasts several strengths, including the capacity to derive
+in-bed awake and asleep times from out-of-bed data, thereby allowing for
+the extraction of vital sleep quality parameters. Furthermore, the
+research benefits from evaluating multiple nights per subject, providing
+valuable information into intra-subject sleep variability. However,
+certain limitations exist. The use of ZM, which isn't recognized as a
+gold standard, could potentially compromise our findings' validity.
+Future research could consider using PSG as a reference for methods
+similar to ours, despite its limitations, for a more accurate
+comparison. Moreover, our models weren't validated using an external
+dataset, a process that would have showcased their broader
+applicability. Hence, our conclusions remain confined primarily to
+children.
 
-It is unclear whether Johansson et al.[@johansson_development_2023]
-provide a method for detecting the in-bed time which hinders its
-applicability for researchers that want to ...
-
-external validation dataset/generalization
-
-expand the feature set (Palotti)
-
-In conclusion, our study has advanced the methods for estimating sleep
-from thigh-worn accelerometers by evaluating various models for
-predicting in-bed and sleep time. Our results indicate that all
-sequential models and the multiclass model performed well at predicting
-in-bed time. However, all models showed difficulty in correctly
-classifying awake epochs during time in bed, highlighting the challenge
-of automating the detection of in-bed awake periods. The application of
-median filters improved the performance metrics of all models,
-suggesting that this approach may be beneficial in future studies.
-Despite these promising results, our findings also underscore the need
-for caution when applying these models to individual-level sleep
-assessments due to the high variability in derived sleep quality
-parameters. Future research should focus on refining these models to
-enhance their precision and reliability for individual sleep
-assessments.
-
-Our open-source algorithm and the inferred sleep parameters will open
-the door to future studies on sleep and sleep architecture using
-large-scale accelerometer databases.
+In conclusion, our study contributes to the ongoing efforts to improve
+sleep estimation methods using thigh-worn accelerometers. We evaluated
+different machine learning models and configurations for predicting
+in-bed and sleep times and their corresponding sleep quality parameters.
+While the sequential models generally demonstrated excellent performance
+in predicting in-bed time, they faced challenges in accurately
+distinguishing between sleep and wake epochs during in-bed time.
+Nevertheless, their performance could be substantially improved through
+the application of median filters. Among all models and configurations
+evaluated, the XGBoost model exhibited superior performance across all
+performance metrics, including epoch-to-epoch predictions and sleep
+quality parameter derivatives. Our research also highlighted the current
+limitations of sleep detection methods, such as challenges in
+effectively detecting wake periods during in-bed time and the need for
+further improvements to increase the precision of individual sleep
+assessments. We believe our work lays the groundwork for future research
+to further refine and improve the performance of these models,
+contributing to a more precise and accurate evaluation of sleep patterns
+and quality using thigh-worn accelerometers.
 

### 2023-07-13 manuscript/manuscript.qmd
fix minor errors in manus text.

@@
   elsevier-pdf:
-    pdf-engine: lualatex
     keep-tex: false
@@
     fig-pos: 'b'
-  simple-article-pdf:
-    indent: false
-    classoption: 
-     - 9pt
-     - lucida
-     - twocolumn
-     - twoside
   docx:
@@
 techniques, including heuristic algorithms, machine learning models,
-regression, and deep learning, were employed to analyze wrist-worn
-accelerometer data[@palotti2019; @cole1992; @sazonov2004; @sadeh1994;
-@hees2015; @sundararajan2021].
+regression, and deep learning, were employed to analyze hip
+andwrist-worn accelerometer data[@palotti2019; @cole1992; @sazonov2004;
+@sadeh1994; @hees2015; @sundararajan2021].
 
@@
 Despite these advancements, the application of machine learning
-techniques in this area is still relatively unexplored. Considering the
-potential of thigh-worn accelerometers for accurate physical behavior
+techniques in this area is still unexplored. Considering the potential
+of thigh-worn accelerometers for accurate physical behavior
 assessment[@skotte_detection_2014; @arvidsson2019], there is a
@@
 To ensure the reliability and effectiveness of our models, we compared
-their outputs with an EEG-based sleep tracking device, which we
-considered the gold standard for measuring sleep. Furthermore, our
-secondary goal was to assess the developed models' performance in
-evaluating important sleep quality parameters, including sleep period
-time (SPT), total sleep time (TST), sleep efficiency (SE), latency until
-persistent sleep (LPS), and wake after sleep onset (WASO).
+their outputs with an EEG-based sleep tracking device, which we, in this
+current study, considered as the gold standard for measuring sleep.
+Furthermore, our secondary goal was to assess the developed models'
+performance in evaluating important sleep quality parameters, including
+sleep period time (SPT), total sleep time (TST), sleep efficiency (SE),
+latency until persistent sleep (LPS), and wake after sleep onset (WASO).
 
@@
 from Axivity AX3 devices attached to the children's thighs, and
-electroencephalography (EEG) data derived from the ZM device. The
-Axivity AX3, an unobtrusive 3-axis accelerometer, was positioned midway
-between the hip and knee on the right anterior thigh, recording
-participant movement data.
+electroencephalography data derived from the ZM device. The Axivity AX3,
+an unobtrusive 3-axis accelerometer, was positioned midway between the
+hip and knee on the right anterior thigh, recording participant movement
+data.
 
@@
 sensors placed outside the hairline for reliable EEG signal acquisition.
-The ZM uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is
-utilized for accurate sleep detection, showcasing its suitability for
-in-home monitoring[@kaplan2014], while the Z-PLUS effectively
-differentiates sleep stages, as evidenced by its alignment with expert
-evaluations using PSG data[@wang2015]. In the current study, we treated
-all sleep stages as a single category effectively deducing the output of
-the ZM to "awake" and "asleep" as the ability to distinguish sleep
-stages are not a necessity to derive sleep quality parameters of
-interest and to simplify the learning process of the models.
+The participants of the SCREENS study were instructed to attach the
+device when they went to bed. The ZM uses two proprietary algorithms:
+Z-ALG and Z-PLUS. The Z-ALG is utilized for accurate sleep detection,
+showcasing its suitability for in-home monitoring[@kaplan2014], while
+the Z-PLUS effectively differentiates sleep stages, as evidenced by its
+alignment with expert evaluations using PSG data[@wang2015]. In the
+current study, we treated all sleep stages as a single category
+effectively deducing the output of the ZM to "awake" and "asleep" as the
+ability to distinguish sleep stages are not a necessity to derive the
+sleep quality parameters of interest and to simplify the learning
+process of the models.
 
@@
 were accompanied by complete accelerometer data and lasted between 7 and
-14 hours. Any night when the ZM reported sensor issues was excluded. The
-children whose recordings were considered had an average age of 9.4
-years, with a standard deviation of 2.1. In their raw form, the ZM
-predictions encompassed 696,779 epochs, each 30 seconds long. Notably,
-approximately 84% of the total ZM recording duration was classified as
-sleep, resulting in an imbalance of the dataset.
+14 hours. Any night when the ZM reported sensor issues was excluded
+yielding 585 nights included in the study. The children whose recordings
+were considered had an average age of 9.4 years, with a standard
+deviation of 2.1. In their raw form, the ZM predictions encompassed
+696,779 epochs, each 30 seconds long. Notably, approximately 84% of the
+total ZM recording duration was classified as sleep, resulting in an
+imbalance of the ZM dataset.
 
@@
 similar to methods described by Skotte et al.[@skotte_detection_2014].
-Any non-wear data was remove using previously described
+Any non-wear data was removed using previously described
 methods[@skovgaard2023] and data was resampled to 30-second epochs so
@@
 both lag and lead values, capturing dynamic data trends by incorporating
-measurements from preceding and upcoming intervals. Furthermore,
-inspired by Walch et al.[@walch2019], we incorporated sensor-independent
-features to encapsulate circadian rhythms. These features offer unique
-insights not directly discernible from sensor outputs and are meant to
+measurements from preceding and upcoming epochs. Furthermore, inspired
+by Walch et al.[@walch2019], we incorporated sensor-independent features
+to encapsulate circadian rhythms. These features offer unique insights
+not directly discernible from sensor outputs and are meant to
 approximate the changing drive of the circadian clock to sleep over the
@@
 for the children in our study, exceeding what would be expected based on
-typical sleep cycle patterns. In particular, many of these brief
-awakenings could be considered as noise, which when present in the data,
-can potentially hinder the learning process of machine learning models
-by obscuring the underlying patterns that the models are trying to
-learn, leading to less accurate predictions. Consequently, we elected to
-train and evaluate our models using not only the raw ZM output, but also
-versions that were subjected to 5-minute and 10-minute median filters.
-This approach, by mitigating this noise, resulted in an anticipated,
-more age-appropriate count of awakenings per night, providing a more
-accurate depiction of children's sleep patterns (see @fig-zm-median).
+typical sleep cycle patterns (see @fig-zm-median). In particular, many
+of these brief awakenings could be considered as noise, which when
+present in the data, can potentially hinder the learning process of
+machine learning models by obscuring the underlying patterns that the
+models are trying to learn, leading to less accurate predictions.
+Consequently, we elected to train and evaluate our models using not only
+the raw ZM output, but also versions that were subjected to 5-minute and
+10-minute median filters. This approach, by mitigating this noise,
+resulted in an anticipated, more age-appropriate count of awakenings per
+night, providing a more accurate depiction of children's sleep patterns
+(see @tbl-zm_overview).
 
@@
 
-In this study, we employed two distinct modeling strategies to analyze
+In this study, we employed two distinct learning strategies to analyze
 sleep patterns from thigh-mounted accelerometer data. We used a
@@
 into a training set and a testing set, each containing approximately 50%
-of the subjects. This division ensured that samples from the same night
-were never simultaneously present in both sets. To optimize
+of the subjects. This division was ensured to not have samples from the
+same night simultaneously present in both sets. To optimize
 hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
@@
 LOAs between our XGBoost model and the algorithm developed for
-thigh-worn devices by Johansson et al. study[@johansson2022], our
-XGBoost model showed narrower LOAs for TST , SE, LPS , and WASO, but not
-SPT. Generally, all methods, both from this study and from the reviewed
-literature, exibit wide LOAs sugesting that there is high variability in
-the derevide sleep quality parameters. These findings imply that the
-current methods, are only reasonbly reliable for assessing sleep quality
-parameters at a group level. However, caution should be exercised when
-applying the models and methods to individual-level sleep assessments.
-Therefore, further improvements and refinements are needed to enhance
-the precision and reliability of these models for individual sleep
-assessments.
+thigh-worn devices by Johansson et al.
+study[@johansson_development_2023], our XGBoost model showed narrower
+LOAs for TST , SE, LPS , and WASO, but not SPT. Generally, all methods,
+both from this study and from the reviewed literature, exibit wide LOAs
+sugesting that there is high variability in the derevide sleep quality
+parameters. These findings imply that the current methods, are only
+reasonbly reliable for assessing sleep quality parameters at a group
+level. However, caution should be exercised when applying the models and
+methods to individual-level sleep assessments. Therefore, further
+improvements and refinements are needed to enhance the precision and
+reliability of these models for individual sleep assessments.
 

### 2023-07-14 manuscript/manuscript.qmd
first draft of manuscript done.

@@
 
-While laboratory-based polysomnography (PSG) is typically considered the
-gold standard for objective sleep measurement, its practicality in
-large-scale epidemiological studies is hindered due to high costs, the
-necessity for professional administration, and it is also subject to
-potential rater bias[@vandewater2011; @lee2022]. As an alternative,
-diaries are commonly employed as cost-effective and low-tech methods for
-sleep assessment in population research. However, reliance on
-diary-based methods may lead to recall bias and other
-limitations[@moore2015]. A more feasible approach in large-scale
-epidemiological studies is to use device-based measurement methods that
-can estimate sleep duration. This approach offers the advantage of being
-less burdensome for participants and eliminates potential biases
-associated with recall.
-
-Body-worn accelerometers have emerged as an effective and affordable
-alternative for objectively assessing sleep patterns at home over
-extended periods. These devices gather continuous, high-resolution data
-for several weeks without the need for recharging, thus reducing
-participant burden. Initial applications of accelerometry for sleep and
-wake stage classification were based on wrist movements, starting with
-an algorithm developed in 1982 and validated with PSG [@webster1982].
-This model was later refined in 1992 [@cole1992], giving rise to the
-widely used Cole-Kripke model. As the field advanced, an array of
-techniques, including heuristic algorithms, machine learning models,
-regression, and deep learning, were employed to analyze hip
-andwrist-worn accelerometer data[@palotti2019; @cole1992; @sazonov2004;
-@sadeh1994; @hees2015; @sundararajan2021].
+The traditional gold standard for objective sleep measurement,
+laboratory-based polysomnography (PSG), has been found to be impractical
+in large-scale epidemiological studies due to its high cost, need for
+professional administration, and susceptibility to rater
+bias[@vandewater2011; @lee2022]. As an alternative, diaries have been
+used due to their cost-effectiveness and simplicity, although they are
+subject to recall bias and other limitations[@moore2015]. An innovative
+approach involves device-based measurement methods. These tools, which
+estimate sleep duration, are advantageous due to their reduced
+participant burden and elimination of potential recall biases. A
+prominent example of such tools is body-worn accelerometers, which offer
+a practical and affordable means of objectively assessing sleep patterns
+at home for extended periods. Accelerometers collect continuous,
+high-resolution data for several weeks without requiring recharging,
+further minimizing participant burden. Their use in sleep and wake
+classification began with a wrist movement-based algorithm developed in
+1982, and validated using PSG[@webster1982]. This algorithm was refined
+in 1992[@cole1992], leading to the widely adopted Cole-Kripke model.
+With advancements in the field, a variety of techniques, including
+heuristic algorithms, machine learning models, regression, and deep
+learning, are now used to analyze data from hip and wrist-worn
+accelerometers[@palotti2019; @cole1992; @sazonov2004; @sadeh1994;
+@hees2015; @sundararajan2021].
 
@@
 
-## Algorithms, Training and Validation
+## Algorithms
 
 In this study, we employed two distinct learning strategies to analyze
-sleep patterns from thigh-mounted accelerometer data. We used a
-sequential strategy, comprising an ensemble of four pairs of models,
-each pair featuring the same algorithm. This strategy aimed to make the
-prediction task more manageable for the algorithms by breaking it down
-into a sequence of two binary classifications: first predicting 'in-bed'
-time, then 'sleep' time. Simultaneously, we also used a multiclass
-approach utilizing a bidirectional Long Short-Term Memory
-(biLSTM)[@hochreiter1997] neural network.
-
-### Models in Sequence
-
-To predict in-bed time and sleep time accurately, we employed an
-ensemble learning strategy based on sequential binary classification
-models. This approach involved constructing a sequence of models using
-multiple machine learning algorithms to improve predictive accuracy. The
-process began with an initial model predicting in-bed time, followed by
-a second model that utilized the output of the initial model to predict
-sleep time. This sequential approach was applied across all four
-algorithms detailed below, with each subsequent model leveraging the
-outputs of the previous models for improved predictions.
+sleep patterns from thigh-mounted accelerometer data. The first strategy
+employed a sequential methodology, using an ensemble of four pairs of
+models, with each pair featuring the same algorithm. This sequential
+strategy was designed to simplify the prediction task by decomposing it
+into two stages: first predicting 'in-bed' time, then 'sleep' time. The
+output from the first set of models, which predicted in-bed time, was
+subjected to a 5-minute median filter to eliminate leading and trailing
+in-bed time blips. This process enabled us to extract a single
+continuous time interval that we identified as the sleep period time
+window (SPT). The SPT, in turn, served as the input to the second models
+in the sequence, enhancing their predictive accuracy for sleep time. We
+applied this sequential strategy across all four algorithms, as detailed
+below.
 
@@
 
-### Multiclass Model
+In parallel, we also employed a multiclass approach using a
+bidirectional Long Short-Term Memory (biLSTM)[@hochreiter1997] neural
+network. This network, which was designed to predict three distinct
+classes: 'out-of-bed-awake', 'in-bed-awake', and 'in-bed-asleep', was
+configured with four layers and 128 hidden units per layer. This balance
+between model complexity and training efficiency was intended to
+facilitate learning of intricate patterns while ensuring feasible
+training times. The bidirectional nature of the LSTM enhanced data
+interpretation and reduced overfitting by doubling the hidden units at
+each time step. The LSTM model used sequences of tensors as input, with
+each sequence spanning 10 minutes and a step size of one. As
+demonstrated by previous studies such as those by Sano et al.
+[@sano2019] and Chen et al. [@chen2021], LSTM models have shown great
+promise in sleep detection using accelerometer data, thanks to their
+ability to capture complex temporal patterns.
 
-We also employed a biLSTM, a multiclass classifier, to predict three
-distinct states: out-of-bed-awake, in-bed-awake, and in-bed-asleep. The
-architecture of the biLSTM was set up with four layers, each equipped
-with 128 hidden units. This configuration was intentionally chosen to
-balance between model complexity and training efficiency: it provided
-the depth necessary for learning intricate patterns while remaining
-feasible for timely training. The bidirectional design of the LSTM
-served to enhance data interpretation and mitigate overfitting by
-doubling the hidden units at each time step. For input, we used tensors
-shaped as sequences, with each sequence spanning 10 minutes and a step
-size of one.This approach follows in the footsteps of previous studies
-that utilized LSTM models for sleep detection. These studies showcased
-the promising potential of LSTMs in capturing complex temporal patterns.
-Particularly, the works of Sano et al. [@sano2019] and Chen et al.
-[@chen2021] demonstrated the effectiveness of LSTM models in improving
-sleep detection using accelerometer data, underscoring the value of this
-modeling approach.
-
-### Model Training
+## Model Training
 
@@
 into a training set and a testing set, each containing approximately 50%
-of the subjects. This division was ensured to not have samples from the
-same night simultaneously present in both sets. To optimize
-hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
-a regular grid, comprising 20 different combinations of hyperparameters.
-The F1 score served as the optimization metric. The best-performing set
-of hyperparameters was then used to fit the models to the full training
-dataset. This approach allowed us to maximize performance by leveraging
-all available data. Moreover, after extracting the in-bed time from the
-initial sequential models, the imbalance on the resulting dataset could
-cause biases during model training, as models may favor predicting the
-majority class. To rectify this, we employed the Synthetic Minority
-Over-sampling Technique (SMOTE)[@chawla2002]. SMOTE generates new
-samples by interpolating random samples with their nearest neighbors. We
-utilized the themis R package[@themis] to implement SMOTE, resulting in
-a balanced distribution of training samples across both classes.
+of the subjects. The splitting of the data was ensured to not have
+samples from the same night simultaneously present in both sets. To
+optimize hyperparameters, we performed a 10-fold Monte Carlo
+cross-validation on a regular grid, comprising 20 different combinations
+of hyperparameters. The F1 score served as the optimization metric. The
+best-performing set of hyperparameters was then used to fit the models
+to the full training dataset. This approach allowed us to maximize
+performance by leveraging all available data. Moreover, after extracting
+the in-bed time from the initial sequential models, the imbalance on the
+resulting dataset could cause biases during model training, as models
+may favor predicting the majority class. To rectify this, we employed
+the Synthetic Minority Over-sampling Technique (SMOTE)[@chawla2002].
+SMOTE generates new samples by interpolating random samples with their
+nearest neighbors. We utilized the themis R package[@themis] to
+implement SMOTE, resulting in a balanced distribution of training
+samples across both classes.
 
@@
 
-### Model Validation
+## Model Validation
 
@@
 
-1.  Sleep Period Time (SPT) - This refers to the total duration of the
-    sleep period, which is defined as the time from the start to the end
-    of the ZM recording.
+1.  Sleep Period Time (SPT) - This refers to the total duration of time
+    in bed with the intention to sleep, which is defined as the time
+    from the start to the end of the ZM recording.
 
@@
 the out-of-bed and in-bed time are presented in @fig-conf_mat. The
-figure shows favorable epoch-to-epoch performance across across all
-sequential models, however, it is evident that the biLSTM is less
-successful in classifying the in-bed-awake class which cannot be deduced
-from the confusion matrices from the sequential models.
+figure shows favorable epoch-to-epoch performance across all sequential
+models, however, it is evident that the biLSTM is less successful in
+classifying the in-bed-awake class which cannot be deduced from the
+confusion matrices from the sequential models.
 
@@
 model had similar trends, with more pronounced underestimation in TST
-and overestimation in LPS. The eed-forward neural network also exhibited
-similar bias as the decision tree and the logistic regression models,
-but with a higher overestimation in WASO. On the other hand, the XGBoost
-model showed least bias among all, especially in its 5-minute median
-predictions. Considering LOA, the decision tree had higher variability
-across different sleep quality parameters and filtering techniques,
-particularly for LPS and WASO, which indicates lower agreement with ZM.
-Other models had comparable LOA but with notable exceptions. For
-example, TST LOA for the logistic regression model was particularly wide
-in the 5-minute median predictions. Correlation-wise, the pearson
-coefficient, revealed that the XGBoost model consistently had the
-highest correlation with ZM across all sleep qualityparameters and
-filtering methods Notably, the XGBoost's 5-minute median predictions
+and overestimation in LPS. The feed-forward neural network also
+exhibited similar bias as the decision tree and the logistic regression
+models, but with a higher overestimation in WASO. On the other hand, the
+XGBoost model showed least bias among all, especially in its 5-minute
+median predictions. Considering LOA, the decision tree had higher
+variability across different sleep quality parameters and filtering
+techniques, particularly for LPS and WASO, which indicates lower
+agreement with ZM. Other models had comparable LOA but with notable
+exceptions. For example, TST LOA for the logistic regression model was
+particularly wide in the 5-minute median predictions. Correlation-wise,
+the pearson coefficient, revealed that the XGBoost model consistently
+had the highest correlation with ZM across all sleep quality parameters
+and filtering methods Notably, the XGBoost's 5-minute median predictions
 showed the strongest correlation (0.66) for TST among all models and
@@
 median-smoothed ZM-derived sleep quality parameters. The Bland-Altman
-plot for the SPT reveals a significant level of agreement with the ZM,
-as evidenced by a positive correlation. However, the presence of extreme
-outliers widens the limits of agreement (LOA). The scatterplot for SPT
-also demonstrates a positive trend, indicating a moderate linear
-correlation between the XGBoost model and the ZM-derived sleep quality
-parameters.In terms of TST, perfect agreement is not observed for a
-substantial number of nights, but there is a positive correlation
-between the XGBoost model and ZM-derived sleep quality parameters. The
-bias and LOA for TST are comparable to those observed for SPT,
-indicating a consistent level of agreement between the two methods. The
-scatterplot for TST also shows a slightly higher correlation, primarily
-driven by the absence of extreme outliers.Furthermore, the remaining
-three sleep quality parameters, SE, LPS, and WASO, exhibit
-heteroscedasticity in contrast to SPT and TST. This outcome is expected
-as achieving 100% sleep efficiency is relatively rare, resulting in less
-disagreement between the methods as values approach the upper limit.
-However, as sleep efficiency decreases, the potential for discrepancies
-and differing interpretations between the methods increases, leading to
-greater heteroscedasticity. A moderate linear correlation is observed
-between the XGBoost model and ZM-derived sleep quality parameters for
-SE, indicating a positive relationship. However, a poor correlation is
-observed for LPS and WASO, suggesting less agreement between the two
-methods for these parameters. Similar plots for all models are available
-in the supplementary materials.
+plot for the SPT and TST reveals a significant level of agreement with
+the ZM, as evidenced by a bias close to zero. Interestingly, a portion
+of the data points are located near the zero line indicating perfect
+agreement. However, the presence of extreme outliers widens the limits
+of agreement (LOA). The scatterplot for SPT also demonstrates a positive
+trend, indicating a moderate linear correlation between the XGBoost
+model and the ZM-derived sleep quality parameters. The bias and LOA for
+TST are comparable to those observed for SPT, indicating a consistent
+level of agreement between the two methods. The scatterplot for TST also
+shows a slightly higher correlation, primarily driven by the absence of
+extreme outliers.Furthermore, the remaining three sleep quality
+parameters, SE, LPS, and WASO, exhibit heteroscedasticity in contrast to
+SPT and TST. This outcome is expected as achieving 100% sleep efficiency
+is relatively rare, resulting in less disagreement between the methods
+as values approach the upper limit. However, as sleep efficiency
+decreases, the potential for discrepancies and differing interpretations
+between the methods increases, leading to greater heteroscedasticity. A
+moderate positive linear correlation is observed between the XGBoost
+model and ZM-derived sleep quality parameters for SE, however, a poor
+correlation is observed for LPS and WASO, suggesting less agreement
+between the two methods for these parameters. Similar plots for all
+models are available in the **supplementary materials (75 plots. Jeg ved
+ikke om nogen orker at bladre igennem dem).**
 
@@
 
-In our evaluation of sleep quality parameters, we found that Latency to
-Persistent Sleep (LPS) had the largest mean error relative to absolute
-time allocated to LPS. This suggests that the initial epochs of Sleep
-Period Time (SPT) are particularly challenging to classify correctly.
-This is also supported by the poor Pearson correlations between LPS
-derived from model predictions and the ZM. The XGBoost model, which was
-the best performer among all models, overestimated LPS by an average of
-26.4 minutes for models trained on raw ZM predictions, 28.5 minutes for
-models trained on 5-minute filtered ZM predictions, and 34.5 minutes for
-models trained on 10-minute filtered ZM predictions. This level of
-discrepancy is comparable to the mean error of sleep latency of 23
-minutes reported by Johansson et al.[@johansson_development_2023].
-Johansson et al. suggest that the discrepancy with the gold standard is
-likely due to the multifaceted nature of the sleep state, which is a
-complex physiological process. Short awakenings or sleep episodes may
-not necessarily correspond to noticeable changes in thigh movement,
-making them difficult to detect and accurately classify. These results
-align with several methods for wrist-worn devices reviewed by Conley and
+Typically, sleep detection methods are applied in two contexts: either
+to night recordings or to 24-hour recordings. In night recordings, it is
+possible to derive sleep quality parameters like SE and LPS because the
+SPT is already known because it is inferred from the length of the
+recording[@conley2019; @patterson_40_2023]. On the other hand, when
+sleep detection methods are applied to 24-hour recordings, most methods
+do not have the ability to infer the SPT with sleep
+diaries[@girschik2012]. Consequently, these methods are unable to
+generate certain sleep quality parameters that rely on the
+SPT[@doherty2017; @anderson2014]. To overcome this limitation, we have
+incorporated models that can differentiate between in-bed awake time and
+in-bed asleep from out-bed awake time over a 24-hour recording. This
+approach allows our models to estimate all commonly used sleep quality
+parameters. Van Hees et al. [@van_hees_estimating_2018] have proposed an
+algorithm to determine SPT from data collected by wrist-worn devices.
+This algorithm was recently validated by Plekhanova and her team
+[@plekhanova2023]. By combining this algorithm with other methods,
+further sleep quality parameters can be inferred based on the identified
+SPT. Van Hees et al.[@van_hees_estimating_2018] reported good agreements
+and low mean differences compared to self-report and PSG on SPT,
+findings later confirmed by Plekhanova and colleagues. However, they
+also observed poor agreement with LPS and Wake After Sleep Onset (WASO).
+They found low reliability with PSG, indicating difficulties in
+detecting wakefulness during in-bed time. These challenges parallel
+those we experienced in our study.
+
+In our evaluation of sleep quality parameters, we found that LPS had the
+largest mean error relative to absolute time allocated to LPS. This
+suggests that the initial epochs of Sleep Period Time (SPT) are
+particularly challenging to classify correctly. This is also supported
+by the poor Pearson correlations between LPS derived from model
+predictions and the ZM. The XGBoost model, which was the best performer
+among all models, overestimated LPS by an average of 26.4 minutes for
+models trained on raw ZM predictions, 28.5 minutes for models trained on
+5-minute filtered ZM predictions, and 34.5 minutes for models trained on
+10-minute filtered ZM predictions. This level of discrepancy is
+comparable to the mean error of sleep latency of 23 minutes reported by
+Johansson et al.[@johansson_development_2023]. Johansson et al. suggest
+that the discrepancy with the gold standard is likely due to the
+multifaceted nature of the sleep state, which is a complex physiological
+process. Short awakenings or sleep episodes may not necessarily
+correspond to noticeable changes in thigh movement, making them
+difficult to detect and accurately classify. These results align with
+several methods for wrist-worn devices reviewed by Conley and
 colleagues[@conley2019]. They reported correlations between
@@
 accelerometry alone.
+
+Our study's XGBoost model demonstrated relatively narrower limits of
+agreements (LOAs) for TST, SE, and WASO, with ranges of -95.5 to 81.4
+min, -15.6 to 13.3%, and -83.4 to 81.7 min, respectively when compared
+with other models such as the Van Hees algorithm[@hees2015], Oakley rsc
+(rescored)[@palotti2019], and LSTM-50[@palotti2019] evaluated in the
+Patterson et al. study[@patterson_40_2023]. Furthermore, comparing the
+LOAs between our XGBoost model and the algorithm developed for
+thigh-worn devices by Johansson et al.
+study[@johansson_development_2023], our XGBoost model showed narrower
+LOAs for TST , SE, LPS , and WASO, but not SPT. Generally, all methods,
+both from this study and from the reviewed literature, exibit wide LOAs
+sugesting that there is high variability in the derevide sleep quality
+parameters. These findings imply that the current methods, are only
+reasonbly reliable for assessing sleep quality parameters at a group
+level. However, caution should be exercised when applying the models and
+methods to individual-level sleep assessments. Therefore, further
+improvements and refinements are needed to enhance the precision and
+reliability of these models for individual sleep assessments.
 
@@
 
-Our study's XGBoost model demonstrated relatively narrower limits of
-agreements (LOAs) for TST, SE, and WASO, with ranges of -95.5 to 81.4
-min, -15.6 to 13.3%, and -83.4 to 81.7 min, respectively when compared
-with other models such as the Van Hees algorithm[@hees2015], Oakley rsc
-(rescored)[@palotti2019], and LSTM-50[@palotti2019] evaluated in the
-Patterson et al. study[@patterson_40_2023]. Furthermore, comparing the
-LOAs between our XGBoost model and the algorithm developed for
-thigh-worn devices by Johansson et al.
-study[@johansson_development_2023], our XGBoost model showed narrower
-LOAs for TST , SE, LPS , and WASO, but not SPT. Generally, all methods,
-both from this study and from the reviewed literature, exibit wide LOAs
-sugesting that there is high variability in the derevide sleep quality
-parameters. These findings imply that the current methods, are only
-reasonbly reliable for assessing sleep quality parameters at a group
-level. However, caution should be exercised when applying the models and
-methods to individual-level sleep assessments. Therefore, further
-improvements and refinements are needed to enhance the precision and
-reliability of these models for individual sleep assessments.
-
-Typically, sleep detection methods are applied in two contexts: either
-to night recordings or to 24-hour recordings. In night recordings, it is
-possible to derive sleep quality parameters like SE and LPS because the
-SPT is already known because it is inferred from the length of the
-recording[@conley2019; @patterson_40_2023]. On the other hand, when
-sleep detection methods are applied to 24-hour recordings, most methods
-do not have the ability to infer the SPT with sleep
-diaries[@girschik2012]. Consequently, these methods are unable to
-generate certain sleep quality parameters that rely on the
-SPT[@doherty2017; @anderson2014]. To overcome this limitation, we have
-incorporated models that can differentiate between in-bed awake time and
-in-bed asleep from out-bed awake time over a 24-hour recording. This
-approach allows our models to estimate all commonly used sleep quality
-parameters. Van Hees et al. [@van_hees_estimating_2018] have proposed an
-algorithm to determine SPT from data collected by wrist-worn devices.
-This algorithm was recently validated by Plekhanova and her team
-[@plekhanova2023]. By combining this algorithm with other methods,
-further sleep quality parameters can be inferred based on the identified
-SPT. Van Hees et al.[@van_hees_estimating_2018] reported good agreements
-and low mean differences compared to self-report and PSG on SPT,
-findings later confirmed by Plekhanova and colleagues. However, they
-also observed poor agreement with LPS and Wake After Sleep Onset (WASO).
-They found low reliability with PSG, indicating difficulties in
-detecting wakefulness during in-bed time. These challenges parallel
-those we experienced in our study.
-
-The study boasts several strengths, including the capacity to derive
-in-bed awake and asleep times from out-of-bed data, thereby allowing for
-the extraction of vital sleep quality parameters. Furthermore, the
+The study boasts several strengths, including the capacity to
+distinguish in-bed awake and asleep from out-of-bed, thereby allowing
+for the extraction of vital sleep quality parameters. Furthermore, the
 research benefits from evaluating multiple nights per subject, providing
@@
 sleep estimation methods using thigh-worn accelerometers. We evaluated
-different machine learning models and configurations for predicting
-in-bed and sleep times and their corresponding sleep quality parameters.
-While the sequential models generally demonstrated excellent performance
-in predicting in-bed time, they faced challenges in accurately
-distinguishing between sleep and wake epochs during in-bed time.
-Nevertheless, their performance could be substantially improved through
-the application of median filters. Among all models and configurations
+different machine learning models for predicting in-bed and sleep times
+and their corresponding sleep quality parameters. While the sequential
+models generally demonstrated excellent performance in predicting in-bed
+time, they faced challenges in accurately distinguishing between sleep
+and wake epochs during in-bed time. Among all models and configurations
 evaluated, the XGBoost model exhibited superior performance across all

### 2023-08-01 manuscript/manuscript.qmd
revise based on comments from Jan

@@
 abstract: "LÆS IKKE!
-Det er volapyk på latin.
+Dette er volapyk på latin.
 Lorem ipsum dolor sit amet, consectetur adipiscing elit.
@@
 Furthermore, our secondary goal was to assess the developed models'
-performance in evaluating important sleep quality parameters, including
+performance in evaluating important sleep quality metrics, including
 sleep period time (SPT), total sleep time (TST), sleep efficiency (SE),
@@
 ability to distinguish sleep stages are not a necessity to derive the
-sleep quality parameters of interest and to simplify the learning
-process of the models.
+sleep quality metrics of interest and to simplify the learning process
+of the models.
 
@@
 
-In this study, we employed two distinct learning strategies to analyze
-sleep patterns from thigh-mounted accelerometer data. The first strategy
-employed a sequential methodology, using an ensemble of four pairs of
-models, with each pair featuring the same algorithm. This sequential
-strategy was designed to simplify the prediction task by decomposing it
-into two stages: first predicting 'in-bed' time, then 'sleep' time. The
-output from the first set of models, which predicted in-bed time, was
-subjected to a 5-minute median filter to eliminate leading and trailing
-in-bed time blips. This process enabled us to extract a single
-continuous time interval that we identified as the sleep period time
-window (SPT). The SPT, in turn, served as the input to the second models
-in the sequence, enhancing their predictive accuracy for sleep time. We
-applied this sequential strategy across all four algorithms, as detailed
-below.
+We employed two different model strategies to assess sleep patterns from
+thigh-mounted accelerometer data. The first model strategy was designed
+as a sequence of two models to simplify the prediction task by
+decomposing it into two stages: first predicting 'in-bed' time, then
+'sleep' time. In the sequential model the output from the first set of
+models, which predicted in-bed time, was subjected to a 5-minute median
+filter to eliminate leading and trailing in-bed time blips. This process
+enabled us to extract a single continuous time interval that we
+identified as the sleep period time window (SPT). The SPT, in turn,
+served as the input to the second models in the sequence, enhancing
+their predictive accuracy for sleep time. We applied this sequential
+strategy using the following four machine learning algorithms:
 
-1.  Logistic Regression (LREG): Logistic regression served as a simple
-    and fast baseline model. However, due to its linear nature, it may
-    struggle with capturing complex relationships and non-linear
-    patterns present in the accelerometer data.
+1.  Logistic Regression: Logistic regression served as a simple and fast
+    baseline model. However, due to its linear nature, it may struggle
+    with capturing complex relationships and non-linear patterns present
+    in the accelerometer data.
 
-2.  Decision Tree (TREE): Decision trees are capable of handling
-    non-linear patterns and are easily interpretable. However, they are
-    prone to overfitting, particularly when dealing with complex
-    patterns that require simultaneous consideration of multiple
-    features.
+2.  Decision Tree: Decision trees are capable of handling non-linear
+    patterns and are easily interpretable. However, they are prone to
+    overfitting, particularly when dealing with complex patterns that
+    require simultaneous consideration of multiple features.
 
-3.  Single-layer Feed-forward Neural Network (SNN): Single-layer
-    feed-forward neural networks can effectively capture non-linear
-    relationships, even with their relatively simple structure. However,
-    they tend to be more challenging to interpret compared to simpler
-    models. Additionally, careful tuning of the network's architecture
-    and training process is required to mitigate the risk of
-    overfitting.
+3.  Single-layer Feed-forward Neural Network: Single-layer feed-forward
+    neural networks can effectively capture non-linear relationships,
+    even with their relatively simple structure. However, they tend to
+    be more challenging to interpret compared to simpler models.
+    Additionally, careful tuning of the network's architecture and
+    training process is required to mitigate the risk of overfitting.
 
-4.  XGBoost (XGB): XGBoost is a powerful algorithm known for its ability
-    to provide highly accurate predictions and handle complex,
-    non-linear patterns in the data. It also incorporates built-in
-    methods to prevent overfitting. However, training XGBoost models can
-    be computationally intensive, and interpreting the predictions it
+4.  XGBoost: XGBoost is a powerful algorithm known for its ability to
+    provide highly accurate predictions and handle complex, non-linear
+    patterns in the data. It also incorporates built-in methods to
+    prevent overfitting. However, training XGBoost models can be
+    computationally intensive, and interpreting the predictions it
     generates can pose challenges.
 
-In parallel, we also employed a multiclass approach using a
-bidirectional Long Short-Term Memory (biLSTM)[@hochreiter1997] neural
-network. This network, which was designed to predict three distinct
-classes: 'out-of-bed-awake', 'in-bed-awake', and 'in-bed-asleep', was
-configured with four layers and 128 hidden units per layer. This balance
-between model complexity and training efficiency was intended to
-facilitate learning of intricate patterns while ensuring feasible
-training times. The bidirectional nature of the LSTM enhanced data
-interpretation and reduced overfitting by doubling the hidden units at
-each time step. The LSTM model used sequences of tensors as input, with
-each sequence spanning 10 minutes and a step size of one. As
+In parallel, we also employed a multiclass algorithm as the second model
+strategy using a bidirectional Long Short-Term Memory
+(biLSTM)[@hochreiter1997] neural network which also incorporates
+temporal aspects of the data. This network, which was designed to
+predict three distinct classes: 'out-of-bed-awake', 'in-bed-awake', and
+'in-bed-asleep', was configured with four layers and 128 hidden units
+per layer. This balance between model complexity and training efficiency
+was intended to facilitate learning of intricate patterns while ensuring
+feasible training times. The bidirectional nature of the LSTM enhanced
+data interpretation and reduced overfitting by doubling the hidden units
+at each time step. The LSTM model used sequences of tensors as input,
+with each sequence spanning 10 minutes and a step size of one. As
 demonstrated by previous studies such as those by Sano et al.
@@
 optimize hyperparameters, we performed a 10-fold Monte Carlo
-cross-validation on a regular grid, comprising 20 different combinations
-of hyperparameters. The F1 score served as the optimization metric. The
-best-performing set of hyperparameters was then used to fit the models
-to the full training dataset. This approach allowed us to maximize
-performance by leveraging all available data. Moreover, after extracting
-the in-bed time from the initial sequential models, the imbalance on the
-resulting dataset could cause biases during model training, as models
-may favor predicting the majority class. To rectify this, we employed
-the Synthetic Minority Over-sampling Technique (SMOTE)[@chawla2002].
-SMOTE generates new samples by interpolating random samples with their
-nearest neighbors. We utilized the themis R package[@themis] to
-implement SMOTE, resulting in a balanced distribution of training
-samples across both classes.
+cross-validation on a regular grid, i.e., for each hyperparameter, a
+range of values at evenly-spaced intervals was selected. comprising 20
+different combinations of hyperparameters. The F1 score served as the
+optimization metric. The best-performing set of hyperparameters was then
+used to fit the models to the full training dataset. This approach
+allowed us to maximize performance by leveraging all available data. An
+imbalance was observed with the in-bed time determined in the initial
+step of the sequential model strategy which after extracting the in-bed
+time from the initial sequential models, the imbalance on the resulting
+dataset could cause biases during model training, as models may favor
+predicting the majority class. To account for this imbalance, we
+employed the Synthetic Minority Over-sampling Technique
+(SMOTE)[@chawla2002]. SMOTE generates new samples by interpolating
+random samples with their nearest neighbors. We utilized the themis R
+package[@themis] to implement SMOTE, resulting in a balanced
+distribution of training samples across both classes.
 
@@
 performance of each model on an epoch-to-epoch basis. These include
-accuracy ($accuracy = \frac{TP+TN}{TP+TN+FP+FN}$), sensitivity
-($sensitivity = \frac{TP}{TP+FN}$), specificity
-($specificity = \frac{TN}{TN+FP}$), precision
-($precision = \frac{TP}{TP+FP}$), negative predictive value (NPV,
-$NPV = \frac{TN}{TN + FN}$), and F1 score
-($F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}$).
+$$accuracy = \frac{TP+TN}{TP+TN+FP+FN}$$
+$$sensitivity = \frac{TP}{TP+FN}$$ $$specificity = \frac{TN}{TN+FP}$$
+$$precision = \frac{TP}{TP+FP}$$ $$NPV = \frac{TN}{TN + FN}$$
+$$F_1 = 2 \cdot \frac{precision \cdot sensitivity}{precision + sensitivity}$$
 
-In the context of our sequential learning strategy, the initial models
-were tasked with the binary classification of in-bed vs. out-of-bed. For
-this task, we assessed performance using the F1-score, accuracy,
-sensitivity, specificity, and precision metrics. The second models in
-our sequential learning strategy focused on the binary classification of
-asleep vs. awake. For these models, we considered the same metrics, in
-addition to the negative predictive rate. The class imbalance in this
-case led us to compute the F1 score as an unweighted macro-average.
-Additionally, we evaluated the multiclass classifier, biLSTM, using the
-same metrics. To do this, we considered the multiclass output as to
-binary classifications, where the first was out-of-bed vs the rest and
-the second binary classification as in-bed-awake vs in-bed-asleep. To
+where NPV is negative predictive value, TP is true positives, FP is
+false positives, TN is true negatives, and FN is false negatives.
+
+In the context of our sequential model strategy, the initial models were
+tasked with the binary classification of in-bed vs. out-of-bed. For this
+task, we assessed performance using the F1-score, accuracy, sensitivity,
+specificity, and precision metrics. The second models in our sequential
+model strategy focused on the binary classification of asleep vs. awake.
+For these models, we considered the same metrics, in addition to the
+negative predictive rate. The class imbalance in this case led us to
+compute the F1 score as an unweighted macro-average. Additionally, we
+evaluated the multiclass classifier, biLSTM, using the same metrics. To
+do this, we considered the multiclass output as to binary
+classifications, where the first was out-of-bed vs the rest and the
+second binary classification as in-bed-awake vs in-bed-asleep. To
 further illustrate model performance, we provide confusion matrices for
@@
 To assess the performance of our models in deriving sleep quality
-parameters, we utilized Bland-Altman plots and Pearson correlations. The
+metrics, we utilized Bland-Altman plots and Pearson correlations. The
 Bland-Altman method was employed specifically to determine the level of
-agreement between two measurement techniques. Considering our dataset
-contained multiple observations per subject, we integrated a bootstrap
-procedure to address this extra source of variability. We calculated the
-mean difference (bias) and defined the LOA as the mean difference plus
-or minus 1.96 times the standard deviation of these differences. To
-ensure our measurements were robust and accounted for intra-subject
-variability, we estimated the 95% confidence intervals for both the bias
-and the LOA using a bias-corrected and accelerated bootstrap method,
-utilizing 10,000 bootstrap replicates. The sleep quality parameteres
-included are defined as follews in accordance with the ZM definitions:
+agreement between two measurement techniques. Given the nature of our
+dataset, which contains multiple observations per subject, we employed a
+bootstrap procedure to account for this added variability. We first
+calculated the mean difference (bias) and then defined the limits of
+agreement (LOA) as the mean difference plus or minus 1.96 times the
+standard deviation of these differences. Acknowledging the possibility
+of non-normality and potential skewness in our data, we chose to apply a
+bias-corrected and accelerated (BCa) bootstrap
+method[@diciccio_bootstrap_1996]. This approach allowed us to better
+address potential bias in our estimates and the inherent intra-subject
+variability. Utilizing 10,000 bootstrap replicates, we estimated the 95%
+confidence intervals for both the bias and the LOA, thus ensuring
+robustness in our measurements. The sleep quality parameteres included
+are defined as follows in accordance with the ZM definitions:
 
@@
 
-As reported in @tbl-zm_overview the sleep quality parameters derived
-from ZM predictions were modified by the implementation of 5-minute and
+As reported in @tbl-zm_overview the sleep quality metrics derived from
+ZM predictions were modified by the implementation of 5-minute and
 10-minute median filters. SPT were consistent across raw and filtered
@@
 from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. LPS
-also elevated, suggesting the filter smooths out brief awakenings at
-sleep onset, leading to a prolonged time to persistent sleep. A
-significant change was seen in WASO, which dropped from 39.0 ± 33.6
-minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in
-the 5-minute and 10-minute filtered data, respectively. The number of
-awakenings was also considerably reduced with the application of
-filters. In the raw data, the average number of awakenings was 34.46 ±
-11.33 per night, which reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the
-5-minute and 10-minute filtered data sets respectively.
+also increased, suggesting the filter removes brief awakenings at sleep
+onset, leading to a prolonged time to persistent sleep. A change was
+seen in WASO, which dropped from 39.0 ± 33.6 minutes in raw data to 30.6
+± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and 10-minute
+filtered data, respectively. The number of awakenings was also
+considerably reduced with the application of filters. In the raw data,
+the average number of awakenings was 34.46 ± 11.33 per night, which
+reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the 5-minute and 10-minute
+filtered data sets respectively.
 
@@
 (Decision Tree) to 96.1% (XGBoost). Sensitivity, Precision, and
-Specificity also demonstrate consistent results across the models. The
-XGBoost model, despite recording the highest metrics with an F1 score of
-95.4% and accuracy of 96.1%, outpaced the others only marginally.
+Specificity also demonstrate consistent results across the different
+models. The XGBoost model provide the best performance with an F1 score
+of 95.4% and accuracy of 96.1%, although only outpacing the other models
+marginally.
 
@@
 
-## Evaluation of Sleep Quality Parameters
+## Evaluation of sleep quality metrics
 
 @tbl-ba_cor presents a comparative analysis of the included models used
-to predict various sleep quality parameters (SPT, TST, SE, LPS, WASO)
-using the 5-minute median filtered ZM predictions. To see the full table
+to predict various sleep quality metrics (SPT, TST, SE, LPS, WASO) using
+the 5-minute median filtered ZM predictions. To see the full table
 including models developed from raw ZM predictions and 10-minute median
@@
 median predictions. Considering LOA, the decision tree had higher
-variability across different sleep quality parameters and filtering
+variability across different sleep quality metrics and filtering
 techniques, particularly for LPS and WASO, which indicates lower
@@
 the pearson coefficient, revealed that the XGBoost model consistently
-had the highest correlation with ZM across all sleep quality parameters
-and filtering methods Notably, the XGBoost's 5-minute median predictions
+had the highest correlation with ZM across all sleep quality metrics and
+filtering methods Notably, the XGBoost's 5-minute median predictions
 showed the strongest correlation (0.66) for TST among all models and
@@
 on 5-minute median filtered ZM predictions, and the 5-minute
-median-smoothed ZM-derived sleep quality parameters. The Bland-Altman
-plot for the SPT and TST reveals a significant level of agreement with
-the ZM, as evidenced by a bias close to zero. Interestingly, a portion
-of the data points are located near the zero line indicating perfect
-agreement. However, the presence of extreme outliers widens the limits
-of agreement (LOA). The scatterplot for SPT also demonstrates a positive
-trend, indicating a moderate linear correlation between the XGBoost
-model and the ZM-derived sleep quality parameters. The bias and LOA for
-TST are comparable to those observed for SPT, indicating a consistent
-level of agreement between the two methods. The scatterplot for TST also
-shows a slightly higher correlation, primarily driven by the absence of
-extreme outliers.Furthermore, the remaining three sleep quality
-parameters, SE, LPS, and WASO, exhibit heteroscedasticity in contrast to
-SPT and TST. This outcome is expected as achieving 100% sleep efficiency
-is relatively rare, resulting in less disagreement between the methods
-as values approach the upper limit. However, as sleep efficiency
+median-smoothed ZM-derived sleep quality metrics. The Bland-Altman plot
+for the SPT and TST reveals a good level of agreement with the ZM, as
+evidenced by a bias close to zero. Interestingly, a portion of the data
+points are located near the zero line indicating perfect agreement. The
+scatterplot for SPT also demonstrates a positive trend, indicating a
+moderate linear correlation between the XGBoost model and the ZM-derived
+sleep quality metrics. The bias and LOA for TST are comparable to those
+observed for SPT, indicating a consistent level of agreement between the
+two methods. The scatterplot for TST also shows a slightly higher
+correlation, primarily driven by the absence of extreme
+outliers.Furthermore, the remaining three sleep quality metrics, SE,
+LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST.
+This outcome is expected as achieving 100% sleep efficiency is
+relatively rare, resulting in less disagreement between the methods as
+values approach the upper limit. Conversely, as sleep efficiency
 decreases, the potential for discrepancies and differing interpretations
@@
 moderate positive linear correlation is observed between the XGBoost
-model and ZM-derived sleep quality parameters for SE, however, a poor
-correlation is observed for LPS and WASO, suggesting less agreement
-between the two methods for these parameters. Similar plots for all
-models are available in the **supplementary materials (75 plots. Jeg ved
-ikke om nogen orker at bladre igennem dem).**
+model and ZM-derived sleep quality metrics for SE, however, a poor
+correlation is observed for LPS and WASO. Similar plots for all models
+are available in the **supplementary materials (75 plots. Jeg ved ikke
+om nogen orker at bladre igennem dem).**
 
-![Comparison of sleep quality parameters derived from the XGBoost model
+![Comparison of sleep quality metrics derived from the XGBoost model
 trained on the 5-minute smoothed ZM predictions. The left column
@@
 displays scatter plots of XGBoost-derived vs ZM-derived sleep quality
-parameters. The dashed line represents the identity line, while the
+metrics. The dashed line represents the identity line, while the
 full-drawn line represents the best linear fit. Pearson's correlations
@@
 
-In an effort to mature the methods for estimating sleep from thigh-worn
+To select the most optimal method for estimating sleep from thigh-worn
 accelerometers, we evaluated various models for predicting in-bed and
-sleep time and their derived sleep quality parameters. Furthermore, we
-trained and evaluated the models using raw and median-filtered gold
-standard predictions from the ZM. In general, all sequential models
-performed well at predicting in-bed time. More challenging was it to
-distinguish wake from sleep on the extracted in-bed time, however, the
-performance of the sequential models were enhanced by the application of
-median filters. Moreover, even though the multiclass biLSTM showed good
-performance across F1 score, precision and NPV, the derived sleep
-quality parameters were not on par with the XGBoost model which
-demonstrated the highest performance metrics across all evaluations,
-including epoch-to-epoch prediction and sleep quality parameter
-derivatives. Despite this, all sequential models showed low specificity
+sleep time and their derived sleep quality metrics. We trained and
+evaluated the models using raw and median-filtered gold standard sleeå
+estimates from the ZM EEG-based sleep monitor. In general, all
+sequential models performed well at predicting in-bed time. More
+challenging was it to distinguish wake from sleep on the extracted
+in-bed time, and the performance of the sequential models were enhanced
+by the application of median filterings. Moreover, even though the
+multiclass biLSTM showed good performance across F1 score, precision and
+NPV, the derived sleep quality metrics were not on par with the XGBoost
+model which demonstrated the highest performance metrics across all
+evaluations, including epoch-to-epoch prediction and all sleep quality
+metrics. Despite this, all sequential models showed low specificity
 values, indicating difficulty in correctly classifying awake epochs
 during time in bed. The application of 5-minute and 10-minute median
-filters improved the performance metrics of all models, with the XGBoost
-model consistently leading. The median filters increased total sleep
-time and sleep efficiency, while reducing wake after sleep onset and the
-number of awakenings. The XGBoost model also showed the least bias and
-highest correlation with ZM across sleep quality parameters.
+filters improved the performance metrics of all models. Median
+filterings increase total sleep time and sleep efficiency, while
+reducing wake after sleep onset and the number of awakenings. The
+XGBoost model provides the smallest bias and highest correlation with
+all ZM sleep quality metrics.
 
@@
 determining in-bed time does not necessarily translate to accurate
-predictions of broader sleep quality parameters. The crucial task of
+predictions of broader sleep quality metrics. The crucial task of
 detecting awake periods during in-bed time, a key factor in assessing
@@
 to night recordings or to 24-hour recordings. In night recordings, it is
-possible to derive sleep quality parameters like SE and LPS because the
-SPT is already known because it is inferred from the length of the
+possible to derive sleep quality metrics like SE and LPS because the SPT
+is already known because it is inferred from the length of the
 recording[@conley2019; @patterson_40_2023]. On the other hand, when
@@
 diaries[@girschik2012]. Consequently, these methods are unable to
-generate certain sleep quality parameters that rely on the
+generate certain sleep quality metrics that rely on the
 SPT[@doherty2017; @anderson2014]. To overcome this limitation, we have
@@
 approach allows our models to estimate all commonly used sleep quality
-parameters. Van Hees et al. [@van_hees_estimating_2018] have proposed an
+metrics. Van Hees et al. [@van_hees_estimating_2018] have proposed an
 algorithm to determine SPT from data collected by wrist-worn devices.
@@
 [@plekhanova2023]. By combining this algorithm with other methods,
-further sleep quality parameters can be inferred based on the identified
+further sleep quality metrics can be inferred based on the identified
 SPT. Van Hees et al.[@van_hees_estimating_2018] reported good agreements
@@
 
-In our evaluation of sleep quality parameters, we found that LPS had the
+In our evaluation of sleep quality metrics, we found that LPS had the
 largest mean error relative to absolute time allocated to LPS. This
@@
 
-Our study's XGBoost model demonstrated relatively narrower limits of
-agreements (LOAs) for TST, SE, and WASO, with ranges of -95.5 to 81.4
-min, -15.6 to 13.3%, and -83.4 to 81.7 min, respectively when compared
-with other models such as the Van Hees algorithm[@hees2015], Oakley rsc
-(rescored)[@palotti2019], and LSTM-50[@palotti2019] evaluated in the
-Patterson et al. study[@patterson_40_2023]. Furthermore, comparing the
-LOAs between our XGBoost model and the algorithm developed for
-thigh-worn devices by Johansson et al.
-study[@johansson_development_2023], our XGBoost model showed narrower
-LOAs for TST , SE, LPS , and WASO, but not SPT. Generally, all methods,
-both from this study and from the reviewed literature, exibit wide LOAs
-sugesting that there is high variability in the derevide sleep quality
-parameters. These findings imply that the current methods, are only
-reasonbly reliable for assessing sleep quality parameters at a group
-level. However, caution should be exercised when applying the models and
-methods to individual-level sleep assessments. Therefore, further
-improvements and refinements are needed to enhance the precision and
-reliability of these models for individual sleep assessments.
+Our study's XGBoost model demonstrated relatively narrower LOAs for TST,
+SE, and WASO, with ranges of -95.5 to 81.4 min, -15.6 to 13.3%, and
+-83.4 to 81.7 min, respectively when compared with other models such as
+the Van Hees algorithm[@hees2015], Oakley rsc (rescored)[@palotti2019],
+and LSTM-50[@palotti2019] evaluated in the Patterson et al.
+study[@patterson_40_2023]. Furthermore, comparing the LOAs between our
+XGBoost model and the algorithm developed for thigh-worn devices by
+Johansson et al. study[@johansson_development_2023], our XGBoost model
+showed narrower LOAs for TST , SE, LPS , and WASO, but not SPT.
+Generally, all methods, both from this study and from the reviewed
+literature, exibit wide LOAs sugesting that there is high variability in
+the derevide sleep quality metrics. In the current study, the presence
+of extreme outliers seem to drive the widening the LOAs. These findings
+imply that the current methods, are only reasonbly reliable for
+assessing sleep quality metrics at a group level. However, caution
+should be exercised when applying the models and methods to
+individual-level sleep assessments. Therefore, further improvements and
+refinements are needed to enhance the precision and reliability of these
+models for individual sleep assessments.
 
@@
 itself mitigates this issue by applying certain filtering processes when
-generating sleep quality parameters. For example, epochs contributing to
+generating sleep quality metrics. For example, epochs contributing to
 WASO must be in contiguous epochs of 3, and sleep only counts towards
-sleep quality parameters if 10 out of 12 minutes are scored as sleep. To
+sleep quality metrics if 10 out of 12 minutes are scored as sleep. To
 improve the prospect of our machine learning algorithms, we applied
 median filters to the ZM raw predictions. This did in fact alter the
-derived sleep quality parameters. Notably, the mean WASO decreased from
-39 minutes in the raw predictions to 30.6 minutes in the 5-minute median
+derived sleep quality metrics. Notably, the mean WASO decreased from 39
+minutes in the raw predictions to 30.6 minutes in the 5-minute median
 filtered predictions, and further decreased to 22.3 minutes in the
@@
 distinguish in-bed awake and asleep from out-of-bed, thereby allowing
-for the extraction of vital sleep quality parameters. Furthermore, the
+for the extraction of vital sleep quality metrics. Furthermore, the
 research benefits from evaluating multiple nights per subject, providing
@@
 different machine learning models for predicting in-bed and sleep times
-and their corresponding sleep quality parameters. While the sequential
+and their corresponding sleep quality metrics. While the sequential
 models generally demonstrated excellent performance in predicting in-bed

### 2023-08-01 manuscript/manuscript.qmd
manus

@@
           postal-code: 5230
-abstract: "LÆS IKKE!
-Dette er volapyk på latin.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse vitae dictum eros, ullamcorper elementum orci.
-Sed laoreet nulla neque, pulvinar fermentum mi iaculis at.
-Nulla ultricies nibh sit amet vestibulum rutrum.
-Nam pharetra nisl sed ipsum maximus suscipit.
-Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante.
-Nunc vitae lectus nisi.
-Aliquam efficitur ut eros ut pellentesque.
-Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus.
-Curabitur sollicitudin lectus sit amet velit pellentesque laoreet.
-Ut posuere diam lobortis nisi eleifend tincidunt.
-Ut at euismod sem, sed dignissim ligula.
-Aliquam lacinia massa libero, id eleifend velit pulvinar ac.
-Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
+abstract: "Studying sleep is vital in health research, but the gold standard, polysomnography, is costly and impractical for large-scale studies.
+An affordable alternative is using wearable accelerometers.
+While wrist and hip-worn devices are commonly used in sleep research, thigh-worn accelerometers have been relatively unexplored.
+Our study evaluated machine learning and deep learning models utilizing data from thigh-worn accelerometers to estimate sleep and sleep quality metrics, comparing them with an EEG-based sleep monitor.
+The dataset consisted of data from 585 days and nights, comprising accelerometry and EEG-based sleep estimates from children aged 6-10.
+We employed both sequential and multiclass model strategies on both raw and filtered data.
+The most effective model was XGBoost, which performed well when applied to 5-minute median filtered data, exhibiting small biases in sleep period time (0.2 minutes), total sleep time (-7 minutes), sleep efficiency (-1.1%), and wake after sleep onset (-0.9 minutes).
+Furthermore, the XGBoost model showed a robust correlation (0.66, 95% CI: 0.61 - 0.7) with total sleep time, indicating its potential.
+However, despite these favorable results in bias, our study revealed large limits of agreements in accordance with previous research on hip- and wrist-worn devices.
+In conclusion, we present promising results in using machine learning technieques to estimate sleep quality metrics, however, accurately classifying awake periods during in-bed time remained challenging.
+Moreover, additional improvements are necessary to precisely assess individual sleep quality metrics due to the notable limits of agreement."
 keywords: 
@@
 encompass vector magnitude, mean crossing rate, skewness, and kurtosis
-for each of the x, y, and z dimensions. All features are summarized in
-table ??? **in the supplementary matrials**. Subsequently, we merged the
-ZM and corresponding accelerometer recordings. Any overlapping time
-between the ZM and accelerometer data was treated as 'in-bed' time, with
-the remaining time considered 'out-of-bed'. This process yielded a
-dataset providing a around the clock temporal view of each participant's
+for each of the x, y, and z dimensions. Subsequently, we merged the ZM
+and corresponding accelerometer recordings. Any overlapping time between
+the ZM and accelerometer data was treated as 'in-bed' time, with the
+remaining time considered 'out-of-bed'. This process yielded a dataset
+providing a around the clock temporal view of each participant's
 activity and sleep patterns.
@@
 thigh-mounted accelerometer data. The first model strategy was designed
-as a sequence of two models to simplify the prediction task by
-decomposing it into two stages: first predicting 'in-bed' time, then
-'sleep' time. In the sequential model the output from the first set of
-models, which predicted in-bed time, was subjected to a 5-minute median
-filter to eliminate leading and trailing in-bed time blips. This process
-enabled us to extract a single continuous time interval that we
-identified as the sleep period time window (SPT). The SPT, in turn,
-served as the input to the second models in the sequence, enhancing
-their predictive accuracy for sleep time. We applied this sequential
-strategy using the following four machine learning algorithms:
+as a sequence of two models, each functioning as a binary classifier.
+This approach aimed to simplify the prediction task by decomposing the
+multiclass problem of classifying 'out-of-bed-awake', 'in-bed-awake',
+and 'in-bed-asleep' into two binary stages: first predicting 'in-bed'
+time, then 'sleep' time. The output from the first set of binary
+classifiers, which predicted in-bed time, was subjected to a 5-minute
+median filter to remove transient in-bed time blips. This process
+enabled us to establish a single continuous time interval that we
+identified as the sleep period time window (SPT). The SPT then served as
+the input for the second stage of binary classifiers in the sequence,
+further enhancing their predictive accuracy for sleep time." We applied
+this sequential strategy using the following four machine learning
+algorithms:
 
@@
     overfitting, particularly when dealing with complex patterns that
-    require simultaneous consideration of multiple features.
+    require simultaneous consideration of multiple features. In the
+    current study, we used a maximum tree depth of 8.
 
@@
 
-For the models in sequence, we trained four pairs of classification
-models. Each pair was designed to distinguish between in-bed/out-of-bed
-and asleep/awake states, respectively. The dataset was randomly split
-into a training set and a testing set, each containing approximately 50%
-of the subjects. The splitting of the data was ensured to not have
-samples from the same night simultaneously present in both sets. To
-optimize hyperparameters, we performed a 10-fold Monte Carlo
-cross-validation on a regular grid, i.e., for each hyperparameter, a
-range of values at evenly-spaced intervals was selected. comprising 20
-different combinations of hyperparameters. The F1 score served as the
-optimization metric. The best-performing set of hyperparameters was then
-used to fit the models to the full training dataset. This approach
-allowed us to maximize performance by leveraging all available data. An
-imbalance was observed with the in-bed time determined in the initial
-step of the sequential model strategy which after extracting the in-bed
-time from the initial sequential models, the imbalance on the resulting
-dataset could cause biases during model training, as models may favor
-predicting the majority class. To account for this imbalance, we
-employed the Synthetic Minority Over-sampling Technique
-(SMOTE)[@chawla2002]. SMOTE generates new samples by interpolating
-random samples with their nearest neighbors. We utilized the themis R
-package[@themis] to implement SMOTE, resulting in a balanced
-distribution of training samples across both classes.
+We trained four pairs of models in sequence, with each pair
+distinguishing between in-bed/out-of-bed and asleep/awake states,
+respectively. We divided our dataset randomly into a training set and a
+testing set, with each containing roughly half of the subjects. The
+splitting of the data was ensured to not have samples from the same
+night simultaneously present in both sets. To optimize hyperparameters,
+we performed a 10-fold Monte Carlo cross-validation on a regular grid
+(i.e., for each hyperparameter, a range of values at evenly-spaced
+intervals was selected) comprising 20 different combinations of
+hyperparameters. The F1 score served as the optimization metric. The
+best-performing set of hyperparameters was then used to fit the models
+to the full training dataset. This approach allowed us to maximize
+performance by leveraging all available training data to estimate the
+model parameters. An imbalance was observed with the in-bed time
+determined in the initial step of the sequential model strategy which
+after extracting the in-bed time from the initial sequential models, the
+imbalance on the resulting dataset could cause biases during model
+training, as models may favor predicting the majority class. To account
+for this imbalance, we employed the Synthetic Minority Over-sampling
+Technique (SMOTE)[@chawla2002]. SMOTE generates new samples by
+interpolating random samples with their nearest neighbors. We utilized
+the themis R package[@themis] to implement SMOTE, resulting in a
+balanced distribution of training samples across both classes.
 
@@
 agreement between two measurement techniques. Given the nature of our
-dataset, which contains multiple observations per subject, we employed a
-bootstrap procedure to account for this added variability. We first
-calculated the mean difference (bias) and then defined the limits of
-agreement (LOA) as the mean difference plus or minus 1.96 times the
-standard deviation of these differences. Acknowledging the possibility
-of non-normality and potential skewness in our data, we chose to apply a
-bias-corrected and accelerated (BCa) bootstrap
-method[@diciccio_bootstrap_1996]. This approach allowed us to better
-address potential bias in our estimates and the inherent intra-subject
-variability. Utilizing 10,000 bootstrap replicates, we estimated the 95%
-confidence intervals for both the bias and the LOA, thus ensuring
-robustness in our measurements. The sleep quality parameteres included
-are defined as follows in accordance with the ZM definitions:
+dataset, which contains multiple observations per subject but
+necessarily equal number, we employed a bootstrap procedure to account
+for this added variability. We first calculated the mean difference
+(bias) and then defined the limits of agreement (LOA) as the mean
+difference plus or minus 1.96 times the standard deviation of these
+differences. Acknowledging the possibility of non-normality and
+potential skewness in our data, we chose to apply a bias-corrected and
+accelerated (BCa) bootstrap method[@diciccio_bootstrap_1996]. This
+approach allowed us to better address potential bias in our estimates
+and the inherent intra-subject variability. Utilizing 10,000 bootstrap
+replicates, we estimated the 95% confidence intervals for both the bias
+and the LOA, thus ensuring robustness in our measurements. The sleep
+quality metrics included are defined as follows in accordance with the
+ZM definitions:
 
@@
 #| label: tbl-in_bed_performance
-#| tbl-cap: In-bed performance metrics
+#| tbl-cap: Performance metrics of the classification of in-bed/out-of-bed time of the included models.
 
@@
 #| label: tbl-sleep_performance
-#| tbl-cap: Performance of the sleep/wake classification of the sequential models.
+#| tbl-cap: Performance metrics of the sleep/wake classification of the included models.
 
@@
 A complete set of confusion matrices generated from data both containing
-the out-of-bed and in-bed time are presented in @fig-conf_mat. The
-figure shows favorable epoch-to-epoch performance across all sequential
-models, however, it is evident that the biLSTM is less successful in
-classifying the in-bed-awake class which cannot be deduced from the
-confusion matrices from the sequential models.
+the out-of-bed and in-bed time are presented in @fig-conf_mat. These
+matrices showcase the epoch-to-epoch performance of all sequential
+models in distinguishing between 'awake' and 'asleep' states, regardless
+of whether the subject is 'in-bed' or 'out-of-bed'. However, it's
+important to note that the binary nature of these sequential models
+means they cannot provide direct information about the classification of
+the 'in-bed-awake' state. In contrast, the biLSTM model, which also
+categorizes the 'in-bed-awake' state as a distinct class, appears to
+have less success in classifying this particular state.
 
 ![Confusion matrices for binary sleep prediction. The middle of each
-tile is the normalized count (overall percentage) and, beneath it, the
-count. The bottom number is the column percentage (target). At the right
-side of each tile is the row percentage (prediction). i) decision tree,
-ii) logistic regression, iii) feed-forward neural net, iv) XGBoost, and
-v) biLSTM.](visuals/all_conf_mats.pdf){#fig-conf_mat}
+tile is the normalized count (overall percentage). The bottom number of
+each tile is the column percentage and the right side of each tile is
+the row percentage. i) decision tree, ii) logistic regression, iii)
+feed-forward neural net, iv) XGBoost, and v)
+biLSTM.](visuals/all_conf_mats.pdf){#fig-conf_mat}
 
@@
 including models developed from raw ZM predictions and 10-minute median
-filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
-decision tree model consistently underestimated SPT, TST, and SE, and
-overestimated LPS and WASO in comparison to ZM. The logistic regression
-model had similar trends, with more pronounced underestimation in TST
-and overestimation in LPS. The feed-forward neural network also
-exhibited similar bias as the decision tree and the logistic regression
-models, but with a higher overestimation in WASO. On the other hand, the
-XGBoost model showed least bias among all, especially in its 5-minute
-median predictions. Considering LOA, the decision tree had higher
-variability across different sleep quality metrics and filtering
-techniques, particularly for LPS and WASO, which indicates lower
-agreement with ZM. Other models had comparable LOA but with notable
-exceptions. For example, TST LOA for the logistic regression model was
-particularly wide in the 5-minute median predictions. Correlation-wise,
-the pearson coefficient, revealed that the XGBoost model consistently
-had the highest correlation with ZM across all sleep quality metrics and
-filtering methods Notably, the XGBoost's 5-minute median predictions
-showed the strongest correlation (0.66) for TST among all models and
-filtering techniques.
+filtered ZM predictions, see table 1 in supplementary materials. In
+terms of bias, the decision tree model consistently underestimated SPT,
+TST, and SE, and overestimated LPS and WASO in comparison to ZM. The
+logistic regression model had similar trends, with more pronounced
+underestimation in TST and overestimation in LPS. The feed-forward
+neural network also exhibited similar bias as the decision tree and the
+logistic regression models, but with a higher overestimation in WASO. On
+the other hand, the XGBoost model showed least bias among all,
+especially in its 5-minute median predictions. Considering LOA, the
+decision tree had higher variability across different sleep quality
+metrics and filtering techniques, particularly for LPS and WASO, which
+indicates lower agreement with ZM. Other models had comparable LOA but
+with notable exceptions. For example, TST LOA for the logistic
+regression model was particularly wide in the 5-minute median
+predictions. Correlation-wise, the pearson coefficient, revealed that
+the XGBoost model consistently had the highest correlation with ZM
+across all sleep quality metrics and filtering methods Notably, the
+XGBoost's 5-minute median predictions showed the strongest correlation
+(0.66) for TST among all models and filtering techniques.
 
@@
 #| label: tbl-ba_cor
-#| tbl-cap: Summary of Bias, Limits of Agreement, and Pearson Correlation for various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using different Machine Learning Models (Decision Tree, Logistic Regression, Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and 10-Min Median as predictors.
-Each value is provided with its 95% Confidence Interval (CI).
+#| tbl-cap: Summary of bias, limits of agreement, and pearson correlation for various sleep parameter predictions (SPT, TST,SE, LPS, WASO) using different machine learning and deep learning models (decision tree, logistic regression, feed-forward neural network, XGBoost) on raw ZM predictions, 5-minute and 10-minute median predictions.
+Each value is provided with its 95% confidence interval (CI).
 
@@
 LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST.
-This outcome is expected as achieving 100% sleep efficiency is
+~~This outcome is expected as achieving 100% sleep efficiency is
 relatively rare, resulting in less disagreement between the methods as
@@
 decreases, the potential for discrepancies and differing interpretations
-between the methods increases, leading to greater heteroscedasticity. A
-moderate positive linear correlation is observed between the XGBoost
+between the methods increases, leading to greater heteroscedasticity.~~
+A moderate positive linear correlation is observed between the XGBoost
 model and ZM-derived sleep quality metrics for SE, however, a poor
-correlation is observed for LPS and WASO. Similar plots for all models
-are available in the **supplementary materials (75 plots. Jeg ved ikke
-om nogen orker at bladre igennem dem).**
+correlation is observed for LPS and WASO.
 
@@
 sleep estimation methods using thigh-worn accelerometers. We evaluated
-different machine learning models for predicting in-bed and sleep times
-and their corresponding sleep quality metrics. While the sequential
-models generally demonstrated excellent performance in predicting in-bed
-time, they faced challenges in accurately distinguishing between sleep
-and wake epochs during in-bed time. Among all models and configurations
-evaluated, the XGBoost model exhibited superior performance across all
-performance metrics, including epoch-to-epoch predictions and sleep
-quality parameter derivatives. Our research also highlighted the current
-limitations of sleep detection methods, such as challenges in
-effectively detecting wake periods during in-bed time and the need for
-further improvements to increase the precision of individual sleep
-assessments. We believe our work lays the groundwork for future research
-to further refine and improve the performance of these models,
-contributing to a more precise and accurate evaluation of sleep patterns
-and quality using thigh-worn accelerometers.
+different machine learning and deep learnng models for predicting in-bed
+and sleep times and their corresponding sleep quality metrics. While the
+sequential models generally demonstrated excellent performance in
+predicting in-bed time, they faced challenges in accurately
+distinguishing between sleep and wake epochs during in-bed time. Among
+all models and configurations evaluated, the XGBoost model exhibited the
+best performance, including epoch-to-epoch predictions and sleep quality
+metrics. Our research also highlighted the current limitations of sleep
+detection methods, such as challenges in effectively detecting wake
+periods during in-bed time and the need for further improvements to
+increase the precision of individual sleep assessments. We believe our
+work lays the groundwork for future research to further refine and
+improve the performance of these models, contributing to a more precise
+and accurate evaluation of sleep patterns and quality using thigh-worn
+accelerometers.
 

### 2023-08-01 manuscript/supp_mat/supp.qmd
manus

--- 2023-07-07 manuscript/supp_mat/supp.qmd
@@
 ---
+
+Table of features? Er det nødvendigt?
 
@@
 
-![all ba and scatterplots](../visuals/all_ba_scatterplots.pdf)
+Plots similar to figure 5 for all model and filtering combinations (75 plots in total) are available upon request.
+
+<!-- ![Decision tree, raw](../visuals/raw_decision_tree_ba_cor.pdf){#tbl-raw_dc} -->
+<!-- ![Decision tree, median_5](../visuals/median_5_decision_tree_ba_cor.pdf){#tbl-m5_dc} -->
+<!-- ![Decision tree, median_10](../visuals/median_10_decision_tree_ba_cor.pdf){#tbl-m10_dc} -->
+
+<!-- ![Logistic regression, raw](../visuals/raw_logistic_regression_ba_cor.pdf) -->
+<!-- ![Logistic regression, median_5](../visuals/median_5_logistic_regression_ba_cor.pdf) -->
+<!-- ![Logistic regression, median_10](../visuals/median_10_logistic_regression_ba_cor.pdf) -->
+
+<!-- ![Neural_network, raw](../visuals/raw_neural_network_ba_cor.pdf) -->
+<!-- ![Neural_network, median_5](../visuals/median_5_neural_network_ba_cor.pdf) -->
+<!-- ![Neural_network, median_10](../visuals/median_10_neural_network_ba_cor.pdf) -->
+
+<!-- ![XGBoost, raw](../visuals/raw_xgboost_ba_cor.pdf) -->
+<!-- ![XGBoost, median_5](../visuals/median_5_xgboost_ba_cor.pdf) -->
+<!-- ![XGBoost, median_10](../visuals/median_10_xgboost_ba_cor.pdf) -->
+
+<!-- ![biLSTM, raw](../visuals/raw_biLSTM_ba_cor.pdf) -->
+<!-- ![biLSTM, median_5](../visuals/median_5_biLSTM_ba_cor.pdf) -->
+<!-- ![biLSTM, median_10](../visuals/median_10_biLSTM_ba_cor.pdf) -->

### 2023-08-02 manuscript/manuscript.qmd
update README.md

@@
 title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
+shorttitle: "Improving Sleep Quality Estimation"
 author:
@@
 bibliography: bibliography.bib
+include-in-header: 
+      text: '\usepackage{lineno}\linenumbers'
 format:
@@
     keep-tex: false
+    # fig-pos: 'b'
     journal:
@@
       cite-style: super
-    include-in-header: 
-      text: '\usepackage{lineno}\linenumbers'
-    fig-pos: 'b'
+  simple-article-pdf:
+    indent: false
+    classoption: 
+     - 9pt
+     - twoside
   docx:
     number-sections: true
+  html: 
+    csl: https://www.zotero.org/styles/nature
+pdf-engine: xelatex
 editor: 
@@
 outliers.Furthermore, the remaining three sleep quality metrics, SE,
-LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST.
-~~This outcome is expected as achieving 100% sleep efficiency is
-relatively rare, resulting in less disagreement between the methods as
-values approach the upper limit. Conversely, as sleep efficiency
-decreases, the potential for discrepancies and differing interpretations
-between the methods increases, leading to greater heteroscedasticity.~~
-A moderate positive linear correlation is observed between the XGBoost
+LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST. A
+moderate positive linear correlation is observed between the XGBoost
 model and ZM-derived sleep quality metrics for SE, however, a poor
